### 2012-02-07 mrfy_experience_report.tex
Initial commit

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

\begin{document}

\conferenceinfo{ICFP '12}{date, City.} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Experience Report: Haskell in Computational Biology}
\subtitle{Subtitle Text, if any}

\authorinfo{Noah M. Daniels \and Andrew Gallant \and Norman Ramsey}
           {Department of Computer Science, Tufts University}
           {\{ndaniels, agallant, nr\}@cs.tufts.edu}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

We address this paper to computational biologists who may have an interest in 
functional programming, but this is not an advocacy paper. We do not argue that 
computational biologists \textit{should} use functional programming. Rather, 
argue that computational biologists \textit{can} successfully develop software 
in Haskell, and what to expect if they do.

We also address this paper to the functional programming community, in the 
hopes that our experiences will be enlightening. We believe that our 
experiences can provide lessons about outreach and education.

\subsection{Problem Domain}

Computational biologists write software to solve problems on large genomic or 
proteomic data sets. Traditionally, these software programs fall into two 
categories: those with high-performance computing needs are typically 
implemented in C++, while the remainder are typically implemented in Perl, 
Python, Ruby, or R. Recently, we have developed a program in Haskell which 
implements a novel algorithm for solving the remote homology detection problem. 
During this process, we have wondered: would the algorithm be easily expressed 
in Haskell? Would we be able to achieve acceptable runtime performance? How 
would our experience compare with our past experiences developing and 
maintaining computational biology software in C++, Ruby, and Python? What would 
be our greatest challenges? In this paper, we briefly explain the remote 
homology detection problem and our algorithm, and then attempt to answer the 
above questions.

As a brief refresher in the standard dogma of genetics, recall that genes 
(strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
into the 20 naturally-occuring amino acids, which form peptide chains -- 
proteins -- which fold into complex structures in the cell. These proteins are 
cellular machinery, performing the varied functions an organism needs to 
survive. Since the dawn of the human genome project in 1990, the cost of 
sequencing an entire genome has dropped exponentially. Recently, the doubling 
time for genomic data has been one year, in contrast to the eighteen month 
doubling time of Moore's law. However, determining what newly sequenced genes 
actually do -- when they become proteins, what structures those proteins fold 
into, and what functions they perform -- has not kept pace. Determining the 
atomic coordinates of a newly discovered protein may require as much as two 
graduate-student-years in a lab!

The problem of taking a newly found protein sequence about which only the 
sequence is known, and determining what existing proteins of known structure 
and function most closely resemble that new protein, is known as homology 
detection. When proteins of similar sequence can be found, this problem is 
largely solved. To solve the homology detection problem, computational 
biologists have developed fast, approximate methods for determining the 
structure and function of proteins based on their sequence. The most popular 
software for homology detection is called HMMER, which uses a hidden Markov 
model to capture evolutionary change.

However, when there are no known proteins of similar sequence, only 
evolutionarily distant proteins that may share similar structure, this problem 
is called remote homology detection. Hidden Markov model approaches begin to 
fail as sequence similarity falls off. Recently, Markov random field 
approaches, which capture non-local interactions in the protein structure, have 
shown to perform well. However, these approaches face the challenge of 
increased computational complexity; the SMURF program exhibits exponential time 
complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
simplifying the Markov random field's dependency graph. We have implemented a 
stochastic search approach to Markov random fields, called MRFy.

The algorithm underlying MRFy grew out of an understanding of the limitations 
of the SMURF (Structural Motifs Using Random Fields) algorithm, which used 
multidimensional dynamic programming to compute the optimal parse of a query 
sequence onto a Markov random field.

TODO: (add explanation of MRFy architecture and algorithm)

\subsection{Our Background}

All three authors are members of the Tufts University Department of Computer 
Science. Noah Daniels has taken a graduate seminar in advanced functional 
programming, which included some Haskell, and spent ten years in industry as a 
professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
has taken a programming language heavily focused on functional programming, but 
with no Haskell. Norman Ramsey is the local expert, but he wrote no code for 
this project.

\section{What we did}

\subsection{Beautiful code}
MRFy contains two distinct algorithms: the classical Viterbi dynamic 
programming algorithm for hidden Markov models, and a stochastic search 
algorithm for placing beta strand positions. This stochastic search algorithm 
implements several search heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations that seek to maximize the probability of an observed sequence (in 
this case, of amino acid ``letters'') begin emitted by a particular 
finite-state automaton (the hidden Markov model). The particular states of this 
finite-state automaton map to the recurrence relations, and in this instance 
derive from what is known in the computational biology community as a 
\textit{profile} hidden Markov model. The Viterbi algorithm in this instance 
recurses on ``match,'' ``insertion,'' and ``deletion'' states. To be efficient, 
the Viterbi algorithm relies on dynamic programming, or memoization, and the 
resulting asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence 
of distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.

We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
bottom-up dynamic programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model. The array containing the paths 
served as the memoization table. We next wrote a top-down dynamic programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both). In this second approach, we used the 
Data.Memocombinators library to provide the memoization container (again, a 
three-dimensional array). We determined that the run-time performances of both 
approaches were virtually indistinguishable, and we also noted that the 
top-down version quite faithfully resembled the recurrence relations found in a 
textbook~\cite{durbin} or the HMMER literature~\cite{eddy}. We chose to keep 
the top-down version, which later proved to be wise when we had to track down a 
bug in one of our base cases. This bug was observed as a missing state in the 
alignment our program displayed, and we quickly discovered that one of our base 
cases was missing this state: it was returning an empty list as the base of the 
path, rather than a list containing a single node. In this, we were grateful 
for the resemblance between the mathematical description of the algorithm and 
the top-down dynamic programming approach in Haskell, which resulted in 
perspicuous code.

breadcrumb: HOFs for search.

\subsection{Ugly code}

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

### 2012-02-07 mrfy_experience_report.tex
a few corrections

@@
 implements a novel algorithm for solving the remote homology detection problem. 
-During this process, we have wondered: would the algorithm be easily expressed 
-in Haskell? Would we be able to achieve acceptable runtime performance? How 
-would our experience compare with our past experiences developing and 
-maintaining computational biology software in C++, Ruby, and Python? What would 
-be our greatest challenges? In this paper, we briefly explain the remote 
-homology detection problem and our algorithm, and then attempt to answer the 
-above questions.
+While designing the algorithm, we wondered: would the algorithm be easily 
+expressed in Haskell? Would we be able to achieve acceptable runtime 
+performance? How would our experience compare with our past experiences 
+developing and maintaining computational biology software in C++, Ruby, and 
+Python? What would be our greatest challenges? In this paper, we briefly 
+explain the remote homology detection problem and our algorithm, and then 
+attempt to answer the above questions.
 
@@
 professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
-has taken a programming language heavily focused on functional programming, but 
-with no Haskell. Norman Ramsey is the local expert, but he wrote no code for 
-this project.
+has taken a programming languages course heavily focused on functional 
+programming, but with no Haskell. Norman Ramsey is the local expert, but he 
+wrote no code for this project.
 
@@
 relations that seek to maximize the probability of an observed sequence (in 
-this case, of amino acid ``letters'') begin emitted by a particular 
+this case, of amino acid ``letters'') being emitted by a particular 
 finite-state automaton (the hidden Markov model). The particular states of this 

### 2012-02-08 mrfy_experience_report.tex
"beautiful code" shitty draft done

@@
 
-breadcrumb: HOFs for search.
+Just as the Viterbi algorithm is one major component of MRFy, stochastic search 
+is the other. In designing MRFy, we did not know in advance what sort of 
+stochastic search technique would converge the fastest, or give us the best 
+alignments with the fastest performance. We decided to implement several 
+different search strategies so that we could experiment with all of them. We 
+found this to be easy and pleasant thanks to Haskell's support of higher order 
+functions.
+
+We wrote a basic search function in only twenty-five lines of code; as one of 
+its parameters we provide a SearchStrategy data type, which we use to define 
+specific search techniques. We require a SearchStrategy to define four 
+functions: mutate, accept, terminate, and initialize; we use the Haskell module 
+system to allow sharing of some of these functions between different search 
+strategies. For example, both \textit{random mutation hill climb} and 
+\textit{simulated annealing} use the same initialize and mutate functions, but 
+very different accept functions. In all, we implemented four search strategies: 
+\textit{random mutation hill climb}, \textit{simulated annealing}, 
+\textit{multistart simulated annealing}, and \textit{genetic algorithm}. We 
+found the implementation of the search function and search strategies to be 
+pleasant, and the resulting code to be simple and elegant.
+
+In MRFy, the Viterbi function is called several times (once for each 
+non-beta-strand region) for each search generation. Furthermore, in a 
+population-based search such as \textit{multistart simulated annealing}, the 
+scoring function (which itself calls Viterbi) is called many times per 
+generation. Beyond traditional profiling, we found parallelization to be an 
+obvious way to improve runtime performance. We found this to be trivial thanks 
+to the Control.Parallel library. We pass the Viterbi function to a map over all 
+non-beta-strand regions of the model and query sequence, and these regions are 
+independent of one another. Similarly, we pass the scoring function to a map 
+over the population of potential solutions, which are also independent. We 
+merely substituted ``parmap rseq'' for map. MRFy now efficiently uses all six 
+cores on an AMD workstation for a non-population-based simulated annealing 
+search, and all forty-eight cores on a compute server for population-based 
+searches.
 
 \subsection{Ugly code}
+
+- profiling
+-
 

### 2012-02-09 mrfy_experience_report.tex
tweak formatting

@@
 node or first observation (or both). In this second approach, we used the 
-Data.Memocombinators library to provide the memoization container (again, a 
+\texttt{Data.Memocombinators} library to provide the memoization container (again, a 
 three-dimensional array). We determined that the run-time performances of both 
@@
 We wrote a basic search function in only twenty-five lines of code; as one of 
-its parameters we provide a SearchStrategy data type, which we use to define 
-specific search techniques. We require a SearchStrategy to define four 
-functions: mutate, accept, terminate, and initialize; we use the Haskell module 
-system to allow sharing of some of these functions between different search 
-strategies. For example, both \textit{random mutation hill climb} and 
-\textit{simulated annealing} use the same initialize and mutate functions, but 
-very different accept functions. In all, we implemented four search strategies: 
-\textit{random mutation hill climb}, \textit{simulated annealing}, 
-\textit{multistart simulated annealing}, and \textit{genetic algorithm}. We 
-found the implementation of the search function and search strategies to be 
-pleasant, and the resulting code to be simple and elegant.
+its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
+define specific search techniques. We require a \texttt{SearchStrategy} to 
+define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
+and \texttt{initialize}; we use the Haskell module system to allow sharing of 
+some of these functions between different search strategies. For example, both 
+\textit{random mutation hill climb} and \textit{simulated annealing} use the 
+same initialize and mutate functions, but very different accept functions. In 
+all, we implemented four search strategies: \textit{random mutation hill 
+climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
+and \textit{genetic algorithm}. We found the implementation of the search 
+function and search strategies to be pleasant, and the resulting code to be 
+simple and elegant.
 
@@
 obvious way to improve runtime performance. We found this to be trivial thanks 
-to the Control.Parallel library. We pass the Viterbi function to a map over all 
+to the \texttt{Control.Parallel} library. We pass the Viterbi function to a map over all 
 non-beta-strand regions of the model and query sequence, and these regions are 
@@
 over the population of potential solutions, which are also independent. We 
-merely substituted ``parmap rseq'' for map. MRFy now efficiently uses all six 
+merely substituted \texttt{parmap rseq} for map. MRFy now efficiently uses all six 
 cores on an AMD workstation for a non-population-based simulated annealing 

### 2012-02-09 mrfy_experience_report.tex
start of 'ugly code'

@@
 
+Despite the beauty of some of the code we were able to write, we found several 
+instances where the easiest path seemed to be to write ugly code. The first of 
+these relates to the otherwise elegant Viterbi function. Our stochastic search 
+approach calls the Viterbi function a tremendous number of times, and most of 
+the program's runtime is spent in this function. However, the Viterbi algorithm 
+retains not just a score (the probability of the sequence given a state path) 
+but also the state path itself. We implemented this state path as a list of 
+\texttt{Int}, and every step in the recursion conses a new element onto the 
+list. However, at each step of the stochastic search, we discard the state 
+path; we are searching to optimize score only. We only need the state path at 
+the termination of the search, when we wish the program to output an alignment 
+of the query sequence to the Markov random field model. We found that 
+optimizing the Viterbi function by removing this state path entirely improved 
+runtime performance by nearly 50\%. However, we still needed the Viterbi 
+function to provide a state path when we called it after search finally 
+terminated. Thus, we wondered how to implement the Viterbi function so that a 
+state path was optional.
+
+In order to determine if this optimization was even worthwhile, we first wrote 
+a quick and dirty approach that is shameful in any language, but easy: we 
+duplicated code. We copied \texttt{viterbi} to produce a \texttt{viterbiF} 
+which did not retain state path. Next, however, we wondered how to avoid code 
+duplication. Clearly, littering the code with conditionals would produce code 
+that was ugly as well as less efficient. We knew how we would solve this 
+problem in other languages; we would have used macros in Lisp, or 
+metaprogramming via \texttt{eval} in Ruby, to produce two versions of the 
+function from one prototype. We struggled, however, to solve this problem 
+idiomatically in Haskell.
+
+breadcrumb: solution
+
+Profiling: only top-level cost centers can be added automatically
+ - had to pollute our code with cost center annotations
+ - perhaps a better way to do this?
+
+Awkwardness of trace and no backtraces led us to write ugly wrappers around Vector.slice and Vector.!
+
+\section{Barriers}
+
 - profiling
--
+- debugging
 

### 2012-02-10 mrfy_experience_report.tex
some notes added

@@
 
+[insert Viterbi recurrence relations]
+
 We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
@@
 
+[part of Viterbi code]
+
 Just as the Viterbi algorithm is one major component of MRFy, stochastic search 
@@
 
-We wrote a basic search function in only twenty-five lines of code; as one of 
+We wrote a basic search function in only twenty-two lines of code; as one of 
 its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
@@
 simple and elegant.
+
+[listing for search function, and constructor for SearchStrategy]
 

### 2012-02-14 mrfy_experience_report.tex
'ugly code' draft written. Need to revise intro.

@@
 \usepackage{amsmath}
+\usepackage{listings}
+\lstset{
+    tabsize = 2
+    }
 
@@
 [part of Viterbi code]
+\begin{figure*}
+\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
+\end{figure*}
 
@@
 
-breadcrumb: solution
+Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
+the Viterbi function, we wrote a higher-order function type \texttt{type 
+ScorePathCons a = a -> [a] -> [a]} and two functions conforming to that type. 
+Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
+or \texttt{consNoPath} as a parameter, the path would or would not be 
+allocated. We found this higher-order function approach to be simple in code 
+and in concept, and it imposes negligible run-time overhead as compared to 
+having two distinct functions.
 
-Profiling: only top-level cost centers can be added automatically
- - had to pollute our code with cost center annotations
- - perhaps a better way to do this?
+\lstinputlisting[label=scorepathcons,caption=ScorePathCons]{scorepathcons.hs}
 
-Awkwardness of trace and no backtraces led us to write ugly wrappers around Vector.slice and Vector.!
+Another source of ugliness resulted from adding cost-center annotations when we 
+wished to profile our code. GHC will happily add cost centers to top-level 
+functions, but much of our program relies on sub-functions. We found that 
+manually adding \texttt{\{-SCC -\}} annotations to dozens of guard clauses and 
+sub-functions harmed the readability of our program, such that we felt 
+compelled to remove all such annotations as soon as we could. In our prior 
+experience with profiling tools such as kcachegrind, simply compiling with 
+debug symbols -- rather than manually annotating many lines of code -- allowed 
+the profiling tools to provide us enough information. Without these 
+annotations, both call-site and allocation profiles are relatively useless in 
+GHC. Perhaps a compile-time flag that would auto-annotate every equation -- 
+even if the resulting cost center is simply named with a line number -- would 
+make this task easier.
+
+The final source of ugliness in our code came from attempts at debugging run-time errors.
+MRFy relies on a source of input data -- the paired beta strands in the Markov random field -- which is represented in an awkward manner by the original SMURF program, and which is outside our control.
+Rather than represent beta strands directly, SMURF represents pairings of residues, leaving us to infer and reconstruct beta strands.
+This task was difficult due to unclear invariants around overlapping, doubly-paired beta strands (and it would have been difficult in any programming language; the difficulty was one of representation).
+However, the debugging tools in GHC caused us some difficulties in determining the sources of run-time errors.
+We used \texttt{trace} extensively, but this style of ``\texttt{printf} debugging'' littered our code.
+Being initially unaware of the backtrace feature of the GHC profiler, we wrote wrappers around \texttt{Vector.slice} and \texttt{Vector.!} which themselves called trace, which at least kept our ``\texttt{printf} debugging'' less cluttered.
+Nonetheless, we struggled with these runtime errors due to an awkward representation, and wished for debugging tools such as we might find in gdb, which would allow us to examine the stack when a backtrace occurs.
 

### 2012-02-14 scorepathcons.hs
some additional files

@@ -1,0 +1,5 @@
+consPath :: ScorePathCons a
+consPath x xs = x:xs
+
+consNoPath :: ScorePathCons a
+consNoPath _ _ = []

### 2012-02-14 viterbi_simple.hs

@@ -1,0 +1,39 @@
+viterbi :: Alphabet -> QuerySequence -> HMM -> (Score, StatePath)
+viterbi alpha query hmm =
+    flipSnd $ DL.minimum $
+    [viterbi' mat (numNodes - 1) (seqlen - 1),
+     viterbi' ins (numNodes - 1) (seqlen - 1),
+     viterbi' del (numNodes - 1) (seqlen - 1)
+    ]
+  where viterbi' state node obs = Memo.memo3 (Memo.arrayRange (mat, end)) 
+                                  (Memo.arrayRange (0, numNodes))
+                                  (Memo.arrayRange (0, seqlen)) 
+                                  viterbi'' state node obs
+        viterbi'' s n o
+          -- consume an observation AND a node
+          | s == mat = let trans t prev = 
+                           (score + tProb hmm (n-1) t + eProb, mat:path)
+                            where (score, path) = viterbi' prev (n - 1) (o - 1)
+                                  eProb = eProb' (mEmissions $ hmm ! n) (res o)
+                        in minimum $ [
+                            transition m_m mat,
+                            transition i_m ins,
+                            transition d_m del
+                            ]
+          -- consume an observation but not a node
+          | s == ins = let trans t prev =
+                           (score + tProb hmm n t + eProb, ins:path)
+                            where (score, path) = viterbi' prev n (o - 1)
+                                  eProb = eProb' (iEmissions $ hmm ! n) (res o)
+                        in minimum [
+                            transition m_i mat,
+                            transition i_i ins
+                            ]
+          -- consume a node but not an observation
+          | s == del = let trans t prev = 
+                              (score + tProb hmm (n-1) t, del:path)
+                              where (score, path) = viterbi' prev (n - 1) o
+                        in minimum [
+                            transition m_d mat,
+                            transition d_d del
+                            ]

### 2012-02-15 mrfy_experience_report.tex
Lots of outlining, included files, math. Little new written content.

@@
 \usepackage{amsmath}
+\usepackage{array}
 \usepackage{listings}
 \lstset{
-    tabsize = 2
+    tabsize = 2,
+    basicstyle = \ttfamily,
+    language = Haskell
     }
@@
 
-[insert Viterbi recurrence relations]
+\begin{eqnarray}    
+V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
+\begin{array}{l l}
+V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
+V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
+V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
+\end{array} \right.\nonumber\\
+V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
+\begin{array}{l l}
+V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}I_{j}},\\
+V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}I_{j}},\\
+\end{array} \right.\nonumber\\
+V_{j}^{D}(i) = max \left\{
+\begin{array}{l l}
+V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
+V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
+\end{array} \right.\nonumber\\
+\end{eqnarray}\label{viterbi_eqn}
 
@@
 node or first observation (or both). In this second approach, we used the 
-\texttt{Data.Memocombinators} library to provide the memoization container (again, a 
-three-dimensional array). We determined that the run-time performances of both 
-approaches were virtually indistinguishable, and we also noted that the 
-top-down version quite faithfully resembled the recurrence relations found in a 
-textbook~\cite{durbin} or the HMMER literature~\cite{eddy}. We chose to keep 
-the top-down version, which later proved to be wise when we had to track down a 
-bug in one of our base cases. This bug was observed as a missing state in the 
-alignment our program displayed, and we quickly discovered that one of our base 
-cases was missing this state: it was returning an empty list as the base of the 
-path, rather than a list containing a single node. In this, we were grateful 
-for the resemblance between the mathematical description of the algorithm and 
-the top-down dynamic programming approach in Haskell, which resulted in 
-perspicuous code.
+\texttt{Data.Memocombinators} library to provide the memoization container 
+(again, a three-dimensional array). We determined that the run-time 
+performances of both approaches were virtually indistinguishable, and we also 
+noted that the top-down version quite faithfully resembled the recurrence 
+relations found in a textbook~\cite{durbin} or the HMMER literature~\cite{eddy} 
+(see Equation \ref{viterbi_eqn}). We chose to keep the top-down version, which 
+later proved to be wise when we had to track down a bug in one of our base 
+cases. This bug was observed as a missing state in the alignment our program 
+displayed, and we quickly discovered that one of our base cases was missing 
+this state: it was returning an empty list as the base of the path, rather than 
+a list containing a single node. In this, we were grateful for the resemblance 
+between the mathematical description of the algorithm and the top-down dynamic 
+programming approach in Haskell, which resulted in perspicuous code.
 
-[part of Viterbi code]
 \begin{figure*}
@@
 function and search strategies to be pleasant, and the resulting code to be 
-simple and elegant.
+simple and elegant. In an imperative language with object orientation like Ruby 
+or Python, we likely would have used a class hierarchy to accomplish this. We 
+feel that this would have required more code, because object orientation is 
+meant for encapsulating data as well as functions on those data; here we only 
+wanted to encapsulate functions, as the data did not change.
 
-[listing for search function, and constructor for SearchStrategy]
+\begin{figure*}
+\lstinputlisting[label=search,caption=Search]{search_simple.hs}
+\end{figure*}
 
@@
 Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
-the Viterbi function, we wrote a higher-order function type \texttt{type 
-ScorePathCons a = a -> [a] -> [a]} and two functions conforming to that type. 
+the Viterbi function, we wrote a higher-order function type 
+\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type. 
 Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
@@
 
-The final source of ugliness in our code came from attempts at debugging run-time errors.
-MRFy relies on a source of input data -- the paired beta strands in the Markov random field -- which is represented in an awkward manner by the original SMURF program, and which is outside our control.
-Rather than represent beta strands directly, SMURF represents pairings of residues, leaving us to infer and reconstruct beta strands.
-This task was difficult due to unclear invariants around overlapping, doubly-paired beta strands (and it would have been difficult in any programming language; the difficulty was one of representation).
-However, the debugging tools in GHC caused us some difficulties in determining the sources of run-time errors.
-We used \texttt{trace} extensively, but this style of ``\texttt{printf} debugging'' littered our code.
-Being initially unaware of the backtrace feature of the GHC profiler, we wrote wrappers around \texttt{Vector.slice} and \texttt{Vector.!} which themselves called trace, which at least kept our ``\texttt{printf} debugging'' less cluttered.
-Nonetheless, we struggled with these runtime errors due to an awkward representation, and wished for debugging tools such as we might find in gdb, which would allow us to examine the stack when a backtrace occurs.
+The final source of ugliness in our code came from attempts at debugging 
+run-time errors. MRFy relies on a source of input data -- the paired beta 
+strands in the Markov random field -- which is represented in an awkward manner 
+by the original SMURF program, and which is outside our control. Rather than 
+represent beta strands directly, SMURF represents pairings of residues, leaving 
+us to infer and reconstruct beta strands. This task was difficult due to 
+unclear invariants around overlapping, doubly-paired beta strands (and it would 
+have been difficult in any programming language; the difficulty was one of 
+representation). However, the debugging tools in GHC caused us some 
+difficulties in determining the sources of run-time errors. We used 
+\texttt{trace} extensively, but this style of ``\texttt{printf} debugging'' 
+littered our code. Being initially unaware of the backtrace feature of the GHC 
+profiler, we wrote wrappers around \texttt{Vector.slice} and \texttt{Vector.!} 
+which themselves called trace, which at least kept our ``\texttt{printf} 
+debugging'' less cluttered. Nonetheless, we struggled with these runtime errors 
+due to an awkward representation, and wished for debugging tools such as we 
+might find in gdb, which would allow us to examine the stack when a backtrace 
+occurs.
 
-\section{Barriers}
+\section{Our previous code base compared}
+
+ - SMURF bad
+ - HMMER good
+
+\section{State of the practice}
+
+\subsection{Our group then}
+ - Modifying Formatt took nearly a year
+ - mutable data structures hard to repurpose
+ - one enhancement deemed infeasible
+ - C++ failures most commonly segfaults, memory errors, heap exhaustion
+
+\subsection{Our group now}
+ - Higher order functions militate towards more flexibility
+ - Haskell failures are usually typecheck failures; at worst runtime errors around bounds checking
+ - MRFy about 3 months of part-time work
+
+\subsection{Computational biology at large}
+ - some good software, like HMMER
+
+\section{What we think about it}
+
+\subsection{Barriers}
 
@@
 - debugging
+ [ talk about profiling and debugging problems where?]
+[ something about quickcheck: how do we know when we have enough QC props? How do we start testing a complex module? ]
+
+\subsection{What you must know to succeed}
+- Ecosystem is loaded with tools that look promising but either are no longer maintained or are not ready for prime time (Pads, Backtrace)
+
+\subsection{Evaluation}
+- What does the local expert do? Do you have to have one?
+
+- To the FP community: Those who want us to abandon our old tools for debugging imperative code have not provided a clear path to follow (both intellectually, and via software tools).
+
+- No obvious body of knowledge on code improvement.
+
+- The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies. How to tell what's good? How did we discover what's good?
+- Profiler can produce backtraces, but this was hard to discover.
 

### 2012-02-15 search_simple.hs
simplified search code

@@ -1,0 +1,40 @@
+module StochasticSearch where
+
+type Scorer = QuerySequence -> [BetaStrand] -> SearchGuess -> SearchSolution
+data SearchStrategy =
+     SearchStrategy { accept :: SearchParameters -> Seed -> History ->
+                       Age -> Bool
+                    , terminate :: SearchParameters -> History -> Age -> Bool
+                    , mutate :: SearchParameters -> Seed -> QuerySequence ->
+                       Scorer -> [BetaStrand] -> [SearchSolution] ->
+                       [SearchSolution]
+                    , initialize :: SearchParameters -> Seed ->
+                       QuerySequence -> [BetaStrand]-> [SearchGuess]
+                    }
+
+search :: QuerySequence -> HMM -> [BetaStrand] -> SearchParameters ->
+          [Seed] -> (SearchSolution, History)
+search query hmm betas searchP seeds =
+        search' (tail seeds) initialGuessScore [] 0
+  where initialGuessScore = map (score hmm query betas) initialGuess
+        initialGuess = initialize strat searchP (head seeds) query betas
+        strat = strategy searchP
+        search' :: [Seed] -> [SearchSolution] -> History -> Age ->
+                   (SearchSolution, History)
+        search' (s1:s2:seeds) oldPop hist age =
+          let newPop = mutate' oldPop
+              score = fst $ minimum newPop
+              newHist = score : hist
+            in if accept' newHist age then
+                  if terminate' newHist age then
+                    (minimum newPop, newHist)
+                  else
+                    search' seeds newPop newHist (age + 1)
+               else
+                 if terminate' hist age then
+                   (minimum oldPop, hist)
+                 else
+                   search' seeds oldPop hist (age + 1)
+            where mutate' = mutate strat searchP s1 query (score hmm) betas
+                  terminate' = terminate strat searchP
+                  accept' = accept strat searchP s2

### 2012-02-15 viterbi_simple.hs
simple viterbi code

@@ -1,10 +1,9 @@
 viterbi :: Alphabet -> QuerySequence -> HMM -> (Score, StatePath)
-viterbi alpha query hmm =
-    flipSnd $ DL.minimum $
-    [viterbi' mat (numNodes - 1) (seqlen - 1),
-     viterbi' ins (numNodes - 1) (seqlen - 1),
-     viterbi' del (numNodes - 1) (seqlen - 1)
-    ]
+viterbi alpha query hmm = flipSnd $ DL.minimum $
+                        [viterbi' mat (numNodes - 1) (seqlen - 1),
+                         viterbi' ins (numNodes - 1) (seqlen - 1),
+                         viterbi' del (numNodes - 1) (seqlen - 1)
+                        ]
   where viterbi' state node obs = Memo.memo3 (Memo.arrayRange (mat, end)) 
                                   (Memo.arrayRange (0, numNodes))
                                   (Memo.arrayRange (0, seqlen)) 

### 2012-02-15 mrfy_experience_report.tex
added diagram, MRFy architecture explanation. Some revisions to introduction.

@@
 \usepackage{listings}
+\usepackage{graphicx}
 \lstset{
@@
 
-Computational biologists write software to solve problems on large genomic or 
-proteomic data sets. Traditionally, these software programs fall into two 
-categories: those with high-performance computing needs are typically 
-implemented in C++, while the remainder are typically implemented in Perl, 
-Python, Ruby, or R. Recently, we have developed a program in Haskell which 
-implements a novel algorithm for solving the remote homology detection problem. 
-While designing the algorithm, we wondered: would the algorithm be easily 
-expressed in Haskell? Would we be able to achieve acceptable runtime 
-performance? How would our experience compare with our past experiences 
-developing and maintaining computational biology software in C++, Ruby, and 
-Python? What would be our greatest challenges? In this paper, we briefly 
-explain the remote homology detection problem and our algorithm, and then 
-attempt to answer the above questions.
+Computational biologists write software that answers questions about large data 
+sets involving sequences of nucleic acids (genomic data) or sequences of amino 
+acids (proteomic data). This software comes in two varieties. The first variety 
+is \textit{high performance software}, where performance is a top priority; 
+this kind of software is usually written in a programming language like C or 
+C++. The second variety is software where performance is not a top priority; 
+this kind of software written in higher-level languages that emphasize 
+convenience and programmer productivity, such as Perl, Python, Ruby, or R.
+
+In this Experience Report, we show that the Haskell programming language is an 
+excellent alternative for computational biologists: a Haskell program can 
+perform comparably to a C++ program while being as flexible and easy to 
+understand as a program written in an interpreted language. Our experience is 
+based on an implementation of a novel algorithm for finding remote homologs to 
+newly discovered proteins. We call this program MRFy.
 
@@
 cellular machinery, performing the varied functions an organism needs to 
-survive. Since the dawn of the human genome project in 1990, the cost of 
-sequencing an entire genome has dropped exponentially. Recently, the doubling 
-time for genomic data has been one year, in contrast to the eighteen month 
-doubling time of Moore's law. However, determining what newly sequenced genes 
-actually do -- when they become proteins, what structures those proteins fold 
-into, and what functions they perform -- has not kept pace. Determining the 
-atomic coordinates of a newly discovered protein may require as much as two 
-graduate-student-years in a lab!
+survive.
+
+However, determining what newly sequenced genes actually do -- when they become 
+proteins, what structures those proteins fold into, and what functions they 
+perform -- has not kept pace. Determining the atomic coordinates of a newly 
+discovered protein may require as much as two graduate-student-years in a lab! 
+Fortunately, determining the full structure of a protein is not always 
+necessary. Often, we can identify well-known proteins of similar sequence or 
+structure to a new protein, and thus make educated guesses as to the structure 
+and function of the new protein.
 
@@
 approaches, which capture non-local interactions in the protein structure, have 
-shown to perform well. However, these approaches face the challenge of 
+been shown to perform well. However, these approaches face the challenge of 
 increased computational complexity; the SMURF program exhibits exponential time 
 complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
-simplifying the Markov random field's dependency graph. We have implemented a 
-stochastic search approach to Markov random fields, called MRFy.
+simplifying the Markov random field's dependency graph. MRFy is a 
+stochastic search approach to Markov random fields.
 
-The algorithm underlying MRFy grew out of an understanding of the limitations 
-of the SMURF (Structural Motifs Using Random Fields) algorithm, which used 
-multidimensional dynamic programming to compute the optimal parse of a query 
-sequence onto a Markov random field.
+SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
+alignment, of a protein sequence onto the directed graph representing a Markov 
+random field (see Figure \ref{mrf}). The algorithm underlying MRFy grew out of 
+an understanding of the limitations of the SMURF (Structural Motifs Using 
+Random Fields) algorithm, which used multidimensional dynamic programming to 
+compute the optimal parse of a query sequence onto a Markov random field. 
+Rather than attempt to compute exactly the optimal parse, MRFy treats the 
+non-local interactions (which correspond to hydrogen-bonded ``beta strands'' in 
+the protein) as ``beads'' which can be slid along a string. In this analogy, 
+the string is the sequence of amino acids of the query protein. A given 
+placement of these beads corresponds to an assignment of particular amino acid 
+residues being hydrogen bonded to other particular amino acid residues, and 
+thus MRFy computes a probability based on observed frequencies of residues 
+hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
+strands, the Markov random field is in fact a traditional hidden Markov model, 
+whose optimal solution can be easily computed via dynamic programming (known as 
+the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
+strands, based on an objective function that combines the probability computed 
+by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
+residues in the beta strands.
 
-TODO: (add explanation of MRFy architecture and algorithm)
+\begin{figure}
+\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}}
+\caption{The directed graph representing the states of a MRFy Markov random field finite state
+automaton}\label{mrf}
+\end{figure}
 

### 2012-02-27 mrfy_experience_report.tex
a few code example improvements. Added transform of Viterbi relations in negative log space.

@@
 
-\conferenceinfo{ICFP '12}{date, City.} 
+\conferenceinfo{ICFP '12}{September 9-15, Copenhagen.} 
 \copyrightyear{2012} 
@@
 \title{Experience Report: Haskell in Computational Biology}
-\subtitle{Subtitle Text, if any}
+% \subtitle{Subtitle Text, if any}
 
@@
 \centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
-\caption{The directed graph representing the states of a MRFy Markov random field finite state
-automaton}\label{mrf}
-\end{figure}
+\caption{The directed graph representing the states of a MRFy Markov random 
+field finite state automaton. Shaded nodes represent beta-strand positions, 
+while the dashed edges represent the long-range dependencies due to hydrogen 
+bonding.}\label{mrf} \end{figure}
 
@@
 The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
-relations that seek to maximize the probability of an observed sequence (in 
-this case, of amino acid ``letters'') being emitted by a particular 
-finite-state automaton (the hidden Markov model). The particular states of this 
-finite-state automaton map to the recurrence relations, and in this instance 
-derive from what is known in the computational biology community as a 
-\textit{profile} hidden Markov model. The Viterbi algorithm in this instance 
-recurses on ``match,'' ``insertion,'' and ``deletion'' states. To be efficient, 
-the Viterbi algorithm relies on dynamic programming, or memoization, and the 
-resulting asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence 
-of distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
-letters in the protein sequence. In our particular problem domain, both $M$ and 
-$N$ routinely contain several hundred to a few thousand elements.
+relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
+probability of an observed sequence (in this case, of amino acid ``letters'') 
+being emitted by a particular finite-state automaton (the hidden Markov model). 
+The particular states of this finite-state automaton map to the recurrence 
+relations, and in this instance derive from what is known in the computational 
+biology community as a \textit{profile} hidden Markov model. The Viterbi 
+algorithm in this instance recurses on ``match,'' ``insertion,'' and 
+``deletion'' states. To be efficient, the Viterbi algorithm relies on dynamic 
+programming, or memoization, and the resulting asymptotic complexity is 
+$O(|M|\times|N|)$, when $M$ is the sequence of distinct states (nodes in the 
+automaton's graph) and $N$ is the sequence of letters in the protein sequence. 
+In our particular problem domain, both $M$ and $N$ routinely contain several 
+hundred to a few thousand elements.
 
@@
 \end{eqnarray}\label{viterbi_eqn}
+
+\begin{eqnarray}    
+V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
+\begin{array}{l l}
+V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
+V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
+V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
+\end{array} \right.\nonumber\\
+V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
+\begin{array}{l l}
+V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}I_{j}},\\
+V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}I_{j}},\\
+\end{array} \right.\nonumber\\
+V_{j}^{\prime D}(i) = min \left\{
+\begin{array}{l l}
+V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
+V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
+\end{array} \right.\nonumber\\
+\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
+\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
+\end{eqnarray}\label{viterbi_log_eqn}
 
@@
 
- - SMURF bad
- - HMMER good
+Haskell makes it difficult to write code in haste, because Haskell militates towards careful type-checking and clearly thought-out function contracts.
+Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the programmer who attempts to write code quickly with the goal of later making it work correctly.
+We have found MRFy to be easier to enhance and maintain than our existing C++ code bases.
 
@@
 
-\appendix
-\section{Appendix Title}
-
-This is the text of the appendix, if you need one.
+% \appendix
+% \section{Appendix Title}
+% 
+% This is the text of the appendix, if you need one.
 
@@
 
-Acknowledgments, if needed.
+We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford Larsen, and Nathan Ricci.
 

### 2012-02-27 search_simple.hs

@@ -1,21 +1,15 @@
 module StochasticSearch where
-
 type Scorer = QuerySequence -> [BetaStrand] -> SearchGuess -> SearchSolution
-data SearchStrategy =
-     SearchStrategy { accept :: SearchParameters -> Seed -> History ->
-                       Age -> Bool
-                    , terminate :: SearchParameters -> History -> Age -> Bool
-                    , mutate :: SearchParameters -> Seed -> QuerySequence ->
-                       Scorer -> [BetaStrand] -> [SearchSolution] ->
-                       [SearchSolution]
-                    , initialize :: SearchParameters -> Seed ->
-                       QuerySequence -> [BetaStrand]-> [SearchGuess]
-                    }
-
-search :: QuerySequence -> HMM -> [BetaStrand] -> SearchParameters ->
-          [Seed] -> (SearchSolution, History)
-search query hmm betas searchP seeds =
-        search' (tail seeds) initialGuessScore [] 0
+data SearchStrategy = SearchStrategy { accept :: SearchParameters -> Seed -> History -> Age -> Bool
+                                     , terminate :: SearchParameters -> History -> Age -> Bool
+                                     , mutate :: SearchParameters -> Seed -> QuerySequence ->
+                                        Scorer -> [BetaStrand] -> [SearchSolution] ->
+                                        [SearchSolution]
+                                     , initialize :: SearchParameters -> Seed ->
+                                        QuerySequence -> [BetaStrand]-> [SearchGuess]
+                                     }
+search :: QuerySequence -> HMM -> [BetaStrand] -> SearchParameters -> [Seed] -> (SearchSolution, History)
+search query hmm betas searchP seeds = search' (tail seeds) initialGuessScore [] 0
   where initialGuessScore = map (score hmm query betas) initialGuess
         initialGuess = initialize strat searchP (head seeds) query betas
         strat = strategy searchP

### 2012-02-27 viterbi_simple.hs

@@ -1,3 +1,22 @@
+type HMM = V.Vector HmmNode
+type EmissionProbabilities = V.Vector Double
+data HmmNode = 
+     HmmNode { nodeNum :: Int
+             , matchEmissions :: EmissionProbabilities
+             , annotations :: Maybe EmissionAnnotationSet
+             , insertionEmissions :: EmissionProbabilities
+             , transitions :: TransitionProbabilities
+             }
+data TransitionProbabilities = 
+     TransitionProbabilities { m_m :: TransitionProbability
+                             , m_i :: TransitionProbability
+                             , m_d :: TransitionProbability
+                             , i_m :: TransitionProbability
+                             , i_i :: TransitionProbability
+                             , d_m :: TransitionProbability
+                             , d_d :: TransitionProbability
+                             }
+
 viterbi :: Alphabet -> QuerySequence -> HMM -> (Score, StatePath)
 viterbi alpha query hmm = flipSnd $ DL.minimum $
                         [viterbi' mat (numNodes - 1) (seqlen - 1),
@@ -15,9 +34,9 @@ viterbi alpha query hmm = flipSnd $ DL.minimum $
                             where (score, path) = viterbi' prev (n - 1) (o - 1)
                                   eProb = eProb' (mEmissions $ hmm ! n) (res o)
                         in minimum $ [
-                            transition m_m mat,
-                            transition i_m ins,
-                            transition d_m del
+                            trans m_m mat,
+                            trans i_m ins,
+                            trans d_m del
                             ]
           -- consume an observation but not a node
           | s == ins = let trans t prev =
@@ -25,14 +44,26 @@ viterbi alpha query hmm = flipSnd $ DL.minimum $
                             where (score, path) = viterbi' prev n (o - 1)
                                   eProb = eProb' (iEmissions $ hmm ! n) (res o)
                         in minimum [
-                            transition m_i mat,
-                            transition i_i ins
+                            trans m_i mat,
+                            trans i_i ins
                             ]
           -- consume a node but not an observation
           | s == del = let trans t prev = 
                               (score + tProb hmm (n-1) t, del:path)
                               where (score, path) = viterbi' prev (n - 1) o
                         in minimum [
-                            transition m_d mat,
-                            transition d_d del
+                            trans m_d mat,
+                            trans d_d del
                             ]
+                            
+        numNodes = Data.Vector.length hmm
+        seqlen = Data.Vector.length query
+        
[…]

### 2012-02-28 viterbi_simple.hs
Break out viterbi types.
Add some notes on connecting the types to the concepts.

@@ -1,22 +1,3 @@
-type HMM = V.Vector HmmNode
-type EmissionProbabilities = V.Vector Double
-data HmmNode = 
-     HmmNode { nodeNum :: Int
-             , matchEmissions :: EmissionProbabilities
-             , annotations :: Maybe EmissionAnnotationSet
-             , insertionEmissions :: EmissionProbabilities
-             , transitions :: TransitionProbabilities
-             }
-data TransitionProbabilities = 
-     TransitionProbabilities { m_m :: TransitionProbability
-                             , m_i :: TransitionProbability
-                             , m_d :: TransitionProbability
-                             , i_m :: TransitionProbability
-                             , i_i :: TransitionProbability
-                             , d_m :: TransitionProbability
-                             , d_d :: TransitionProbability
-                             }
-
 viterbi :: Alphabet -> QuerySequence -> HMM -> (Score, StatePath)
 viterbi alpha query hmm = flipSnd $ DL.minimum $
                         [viterbi' mat (numNodes - 1) (seqlen - 1),
@@ -55,15 +36,15 @@ viterbi alpha query hmm = flipSnd $ DL.minimum $
                             trans m_d mat,
                             trans d_d del
                             ]
-                            
+
         numNodes = Data.Vector.length hmm
         seqlen = Data.Vector.length query
-        
+
         eProb' :: Vector a -> Int -> a
         eProb' emissions residue = emissions ! residue
 
         tProb :: HMM -> Int -> StateAcc -> Double
-        tProb hmm nodenum state = case logProbability $ state (transitions (hmm ! nodenum)) of
-                                           NonZero p -> p
-                                           LogZero -> maxProb
-        
+        tProb hmm nodenum state =
+            case logProbability $ state (transitions (hmm ! nodenum)) of
+                   NonZero p -> p
+                   LogZero -> maxProb

### 2012-02-28 viterbi_types.hs

@@ -1,0 +1,27 @@
+type HMMState = Int
+type StatePath = [ HMMState ]
+mat = 0 :: HMMState
+ins = 1 :: HMMState
+del = 2 :: HMMState
+type EmissionProbabilities = V.Vector Double
+data TransitionProbability (fState::HMMState, tState::HMMState) = 
+     TransitionProbability { logProbability::LogProbability
+                           , fromState = value fState::HMMState
+                           , toState = value tState::HMMState
+                           }
+data TransitionProbabilities =
+     TransitionProbabilities { m_m :: TransitionProbability
+                             , m_i :: TransitionProbability
+                             , m_d :: TransitionProbability
+                             , i_m :: TransitionProbability
+                             , i_i :: TransitionProbability
+                             , d_m :: TransitionProbability
+                             , d_d :: TransitionProbability
+                             }
+data HmmNode =
+     HmmNode { nodeNum :: Int
+             , matchEmissions :: EmissionProbabilities
+             , insertionEmissions :: EmissionProbabilities
+             , transitions :: TransitionProbabilities
+             }
+type HMM = V.Vector HmmNode

### 2012-02-28 mrfy_experience_report.tex
Break out viterbi types.
Add some notes on connecting the types to the concepts.
Another figure and further explanations. We'll need to shrink some code.

@@
 alignment, of a protein sequence onto the directed graph representing a Markov 
-random field (see Figure \ref{mrf}). The algorithm underlying MRFy grew out of 
-an understanding of the limitations of the SMURF (Structural Motifs Using 
-Random Fields) algorithm, which used multidimensional dynamic programming to 
-compute the optimal parse of a query sequence onto a Markov random field. 
-Rather than attempt to compute exactly the optimal parse, MRFy treats the 
-non-local interactions (which correspond to hydrogen-bonded ``beta strands'' in 
-the protein) as ``beads'' which can be slid along a string. In this analogy, 
-the string is the sequence of amino acids of the query protein. A given 
-placement of these beads corresponds to an assignment of particular amino acid 
-residues being hydrogen bonded to other particular amino acid residues, and 
-thus MRFy computes a probability based on observed frequencies of residues 
+random field (see Figure \ref{mrf}). A parse is a mapping of the letters of a 
+query sequence onto the states of a markov Random field, while an optimal parse 
+is the parse exhibiting the greatest probability once. The algorithm underlying 
+MRFy grew out of an understanding of the limitations of the SMURF (Structural 
+Motifs Using Random Fields) algorithm, which used multidimensional dynamic 
+programming to compute the optimal parse of a query sequence onto a Markov 
+random field. Rather than attempt to compute exactly the optimal parse, MRFy 
+treats the non-local interactions (which correspond to hydrogen-bonded ``beta 
+strands'' in the protein) as ``beads'' which can be slid along a string. In 
+this analogy, the string is the sequence of amino acids of the query protein. A 
+given placement of these beads corresponds to an assignment of particular amino 
+acid residues being hydrogen bonded to other particular amino acid residues, 
+and thus MRFy computes a probability based on observed frequencies of residues 
 hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
@@
 \subsection{Beautiful code}
-MRFy contains two distinct algorithms: the classical Viterbi dynamic 
-programming algorithm for hidden Markov models, and a stochastic search 
-algorithm for placing beta strand positions. This stochastic search algorithm 
+MRFy contains two distinct algorithms: the classical Viterbi 
+dynamic-programming algorithm for hidden Markov models, and a stochastic search 
+algorithm for placing beta-strand positions. This stochastic search algorithm 
 implements several search heuristics: random-mutation hill climbing, simulated 
@@
 biology community as a \textit{profile} hidden Markov model. The Viterbi 
-algorithm in this instance recurses on ``match,'' ``insertion,'' and 
-``deletion'' states. To be efficient, the Viterbi algorithm relies on dynamic 
-programming, or memoization, and the resulting asymptotic complexity is 
-$O(|M|\times|N|)$, when $M$ is the sequence of distinct states (nodes in the 
-automaton's graph) and $N$ is the sequence of letters in the protein sequence. 
-In our particular problem domain, both $M$ and $N$ routinely contain several 
-hundred to a few thousand elements.
+algorithm in this instance recurses on ``insertion,'' ``deletion,'' and 
+``match'' states (Figure~\ref{plan7}). Insertion states represent mutations 
+that gain an amino acid letter in the query sequence, deletion states represent 
+mutations that lose an amino acid letter in the query sequence, and match 
+states represent either conserved positions (no change) or mutations that 
+substitute one amino acid for another. The match and insertion states each have 
+an associated table of emission probabilities, determined during training, 
+indicating how likely that state is to emit a particular amino acid letter. 
+Edges from state to state represent state transitions, and have associated 
+probabilities determined during training. To be efficient, the Viterbi 
+algorithm relies on dynamic programming, or memoization, and the resulting 
+asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
+distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
+letters in the protein sequence. In our particular problem domain, both $M$ and 
+$N$ routinely contain several hundred to a few thousand elements. The types of 
+these states and probabilities are provided in Figure~\ref{viterbi_types}.
+
+\begin{figure*}
+\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
+\end{figure*}
+
+\begin{figure} \centerline{\includegraphics[width=8cm]{Plan7.pdf}} \caption{The 
+directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
+Markov model. Only 7 of the possible 9 state transitions are allowed, which 
+simplifies the dynamic programming. Insertion states represent mutations that 
+insert an amino acid, deletion states represent mutations that lose an amino 
+acid, and match states represent either conserved amino acids (those with no 
+changes) or mutations that substitute one amino acid for another. Each match or 
+insertion state has an associated table of emission probabilities (here 
+illustrated only for the first two states)}\label{plan7} \end{figure}
+
+The textbook Viterbi recurrence relations refer to logs of probabilities, so 
+that the terms may be added rather than multiplying raw probabilities. For 
+reasons including floating point underflow, the HMMER software (with which we 
+maintain file format compatibility) stores all probabilities in a trained HMM 
+file as negative natural logs. Thus, the Viterbi recurrence relations are 
+simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
+and because they are \textit{negative} logs, the problem transforms from 
+maximization to minimization.
 
@@
 
+\begin{figure*}
+\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
+\end{figure*}
+
 We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
-bottom-up dynamic programming version, which began with the first node of the 
+bottom-up dynamic-programming version, which began with the first node of the 
 hidden Markov model and first observation and built up the three-dimensional 
 array of possible paths through the model. The array containing the paths 
-served as the memoization table. We next wrote a top-down dynamic programming 
+served as the memoization table. We next wrote a top-down dynamic-programming 
 version, which began with the last node of the model and the last observation, 
@@
 a list containing a single node. In this, we were grateful for the resemblance 
-between the mathematical description of the algorithm and the top-down dynamic 
-programming approach in Haskell, which resulted in perspicuous code.
-
-\begin{figure*}
-\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
-\end{figure*}
+between the mathematical description of the algorithm and the top-down 
+dynamic-programming approach in Haskell, which resulted in perspicuous code.
 
@@
 wanted to encapsulate functions, as the data did not change.
+
+We note that \texttt{seed} in \texttt{search} comes from a monadic random 
+number generator in Main, but we wished for search itself to be pure at this 
+level. One advantage is that this allows for testing of the \texttt{search} 
+function with no randomness at all: if we pass in a known number as the seed, 
+behavior is entirely deterministic.
 

### 2012-03-02 mrfy_experience_report.tex
clean up types, rewrite intro, tweak diagrams, add alignment diagram

@@
 
-As a brief refresher in the standard dogma of genetics, recall that genes 
-(strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
-adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
-and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
-into the 20 naturally-occuring amino acids, which form peptide chains -- 
-proteins -- which fold into complex structures in the cell. These proteins are 
-cellular machinery, performing the varied functions an organism needs to 
-survive.
+Proteins are cellular machinery. They interact with one another and with other 
+molecules to carry out the functions of living cells: metabolism, regulation, 
+signalling, and more. The function of a protein is determined by its structure, 
+and this structure in turn is determined by the specific sequence of the twenty 
+different amino acids that make up the protein, but proteins with significantly 
+different amino acid sequences may share similar structure and function. This 
+amino acid sequence is ultimately determined by the DNA sequence we call a 
+gene. Given the genetic sequence of a protein, biologists wish to know its 
+specific cellular function. The best path to this is to find other proteins of 
+similar structure, which likely share similar function. This problem is called 
+\textit{homology detection}.
 
-However, determining what newly sequenced genes actually do -- when they become 
-proteins, what structures those proteins fold into, and what functions they 
-perform -- has not kept pace. Determining the atomic coordinates of a newly 
-discovered protein may require as much as two graduate-student-years in a lab! 
-Fortunately, determining the full structure of a protein is not always 
-necessary. Often, we can identify well-known proteins of similar sequence or 
-structure to a new protein, and thus make educated guesses as to the structure 
-and function of the new protein.
+Computational biologists address the homology detection problem by building 
+algorithms to find similar proteins given an amino acid sequence. When there 
+are nearly-identical sequences for proteins of known structure, this is a 
+simple string-matching problem. When there are no identical sequences, but 
+still some reasonably similar sequences, one family of algorithms called 
+\textit{hidden Markov models} excels at homology detection. We will explain an 
+algorithm for detecting reasonably similar sequences, and then move on to 
+explain MRFy, our novel approach for detecting more distantly related sequences 
+for proteins that share similar structure and function.
 
-The problem of taking a newly found protein sequence about which only the 
-sequence is known, and determining what existing proteins of known structure 
-and function most closely resemble that new protein, is known as homology 
-detection. When proteins of similar sequence can be found, this problem is 
-largely solved. To solve the homology detection problem, computational 
-biologists have developed fast, approximate methods for determining the 
-structure and function of proteins based on their sequence. The most popular 
-software for homology detection is called HMMER, which uses a hidden Markov 
-model to capture evolutionary change.
+% As a brief refresher in the standard dogma of genetics, recall that genes 
+% (strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
+% adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
+% and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
+% into the 20 naturally-occuring amino acids, which form peptide chains -- 
+% proteins -- which fold into complex structures in the cell. These proteins are 
+% cellular machinery, performing the varied functions an organism needs to 
+% survive.
 
-However, when there are no known proteins of similar sequence, only 
-evolutionarily distant proteins that may share similar structure, this problem 
-is called remote homology detection. Hidden Markov model approaches begin to 
-fail as sequence similarity falls off. Recently, Markov random field 
-approaches, which capture non-local interactions in the protein structure, have 
-been shown to perform well. However, these approaches face the challenge of 
-increased computational complexity; the SMURF program exhibits exponential time 
-complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
-simplifying the Markov random field's dependency graph. MRFy is a 
-stochastic search approach to Markov random fields.
+% However, determining what newly sequenced genes actually do -- when they become 
+% proteins, what structures those proteins fold into, and what functions they 
+% perform -- has not kept pace. Determining the atomic coordinates of a newly 
+% discovered protein may require as much as two graduate-student-years in a lab! 
+% Fortunately, determining the full structure of a protein is not always 
+% necessary. Often, we can identify well-known proteins of similar sequence or 
+% structure to a new protein, and thus make educated guesses as to the structure 
+% and function of the new protein.
+% 
+% The problem of taking a newly found protein sequence about which only the 
+% sequence is known, and determining what existing proteins of known structure 
+% and function most closely resemble that new protein, is known as homology 
+% detection. When proteins of similar sequence can be found, this problem is 
+% largely solved. To solve the homology detection problem, computational 
+% biologists have developed fast, approximate methods for determining the 
+% structure and function of proteins based on their sequence. The most popular 
+% software for homology detection is called HMMER, which uses a hidden Markov 
+% model to capture evolutionary change.
+
+% However, when there are no known proteins of similar sequence, only 
+% evolutionarily distant proteins that may share similar structure, this problem 
+% is called remote homology detection. Hidden Markov model approaches begin to 
+% fail as sequence similarity falls off. Recently, Markov random field 
+% approaches, which capture non-local interactions in the protein structure, have 
+% been shown to perform well. However, these approaches face the challenge of 
+% increased computational complexity; the SMURF program exhibits exponential time 
+% complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
+% simplifying the Markov random field's dependency graph. MRFy is a 
+% stochastic search approach to Markov random fields.
+
+\subsection{Our Background}
+
+All three authors are members of the Tufts University Department of Computer 
+Science. Noah Daniels has taken a graduate seminar in advanced functional 
+programming, which included some Haskell, and spent ten years in industry as a 
+professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
+has taken a programming languages course heavily focused on functional 
+programming, but with no Haskell. Norman Ramsey is the local expert, but he 
+wrote no code for this project.
+
+\section{What we did}
+
+\subsection{The algorithms}
+
+A biologist may use software such as HMMER or MRFy to test a hypothesis about 
+the function of a single, newly discovered protein, or she may use the software 
+to scan an entire newly sequenced genome against a library of known protein 
+structures. This latter case demands particularly high-performance software.
+
+%% move into later section
+
+%% add something about groups of related proteins aligned (put into correspondence)
 
@@
 random field (see Figure \ref{mrf}). A parse is a mapping of the letters of a 
-query sequence onto the states of a markov Random field, while an optimal parse 
-is the parse exhibiting the greatest probability once. The algorithm underlying 
+query sequence onto the states of a markov Random field (and the corresponding path through the finite state automaton), while an optimal parse 
+is the parse exhibiting the greatest probability. The algorithm underlying 
 MRFy grew out of an understanding of the limitations of the SMURF (Structural 
@@
 
+%% "nailing down" certain subsequences of the model; between beta strands we have HMMs
+
 \begin{figure} 
@@
 \caption{The directed graph representing the states of a MRFy Markov random 
-field finite state automaton. Shaded nodes represent beta-strand positions, 
+field finite state automaton. Shaded nodes represent beta-strand positions which have been fixed into match states, 
 while the dashed edges represent the long-range dependencies due to hydrogen 
-bonding.}\label{mrf} \end{figure}
+bonding. Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}
 
-\subsection{Our Background}
-
-All three authors are members of the Tufts University Department of Computer 
-Science. Noah Daniels has taken a graduate seminar in advanced functional 
-programming, which included some Haskell, and spent ten years in industry as a 
-professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
-has taken a programming languages course heavily focused on functional 
-programming, but with no Haskell. Norman Ramsey is the local expert, but he 
-wrote no code for this project.
-
-\section{What we did}
-
-\subsection{Beautiful code}
 MRFy contains two distinct algorithms: the classical Viterbi 
@@
 
+%% break the viterbi types up and insert into the Plan7 explanation.
+
 \begin{figure*}
@@
 \end{figure*}
+
+%% group nodes, illustrate an alignment
+
+\begin{figure}\centerline{\includegraphics[width=6cm]{alignment.pdf}} \caption{Protein sequences are put into alignment based on structural or sequence similarities.
+Each column of the alignment corresponds to a node of the hidden Markov model; columns with many gaps will induce highly probable insertion states in the HMM, while columns with fewer gaps will induce more-probable deletion and match states.
+Columns with strong consensus towards a particular residue will induce match states with high emission probability for that residue.}\label{alignment} \end{figure}
+
+%% how do proteins change due to evolution?
+%% explain insertion, deletion, substitution.
+
+%% what looks like an intimidating geometry is completely explained by the fact
+%% that we have insertions, deletions, and substitutions.
 
@@
 insertion state has an associated table of emission probabilities (here 
-illustrated only for the first two states)}\label{plan7} \end{figure}
+illustrated only for the first two states).
+Triplets of insertion, deletion, and match states are grouped into ``nodes,'' represented here with dashed ovals; each node corresponds to a column in the original alignment used to train the HMM.}\label{plan7} \end{figure}
 

### 2012-03-02 viterbi_types.hs

@@ -1,27 +1,25 @@
 type HMMState = Int
 type StatePath = [ HMMState ]
-mat = 0 :: HMMState
-ins = 1 :: HMMState
-del = 2 :: HMMState
-type EmissionProbabilities = V.Vector Double
-data TransitionProbability (fState::HMMState, tState::HMMState) = 
-     TransitionProbability { logProbability::LogProbability
-                           , fromState = value fState::HMMState
-                           , toState = value tState::HMMState
-                           }
-data TransitionProbabilities =
-     TransitionProbabilities { m_m :: TransitionProbability
-                             , m_i :: TransitionProbability
-                             , m_d :: TransitionProbability
-                             , i_m :: TransitionProbability
-                             , i_i :: TransitionProbability
-                             , d_m :: TransitionProbability
-                             , d_d :: TransitionProbability
-                             }
+data HMMState = Match | Insertion | Deletion
+type EProbs = V.Vector Double
+data TProb  = 
+     TProb { logProbability::LogProbability
+           , fromState::HMMState
+           , toState::HMMState
+           }
+data TProbs =
+     TProbs { m_m :: TProb
+            , m_i :: TProb
+            , m_d :: TProb
+            , i_m :: TProb
+            , i_i :: TProb
+            , d_m :: TProb
+            , d_d :: TProb
+            }
 data HmmNode =
      HmmNode { nodeNum :: Int
-             , matchEmissions :: EmissionProbabilities
-             , insertionEmissions :: EmissionProbabilities
-             , transitions :: TransitionProbabilities
+             , matchEmissions :: EProbs
+             , insertionEmissions :: EProbs
+             , transitions :: TProbs
              }
 type HMM = V.Vector HmmNode

### 2012-03-05 mrfy_experience_report.tex
further changes to viterbi section. more to come.

@@
 
-%% move into later section
+In typical use, HMMER builds its structural template -- its Markov random field 
+-- from a group of proteins that share common function and structure. A 
+biologist puts these protein sequences into a correspondence, or alignment, 
+typically using structural alignment software. The columns of the alignment 
+(Figure~\ref{alignment}) may contain gaps, indicating that a particular protein 
+does not have an amino acid, or residue, participating in that position. Gaps 
+in a few sequences at a position likely correspond to evolutionary deletions, 
+while gaps in many sequences at a position may correspond to evolutionary 
+insertions. A position that has residues in a majority of the sequences becomes 
+a \textit{match} state in the hidden Markov model; this position may see a 
+variety of amino acids due to evolutionary substitution, but it still 
+corresponds to a position that a homologous protein should most likely fill.
 
-%% add something about groups of related proteins aligned (put into correspondence)
+\begin{figure}\centerline{\includegraphics[width=6cm]{alignment.pdf}} 
+\caption{Protein sequences are put into alignment based on structural or 
+sequence similarities. Each column of the alignment corresponds to a node of 
+the hidden Markov model; columns with many gaps will induce highly probable 
+insertion states in the HMM, while columns with fewer gaps will induce 
+more-probable deletion and match states. Columns with strong consensus towards 
+a particular residue will induce match states with high emission probability 
+for that residue.}\label{alignment} \end{figure}
+
+The resulting match states, as well as \textit{insertion} states that allow 
+additional residues to appear in a protein sequence, and \textit{deletion} 
+states that allow match states to be skipped, form a directed graph, the hidden 
+Markov model (Figure~\ref{plan7}). Each match or insertion state has an 
+associated table of emission probabilities, which correspond to the probability 
+of seeing each specific amino acid letter at a given position. Triplets of 
+insertion, deletion, and match states are grouped into ``nodes,'' which 
+correspond to columns of the original alignment. What looks like an 
+intimidating geometry is thus completely explained by the fact that we have 
+insertions, deletions, and substitutions due to evolution.
+
+\begin{figure} \centerline{\includegraphics[width=8cm]{Plan7.pdf}} \caption{The 
+directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
+Markov model. Only 7 of the possible 9 state transitions are allowed, which 
+simplifies the dynamic programming. Insertion states represent mutations that 
+insert an amino acid, deletion states represent mutations that lose an amino 
+acid, and match states represent either conserved amino acids (those with no 
+changes) or mutations that substitute one amino acid for another. Each match or 
+insertion state has an associated table of emission probabilities (here 
+illustrated only for the first two states). Triplets of insertion, deletion, 
+and match states are grouped into ``nodes,'' represented here with dashed 
+ovals; each node corresponds to a column in the original alignment used to 
+train the HMM.}\label{plan7} \end{figure}
+
+%% now explain computing a *parse* (HMMER) leading into Viterbi.
 
@@
 %% group nodes, illustrate an alignment
-
-\begin{figure}\centerline{\includegraphics[width=6cm]{alignment.pdf}} \caption{Protein sequences are put into alignment based on structural or sequence similarities.
-Each column of the alignment corresponds to a node of the hidden Markov model; columns with many gaps will induce highly probable insertion states in the HMM, while columns with fewer gaps will induce more-probable deletion and match states.
-Columns with strong consensus towards a particular residue will induce match states with high emission probability for that residue.}\label{alignment} \end{figure}
-
-%% how do proteins change due to evolution?
-%% explain insertion, deletion, substitution.
-
-%% what looks like an intimidating geometry is completely explained by the fact
-%% that we have insertions, deletions, and substitutions.
-
-\begin{figure} \centerline{\includegraphics[width=8cm]{Plan7.pdf}} \caption{The 
-directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
-Markov model. Only 7 of the possible 9 state transitions are allowed, which 
-simplifies the dynamic programming. Insertion states represent mutations that 
-insert an amino acid, deletion states represent mutations that lose an amino 
-acid, and match states represent either conserved amino acids (those with no 
-changes) or mutations that substitute one amino acid for another. Each match or 
-insertion state has an associated table of emission probabilities (here 
-illustrated only for the first two states).
-Triplets of insertion, deletion, and match states are grouped into ``nodes,'' represented here with dashed ovals; each node corresponds to a column in the original alignment used to train the HMM.}\label{plan7} \end{figure}
 

### 2012-03-05 mrfy_experience_report.tex
tweak viterbi types, rearrange implementation section

@@
 
-\subsection{The algorithms}
+\subsection{The Viterbi Algorithm}
 
@@
 associated table of emission probabilities, which correspond to the probability 
-of seeing each specific amino acid letter at a given position. Triplets of 
-insertion, deletion, and match states are grouped into ``nodes,'' which 
-correspond to columns of the original alignment. What looks like an 
+of seeing each specific amino acid letter at a given position. Each edge in the 
+graph has a weight which is the probability of the associated state transition. 
+Triplets of insertion, deletion, and match states are grouped into ``nodes,'' 
+which correspond to columns of the original alignment. What looks like an 
 intimidating geometry is thus completely explained by the fact that we have 
@@
 
-%% now explain computing a *parse* (HMMER) leading into Viterbi.
+If a biologist has trained a hidden Markov model on a group of related 
+proteins, she may next use HMMER to determine the likelihood that a new 
+protein, whose sequence is known but structure and function are not, shares 
+structure and evolutionary history with the proteins that comprise the hidden 
+Markov model. HMMER attempts to compute the optimal \textit{parse}, or 
+alignment, of a protein sequence onto the directed graph of the hidden Markov 
+model. A parse is a mapping of the letters of a query sequence onto the states 
+of the model; equivalently, a parse is a \textit{path} through the edges of the 
+graph. Since each edge has a weight which is a probability, and each emitting 
+state (match or insertion) has a table of emission probabilities, the 
+probability of a particular path through the model is simply the product of the 
+associated probabilities. This path and its associated probability are computed 
+using the Viterbi algorithm~\cite{Viterbi:1967}.
 
-SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
-alignment, of a protein sequence onto the directed graph representing a Markov 
-random field (see Figure \ref{mrf}). A parse is a mapping of the letters of a 
-query sequence onto the states of a markov Random field (and the corresponding path through the finite state automaton), while an optimal parse 
-is the parse exhibiting the greatest probability. The algorithm underlying 
-MRFy grew out of an understanding of the limitations of the SMURF (Structural 
-Motifs Using Random Fields) algorithm, which used multidimensional dynamic 
-programming to compute the optimal parse of a query sequence onto a Markov 
-random field. Rather than attempt to compute exactly the optimal parse, MRFy 
-treats the non-local interactions (which correspond to hydrogen-bonded ``beta 
-strands'' in the protein) as ``beads'' which can be slid along a string. In 
-this analogy, the string is the sequence of amino acids of the query protein. A 
-given placement of these beads corresponds to an assignment of particular amino 
-acid residues being hydrogen bonded to other particular amino acid residues, 
-and thus MRFy computes a probability based on observed frequencies of residues 
-hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
-strands, the Markov random field is in fact a traditional hidden Markov model, 
-whose optimal solution can be easily computed via dynamic programming (known as 
-the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
-strands, based on an objective function that combines the probability computed 
-by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
-residues in the beta strands.
-
-%% "nailing down" certain subsequences of the model; between beta strands we have HMMs
-
-\begin{figure} 
-\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
-\caption{The directed graph representing the states of a MRFy Markov random 
-field finite state automaton. Shaded nodes represent beta-strand positions which have been fixed into match states, 
-while the dashed edges represent the long-range dependencies due to hydrogen 
-bonding. Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}
-
-MRFy contains two distinct algorithms: the classical Viterbi 
-dynamic-programming algorithm for hidden Markov models, and a stochastic search 
-algorithm for placing beta-strand positions. This stochastic search algorithm 
-implements several search heuristics: random-mutation hill climbing, simulated 
-annealing, multistart simulated annealing, and a genetic algorithm.
-
-The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
+The Viterbi algorithm comprises a set of recurrence 
 relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
@@
 being emitted by a particular finite-state automaton (the hidden Markov model). 
-The particular states of this finite-state automaton map to the recurrence 
-relations, and in this instance derive from what is known in the computational 
-biology community as a \textit{profile} hidden Markov model. The Viterbi 
-algorithm in this instance recurses on ``insertion,'' ``deletion,'' and 
-``match'' states (Figure~\ref{plan7}). Insertion states represent mutations 
-that gain an amino acid letter in the query sequence, deletion states represent 
-mutations that lose an amino acid letter in the query sequence, and match 
-states represent either conserved positions (no change) or mutations that 
-substitute one amino acid for another. The match and insertion states each have 
-an associated table of emission probabilities, determined during training, 
-indicating how likely that state is to emit a particular amino acid letter. 
-Edges from state to state represent state transitions, and have associated 
-probabilities determined during training. To be efficient, the Viterbi 
+To be efficient, the Viterbi 
 algorithm relies on dynamic programming, or memoization, and the resulting 
@@
 %% break the viterbi types up and insert into the Plan7 explanation.
+%% go over this with NR; I can't make it flow.
 
-\begin{figure*}
+\begin{figure}
 \lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
-\end{figure*}
-
-%% group nodes, illustrate an alignment
+\end{figure}
 
@@
 
-\begin{figure*}
-\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
-\end{figure*}
-
 We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
@@
 dynamic-programming approach in Haskell, which resulted in perspicuous code.
+
+\begin{figure*}
+\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
+\end{figure*}
+
+\subsection{MRFy Implementation}
+
+HMMER simply implements the Viterbi algorithm for computing the optimal parse 
+of a protein sequence onto a hidden Markov model. Thus, the graphical model of 
+a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}. In 
+contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
+SMURFLite~\cite{smurflite}, our previous work. Like SMURF and SMURFLite, MRFy 
+pays attention to longer-range structural interactions than a simple hidden 
+Markov model can understand. In particular, these approaches ``nail down'' 
+certain subsequences of the model, corresponding to the non-local hydrogen-bond 
+interactions between beta strands. Beta strands are regions of protein 
+structure that ``stick'' to one another. Because of their pairwise 
+interactions, they are highly conserved over the course of evolution. Moreover, 
+again because of these pairwise interactions, a change in one amino acid in a 
+beta strand may require a change in a corresponding, hydrogen-bonded residue in 
+another beta strand if the protein is to remain stable. The architecture of the 
+resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
+illustrated in Figure~\ref{mrf}.
+
+These non-local interactions cause what had been a simple hidden Markov model 
+to become a Markov random field. SMURF and SMURFLite attempted to compute the 
+optimal parse of a protein sequence onto the directed graph representing a 
+Markov random field using multidimensional dynamic programming. SMURF's 
+computational complexity quickly exploded with complex protein structures, 
+while SMURFLite limited its computational complexity by also limiting the 
+complexity of protein structure it would understand. The algorithm underlying 
+MRFy grew out of an understanding of these limitations.
+
+Rather than attempt to compute exactly the optimal parse, MRFy treats the beta 
+strands as ``beads'' which can be slid along a string. In this analogy, the 
+string is the sequence of amino acids of the query protein. A given placement 
+of these beads corresponds to an assignment of particular amino acid residues 
+being hydrogen bonded to other particular amino acid residues, and thus MRFy 
+computes a probability based on observed frequencies of residues hydrogen 
+bonding in beta strands~\cite{betawrappro}. In between these beta strands, the 
+Markov random field is in fact a traditional hidden Markov model, whose optimal 
+solution can be easily computed via the Viterbi algorithm. MRFy performs a 
+stochastic search to place these beta strands, based on an objective function 
+that combines the probability computed by the Viterbi algorithm with that 
+computed from the paired, hydrogen-bonded residues in the beta strands. If a 
+structure has many beta strands, then it also has many miniature hidden Markov 
+models which can be solved via the Viterbi algorithm.
+
+\begin{figure}[h!] 
+\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
+\caption{The directed graph representing the states of a MRFy Markov random 
+field finite state automaton. Shaded nodes represent beta-strand positions which have been fixed into match states, 
+while the dashed edges represent the long-range dependencies due to hydrogen 
+bonding. Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}
+
+MRFy contains two distinct algorithms: the classical Viterbi 
+dynamic-programming algorithm for hidden Markov models, and a stochastic search 
+algorithm for placing beta-strand positions. This stochastic search algorithm 
+implements several search heuristics: random-mutation hill climbing, simulated 
+annealing, multistart simulated annealing, and a genetic algorithm.
 

### 2012-03-05 viterbi_types.hs

@@ -1,9 +1,10 @@
 type HMMState = Int
 type StatePath = [ HMMState ]
-data HMMState = Match | Insertion | Deletion
+data HMMState = Match | Insert | Delete
 type EProbs = V.Vector Double
-data TProb  = 
-     TProb { logProbability::LogProbability
+data LogProb = LogZero | NonZero Double
+data TProb  =
+     TProb { logProbability::LogProb
            , fromState::HMMState
            , toState::HMMState
            }
@@ -18,8 +19,8 @@ data TProbs =
             }
 data HmmNode =
      HmmNode { nodeNum :: Int
-             , matchEmissions :: EProbs
-             , insertionEmissions :: EProbs
+             , matEmissions :: EProbs
+             , insEmissions :: EProbs
              , transitions :: TProbs
              }
 type HMM = V.Vector HmmNode

### 2012-03-06 mrfy_experience_report.tex
some conclusion-ish stuff written

@@
 
-Haskell makes it difficult to write code in haste, because Haskell militates towards careful type-checking and clearly thought-out function contracts.
-Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the programmer who attempts to write code quickly with the goal of later making it work correctly.
-We have found MRFy to be easier to enhance and maintain than our existing C++ code bases.
+Our previous computational biology projects include an enhancement to a protein 
+structural aligner that previously came from our group, Matt~\cite{matt}. 
+Matt's code base contains approximately 12,000 lines of C++ code, and relies on 
+data structures such as mutable oct-trees, vectors, and arrays, with 
+significant amounts of clever pointer arithmetic. Our enhancement involved 
+modifying Matt to incorporate sequence information along with the structural 
+aligner. While we initially estimated this would be a three-month project, it 
+took most of a year because the mutable data structures were difficult to 
+repurpose, and the pointer arithmetic was too clever for its own good: since 
+invariants were not clear, nearly every change resulted in new segfaults. One 
+enhancement, partial alignments (essentially a cosmetic manipulation of the 
+output, but requiring deep information about some of the data structures) was 
+deemed infeasible. We believe we could have completed a ground-up rewrite in 
+Haskell, with most of the run-time performance, in fewer than nine months.
+
+In general, the failure modes we encounter most in C++ programming are 
+segmentation faults, memory errors (due to pointer arithmetic and uninitialized 
+memory) and heap exhaustion. In contrast, the failure modes we most commonly 
+see in Haskell code are type-check failures, or at worst, runtime errors around 
+bounds checking of arrays and vectors. While bounds checking failures have 
+proved difficult to debug, they are still less so than C++ memory errors.
+
+Haskell makes it difficult to write code in haste, because Haskell militates 
+towards careful type-checking and clearly thought-out function contracts. 
+Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the 
+programmer who attempts to write code quickly with the goal of later making it 
+work correctly. We have found MRFy to be easier to enhance and maintain than 
+our existing C++ code bases. Despite the learning curve of Haskell, we 
+implemented most of MRFy in roughly three months of part-time work.
 

### 2012-03-06 mrfy_experience_report.tex
a little wordsmithing in the introduction

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[preprint,nocopyrightspace,times]{sigplanconf}
 
@@
 
+\renewcommand\ttdefault{aett}
+
+  \long\def\remark#1{%
+      \ifvmode
+         \marginpar{\raggedright\hbadness=10000
+         \parindent=8pt \parskip=2pt
+         \def\baselinestretch{0.8}\tiny
+         \itshape\noindent #1\par}%
+      \else
+          \unskip\raisebox{-3.5pt}{\rlap{$\scriptstyle\diamond$}}%
+          \marginpar{\raggedright\hbadness=10000
+         \parindent=8pt \parskip=2pt
+         \def\baselinestretch{0.8}\tiny
+         \itshape\noindent #1\par}%
+      \fi}
+
 \begin{document}
@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
+%\titlebanner{banner above paper title}        % These are ignored unless
+\preprintfooter{Haskell in Computational Biology}   % 'preprint' option specified.
 
@@
 \begin{abstract}
-This is the text of the abstract.
+HELLO?  DRAFT OF AN ABSTRACT?
 \end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
-
-\terms
-term1, term2
-
-\keywords
-keyword1, keyword2
+% \category{CR-number}{subcategory}{third-level}
+% 
+% \terms
+% term1, term2
+% 
+% \keywords
+% keyword1, keyword2
 
@@
 
-We address this paper to computational biologists who may have an interest in 
-functional programming, but this is not an advocacy paper. We do not argue that 
-computational biologists \textit{should} use functional programming. Rather, 
-argue that computational biologists \textit{can} successfully develop software 
-in Haskell, and what to expect if they do.
+Computational biologists write software that answers questions about 
+sequences of nucleic acids (genomic data) or sequences of amino 
+acids (proteomic data). 
+For some software, considerations of performance are paramount; this
+software is usually written C~or~C++. 
+For other software, considerations of convenience, readability, and
+productivity are more important;
+this software is usually written in a ``scripting'' language like
+Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
+In~this Experience Report, we show that computational biologists can
+successfully replace both these families of languages with Haskell:
+\begin{itemize}
+\item
+Our experimental Haskell programs are as easy to understand and change
+ as the programs our group writes using scripting languages.
+\item
+The performance of our Haskell code is comparable to the performance
+of~C++.
+\item
+We have experienced barriers to entry, of which the most salient has been
+the difficulty of understanding and tuning  the performance of Haskell
+programs. 
+We found the other barriers 
+relatively easy to overcome.
+\end{itemize}
+Our experience is based on MRFy (pronounced ``Murphy''), which
+implements a novel family of algorithms for finding remote homologs to 
+newly discovered proteins.
 
-We also address this paper to the functional programming community, in the 
-hopes that our experiences will be enlightening. We believe that our 
-experiences can provide lessons about outreach and education.
-
-\subsection{Problem Domain}
-
-Computational biologists write software that answers questions about large data 
-sets involving sequences of nucleic acids (genomic data) or sequences of amino 
-acids (proteomic data). This software comes in two varieties. The first variety 
-is \textit{high performance software}, where performance is a top priority; 
-this kind of software is usually written in a programming language like C or 
-C++. The second variety is software where performance is not a top priority; 
-this kind of software written in higher-level languages that emphasize 
-convenience and programmer productivity, such as Perl, Python, Ruby, or R.
-
-In this Experience Report, we show that the Haskell programming language is an 
-excellent alternative for computational biologists: a Haskell program can 
-perform comparably to a C++ program while being as flexible and easy to 
-understand as a program written in an interpreted language. Our experience is 
-based on an implementation of a novel algorithm for finding remote homologs to 
-newly discovered proteins. We call this program MRFy.
+\section{The biology}
 
@@
 
-\subsection{Our Background}
+%%  \subsection{Our Background}
+%%  
+%%  All three authors are members of the Tufts University Department of Computer 
+%%  Science. Noah Daniels has taken a graduate seminar in advanced functional 
+%%  programming, which included some Haskell, and spent ten years in industry as a 
+%%  professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
+%%  has taken a programming languages course heavily focused on functional 
+%%  programming, but with no Haskell. Norman Ramsey is the local expert, but he 
+%%  wrote no code for this project.
 
-All three authors are members of the Tufts University Department of Computer 
-Science. Noah Daniels has taken a graduate seminar in advanced functional 
-programming, which included some Haskell, and spent ten years in industry as a 
-professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
-has taken a programming languages course heavily focused on functional 
-programming, but with no Haskell. Norman Ramsey is the local expert, but he 
-wrote no code for this project.
+\remark{Dropped the section on ``our background,'' but the information
+  should reappear when and where we discuss barriers to entry in more
+  detail. ---NR}
 

### 2012-03-06 mrfy_experience_report.tex
make the code work with plain latex

@@
+\newif\ifpdfmadness
+
+\ifx\pdfoutput\undefined
+   \pdfmadnessfalse
+\else
+   \ifnum\pdfoutput=1
+     \pdfmadnesstrue
+   \else
+     \pdfmadnessfalse
+   \fi
+\fi
+
 \documentclass[preprint,nocopyrightspace,times]{sigplanconf}
@@
 
-\begin{figure}\centerline{\includegraphics[width=6cm]{alignment.pdf}} 
+\begin{figure}
+\ifpdfmadness
+\centerline{\includegraphics[width=6cm]{alignment.pdf}} 
+\else
+\centerline{\includegraphics[width=6cm]{alignment.eps}} 
+\fi
+
 \caption{Protein sequences are put into alignment based on structural or 
@@
 a particular residue will induce match states with high emission probability 
-for that residue.}\label{alignment} \end{figure}
+for that residue.}\label{alignment} 
+\end{figure}
 
@@
 
-\begin{figure} \centerline{\includegraphics[width=8cm]{Plan7.pdf}} \caption{The 
+\begin{figure} 
+\ifpdfmadness
+\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
+\else
+\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
+\fi
+
+\caption{The 
 directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
@@
 \begin{figure}[h!] 
+\ifpdfmadness
 \centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
+\else
+\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
+\fi
 \caption{The directed graph representing the states of a MRFy Markov random 

### 2012-03-06 mrfy_experience_report.tex
newlines after sentences
intro section receives the Ramsey treatment

@@
 molecules to carry out the functions of living cells: metabolism, regulation, 
-signalling, and more. The function of a protein is determined by its structure, 
-and this structure in turn is determined by the specific sequence of the twenty 
-different amino acids that make up the protein, but proteins with significantly 
-different amino acid sequences may share similar structure and function. This 
-amino acid sequence is ultimately determined by the DNA sequence we call a 
-gene. Given the genetic sequence of a protein, biologists wish to know its 
-specific cellular function. The best path to this is to find other proteins of 
-similar structure, which likely share similar function. This problem is called 
+signalling, and more.
+A~protein's function is determined by its structure, 
+and its structure is determined by the sequence of amino acids that
+form the protein.
+\remark{Lost: there are only 20-odd amino acids}
+The amino-acid sequence is ultimately determined by a sequence of
+nucleic acids in DNA, which we call a gene.
+Given a genetic sequence, biologists wish to know the cellular
+function of the protein that gene codes for.
+The best known method of discovering such function is
+to find other proteins of 
+similar structure, which likely share similar function.
+This problem is called 
 \textit{homology detection}.
 
-Computational biologists address the homology detection problem by building 
-algorithms to find similar proteins given an amino acid sequence. When there 
-are nearly-identical sequences for proteins of known structure, this is a 
-simple string-matching problem. When there are no identical sequences, but 
-still some reasonably similar sequences, one family of algorithms called 
-\textit{hidden Markov models} excels at homology detection. We will explain an 
-algorithm for detecting reasonably similar sequences, and then move on to 
-explain MRFy, our novel approach for detecting more distantly related sequences 
-for proteins that share similar structure and function.
+Computational biologists detect homologies by building 
+algorithms which, given an amino-acid sequence,\remark{This sequence
+  (the sequence ``under test'') needs a name. ---NR}
+find known proteins of
+similar structure.
+When the similar proteins are formed from amino-acid sequences that
+are nearly identical to the sequence under test, the homologies can be
+detected by  
+simple string-matching algorithms.\remark{Can we just drop the
+  sentence about string matching?}
+When the similar proteins are formed from amino-acid sequences that
+are not too far in edit distance from the sequence under test, the homologies can be
+detected by
+a family of algorithms called 
+\textit{hidden Markov models}.\remark{citation needed?}
+But in real biological systems,
+proteins with similar structure and function may be formed from significantly 
+different amino-acid sequences, which are not close in edit distance.
+Our~research contribution---the MRFy algorithm, can detect homologies
+in amino-acid sequences that are only distantly related.
+%
+%We will explain an 
+%algorithm for detecting reasonably similar sequences, and then move on to 
+%explain MRFy, our novel approach for detecting more distantly related sequences 
+%for proteins that share similar structure and function.
+
+\emph{[We need an argument here saying why an understanding of the
+    algorithms is relevant to the claims made in the experience
+    report, then say how we present the algorithms, then say what
+    experiential conclusions this presentation supports.]}
 
@@
 to scan an entire newly sequenced genome against a library of known protein 
-structures. This latter case demands particularly high-performance software.
+structures.
+This latter case demands particularly high-performance software.
 
 In typical use, HMMER builds its structural template -- its Markov random field 
--- from a group of proteins that share common function and structure. A 
+-- from a group of proteins that share common function and structure.
+A 
 biologist puts these protein sequences into a correspondence, or alignment, 
-typically using structural alignment software. The columns of the alignment 
+typically using structural alignment software.
+The columns of the alignment 
 (Figure~\ref{alignment}) may contain gaps, indicating that a particular protein 
-does not have an amino acid, or residue, participating in that position. Gaps 
+does not have an amino acid, or residue, participating in that position.
+Gaps 
 in a few sequences at a position likely correspond to evolutionary deletions, 
 while gaps in many sequences at a position may correspond to evolutionary 
-insertions. A position that has residues in a majority of the sequences becomes 
+insertions.
+A position that has residues in a majority of the sequences becomes 
 a \textit{match} state in the hidden Markov model; this position may see a 
@@
 \caption{Protein sequences are put into alignment based on structural or 
-sequence similarities. Each column of the alignment corresponds to a node of 
+sequence similarities.
+Each column of the alignment corresponds to a node of 
 the hidden Markov model; columns with many gaps will induce highly probable 
 insertion states in the HMM, while columns with fewer gaps will induce 
-more-probable deletion and match states. Columns with strong consensus towards 
+more-probable deletion and match states.
+Columns with strong consensus towards 
 a particular residue will induce match states with high emission probability 
@@
 states that allow match states to be skipped, form a directed graph, the hidden 
-Markov model (Figure~\ref{plan7}). Each match or insertion state has an 
+Markov model (Figure~\ref{plan7}).
+Each match or insertion state has an 
 associated table of emission probabilities, which correspond to the probability 
-of seeing each specific amino acid letter at a given position. Each edge in the 
+of seeing each specific amino acid letter at a given position.
+Each edge in the 
 graph has a weight which is the probability of the associated state transition.
+
 Triplets of insertion, deletion, and match states are grouped into ``nodes,'' 
-which correspond to columns of the original alignment. What looks like an 
+which correspond to columns of the original alignment.
+What looks like an 
 intimidating geometry is thus completely explained by the fact that we have 
@@
 directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
-Markov model. Only 7 of the possible 9 state transitions are allowed, which 
-simplifies the dynamic programming. Insertion states represent mutations that 
+Markov model.
+Only 7 of the possible 9 state transitions are allowed, which 
+simplifies the dynamic programming.
+Insertion states represent mutations that 
 insert an amino acid, deletion states represent mutations that lose an amino 
 acid, and match states represent either conserved amino acids (those with no 
-changes) or mutations that substitute one amino acid for another. Each match or 
+changes) or mutations that substitute one amino acid for another.
+Each match or 
 insertion state has an associated table of emission probabilities (here 
-illustrated only for the first two states). Triplets of insertion, deletion, 
+illustrated only for the first two states).
+Triplets of insertion, deletion, 
 and match states are grouped into ``nodes,'' represented here with dashed 
@@
 structure and evolutionary history with the proteins that comprise the hidden 
-Markov model. HMMER attempts to compute the optimal \textit{parse}, or 
+Markov model.
+HMMER attempts to compute the optimal \textit{parse}, or 
 alignment, of a protein sequence onto the directed graph of the hidden Markov 
-model. A parse is a mapping of the letters of a query sequence onto the states 
+model.
+A parse is a mapping of the letters of a query sequence onto the states 
 of the model; equivalently, a parse is a \textit{path} through the edges of the 
-graph. Since each edge has a weight which is a probability, and each emitting 
+graph.
+Since each edge has a weight which is a probability, and each emitting 
 state (match or insertion) has a table of emission probabilities, the 
 probability of a particular path through the model is simply the product of the 
-associated probabilities. This path and its associated probability are computed 
+associated probabilities.
+This path and its associated probability are computed 
 using the Viterbi algorithm~\cite{Viterbi:1967}.
@@
 being emitted by a particular finite-state automaton (the hidden Markov model).
+
 To be efficient, the Viterbi 
@@
 distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
-letters in the protein sequence. In our particular problem domain, both $M$ and 
-$N$ routinely contain several hundred to a few thousand elements. The types of 
+letters in the protein sequence.
+In our particular problem domain, both $M$ and 
+$N$ routinely contain several hundred to a few thousand elements.
+The types of 
 these states and probabilities are provided in Figure~\ref{viterbi_types}.
@@
 The textbook Viterbi recurrence relations refer to logs of probabilities, so 
-that the terms may be added rather than multiplying raw probabilities. For 
+that the terms may be added rather than multiplying raw probabilities.
+For 
 reasons including floating point underflow, the HMMER software (with which we 
 maintain file format compatibility) stores all probabilities in a trained HMM 
-file as negative natural logs. Thus, the Viterbi recurrence relations are 
+file as negative natural logs.
+Thus, the Viterbi recurrence relations are 
 simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
@@
 
-We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
+We implemented the Viterbi algorithm in two distinct ways.
+We first wrote a 
 bottom-up dynamic-programming version, which began with the first node of the 
 hidden Markov model and first observation and built up the three-dimensional 
-array of possible paths through the model. The array containing the paths 
-served as the memoization table. We next wrote a top-down dynamic-programming 
+array of possible paths through the model.
+The array containing the paths 
+served as the memoization table.
+We next wrote a top-down dynamic-programming 
 version, which began with the last node of the model and the last observation, 
 and applied structural recursion down to the base cases involving the first 
-node or first observation (or both). In this second approach, we used the 
+node or first observation (or both).
+In this second approach, we used the 
 \texttt{Data.Memocombinators} library to provide the memoization container 
-(again, a three-dimensional array). We determined that the run-time 
+(again, a three-dimensional array).
+We determined that the run-time 
 performances of both approaches were virtually indistinguishable, and we also 
@@
 relations found in a textbook~\cite{durbin} or the HMMER literature~\cite{eddy} 
-(see Equation \ref{viterbi_eqn}). We chose to keep the top-down version, which 
+(see Equation \ref{viterbi_eqn}).
+We chose to keep the top-down version, which 
 later proved to be wise when we had to track down a bug in one of our base 
-cases. This bug was observed as a missing state in the alignment our program 
+cases.
+This bug was observed as a missing state in the alignment our program 
 displayed, and we quickly discovered that one of our base cases was missing 
 this state: it was returning an empty list as the base of the path, rather than 
-a list containing a single node. In this, we were grateful for the resemblance 
+a list containing a single node.
+In this, we were grateful for the resemblance 
 between the mathematical description of the algorithm and the top-down 
@@
 HMMER simply implements the Viterbi algorithm for computing the optimal parse 
-of a protein sequence onto a hidden Markov model. Thus, the graphical model of 
-a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}. In 
+of a protein sequence onto a hidden Markov model.
+Thus, the graphical model of 
+a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
+In 
 contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
-SMURFLite~\cite{smurflite}, our previous work. Like SMURF and SMURFLite, MRFy 
+SMURFLite~\cite{smurflite}, our previous work.
+Like SMURF and SMURFLite, MRFy 
 pays attention to longer-range structural interactions than a simple hidden 
-Markov model can understand. In particular, these approaches ``nail down'' 
+Markov model can understand.
+In particular, these approaches ``nail down'' 
 certain subsequences of the model, corresponding to the non-local hydrogen-bond 
-interactions between beta strands. Beta strands are regions of protein 
-structure that ``stick'' to one another. Because of their pairwise 
-interactions, they are highly conserved over the course of evolution. Moreover, 
+interactions between beta strands.
+Beta strands are regions of protein 
+structure that ``stick'' to one another.
+Because of their pairwise 
+interactions, they are highly conserved over the course of evolution.
+Moreover, 
 again because of these pairwise interactions, a change in one amino acid in a 
 beta strand may require a change in a corresponding, hydrogen-bonded residue in 
-another beta strand if the protein is to remain stable. The architecture of the 
+another beta strand if the protein is to remain stable.
+The architecture of the 
 resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
@@
 These non-local interactions cause what had been a simple hidden Markov model 
-to become a Markov random field. SMURF and SMURFLite attempted to compute the 
+to become a Markov random field.
+SMURF and SMURFLite attempted to compute the 
 optimal parse of a protein sequence onto the directed graph representing a 
-Markov random field using multidimensional dynamic programming. SMURF's 
+Markov random field using multidimensional dynamic programming.
+SMURF's 
 computational complexity quickly exploded with complex protein structures, 
 while SMURFLite limited its computational complexity by also limiting the 
-complexity of protein structure it would understand. The algorithm underlying 
+complexity of protein structure it would understand.
+The algorithm underlying 
 MRFy grew out of an understanding of these limitations.
@@
 Rather than attempt to compute exactly the optimal parse, MRFy treats the beta 
-strands as ``beads'' which can be slid along a string. In this analogy, the 
-string is the sequence of amino acids of the query protein. A given placement 
+strands as ``beads'' which can be slid along a string.
+In this analogy, the 
+string is the sequence of amino acids of the query protein.
+A given placement 
 of these beads corresponds to an assignment of particular amino acid residues 
@@
 computes a probability based on observed frequencies of residues hydrogen 
-bonding in beta strands~\cite{betawrappro}. In between these beta strands, the 
+bonding in beta strands~\cite{betawrappro}.
+In between these beta strands, the 
 Markov random field is in fact a traditional hidden Markov model, whose optimal 
-solution can be easily computed via the Viterbi algorithm. MRFy performs a 
+solution can be easily computed via the Viterbi algorithm.
+MRFy performs a 
 stochastic search to place these beta strands, based on an objective function 
 that combines the probability computed by the Viterbi algorithm with that 
-computed from the paired, hydrogen-bonded residues in the beta strands. If a 
+computed from the paired, hydrogen-bonded residues in the beta strands.
+If a 
 structure has many beta strands, then it also has many miniature hidden Markov 
@@
 \caption{The directed graph representing the states of a MRFy Markov random 
-field finite state automaton. Shaded nodes represent beta-strand positions which have been fixed into match states, 
+field finite state automaton.
+Shaded nodes represent beta-strand positions which have been fixed into match states, 
 while the dashed edges represent the long-range dependencies due to hydrogen 
-bonding. Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}
+bonding.
+Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}
 
@@
 dynamic-programming algorithm for hidden Markov models, and a stochastic search 
-algorithm for placing beta-strand positions. This stochastic search algorithm 
+algorithm for placing beta-strand positions.
+This stochastic search algorithm 
 implements several search heuristics: random-mutation hill climbing, simulated 
@@
 Just as the Viterbi algorithm is one major component of MRFy, stochastic search 
-is the other. In designing MRFy, we did not know in advance what sort of 
+is the other.
+In designing MRFy, we did not know in advance what sort of 
 stochastic search technique would converge the fastest, or give us the best 
-alignments with the fastest performance. We decided to implement several 
-different search strategies so that we could experiment with all of them. We 
+alignments with the fastest performance.
+We decided to implement several 
+different search strategies so that we could experiment with all of them.
+We 
 found this to be easy and pleasant thanks to Haskell's support of higher order 
@@
 its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
-define specific search techniques. We require a \texttt{SearchStrategy} to 
+define specific search techniques.
+We require a \texttt{SearchStrategy} to 
 define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
 and \texttt{initialize}; we use the Haskell module system to allow sharing of 
-some of these functions between different search strategies. For example, both 
+some of these functions between different search strategies.
+For example, both 
 \textit{random mutation hill climb} and \textit{simulated annealing} use the 
-same initialize and mutate functions, but very different accept functions. In 
+same initialize and mutate functions, but very different accept functions.
+In 
 all, we implemented four search strategies: \textit{random mutation hill 
 climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
-and \textit{genetic algorithm}. We found the implementation of the search 
+and \textit{genetic algorithm}.
+We found the implementation of the search 
 function and search strategies to be pleasant, and the resulting code to be 
-simple and elegant. In an imperative language with object orientation like Ruby 
-or Python, we likely would have used a class hierarchy to accomplish this. We 
+simple and elegant.
+In an imperative language with object orientation like Ruby 
+or Python, we likely would have used a class hierarchy to accomplish this.
+We 
 feel that this would have required more code, because object orientation is 
@@
 number generator in Main, but we wished for search itself to be pure at this 
-level. One advantage is that this allows for testing of the \texttt{search} 
+level.
+One advantage is that this allows for testing of the \texttt{search} 
 function with no randomness at all: if we pass in a known number as the seed, 
@@
 In MRFy, the Viterbi function is called several times (once for each 
-non-beta-strand region) for each search generation. Furthermore, in a 
+non-beta-strand region) for each search generation.
+Furthermore, in a 
 population-based search such as \textit{multistart simulated annealing}, the 
 scoring function (which itself calls Viterbi) is called many times per 
-generation. Beyond traditional profiling, we found parallelization to be an 
-obvious way to improve runtime performance. We found this to be trivial thanks 
-to the \texttt{Control.Parallel} library. We pass the Viterbi function to a map over all 
+generation.
+Beyond traditional profiling, we found parallelization to be an 
+obvious way to improve runtime performance.
+We found this to be trivial thanks 
+to the \texttt{Control.Parallel} library.
+We pass the Viterbi function to a map over all 
 non-beta-strand regions of the model and query sequence, and these regions are 
-independent of one another. Similarly, we pass the scoring function to a map 
-over the population of potential solutions, which are also independent. We 
-merely substituted \texttt{parmap rseq} for map. MRFy now efficiently uses all six 
+independent of one another.
+Similarly, we pass the scoring function to a map 
+over the population of potential solutions, which are also independent.
+We 
+merely substituted \texttt{parmap rseq} for map.
+MRFy now efficiently uses all six 
 cores on an AMD workstation for a non-population-based simulated annealing 
@@
 Despite the beauty of some of the code we were able to write, we found several 
-instances where the easiest path seemed to be to write ugly code. The first of 
-these relates to the otherwise elegant Viterbi function. Our stochastic search 
+instances where the easiest path seemed to be to write ugly code.
+The first of 
+these relates to the otherwise elegant Viterbi function.
+Our stochastic search 
 approach calls the Viterbi function a tremendous number of times, and most of 
-the program's runtime is spent in this function. However, the Viterbi algorithm 
+the program's runtime is spent in this function.
+However, the Viterbi algorithm 
 retains not just a score (the probability of the sequence given a state path) 
-but also the state path itself. We implemented this state path as a list of 
+but also the state path itself.
+We implemented this state path as a list of 
 \texttt{Int}, and every step in the recursion conses a new element onto the 
-list. However, at each step of the stochastic search, we discard the state 
-path; we are searching to optimize score only. We only need the state path at 
+list.
+However, at each step of the stochastic search, we discard the state 
+path; we are searching to optimize score only.
+We only need the state path at 
 the termination of the search, when we wish the program to output an alignment 
-of the query sequence to the Markov random field model. We found that 
+of the query sequence to the Markov random field model.
+We found that 
 optimizing the Viterbi function by removing this state path entirely improved 
-runtime performance by nearly 50\%. However, we still needed the Viterbi 
+runtime performance by nearly 50\%.
+However, we still needed the Viterbi 
 function to provide a state path when we called it after search finally 
-terminated. Thus, we wondered how to implement the Viterbi function so that a 
+terminated.
+Thus, we wondered how to implement the Viterbi function so that a 
 state path was optional.
@@
 a quick and dirty approach that is shameful in any language, but easy: we 
-duplicated code. We copied \texttt{viterbi} to produce a \texttt{viterbiF} 
-which did not retain state path. Next, however, we wondered how to avoid code 
-duplication. Clearly, littering the code with conditionals would produce code 
-that was ugly as well as less efficient. We knew how we would solve this 
+duplicated code.
+We copied \texttt{viterbi} to produce a \texttt{viterbiF} 
+which did not retain state path.
+Next, however, we wondered how to avoid code 
+duplication.
+Clearly, littering the code with conditionals would produce code 
+that was ugly as well as less efficient.
+We knew how we would solve this 
 problem in other languages; we would have used macros in Lisp, or 
 metaprogramming via \texttt{eval} in Ruby, to produce two versions of the 
-function from one prototype. We struggled, however, to solve this problem 
+function from one prototype.
+We struggled, however, to solve this problem 
 idiomatically in Haskell.
@@
 \lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type.
+
 Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
 or \texttt{consNoPath} as a parameter, the path would or would not be 
-allocated. We found this higher-order function approach to be simple in code 
+allocated.
+We found this higher-order function approach to be simple in code 
 and in concept, and it imposes negligible run-time overhead as compared to 
@@
 Another source of ugliness resulted from adding cost-center annotations when we 
-wished to profile our code. GHC will happily add cost centers to top-level 
-functions, but much of our program relies on sub-functions. We found that 
+wished to profile our code.
+GHC will happily add cost centers to top-level 
+functions, but much of our program relies on sub-functions.
+We found that 
 manually adding \texttt{\{-SCC -\}} annotations to dozens of guard clauses and 
 sub-functions harmed the readability of our program, such that we felt 
-compelled to remove all such annotations as soon as we could. In our prior 
+compelled to remove all such annotations as soon as we could.
+In our prior 
 experience with profiling tools such as kcachegrind, simply compiling with 
 debug symbols -- rather than manually annotating many lines of code -- allowed 
-the profiling tools to provide us enough information. Without these 
+the profiling tools to provide us enough information.
+Without these 
 annotations, both call-site and allocation profiles are relatively useless in 
-GHC. Perhaps a compile-time flag that would auto-annotate every equation -- 
+GHC.
+Perhaps a compile-time flag that would auto-annotate every equation -- 
 even if the resulting cost center is simply named with a line number -- would 
@@
 The final source of ugliness in our code came from attempts at debugging 
-run-time errors. MRFy relies on a source of input data -- the paired beta 
+run-time errors.
+MRFy relies on a source of input data -- the paired beta 
 strands in the Markov random field -- which is represented in an awkward manner 
-by the original SMURF program, and which is outside our control. Rather than 
+by the original SMURF program, and which is outside our control.
+Rather than 
 represent beta strands directly, SMURF represents pairings of residues, leaving 
-us to infer and reconstruct beta strands. This task was difficult due to 
+us to infer and reconstruct beta strands.
+This task was difficult due to 
 unclear invariants around overlapping, doubly-paired beta strands (and it would 
 have been difficult in any programming language; the difficulty was one of 
-representation). However, the debugging tools in GHC caused us some 
-difficulties in determining the sources of run-time errors. We used 
+representation).
+However, the debugging tools in GHC caused us some 
+difficulties in determining the sources of run-time errors.
+We used 
 \texttt{trace} extensively, but this style of ``\texttt{printf} debugging'' 
-littered our code. Being initially unaware of the backtrace feature of the GHC 
+littered our code.
+Being initially unaware of the backtrace feature of the GHC 
 profiler, we wrote wrappers around \texttt{Vector.slice} and \texttt{Vector.!} 
 which themselves called trace, which at least kept our ``\texttt{printf} 
-debugging'' less cluttered. Nonetheless, we struggled with these runtime errors 
+debugging'' less cluttered.
+Nonetheless, we struggled with these runtime errors 
 due to an awkward representation, and wished for debugging tools such as we 
@@
 structural aligner that previously came from our group, Matt~\cite{matt}.
+
 Matt's code base contains approximately 12,000 lines of C++ code, and relies on 
 data structures such as mutable oct-trees, vectors, and arrays, with 
-significant amounts of clever pointer arithmetic. Our enhancement involved 
+significant amounts of clever pointer arithmetic.
+Our enhancement involved 
 modifying Matt to incorporate sequence information along with the structural 
-aligner. While we initially estimated this would be a three-month project, it 
+aligner.
+While we initially estimated this would be a three-month project, it 
 took most of a year because the mutable data structures were difficult to 
 repurpose, and the pointer arithmetic was too clever for its own good: since 
-invariants were not clear, nearly every change resulted in new segfaults. One 
+invariants were not clear, nearly every change resulted in new segfaults.
+One 
 enhancement, partial alignments (essentially a cosmetic manipulation of the 
 output, but requiring deep information about some of the data structures) was 
-deemed infeasible. We believe we could have completed a ground-up rewrite in 
+deemed infeasible.
+We believe we could have completed a ground-up rewrite in 
 Haskell, with most of the run-time performance, in fewer than nine months.
@@
 segmentation faults, memory errors (due to pointer arithmetic and uninitialized 
-memory) and heap exhaustion. In contrast, the failure modes we most commonly 
+memory) and heap exhaustion.
+In contrast, the failure modes we most commonly 
 see in Haskell code are type-check failures, or at worst, runtime errors around 
-bounds checking of arrays and vectors. While bounds checking failures have 
+bounds checking of arrays and vectors.
+While bounds checking failures have 
 proved difficult to debug, they are still less so than C++ memory errors.
@@
 towards careful type-checking and clearly thought-out function contracts.
+
 Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the 
 programmer who attempts to write code quickly with the goal of later making it 
-work correctly. We have found MRFy to be easier to enhance and maintain than 
-our existing C++ code bases. Despite the learning curve of Haskell, we 
+work correctly.
+We have found MRFy to be easier to enhance and maintain than 
+our existing C++ code bases.
+Despite the learning curve of Haskell, we 
 implemented most of MRFy in roughly three months of part-time work.
@@
 
-- The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies. How to tell what's good? How did we discover what's good?
+- The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies.
+How to tell what's good? How did we discover what's good?
 - Profiler can produce backtraces, but this was hard to discover.
@@
 
-% We recommend abbrvnat bibliography style.
-
-\bibliographystyle{abbrvnat}
+\bibliographystyle{plainnat}
 

### 2012-03-07 mrfy_experience_report.tex
edits to roughly first 3 columns

@@
 
-\documentclass[preprint,nocopyrightspace,times]{sigplanconf}
+\documentclass[preprint,nonatbib,blockstyle,nocopyrightspace,times]{sigplanconf}
 
@@
 
+\newcommand\figref[1]{Figure~\ref{#1}}
+
 \renewcommand\ttdefault{aett}
@@
       \fi}
+
+\usepackage[authoryear]{natbib}
+\bibpunct();A{},
+\let\cite\citep
+\let\citeyearnopar=\citeyear
+\let\citeyear=\citeyearpar
 
@@
 algorithms which, given an amino-acid sequence,\remark{This sequence
-  (the sequence ``under test'') needs a name. ---NR}
+  (the sequence ``under test'') needs a name.  (Urk! I see below it is
+  called the ``query sequence.''  This name needs to come in earlier.) ---NR}
 find known proteins of
@@
 different amino-acid sequences, which are not close in edit distance.
-Our~research contribution---the MRFy algorithm, can detect homologies
+Our~research contribution---the MRFy algorithm---can detect homologies
 in amino-acid sequences that are only distantly related.
@@
 
-\section{What we did}
+\section{The Software}
 
-\subsection{The Viterbi Algorithm}
+%% \subsection{The Viterbi Algorithm}
 
-A biologist may use software such as HMMER or MRFy to test a hypothesis about 
-the function of a single, newly discovered protein, or she may use the software 
-to scan an entire newly sequenced genome against a library of known protein 
+Homology-detection software most often used in one of two ways:
+to test a hypothesis about 
+the function of a single, newly discovered protein, or 
+to compare every protein in a genome against a library of known protein 
 structures.
-This latter case demands particularly high-performance software.
+\remark{Drop this sentence? This latter case demands particularly high-performance software.}
+In~both cases, the software is ``trained'' \remark{do we want to talk ``Training''?}
+on a group of proteins that share function and structure.
+These proteins are identified by a biologist, who puts
+their amino-acid sequences into a correspondence, also called an
+\emph{alignment}.
+\remark{What does the alignment signify?  Biological significance?
+  Computational significance?}
+An~alignment may be represented as a matrix
+(Figure~\ref{alignment}) 
+in which each row corresponds to the amino-acid sequence of a protein,
+and each column groups amino acids that play similar roles in
+different proteins.
+A~column
+may contain gaps, which indicate
+indicating certain protein have no amino acid participating in that
+position.
+\remark{The word ``position'' hasn't been defined, and can we get away
+  without ``residue''?  (If~we need ``residue,'' let's give it a
+  proper introduction.}
+If~a column contains just one or two gaps, the most likely explanation
+is that in one or two proteins, mutations in the underlying DNA
+deleted an amino acid from that position (\figref{alignment}).
+If~a column contains \emph{mostly} gaps, 
+the most likely explanation
+is that in one or two proteins, mutations in the underlying DNA
+\emph{inserted} an amino acid in that position (\figref{alignment}).
 
-In typical use, HMMER builds its structural template -- its Markov random field 
--- from a group of proteins that share common function and structure.
-A 
-biologist puts these protein sequences into a correspondence, or alignment, 
-typically using structural alignment software.
-The columns of the alignment 
-(Figure~\ref{alignment}) may contain gaps, indicating that a particular protein 
-does not have an amino acid, or residue, participating in that position.
-Gaps 
-in a few sequences at a position likely correspond to evolutionary deletions, 
-while gaps in many sequences at a position may correspond to evolutionary 
-insertions.
-A position that has residues in a majority of the sequences becomes 
-a \textit{match} state in the hidden Markov model; this position may see a 
-variety of amino acids due to evolutionary substitution, but it still 
-corresponds to a position that a homologous protein should most likely fill.
+The protein alignment is used to train a \emph{hidden-Markov model}.
+A~hidden-Markov model is a probabilistic finite-state machine which 
+can assign a probability to any amino-acid sequence under test.
+\remark{What is the biological or computational meaning of the
+  probability?}
+Proteins whose sequences have higher probabilities are more likely to
+be structurally related to the proteins in the alignment.
+\remark{Should we say ``homologous'' rather than  ``structurally
+  related''?
+  If~so, where should this jargon be introduced?}
+The states of the hidden-Markov model are determined by the alignment:
+\remark{I've tried for lots of parallel structure.
+Somebody let me know if any of this is right.  ---NR}
+\begin{itemize}
+\item
+For a given column~$C$ of the alignment, the hidden-Markov model has a
+\emph{match state}.
+The match state determines the probability that the sequence under
+test has a particular amino acid in column~$C$.
+The match state assigns high probability to the amino acids that
+appear in column~$C$, and a low but nonzero probability to other amino
+acids.
+\item
+For every column of the alignment, the hidden-Markov model has a
+\emph{deletion state}.
+The deletion state determines the probability that the sequence under
+test has \emph{no} amino acid  in column~$C$, i.e., that it has a gap.
+The deletion state has high probability if there are lots of gaps in
+column~$C$,
+and low probability if there are few gaps in column~$C$.
+\item
+Between every adjacent pair of columns in the alignment,\remark{This isn't right}
+the hidden-Markov model has an
+\emph{insertion state}.
+The insertion state determines the probability that the sequence under
+test has an ``extra'' amino acid that does not correspond to any
+column in the alignment.
+The insertion state assigns high probability to \emph{[some amino
+    acids]}, and low probability to \emph{[other amino acids]}.
+\end{itemize}
+%%  A position that has residues in a majority of the sequences becomes 
+%%  a \textit{match} state in the hidden Markov model; this position may see a 
+%%  variety of amino acids due to evolutionary substitution, but it still 
+%%  corresponds to a position that a homologous protein should most likely fill.
+The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
+\figref{plan7} shows a hidden-Markov model derived from an alignment
+with just four columns.
+\emph{[I'm sure I don't have the proper explanation of the probabilities.]}
+\remark{LOST: Each edge in the 
+graph has a weight which is the probability of the associated state transition.}
+Each column in the original alignment corresponds to a \emph{node},
+which is a triplet holding an insertion state, a deletion state, and a
+match state.
+What looks like an 
+intimidating geometry is thus completely explained by the fact that we have 
+insertions, deletions, and substitutions due to evolution.
+\remark{The evolution seems like a distraction to me here, and we hit
+  it above.  Can we simply drop this sentence?}
+\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
+    I THINK WE'RE READY.]}
 
@@
 
-\caption{Protein sequences are put into alignment based on structural or 
+\emph{[Not sure we need this text: 
+Protein sequences are put into alignment based on structural or 
 sequence similarities.
@@
 a particular residue will induce match states with high emission probability 
-for that residue.}\label{alignment} 
+for that residue.]}
+
+\caption{A~structural alignment of four proteins}
+\label{alignment} 
 \end{figure}
-
-The resulting match states, as well as \textit{insertion} states that allow 
-additional residues to appear in a protein sequence, and \textit{deletion} 
-states that allow match states to be skipped, form a directed graph, the hidden 
-Markov model (Figure~\ref{plan7}).
-Each match or insertion state has an 
-associated table of emission probabilities, which correspond to the probability 
-of seeing each specific amino acid letter at a given position.
-Each edge in the 
-graph has a weight which is the probability of the associated state transition.
-
-Triplets of insertion, deletion, and match states are grouped into ``nodes,'' 
-which correspond to columns of the original alignment.
-What looks like an 
-intimidating geometry is thus completely explained by the fact that we have 
-insertions, deletions, and substitutions due to evolution.
 
@@
 
-\caption{The 
-directed graph representing the states of a traditional HMMER ``Plan7'' hidden 
-Markov model.
-Only 7 of the possible 9 state transitions are allowed, which 
-simplifies the dynamic programming.
-Insertion states represent mutations that 
-insert an amino acid, deletion states represent mutations that lose an amino 
-acid, and match states represent either conserved amino acids (those with no 
-changes) or mutations that substitute one amino acid for another.
-Each match or 
-insertion state has an associated table of emission probabilities (here 
-illustrated only for the first two states).
-Triplets of insertion, deletion, 
-and match states are grouped into ``nodes,'' represented here with dashed 
-ovals; each node corresponds to a column in the original alignment used to 
-train the HMM.}\label{plan7} \end{figure}
+This directed graph represents the states of a
+hidden-Markov model in the ``Plan7'' format defined by the HMMER
+software package.
+%% Only 7 of the possible 9 state transitions are allowed, which 
+%% simplifies the dynamic programming.
+Each insertion state~($I$) represents a potential mutation that 
+inserts an amino acid;
+each deletion state~($D$) represents a potental mutation that
+deletes an amino acid;
+and each match state~($M$) represents either
+an unmutated amino acid or a potential mutation that
+substitutes one amino acid for another.
+A~match state or an insertion state includes a table that assigns a
+nonzero probability to each amino acid.
+(The figure shows tables for the first two states only.)
+A~triplet of an insertion state, deletion state, 
+and match state formas a \emph{node}, which corresponds to a column in
+the alignment used to train the model.
+(Each node is shown in a dashed oval.)
 
-If a biologist has trained a hidden Markov model on a group of related 
-proteins, she may next use HMMER to determine the likelihood that a new 
-protein, whose sequence is known but structure and function are not, shares 
-structure and evolutionary history with the proteins that comprise the hidden 
-Markov model.
-HMMER attempts to compute the optimal \textit{parse}, or 
-alignment, of a protein sequence onto the directed graph of the hidden Markov 
+\caption{A hidden-Markov model trained from a four-column alignment}
+
+\label{plan7} \end{figure}
+
+A hidden-Markov model like that shown in \figref{plan7} can be used to
+determine the likelihood that a new protein shares structure and
+evolutionary history with the proteins used to train the model.
+(The new protein's amino-acid sequence is known, but its structure and
+function are unknown.)
+Such a likelihood can be computed by
+a~widely used software package called HMMER (``hammer'').
+HMMER computes the optimal \textit{parse}, or 
+alignment, of a protein sequence onto the directed graph of the hidden-Markov 
 model.
-A parse is a mapping of the letters of a query sequence onto the states 
-of the model; equivalently, a parse is a \textit{path} through the edges of the 
-graph.
-Since each edge has a weight which is a probability, and each emitting 
-state (match or insertion) has a table of emission probabilities, the 
-probability of a particular path through the model is simply the product of the 
-associated probabilities.
-This path and its associated probability are computed 
-using the Viterbi algorithm~\cite{Viterbi:1967}.
+A parse maps each amino acid of the sequence under test onto a state
+of the model; equivalently, a parse is a \textit{path} through the model.
+Each edge is weighted with a probability, and each emitting 
+state (match or insertion) maps each amino acid to a probability,
+and the 
+probability of a given path is the product of these individual probabilities.
+The most likely path, and its associated probability, are computed 
+using an algorithm developed by~\citet{Viterbi:1967}.
 

### 2012-03-07 mrfy_experience_report.tex
a few notes and a cut at the argument in the intro

@@
 a family of algorithms called 
-\textit{hidden Markov models}.\remark{citation needed?}
+\textit{hidden Markov models}~\cite{hmmer}.\remark{citation needed?}
 But in real biological systems,
@@
     experiential conclusions this presentation supports.]}
+
+We present an explanation of the algorithms relevant to MRFy, because there 
+exists a natural fit between the mathematics behind the algorithms and the 
+Haskell implementation of those algorithms.
+If the reader understands the 
+algorithms, he or she is more likely to appreciate the benefits of implementing 
+them in Haskell.
+We will first present the more traditional algorithm used in 
+hidden Markov models, followed by our variation on it and the stochastic search 
+approach.
+We conclude that implementations of mathematically-intensive 
+algorithms such as these lend themselves to the Haskell language, particularly 
+when the programmer needs a flexible implementation that simplifies 
+experimenting with variants of the algorithm.
 
@@
 The protein alignment is used to train a \emph{hidden-Markov model}.
+\remark{I've never seen this hypenated, in ANY literature.}
 A~hidden-Markov model is a probabilistic finite-state machine which 
@@
 the hidden-Markov model has an
-\emph{insertion state}.
+\emph{insertion state}.\remark{concept of consensus columns is what's missing}
 The insertion state determines the probability that the sequence under

### 2012-03-08 mrfy_experience_report.tex
added a barriers section

@@
 % stochastic search approach to Markov random fields.
-
-%%  \subsection{Our Background}
-%%  
-%%  All three authors are members of the Tufts University Department of Computer 
-%%  Science. Noah Daniels has taken a graduate seminar in advanced functional 
-%%  programming, which included some Haskell, and spent ten years in industry as a 
-%%  professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
-%%  has taken a programming languages course heavily focused on functional 
-%%  programming, but with no Haskell. Norman Ramsey is the local expert, but he 
-%%  wrote no code for this project.
 
@@
 
+\subsection{Our Background}
+
+\remark{I moved this section from the Intro as per NR's suggestion. ---AG}
+All three authors are members of the Tufts University Department of Computer
+Science.
+Noah Daniels has taken a graduate seminar in advanced functional
+programming, which included some Haskell, and spent ten years in industry as a
+professional programmer in languages such as Ruby, C, and C++.
+Andrew Gallant has taken a programming languages course heavily focused on 
+functional programming, but with no Haskell.
+Norman Ramsey is the local expert, but he wrote no code for this project.
+
 \subsection{Barriers}
+\remark{I feel like a lot of this is a summary of some of the problems we've
+  discussed previously in the paper. Is that okay? A summary may be worth it. 
+---AG}
 
-- profiling
-- debugging
- [ talk about profiling and debugging problems where?]
-[ something about quickcheck: how do we know when we have enough QC props? How do we start testing a complex module? ]
+As inexperienced functional programmers, developing a complex piece of
+software in Haskell posed several barriers.
+\remark{NOAH: You did the profiling, could you make sure what I've said here
+is accurate and/or relevant? I didn't know what else to say. ---AG}
+The first of which was profiling with GHC.
+We previously mentioned that profiling with GHC produced ugly code with
+cost center annotations scattered throughout our code, but we also found it
+very difficult to profile code at all.
+In particular, profiling required that all installed libraries be re-compiled
+with profiling support---which was a process that we could not replicate
+consistently.
+
+Another barrier we faced was how to test our code using QuickCheck.
+Indeed, the barrier was so formidable in our case that we still have not been 
+able to use it.
+The problems we faced specifically were two-fold: How do we start testing
+a complex module? and How can we use QuickCheck to generate random inputs with 
+complex constraints?
+
+Finally, one last barrier we faced was debugging runtime errors.
+Runtime errors in the form of out-of-bounds indexing were particularly 
+difficult to track down because they were not accompanied by a stack trace.
+(While profiling with GHC could help remedy this, profiling was itself a 
+barrier, as previous mentioned.)
+We had to resort to polluting our code with cals to the \texttt{trace} function 
+in order to track down where the out-of-bounds errors were occurring.
+
+% - profiling 
+% - debugging 
+ % [ talk about profiling and debugging problems where?] 
+% [ something about quickcheck: how do we know when we have enough QC props? How do we start testing a complex module? ]
 

### 2012-03-08 mrfy_experience_report.tex
add something to 'what you must know to succeed' and filled in a bit of 'evaluation'

@@
 \subsection{What you must know to succeed}
-- Ecosystem is loaded with tools that look promising but either are no longer maintained or are not ready for prime time (Pads, Backtrace)
+\remark{I don't like this. I really struggled with this section. ---AG}
+If others would like to follow in our footsteps, there are a few
+things we believe one must know in order to succeed.
+Firstly, some experience with functional programming concepts is a necessity. 
+Specifically, coming from an imperative background, writing algorithms and 
+modeling data without the ability to mutate state were the most difficult 
+concepts to adapt to.
+NEED MORE.
 
 \subsection{Evaluation}
-- What does the local expert do? Do you have to have one?
+\remark{I think we need more in this section. ---AG}
+In hindsight, we feel that there are several key facets of our project in 
+Haskell that deserve evaluation.
+Firstly, we found that existing tools and libraries were plentiful and 
+typically helped us greatly.
+In particular, Data.Memocombinators made dynamic 
+programming with Viterbi exquisitely simple and elegant.
+Control.Parallel.Strategies also made the parallelization of Viterbi trivial.
+Most of the tools we used came straight from Hackage, and although 
+documentation of some form was almost always present, it was sometimes 
+difficult to tell if a particular tool was right for us.
+In particular, parts of the documentation were inaccessible to us as beginning 
+functional programmers.
+We found that documentation with straight-forward examples to be the most 
+helpful.
 
-- To the FP community: Those who want us to abandon our old tools for debugging imperative code have not provided a clear path to follow (both intellectually, and via software tools).
+Another component worthy of evaluation is the debugging process.
+Debugging in Haskell often takes the form of fixing type errors; which we think 
+is a great thing.
+However, in the case where debugging extends into runtime behavior, there 
+doesn't appear to be any clearly defined approach to effective debugging.
+Stack traces are available via profiling, but this was difficult to find and 
+as we mentioned, profiling has its own problems.
 
-- No obvious body of knowledge on code improvement.
-
-- The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies.
-How to tell what's good? How did we discover what's good?
-- Profiler can produce backtraces, but this was hard to discover.
+% - What does the local expert do? Do you have to have one? 
+%  
+% - To the FP community: Those who want us to abandon our old tools for debugging imperative code have not provided a clear path to follow (both intellectually, and via software tools). 
+%  
+% - No obvious body of knowledge on code improvement. 
+%  
+% - The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies. 
+% How to tell what's good? How did we discover what's good? 
+% - Profiler can produce backtraces, but this was hard to discover.
 

### 2012-03-08 mrfy_experience_report.tex
a few more edits. still working through NR's comments.

@@
 nucleic acids in DNA, which we call a gene.
-Given a genetic sequence, biologists wish to know the cellular
+Given a genetic sequence (the \textit{query sequence}), biologists wish to know the cellular
 function of the protein that gene codes for.
@@
 similar structure, which likely share similar function.
-This problem is called 
+Proteins that share structure and function are \textit{homologous}, and thus
+the process of identifying similar proteins to a query sequence is called 
 \textit{homology detection}.
@@
 Computational biologists detect homologies by building 
-algorithms which, given an amino-acid sequence,\remark{This sequence
-  (the sequence ``under test'') needs a name.  (Urk! I see below it is
-  called the ``query sequence.''  This name needs to come in earlier.) ---NR}
-find known proteins of
-similar structure.
+algorithms which, given an amino-acid query sequence,
+find known proteins of similar structure.
 When the similar proteins are formed from amino-acid sequences that
@@
 a family of algorithms called 
-\textit{hidden Markov models}~\cite{hmmer}.\remark{citation needed?}
+\textit{hidden Markov models}~\cite{hmmer}.
 But in real biological systems,
@@
 structures.
-\remark{Drop this sentence? This latter case demands particularly high-performance software.}
-In~both cases, the software is ``trained'' \remark{do we want to talk ``Training''?}
+The software is ``trained'' \remark{do we want to talk ``Training''?}
 on a group of proteins that share function and structure.
@@
 their amino-acid sequences into a correspondence, also called an
-\emph{alignment}.
-\remark{What does the alignment signify?  Biological significance?
-  Computational significance?}
+\emph{alignment}. This alignment signifies structural and evolutionary
+correspondence among individual amino acids in a set of homologous proteins.
 An~alignment may be represented as a matrix
@@
 different proteins.
-A~column
-may contain gaps, which indicate
-indicating certain protein have no amino acid participating in that
-position.
-\remark{The word ``position'' hasn't been defined, and can we get away
-  without ``residue''?  (If~we need ``residue,'' let's give it a
-  proper introduction.}
+A protein may have a gap in a column of the alignment, which indicates
+that during the course of evolution, either that protein lost that amino
+acid, or other proteins gained an amino acid.
 If~a column contains just one or two gaps, the most likely explanation
@@
 Proteins whose sequences have higher probabilities are more likely to
-be structurally related to the proteins in the alignment.
+be homologous to the proteins in the alignment.
 \remark{Should we say ``homologous'' rather than  ``structurally
@@
 \item
-For a given column~$C$ of the alignment, the hidden-Markov model has a
+If a given column~$j$ of the alignment contains amino acids in a majority
+of the protein sequences, it is considered a \textit{consensus column}.
+\item
+For a given column~$j$ of the alignment, the hidden-Markov model has a
 \emph{match state}.
 The match state determines the probability that the sequence under
-test has a particular amino acid in column~$C$.
+test has a particular amino acid in column~$j$.
 The match state assigns high probability to the amino acids that
-appear in column~$C$, and a low but nonzero probability to other amino
+appear in column~$j$, and a low but nonzero probability to other amino
 acids.
+The match state is more probable if it corresponds to a consensus column.
 \item
@@
 The deletion state determines the probability that the sequence under
-test has \emph{no} amino acid  in column~$C$, i.e., that it has a gap.
+test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
 The deletion state has high probability if there are lots of gaps in
-column~$C$,
-and low probability if there are few gaps in column~$C$.
+column~$j$, and low probability if there are few gaps in column~$j$.
+Thus, the deletion state is more likely in non-consensus columns.
 \item
-Between every adjacent pair of columns in the alignment,\remark{This isn't right}
-the hidden-Markov model has an
-\emph{insertion state}.\remark{concept of consensus columns is what's missing}
-The insertion state determines the probability that the sequence under
-test has an ``extra'' amino acid that does not correspond to any
-column in the alignment.
-The insertion state assigns high probability to \emph{[some amino
-    acids]}, and low probability to \emph{[other amino acids]}.
+For every column of the alignment, the hidden-Markov model has an
+\emph{insertion state}.
+The insertion state determines the probability that a query sequence
+has an ``extra'' amino acid that does not correspond to any
+consensus column in the alignment.
+Thus, the insertion state is more likely in non-consensus columns.
+The insertion state assigns high probability to amino acids that are
+frequently observed among all proteins, and low probability to amino acids
+that are rarely observed.
 \end{itemize}

### 2012-03-08 mrfy_experience_report.tex
a few more tweaks corresponding to NR's comments. Still need to add types.

@@
 The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
+The probability associated with a specific state are conditional upon the previous
+state in the model.
+Thus, the hidden Markov model can be represented as a directed graph, with each
+edge weight corresponding to the probability of a specific state transition.
 \figref{plan7} shows a hidden-Markov model derived from an alignment
 with just four columns.
-\emph{[I'm sure I don't have the proper explanation of the probabilities.]}
-\remark{LOST: Each edge in the 
-graph has a weight which is the probability of the associated state transition.}
 Each column in the original alignment corresponds to a \emph{node},

### 2012-03-08 mrfy_experience_report.tex
minor notes and tweaks.

@@
 match state.
-What looks like an 
-intimidating geometry is thus completely explained by the fact that we have 
-insertions, deletions, and substitutions due to evolution.
-\remark{The evolution seems like a distraction to me here, and we hit
-  it above.  Can we simply drop this sentence?}
 \emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
     I THINK WE'RE READY.]}
+
+% time for types.
+% first, ideal: amino acids as an algebraic DT
+% next, back off this: we're working with an existing input format
+% so, how is the graph represented (vector of state-transitions, 
+% vector of emissions)
+% so show those types
 
@@
 \fi
-
-\emph{[Not sure we need this text: 
-Protein sequences are put into alignment based on structural or 
-sequence similarities.
-Each column of the alignment corresponds to a node of 
-the hidden Markov model; columns with many gaps will induce highly probable 
-insertion states in the HMM, while columns with fewer gaps will induce 
-more-probable deletion and match states.
-Columns with strong consensus towards 
-a particular residue will induce match states with high emission probability 
-for that residue.]}
 

### 2012-03-08 mrfy_experience_report.tex
added a few types.

@@
 The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
-The probability associated with a specific state are conditional upon the previous
+The probability associated with a specific state is conditional upon the previous
 state in the model.
@@
 match state.
+
+We employed a variety of Haskell data types to represent the hidden Markov
+model.
+Ideally, if we were writing something like the BioHaskell library, we would
+have begun with an algebraic datatype for amino acids deriving
+\texttt{Ix}, something like
+\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
+which would still allow us to index into an array or vector.
+However, since we wanted to maintain compatibility with the file format
+shared with previous software, we broke with idiomatic Haskell and represented
+our amino acids as \texttt{Int}.
+Similarly, to avoid a great deal of data transformation, each state's emission
+and insertion probabilities were simply
+\begin{verbatim}type EProbs = V.Vector Double \end{verbatim} while the states themselves
+were algebraic datatypes: 
+\begin{verbatim}data HMMState = Match | Insert | Delete \end{verbatim}
+
 \emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?

### 2012-03-08 mrfy_experience_report.tex
flushing our section 6

@@
 concepts to adapt to.
-NEED MORE.
+
+Secondly, identifying existing tools and libraries from Hackage that are mature 
+is also important.
+We found that some packages that could have been immensely useful---like one 
+that provides backtraces---were either not maintained or contained bugs.
+Some other packages like Pads were functional but immature, which resulted in 
+additional code to work around the parts of Pads that had not been implemented 
+yet.
+
+Finally, a ``local expert'' or someone available to talk to with functional 
+programming expertise, can make the transition to a language like Haskell 
+easier.
+We don't believe a local expert is absolutely necessary, but answers 
+specifically related to functional programming idioms were easier to find when 
+a local expert was available.
 
@@
 functional programmers.
-We found that documentation with straight-forward examples to be the most 
-helpful.
+Documentation that was accessible to us typically contained straight-forward 
+examples showing how to use that particular tool or library.
 
@@
 is a great thing.
-However, in the case where debugging extends into runtime behavior, there 
-doesn't appear to be any clearly defined approach to effective debugging.
+However, in the case where debugging extends into runtime behavior, we couldn't 
+find any clearly defined approach to effective debugging.
 Stack traces are available via profiling, but this was difficult to find and 
-as we mentioned, profiling has its own problems.
+as we mentioned, profiling had its own barriers.
+
+Finally, we believe that we could have achieved a better division of modules in 
+our project, but that conflicted with one our goals: to get publishable results
+quickly. Despite being hasty programmers, we were still able to produce code 
+that is easy to understand and easily extensible. Namely, in our experience, 
+hasty Haskell code beats hasty code in a language like C or C++.
+
+In summary, with Haskell, we didn't have to master a completely different 
+design and development style but we did have to wrap our heads around 
+functional programming.
 

### 2012-03-08 mrfy_experience_report.tex
smallverbatim environment, minor edits

--- 2012-03-08 mrfy_experience_report.tex
@@
 \newcommand\figref[1]{Figure~\ref{#1}}
+\newcommand\secref[1]{Section~\ref{sec:#1}}
+\newcommand\seclabel[1]{\label{sec:#1}}
+
+\usepackage{verbatim}
+
+\newenvironment{smallverbatim}{\par\small\verbatim}{\endverbatim}
 
@@
 For some software, considerations of performance are paramount; this
-software is usually written C~or~C++. 
+software is usually written in C~or~C++. 
 For other software, considerations of convenience, readability, and
@@
 of~C++.
+\item
+Although we have had to write a new implementation of a standard
+algorithm, the cost was low, and the connection between code and
+algorithm is unusually strong (\secref{viterbi}).
 \item
@@
 function of the protein that gene codes for.
+\remark{I thought the query sequence was an amino-acid sequence.}
 The best known method of discovering such function is
@@
 
+\seclabel{viterbi}
+
 Homology-detection software most often used in one of two ways:
@@
 \emph{match state}.
+\remark{Time to introduce the elements of Viterbi's equations!}
 The match state determines the probability that the sequence under
@@
 \end{figure*}
+
+\emph{[This might be a place for comparison with standard
+implementations of Viterbi.]}
 

### 2012-03-08 mrfy_experience_report.tex
do not hyphenate hidden Markov model

@@
 
-The protein alignment is used to train a \emph{hidden-Markov model}.
+The protein alignment is used to train a \emph{hidden Markov model}.
 \remark{I've never seen this hypenated, in ANY literature.}
-A~hidden-Markov model is a probabilistic finite-state machine which 
+A~hidden Markov model is a probabilistic finite-state machine which 
 can assign a probability to any amino-acid sequence under test.
@@
   If~so, where should this jargon be introduced?}
-The states of the hidden-Markov model are determined by the alignment:
+The states of the hidden Markov model are determined by the alignment:
 \remark{I've tried for lots of parallel structure.
@@
 \item
-For a given column~$j$ of the alignment, the hidden-Markov model has a
+For a given column~$j$ of the alignment, the hidden Markov model has a
 \emph{match state}.
@@
 \item
-For every column of the alignment, the hidden-Markov model has a
+For every column of the alignment, the hidden Markov model has a
 \emph{deletion state}.
@@
 \item
-For every column of the alignment, the hidden-Markov model has an
+For every column of the alignment, the hidden Markov model has an
 \emph{insertion state}.
@@
 %%  corresponds to a position that a homologous protein should most likely fill.
-The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
+The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
 The probability associated with a specific state is conditional upon the previous
@@
 edge weight corresponding to the probability of a specific state transition.
-\figref{plan7} shows a hidden-Markov model derived from an alignment
+\figref{plan7} shows a hidden Markov model derived from an alignment
 with just four columns.
@@
 This directed graph represents the states of a
-hidden-Markov model in the ``Plan7'' format defined by the HMMER
+hidden Markov model in the ``Plan7'' format defined by the HMMER
 software package.
@@
 
-\caption{A hidden-Markov model trained from a four-column alignment}
+\caption{A hidden Markov model trained from a four-column alignment}
 
@@
 
-A hidden-Markov model like that shown in \figref{plan7} can be used to
+A hidden Markov model like that shown in \figref{plan7} can be used to
 determine the likelihood that a new protein shares structure and
@@
 HMMER computes the optimal \textit{parse}, or 
-alignment, of a protein sequence onto the directed graph of the hidden-Markov 
+alignment, of a protein sequence onto the directed graph of the hidden Markov 
 model.

### 2012-03-08 mrfy_experience_report.tex
starting working on a top-down explanation of beta strands

@@
 
-HMMER simply implements the Viterbi algorithm for computing the optimal parse 
+Our implementations of hidden Markov models and Viterbi's algorithm
+simply duplicate 
+functionality that is already implemented in the C++~package
+HMMER.
+These implementations form is just one component of our new tool, MRFy.
+
+A hidden Markov model effectively uses local information about
+amino-acid sequences.
+But because real proteins are three-dimensional structures,
+their amino-acid sequences are subject to \emph{nonlocal}
+constraints. 
+In~particular, when a real protein folds, groups of amino acids
+that are far away in the one-dimensional sequence can nevertheless be
+adjacent in three-dimensional space.
+Moreover, two such groups can become hydrogen-bonded to one another,
+effectively ``stuck together.''
+Such groups are called \emph{beta strands}.
+\remark{I hope this is relatively accurate. --NR}
+
+\emph{[Next up: how does this bonding phenomenon constrain sequences
+of homologous proteins?]}
+
+ simply implements the Viterbi algorithm for computing the optimal parse 
 of a protein sequence onto a hidden Markov model.

### 2012-03-08 mrfy_experience_report.tex
spelling

@@
 (While profiling with GHC could help remedy this, profiling was itself a 
-barrier, as previous mentioned.)
-We had to resort to polluting our code with cals to the \texttt{trace} function 
+barrier, as previously mentioned.)
+We had to resort to polluting our code with calls to the \texttt{trace} function 
 in order to track down where the out-of-bounds errors were occurring.

### 2012-03-08 mrfy_experience_report.tex
haskell types!

@@
 \remark{I've never seen this hypenated, in ANY literature.}
-A~hidden Markov model is a probabilistic finite-state machine which 
-can assign a probability to any amino-acid sequence under test.
-\remark{What is the biological or computational meaning of the
-  probability?}
+A~hidden-Markov model is a probabilistic finite-state machine which 
+can assign a probability to any query sequence of amino acids.
 Proteins whose sequences have higher probabilities are more likely to
 be homologous to the proteins in the alignment.
-\remark{Should we say ``homologous'' rather than  ``structurally
-  related''?
-  If~so, where should this jargon be introduced?}
-The states of the hidden Markov model are determined by the alignment:
+Mathematically, we represent a query sequence as
+$\left\{x_{1}...x_{n}\right\}$.
+Ideally, if we were writing something like the BioHaskell library, we would
+have begun with an algebraic datatype for amino acids deriving
+\texttt{Ix}, something like
+\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
+which would still allow us to index into an array or vector.
+However, since we wanted to maintain compatibility with the file format
+shared with previous software, we broke with idiomatic Haskell and represented
+our amino acids as 
+\begin{verbatim}
+type AA = Int
+\end{verbatim}
+and thus our query sequence as
+\begin{verbatim}
+type querySeq = Vector AA
+\end{verbatim}
+
+The states of the hidden-Markov model are determined by the alignment:
 \remark{I've tried for lots of parallel structure.
@@
 \emph{match state}.
-\remark{Time to introduce the elements of Viterbi's equations!}
-The match state determines the probability that the sequence under
-test has a particular amino acid in column~$j$.
+The match state determines the probability that the query sequence
+has a particular amino acid in column~$j$.
 The match state assigns high probability to the amino acids that
@@
 \emph{deletion state}.
-The deletion state determines the probability that the sequence under
-test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
+The deletion state determines the probability that the query sequence
+has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
 The deletion state has high probability if there are lots of gaps in
@@
 The probability associated with a specific state is conditional upon the previous
-state in the model.
+state in the model, so it is a \textit{transition probability}.
+Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
+for a state $s_{j}$.
+In Haskell, we use 
+\begin{verbatim}
+type logProb = Double
+data TProb  = TProb { logProbability::LogProb
+                    , fromState::HMMState
+                    , toState::HMMState
+                    }
+\end{verbatim}
+to represent the seven allowed state transitions.
+
 Thus, the hidden Markov model can be represented as a directed graph, with each
@@
 match state.
-
-We employed a variety of Haskell data types to represent the hidden Markov
-model.
-Ideally, if we were writing something like the BioHaskell library, we would
-have begun with an algebraic datatype for amino acids deriving
-\texttt{Ix}, something like
-\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
-which would still allow us to index into an array or vector.
-However, since we wanted to maintain compatibility with the file format
-shared with previous software, we broke with idiomatic Haskell and represented
-our amino acids as \texttt{Int}.
-Similarly, to avoid a great deal of data transformation, each state's emission
-and insertion probabilities were simply
-\begin{verbatim}type EProbs = V.Vector Double \end{verbatim} while the states themselves
-were algebraic datatypes: 
-\begin{verbatim}data HMMState = Match | Insert | Delete \end{verbatim}
-
-\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
-    I THINK WE'RE READY.]}
-
-% time for types.
-% first, ideal: amino acids as an algebraic DT
-% next, back off this: we're working with an existing input format
-% so, how is the graph represented (vector of state-transitions, 
-% vector of emissions)
-% so show those types
+Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
+for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
+query sequence element $x_{i}$.
+In Haskell, we represent a node as a group of transition probabilities
+\begin{verbatim}
+data TProbs = TProbs { m_m :: TProb
+                     , m_i :: TProb
+                     , m_d :: TProb
+                     , i_m :: TProb
+                     , i_i :: TProb
+                     , d_m :: TProb
+                     , d_d :: TProb
+                     }
+\end{verbatim}
+and a vector of emission probabilities for the match and insertion states.
+\begin{verbatim}
+type EProbs = V.Vector logProb
+data HmmNode = HmmNode { nodeNum :: Int
+                       , matEmissions :: EProbs
+                       , insEmissions :: EProbs
+                       , transitions :: TProbs
+                       }
+\end{verbatim}
 
@@
 %% go over this with NR; I can't make it flow.
-
-\begin{figure}
-\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
-\end{figure}
+% 
+% \begin{figure}
+% \lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
+% \end{figure}
+%
 
@@
 \begin{array}{l l}
-V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}I_{j}},\\
-V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}I_{j}},\\
+V_{j-1}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
+V_{j-1}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
 \end{array} \right.\nonumber\\
@@
 \begin{array}{l l}
-V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}I_{j}},\\
-V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}I_{j}},\\
+V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
+V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
 \end{array} \right.\nonumber\\

### 2012-03-08 mrfy_experience_report.tex
Some further edits to AG's writing

@@
 
-As inexperienced functional programmers, developing a complex piece of
-software in Haskell posed several barriers.
-\remark{NOAH: You did the profiling, could you make sure what I've said here
-is accurate and/or relevant? I didn't know what else to say. ---AG}
-The first of which was profiling with GHC.
-We previously mentioned that profiling with GHC produced ugly code with
-cost center annotations scattered throughout our code, but we also found it
-very difficult to profile code at all.
+As inexperienced functional programmers developing a complex piece of
+software in Haskell, we encountered some barriers.
+The first of these was profiling with GHC.
+We previously mentioned that profiling with GHC required annotating
+cost centers by hand, producing ugly code.
+Beyond that, we also found it very difficult to profile code at all.
 In particular, profiling required that all installed libraries be re-compiled
-with profiling support---which was a process that we could not replicate
+with profiling support,which was a process that we could not replicate
 consistently.
@@
 The problems we faced specifically were two-fold: How do we start testing
-a complex module? and How can we use QuickCheck to generate random inputs with 
+a complex module? How can we use QuickCheck to generate random inputs with 
 complex constraints? 
+We wished to test invariants around beta-strand placement 
+(potential solutions in our stochastic search), but found the QuickCheck generator
+to be difficult to write.
 
@@
 \subsection{What you must know to succeed}
-\remark{I don't like this. I really struggled with this section. ---AG}
 If others would like to follow in our footsteps, there are a few
@@
 Firstly, some experience with functional programming concepts is a necessity. 
-Specifically, coming from an imperative background, writing algorithms and 
-modeling data without the ability to mutate state were the most difficult 
-concepts to adapt to.
+Specifically, coming from an imperative background, we found transforming input data 
+from the HMMER file format without the ability to mutate state was difficult.
+In contrast, writing ``pure math'' like the Viterbi algorithm felt easier
+and more natural than it would have been in C++.
 
@@
 yet. 
+In contrast, other packages such as \texttt{Data.Memocombinators} were
+useful and mature.
 
@@
 typically helped us greatly.
-In particular, Data.Memocombinators made dynamic 
+In particular, \texttt{Data.Memocombinators} made dynamic 
 programming with Viterbi exquisitely simple and elegant.
-Control.Parallel.Strategies also made the parallelization of Viterbi trivial.
+\texttt{Control.Parallel.Strategies} also made the parallelization of Viterbi trivial.
 Most of the tools we used came straight from Hackage, and although 

### 2012-03-09 mrfy_experience_report.tex
bib file included, tweaked a couple code listings (now using verbatimsmall). Bib-related changes to .gitignore.

@@
 a family of algorithms called 
-\textit{hidden Markov models}~\cite{hmmer}.
+\textit{hidden Markov models}~\cite{Eddy:1998ut}.
 But in real biological systems,
@@
 In Haskell, we represent a node as a group of transition probabilities
-\begin{verbatim}
-data TProbs = TProbs { m_m :: TProb
-                     , m_i :: TProb
-                     , m_d :: TProb
-                     , i_m :: TProb
-                     , i_i :: TProb
-                     , d_m :: TProb
-                     , d_d :: TProb
-                     }
-\end{verbatim}
+\begin{smallverbatim}
+data TProbs = 
+     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
+            , i_m :: TProb, i_i :: TProb, d_m :: TProb
+            , d_d :: TProb}
+\end{smallverbatim}
 and a vector of emission probabilities for the match and insertion states.
-\begin{verbatim}
+\begin{smallverbatim}
 type EProbs = V.Vector logProb
@@
                        }
-\end{verbatim}
+\end{smallverbatim}
 
@@
 The most likely path, and its associated probability, are computed 
-using an algorithm developed by~\citet{Viterbi:1967}.
+using an algorithm developed by~\citet{Viterbi:1967hq}.
 
@@
 $N$ routinely contain several hundred to a few thousand elements.
-The types of 
-these states and probabilities are provided in Figure~\ref{viterbi_types}.
 
@@
 noted that the top-down version quite faithfully resembled the recurrence 
-relations found in a textbook~\cite{durbin} or the HMMER literature~\cite{eddy} 
+relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
 (see Equation \ref{viterbi_eqn}).
@@
 In 
-contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
-SMURFLite~\cite{smurflite}, our previous work.
+contrast, MRFy builds on the approach of SMURF~\cite{Menke:2010ti} and 
+SMURFLite~\cite{Daniels:2012}, our previous work.
 Like SMURF and SMURFLite, MRFy 
@@
 computes a probability based on observed frequencies of residues hydrogen 
-bonding in beta strands~\cite{betawrappro}.
+bonding in beta strands~\cite{Cowen:2002p588}.
 In between these beta strands, the 
@@
 Our previous computational biology projects include an enhancement to a protein 
-structural aligner that previously came from our group, Matt~\cite{matt}.
+structural aligner that previously came from our group, Matt~\cite{Menke:2008wu}.
 
@@
 
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
+\bibliography{mrfy_experience_report}
 

### 2012-03-09 mrfy_experience_report.tex
minor bib & manuscript tweaks (notes)

@@
 In Haskell, we use 
-\begin{verbatim}
+\begin{smallverbatim}
 type logProb = Double
@@
                     }
-\end{verbatim}
+\end{smallverbatim}
 to represent the seven allowed state transitions.
@@
 \subsection{Our group then}
+ - reimplementing algorithms
+  - we reimplemented and yet it was easy and pleasant
+  - past reimplementation of a variant of Viterbi DP, in haste, -> 700 line C function
+  - building on that reimplementation
+ - new implementations (Matt)
   - Modifying Formatt took nearly a year
@@
 design and development style but we did have to wrap our heads around 
-functional programming.
+functional programming.\remark{Andrew, look at the comments below this
+part of the manuscript. See what you can do with them.}
 

### 2012-03-09 mrfy_experience_report.tex
edits around beta strands

--- 2012-03-08 mrfy_experience_report.tex
@@
 newly discovered proteins. 
+\remark{``Homology'' means ``share a common ancestor.  I hope we can
+avoid this story. ---NR}
 
@@
 \remark{I've never seen this hypenated, in ANY literature.}
-A~hidden-Markov model is a probabilistic finite-state machine which 
-can assign a probability to any query sequence of amino acids.
+A~hidden Markov model is a probabilistic finite-state machine which 
+can assign a probability to any amino-acid sequence under test.
+\remark{What is the biological or computational meaning of the
+  probability?}
 Proteins whose sequences have higher probabilities are more likely to
 be homologous to the proteins in the alignment.
-Mathematically, we represent a query sequence as
-$\left\{x_{1}...x_{n}\right\}$.
-Ideally, if we were writing something like the BioHaskell library, we would
-have begun with an algebraic datatype for amino acids deriving
-\texttt{Ix}, something like
-\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
-which would still allow us to index into an array or vector.
-However, since we wanted to maintain compatibility with the file format
-shared with previous software, we broke with idiomatic Haskell and represented
-our amino acids as 
-\begin{verbatim}
-type AA = Int
-\end{verbatim}
-and thus our query sequence as
-\begin{verbatim}
-type querySeq = Vector AA
-\end{verbatim}
-
-The states of the hidden-Markov model are determined by the alignment:
+\remark{Should we say ``homologous'' rather than  ``structurally
+  related''?
+  If~so, where should this jargon be introduced?}
+The states of the hidden Markov model are determined by the alignment:
 \remark{I've tried for lots of parallel structure.
@@
 \emph{match state}.
-The match state determines the probability that the query sequence
-has a particular amino acid in column~$j$.
+\remark{Time to introduce the elements of Viterbi's equations!}
+The match state determines the probability that the sequence under
+test has a particular amino acid in column~$j$.
 The match state assigns high probability to the amino acids that
@@
 \emph{deletion state}.
-The deletion state determines the probability that the query sequence
-has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
+The deletion state determines the probability that the sequence under
+test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
 The deletion state has high probability if there are lots of gaps in
@@
 The probability associated with a specific state is conditional upon the previous
-state in the model, so it is a \textit{transition probability}.
-Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
-for a state $s_{j}$.
-In Haskell, we use 
-\begin{verbatim}
-type logProb = Double
-data TProb  = TProb { logProbability::LogProb
-                    , fromState::HMMState
-                    , toState::HMMState
-                    }
-\end{verbatim}
-to represent the seven allowed state transitions.
-
+state in the model.
 Thus, the hidden Markov model can be represented as a directed graph, with each
@@
 match state.
-Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
-for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
-query sequence element $x_{i}$.
-In Haskell, we represent a node as a group of transition probabilities
-\begin{verbatim}
-data TProbs = TProbs { m_m :: TProb
-                     , m_i :: TProb
-                     , m_d :: TProb
-                     , i_m :: TProb
-                     , i_i :: TProb
-                     , d_m :: TProb
-                     , d_d :: TProb
-                     }
-\end{verbatim}
-and a vector of emission probabilities for the match and insertion states.
-\begin{verbatim}
-type EProbs = V.Vector logProb
-data HmmNode = HmmNode { nodeNum :: Int
-                       , matEmissions :: EProbs
-                       , insEmissions :: EProbs
-                       , transitions :: TProbs
-                       }
-\end{verbatim}
+
+We employed a variety of Haskell data types to represent the hidden Markov
+model.
+Ideally, if we were writing something like the BioHaskell library, we would
+have begun with an algebraic datatype for amino acids deriving
+\texttt{Ix}, something like
+\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
+which would still allow us to index into an array or vector.
+However, since we wanted to maintain compatibility with the file format
+shared with previous software, we broke with idiomatic Haskell and represented
+our amino acids as \texttt{Int}.
+Similarly, to avoid a great deal of data transformation, each state's emission
+and insertion probabilities were simply
+\begin{verbatim}type EProbs = V.Vector Double \end{verbatim} while the states themselves
+were algebraic datatypes: 
+\begin{verbatim}data HMMState = Match | Insert | Delete \end{verbatim}
+
+\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
+    I THINK WE'RE READY.]}
+
+% time for types.
+% first, ideal: amino acids as an algebraic DT
+% next, back off this: we're working with an existing input format
+% so, how is the graph represented (vector of state-transitions, 
+% vector of emissions)
+% so show those types
 
@@
 %% go over this with NR; I can't make it flow.
-% 
-% \begin{figure}
-% \lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
-% \end{figure}
-%
+
+\begin{figure}
+\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
+\end{figure}
 
@@
 \begin{array}{l l}
-V_{j-1}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
-V_{j-1}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
+V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}I_{j}},\\
+V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}I_{j}},\\
 \end{array} \right.\nonumber\\
@@
 \begin{array}{l l}
-V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
-V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
+V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}I_{j}},\\
+V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}I_{j}},\\
 \end{array} \right.\nonumber\\
@@
 amino-acid sequences.
-But because real proteins are three-dimensional structures,
-their amino-acid sequences are subject to \emph{nonlocal}
+But real proteins are three-dimensional structures,
+and their amino-acid sequences are subject to \emph{nonlocal}
 constraints. 
-In~particular, when a real protein folds, groups of amino acids
-that are far away in the one-dimensional sequence can nevertheless be
+And when a real protein folds, groups of amino acids
+that are far away in the one-dimensional sequence can be
 adjacent in three-dimensional space.
-Moreover, two such groups can become hydrogen-bonded to one another,
-effectively ``stuck together.''
-Such groups are called \emph{beta strands}.
-\remark{I hope this is relatively accurate. --NR}
+We~are studying adjacent groups called \emph{beta strands}, which
+can become hydrogen-bonded to each other,
+and are effectively ``stuck together.''
 
@@
 
- simply implements the Viterbi algorithm for computing the optimal parse 
-of a protein sequence onto a hidden Markov model.
-Thus, the graphical model of 
-a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
-In 
-contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
-SMURFLite~\cite{smurflite}, our previous work.
-Like SMURF and SMURFLite, MRFy 
-pays attention to longer-range structural interactions than a simple hidden 
-Markov model can understand.
 In particular, these approaches ``nail down'' 
@@
 interactions between beta strands.
-Beta strands are regions of protein 
-structure that ``stick'' to one another.
 Because of their pairwise 
@@
 
-As inexperienced functional programmers developing a complex piece of
-software in Haskell, we encountered some barriers.
-The first of these was profiling with GHC.
-We previously mentioned that profiling with GHC required annotating
-cost centers by hand, producing ugly code.
-Beyond that, we also found it very difficult to profile code at all.
+As inexperienced functional programmers, developing a complex piece of
+software in Haskell posed several barriers.
+\remark{NOAH: You did the profiling, could you make sure what I've said here
+is accurate and/or relevant? I didn't know what else to say. ---AG}
+The first of which was profiling with GHC.
+We previously mentioned that profiling with GHC produced ugly code with
+cost center annotations scattered throughout our code, but we also found it
+very difficult to profile code at all.
 In particular, profiling required that all installed libraries be re-compiled
-with profiling support,which was a process that we could not replicate
+with profiling support---which was a process that we could not replicate
 consistently.
@@
 The problems we faced specifically were two-fold: How do we start testing
-a complex module? How can we use QuickCheck to generate random inputs with 
+a complex module? and How can we use QuickCheck to generate random inputs with 
 complex constraints?
-We wished to test invariants around beta-strand placement 
-(potential solutions in our stochastic search), but found the QuickCheck generator
-to be difficult to write.
 
@@
 (While profiling with GHC could help remedy this, profiling was itself a 
-barrier, as previously mentioned.)
-We had to resort to polluting our code with calls to the \texttt{trace} function 
+barrier, as previous mentioned.)
+We had to resort to polluting our code with cals to the \texttt{trace} function 
 in order to track down where the out-of-bounds errors were occurring.
@@
 \subsection{What you must know to succeed}
+\remark{I don't like this. I really struggled with this section. ---AG}
 If others would like to follow in our footsteps, there are a few
@@
 Firstly, some experience with functional programming concepts is a necessity. 
-Specifically, coming from an imperative background, we found transforming input data 
-from the HMMER file format without the ability to mutate state was difficult.
-In contrast, writing ``pure math'' like the Viterbi algorithm felt easier
-and more natural than it would have been in C++.
+Specifically, coming from an imperative background, writing algorithms and 
+modeling data without the ability to mutate state were the most difficult 
+concepts to adapt to.
 
@@
 yet.
-In contrast, other packages such as \texttt{Data.Memocombinators} were
-useful and mature.
 
@@
 typically helped us greatly.
-In particular, \texttt{Data.Memocombinators} made dynamic 
+In particular, Data.Memocombinators made dynamic 
 programming with Viterbi exquisitely simple and elegant.
-\texttt{Control.Parallel.Strategies} also made the parallelization of Viterbi trivial.
+Control.Parallel.Strategies also made the parallelization of Viterbi trivial.
 Most of the tools we used came straight from Hackage, and although 

### 2012-03-09 mrfy_experience_report.tex

--- 2012-03-09 mrfy_experience_report.tex
@@
 
+\emph{[GIT CONFLICT HERE:
  simply implements the Viterbi algorithm for computing the optimal parse 
@@
 Markov model can understand.
+]}
+
 In particular, these approaches ``nail down'' 

### 2012-03-09 mrfy_experience_report.tex
fixed git conflict

@@
 newly discovered proteins. 
-\remark{``Homology'' means ``share a common ancestor.  I hope we can
+\remark{``Homology'' means ``share a common ancestor.''  I hope we can
 avoid this story. ---NR}
@@
 
-\emph{[GIT CONFLICT HERE:
- simply implements the Viterbi algorithm for computing the optimal parse 
-of a protein sequence onto a hidden Markov model.
-Thus, the graphical model of 
-a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
-In 
-contrast, MRFy builds on the approach of SMURF~\cite{Menke:2010ti} and 
-SMURFLite~\cite{Daniels:2012}, our previous work.
-Like SMURF and SMURFLite, MRFy 
-pays attention to longer-range structural interactions than a simple hidden 
-Markov model can understand.
-]}
-
 In particular, these approaches ``nail down'' 
@@
 to become a Markov random field.
-SMURF and SMURFLite attempted to compute the 
+SMURF~\cite{Menke:2010ti} and SMURFLite~\cite{Daniels:2012}, our previous work,
+attempted to compute the 
 optimal parse of a protein sequence onto the directed graph representing a 

### 2012-03-09 mrfy_experience_report.tex
no hyphen in hidden Markov

@@
 \documentclass[preprint,nonatbib,blockstyle,nocopyrightspace,times]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
 
@@
 \remark{I've never seen this hypenated, in ANY literature.}
-A~hidden-Markov model is a probabilistic finite-state machine which 
+A~hidden Markov model is a probabilistic finite-state machine which 
 can assign a probability to any query sequence of amino acids.
@@
 
-The states of the hidden-Markov model are determined by the alignment:
+The states of the hidden Markov model are determined by the alignment:
 \remark{I've tried for lots of parallel structure.

### 2012-03-09 mrfy_experience_report.tex
more math

@@
 Markov random field using multidimensional dynamic programming.
+Mathematically, we can represent the probability of a query sequence 
+$x_{1}..x_{n}$ given a Markov random field with an additional term in the
+Viterbi relations.
+Specifically, each term $V_{j}^{\prime M}(i)$ becomes
+$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log Pr(x_{i} | x_{\pi(j)})$, where
+$x_{\pi(j)}$ is the amino acid in the query sequence that will be placed
+in the beta-strand position paired with position $j$.
+This nonlocal dependency prevents the problem from exhibiting optimal substructure,
+preventing dynamic programming from solving the problem quickly.
 SMURF's 

### 2012-03-09 mrfy_experience_report.tex
section 5 done, but it sucks.

@@
 \subsection{Our group then}
- - reimplementing algorithms
-  - we reimplemented and yet it was easy and pleasant
-  - past reimplementation of a variant of Viterbi DP, in haste, -> 700 line C function
-  - building on that reimplementation
- - new implementations (Matt)
-  - Modifying Formatt took nearly a year
-  - mutable data structures hard to repurpose
-  - one enhancement deemed infeasible
- - C++ failures most commonly segfaults, memory errors, heap exhaustion
+Programmers generally avoid reimplementing algorithms, but there may be good reasons for doing so.
+Although the HMMER software contains an mature C++ implementation of the Viterbi algorithm,
+a previous member of our group implemented \textit{multidimensional} dynamic programming for SMURF
+in haste, resulting in 12,000 lines of undocumented code (including one function of 700 lines).
+Though we are using the same Viterbi algorithm HMMER implements, we chose to reimplement it because
+we wished to explore the design space of stochastic search techniques using higher order functions
+in Haskell.
+We found that reimplementing the Viterbi algorithm in Haskell was easy and pleasant.
+In contrast, the programming effort to transform SMURF into SMURFLite, a minor evolution, required
+several months of research to simply understand the SMURF codebase.
+
+A set of enhancements to another large C++ implementation, the Matt~\cite{Menke:2008wu} structural
+aligner, required nearly a year of programming due to complex mutable data structures being 
+hard to repurpose.
+In fact, we deemed one enhancement infeasible.
+Our experience working with these C++ codebases largely involved the failure modes of
+segmentation faults, memory errors, and heap exhaustion.
 
 \subsection{Our group now}
- - Higher order functions militate towards more flexibility
- - Haskell failures are usually typecheck failures; at worst runtime errors around bounds checking
- - MRFy about 3 months of part-time work
+Implementing MRFy in Haskell allowed us great flexibility to explore the space of stochastic search
+techniques, initial guess heuristics, and search termination.
+Higher order functions militate towards more flexibility, so that adding several planned features 
+to MRFy should be easy.
+The first functioning version of MRFy took roughly three months of part-time work.
+The Haskell failure modes we have experienced have largely been typecheck errors; at the worst,
+we have seen runtime errors around bounds checking of arrays and vectors, though we wanted
+better debugging tools for these runtime errors.
 
 \subsection{Computational biology at large}
- - some good software, like HMMER
-
+There exists some excellent software in the computational biology community, though little of it
+has been written in Haskell.
+The training component of MRFy was modified from that of HMMER, and working with the HMMER
+codebase was relatively pleasant, as data structures and their mutators are well documented.
+BioHaskell exists (and we use \texttt{Bio.Sequence.Fasta} as a parser) but is not nearly as
+complete as BioPython or BioRuby, which are heavily used in the community.
+We hope that BioHaskell and other tools for computational biology in Haskell continue to mature.
 \section{What we think about it}

### 2012-03-09 mrfy_experience_report.tex
trying to restore parallel struture in the intro

--- 2012-03-09 mrfy_experience_report.tex
@@
 newly discovered proteins.
-\remark{``Homology'' means ``share a common ancestor.''  I hope we can
-avoid this story. ---NR}
 
@@
 nucleic acids in DNA, which we call a gene.
-Given a genetic sequence (the \textit{query sequence}), biologists wish to know the cellular
+Given a genetic sequence, biologists wish to know the cellular
 function of the protein that gene codes for.
-\remark{I thought the query sequence was an amino-acid sequence.}
 The best known method of discovering such function is
@@
 similar structure, which likely share similar function.
-Proteins that share structure and function are \textit{homologous}, and thus
-the process of identifying similar proteins to a query sequence is called 
+Proteins that share structure and function are expected to be
+descended from a common ancestor---in biological terms, \emph{homologous}---and
+thus
+the problem of identifying proteins similar to a query sequence is called 
 \textit{homology detection}.
@@
 Computational biologists detect homologies by building 
-algorithms which, given an amino-acid query sequence,
+algorithms which, given a \emph{query sequence} of amino acids,
 find known proteins of similar structure.
 When the similar proteins are formed from amino-acid sequences that
-are nearly identical to the sequence under test, the homologies can be
-detected by  
-simple string-matching algorithms.\remark{Can we just drop the
-  sentence about string matching?}
-When the similar proteins are formed from amino-acid sequences that
-are not too far in edit distance from the sequence under test, the homologies can be
+are not too different from the query sequence, the homology can be
 detected by
@@
 %for proteins that share similar structure and function.
+%
+To~enable you to understand our experience with Haskell,
+we~sketch not only the MRFy algorithm (\secref{mrfy})
+but also one of the older hidden-Markov algorithms (\secref{viterbi}),
+which is incorporated into~MRFy.
 
-\emph{[We need an argument here saying why an understanding of the
-    algorithms is relevant to the claims made in the experience
-    report, then say how we present the algorithms, then say what
-    experiential conclusions this presentation supports.]}
-
-We present an explanation of the algorithms relevant to MRFy, because there 
-exists a natural fit between the mathematics behind the algorithms and the 
-Haskell implementation of those algorithms.
-If the reader understands the 
-algorithms, he or she is more likely to appreciate the benefits of implementing 
-them in Haskell.
-We will first present the more traditional algorithm used in 
-hidden Markov models, followed by our variation on it and the stochastic search 
-approach.
-We conclude that implementations of mathematically-intensive 
+\remark{LOST: We conclude that implementations of mathematically-intensive 
 algorithms such as these lend themselves to the Haskell language, particularly 
 when the programmer needs a flexible implementation that simplifies 
-experimenting with variants of the algorithm.
+experimenting with variants of the algorithm.}
 
@@
 structures.
-The software is ``trained'' \remark{do we want to talk ``Training''?}
+Either way, 
+the software is \emph{trained}
 on a group of proteins that share function and structure.
@@
 their amino-acid sequences into a correspondence, also called an
-\emph{alignment}. This alignment signifies structural and evolutionary
+\emph{alignment}. 
+This alignment represents structural
 correspondence among individual amino acids in a set of homologous proteins.
@@
 different proteins.
-A protein may have a gap in a column of the alignment, which indicates
-that during the course of evolution, either that protein lost that amino
-acid, or other proteins gained an amino acid.
-If~a column contains just one or two gaps, the most likely explanation
-is that in one or two proteins, mutations in the underlying DNA
-deleted an amino acid from that position (\figref{alignment}).
-If~a column contains \emph{mostly} gaps, 
-the most likely explanation
-is that in one or two proteins, mutations in the underlying DNA
-\emph{inserted} an amino acid in that position (\figref{alignment}).
 
-The protein alignment is used to train a \emph{hidden Markov model}.
-\remark{I've never seen this hypenated, in ANY literature.}
+An alignment may contain \emph{gaps} (notated as dashes in
+\figref{alignment}) for some proteins.
+A~gap in row~2, column~$j$ indicates that as proteins evolved, either 
+protein~2 lost its amino acid in position~$j$, or 
+other proteins gained an amino acid in position~$j$.
+If~column~$j$ contains few gaps, 
+it~is considered a \emph{consensus column},
+and the few proteins with gaps probably lost amino acids via
+\emph{deletions}. 
+If~column~$j$ contains \emph{mostly} gaps, 
+it~is considered a \emph{non-consensus column},
+and the few proteins without gaps probably gained amino acids via
+\emph{insertions}.
+
+Once a protein alignment is constructed, it~is used to train a
+\emph{hidden Markov model}. 
 A~hidden Markov model is a probabilistic finite-state machine which 
-can assign a probability to any query sequence of amino acids.
-Proteins whose sequences have higher probabilities are more likely to
+can assign a probability to any query sequence.
+A~protein whose query sequence has a higher probability is more likely to %
 be homologous to the proteins in the alignment.
-Mathematically, we represent a query sequence as
-$\left\{x_{1}...x_{n}\right\}$.
-Ideally, if we were writing something like the BioHaskell library, we would
-have begun with an algebraic datatype for amino acids deriving
-\texttt{Ix}, something like
-\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
-which would still allow us to index into an array or vector.
-However, since we wanted to maintain compatibility with the file format
-shared with previous software, we broke with idiomatic Haskell and represented
-our amino acids as 
-\begin{verbatim}
-type AA = Int
-\end{verbatim}
-and thus our query sequence as
-\begin{verbatim}
-type querySeq = Vector AA
-\end{verbatim}
+We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
+where each $x_i$~is 
+an amino acid.
 
@@
 \item
-If a given column~$j$ of the alignment contains amino acids in a majority
-of the protein sequences, it is considered a \textit{consensus column}.
-\item
-For a given column~$j$ of the alignment, the hidden Markov model has a
+For each column~$j$ of the alignment, the hidden Markov model has a
 \emph{match state}.
@@
 acids.
-The match state is more probable if it corresponds to a consensus column.
+A~match state is more probable if it corresponds to a consensus column.
 \item
-For every column of the alignment, the hidden Markov model has a
+For each column~$j$ of the alignment, the hidden Markov model has a
 \emph{deletion state}.
@@
 has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
-The deletion state has high probability if there are lots of gaps in
-column~$j$, and low probability if there are few gaps in column~$j$.
-Thus, the deletion state is more likely in non-consensus columns.
+A~deletion state is more probable if it corresponds to a
+non-consensus column.
+% has high probability if there are lots of gaps in
+%column~$j$, and low probability if there are few gaps in column~$j$.
+%Thus, the deletion state is more likely in non-consensus columns.
 \item
-For every column of the alignment, the hidden Markov model has an
+For each column~$j$ of the alignment, the hidden Markov model has an
 \emph{insertion state}.
 The insertion state determines the probability that a query sequence
-has an ``extra'' amino acid that does not correspond to any
-consensus column in the alignment.
-Thus, the insertion state is more likely in non-consensus columns.
-The insertion state assigns high probability to amino acids that are
-frequently observed among all proteins, and low probability to amino acids
-that are rarely observed.
+has an ``extra'' amino acid.
+Like a match state, the insertion state assigns high probability to
+amino acids appearing in column~$j$.
+An~insertion state is more probable if it corresponds to a
+non-consensus column.
 \end{itemize}
@@
 to represent the seven allowed state transitions.
+
+We~write a query sequence as
+$x_{1}, \ldots, x_{n}$, where each~$x_i$ is an amino acid.
+In~idiomatic Haskell, we might represent each individual amino acid
+using a value of algebraic data type, thus:
+\begin{smallverbatim}
+data Amino = Ala | Cys | Asp | Glu | Gly | ...
+\end{smallverbatim}
+But our primary goal was to get something working quickly with a
+legacy file format that represents each amino acid as a small integer.
+Our types are therefore
+\begin{smallverbatim}
+type AA       = Int -- amino acid using HMMER coding
+type QuerySeq = Vector AA
+\end{smallverbatim}
 

### 2012-03-09 mrfy_experience_report.tex
fix math
huge code listings slain.

@@
 \begin{array}{l l}
-V_{j-1}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
-V_{j-1}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
+V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
+V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
 \end{array} \right.\nonumber\\
@@
 \begin{array}{l l}
-V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
-V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
+V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}},\\
+V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
 \end{array} \right.\nonumber\\
@@
 \begin{array}{l l}
-V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
-V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
+V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
+V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
 \end{array} \right.\nonumber\\
@@
 \begin{array}{l l}
-V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
-V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
+V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}},\\
+V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
 \end{array} \right.\nonumber\\
@@
 node or first observation (or both).
+The general case of this top-down, structurally recursive implementation closely
+resembles the Viterbi equations above.
+The match equation is thus:
+\begin{verbatim}
+viterbi s n o
+  -- consume an observation AND a node
+  | s == mat = let trans t prev = 
+                   (score + tProb (n-1) t + (eProb n o), mat:path)
+                    where (score, path) = viterbi prev (n - 1) (o - 1)
+                in minimum $ [
+                    trans m_m mat,
+                    trans i_m ins,
+                    trans d_m del
+                    ]
+\end{verbatim}
+% Similarly, the insertion equation is:
+% \begin{verbatim}                    
+%   -- consume an observation but not a node
+%   | s == ins = let trans t prev =
+%                    (score + tProb n t + (eProb n o), ins:path)
+%                     where (score, path) = viterbi prev n (o - 1)
+%                 in minimum [
+%                     trans m_i mat,
+%                     trans i_i ins
+%                     ]
+% \end{verbatim}
+% Finally, the deletion equation is:
+% \begin{verbatim}                    
+%   -- consume a node but not an observation
+%   | s == del = let trans t prev = 
+%                       (score + tProb (n-1) t, del:path)
+%                       where (score, path) = viterbi prev (n - 1) o
+%                 in minimum [
+%                     trans m_d mat,
+%                     trans d_d del
+%                     ]
+% \end{verbatim}
+
 In this second approach, we used the 
@@
 dynamic-programming approach in Haskell, which resulted in perspicuous code.
-
-\begin{figure*}
-\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
-\end{figure*}
 
@@
 
-We wrote a basic search function in only twenty-two lines of code; as one of 
-its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
+We wrote a basic search function in only twenty-two lines of code.
+The central logic of this search function is simply:
+\begin{verbatim}
+search' (s1:s2:seeds) oldPop hist age =
+  let newPop = mutate oldPop
+      score = fst $ minimum newPop
+      newHist = score : hist
+    in if accept newHist age then
+          if terminate newHist age then
+            (minimum newPop, newHist)
+          else
+            search' seeds newPop newHist (age + 1)
+       else
+         if terminate hist age then
+           (minimum oldPop, hist)
+         else
+           search' seeds oldPop hist (age + 1)
+\end{verbatim}
+As one of the parameters to the top-level \texttt{search} function,
+we provide a \texttt{SearchStrategy} data type, which we use to 
 define specific search techniques.
@@
 behavior is entirely deterministic.
-
-\begin{figure*}
-\lstinputlisting[label=search,caption=Search]{search_simple.hs}
-\end{figure*}
 

### 2012-03-09 mrfy_experience_report.tex
working on 'perspicuous haskell'

--- 2012-03-09 mrfy_experience_report.tex
@@
 \documentclass[preprint,nonatbib,blockstyle,nocopyrightspace,times]{sigplanconf}
+
+\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
+\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
 
@@
 
-The states of the hidden Markov model are determined by the alignment:
-\remark{I've tried for lots of parallel structure.
-Somebody let me know if any of this is right.  ---NR}
+A~hidden Markov model carries probabilities on its states and on its
+state transitions.
+Both the probabilities and the
+the states are determined by the alignment:
 \begin{itemize}
@@
 \emph{match state}.
-The match state determines the probability that the query sequence
-has a particular amino acid in column~$j$.
-The match state assigns high probability to the amino acids that
-appear in column~$j$, and a low but nonzero probability to other amino
-acids.
-A~match state is more probable if it corresponds to a consensus column.
+The match state contains a table $e_{M_j}(x)$ which gives the
+ probability that a homologous protein has amino acid~$x$ in
+ column~$j$.
+This probability is called an \emph{emission} probability.
 \item
@@
 has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
-A~deletion state is more probable if it corresponds to a
-non-consensus column.
-% has high probability if there are lots of gaps in
-%column~$j$, and low probability if there are few gaps in column~$j$.
-%Thus, the deletion state is more likely in non-consensus columns.
 \item 
@@
 \emph{insertion state}.
-The insertion state determines the probability that a query sequence
-has an ``extra'' amino acid.
-Like a match state, the insertion state assigns high probability to
-amino acids appearing in column~$j$.
-An~insertion state is more probable if it corresponds to a
-non-consensus column.
+The insertion state contains a table $e_{I_j}(x)$ which gives the
+probability that a homologous protein has gained amino acid~$x$ by
+insertion at column~$j$.
 \end{itemize}
-%%  A position that has residues in a majority of the sequences becomes 
-%%  a \textit{match} state in the hidden Markov model; this position may see a 
-%%  variety of amino acids due to evolutionary substitution, but it still 
-%%  corresponds to a position that a homologous protein should most likely fill.
 The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
-The probability associated with a specific state is conditional upon the previous
-state in the model, so it is a \textit{transition probability}.
-Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
-for a state $s_{j}$.
-In Haskell, we use 
-\begin{smallverbatim}
-type logProb = Double
-data TProb  = TProb { logProbability::LogProb
-                    , fromState::HMMState
-                    , toState::HMMState
-                    }
-\end{smallverbatim}
-to represent the seven allowed state transitions.
-
-We~write a query sequence as
-$x_{1}, \ldots, x_{n}$, where each~$x_i$ is an amino acid.
-In~idiomatic Haskell, we might represent each individual amino acid
+
+The particular hidden Markov model we use requires that every nonempty
+sequence of insertions be followed by a match, and similarly for
+deletions.
+\remark{citation needed?}
+In~other words, insertions and deletions cannot be mixed without an
+intervening match.
+This restriction means that between two adjacent columns, six
+transitions are possible.
+An~insertion state may also transition to itself.
+\begin{itemize} 
+\item
+A~transition into a match state in
+column~$j$ is more likely when column~$j$ is a consensus column.
+Depending on the predecessor state, 
+the probability of such a transition is notated
+$\txprobj M M$, $\txprobj I M$, or $\txprobj D M$.
+\item
+A~transition into a deletion state in
+ column~$j$ is more likely when column~$j$ is a non-consensus column.
+The probability of such a transition is notated
+$\txprobj M D$ or $\txprobj D D$.
+\item
+A~transition into an insertion state in
+ column~$j$ is relatively unlikely,
+but it is more likely when column~$j$ is a non-consensus column.
+The probability of such a transition is notated
+$\txprobjj M I$ or $\txprobjj I I$.
+\end{itemize}
+
+\subsection{Perspicuous Haskell}
+
+Computation of probabilities from hidden Markov models is well
+understood and is implemented in standard software packages.
+To~be able to use Haskell, we had to reimplement this standard
+algorithm.
+Haskell made it possible for us to write code that looks like the
+math,
+which made the code easy to write and gives us confidence that it is
+correct.
+
+We~begin with our representations.
+In~idiomatic Haskell, a query sequence is an immutable vector, and
+we might represent an individual amino acid~$x$
 using a value of algebraic data type, thus:
@@
 legacy file format that represents each amino acid as a small integer.
-Our types are therefore
+Our representations are therefore
 \begin{smallverbatim}
@@
 
-Thus, the hidden Markov model can be represented as a directed graph, with each
-edge weight corresponding to the probability of a specific state transition.
-\figref{plan7} shows a hidden Markov model derived from an alignment
-with just four columns.
-Each column in the original alignment corresponds to a \emph{node},
-which is a triplet holding an insertion state, a deletion state, and a
-match state.
-Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
-for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
-query sequence element $x_{i}$.
-In Haskell, we represent a node as a group of transition probabilities
+A~transition probability is simply a number.
+For accuracy in computation, we add log probabilities instead of
+multiplying probabilities.
+And to distinguish a transition probability from other numbers, we wrap it in
+the value constructor \texttt{TProb}:
+\begin{smallverbatim}
+type LogProb = Double
+data TProb  = TProb { logProbability :: LogProb }
+\end{smallverbatim}
+
+We group transition probabilities into 7-tuples.
+Each tuple corresponds to a ``node'' in the hidden Markov model
+(a~dashed ovel in \figref{plan7}), and to a column of the original
+alignment.
+As~noted above, 
+a match state may transition to any state in the successor node;
+an insertion state may transition to itself or to its successor match
+state;
+and
+a~deletion may transition to its successor match
+state or deletion state.
 \begin{smallverbatim}
@@
      TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
-            , i_m :: TProb, i_i :: TProb, d_m :: TProb
-            , d_d :: TProb}
+            , i_m :: TProb, i_i :: TProb
+            , d_m :: TProb, d_d :: TProb }
 \end{smallverbatim}
+
+The
 and a vector of emission probabilities for the match and insertion states.
 \begin{smallverbatim}
-type EProbs = V.Vector logProb
+type EProbs = V.Vector LogProb
 data HmmNode = HmmNode { nodeNum :: Int

### 2012-03-09 mrfy_experience_report.tex
wordsmtihing, formatting, subsectino headers

@@
 
+\begin{figure}
+\ifpdfmadness
+\centerline{\includegraphics[width=6cm]{alignment.pdf}} 
+\else
+\centerline{\includegraphics[width=6cm]{alignment.eps}} 
+\fi
+
+\caption{A~structural alignment of four proteins}
+\label{alignment} 
+\end{figure}
+
 A~hidden Markov model carries probabilities on its states and on its
@@
 
+\begin{figure} 
+\ifpdfmadness
+\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
+\else
+\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
+\fi
+
+This model, trained from a four-column alignment,
+has begin and end states $B$~and~$E$,
+as well as four
+nodes, each containing 
+an insertion state~$I$, 
+a match state~$M$, and a
+deletion state~$D$.
+
+\caption{A hidden Markov model}
+
+\label{plan7} \end{figure}
+
 \subsection{Perspicuous Haskell}
@@
 \end{smallverbatim}
-
-\begin{figure}
-\ifpdfmadness
-\centerline{\includegraphics[width=6cm]{alignment.pdf}} 
-\else
-\centerline{\includegraphics[width=6cm]{alignment.eps}} 
-\fi
-
-\caption{A~structural alignment of four proteins}
-\label{alignment} 
-\end{figure}
-
-\begin{figure} 
-\ifpdfmadness
-\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
-\else
-\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
-\fi
-
-This directed graph represents the states of a
-hidden Markov model in the ``Plan7'' format defined by the HMMER
-software package.
-%% Only 7 of the possible 9 state transitions are allowed, which 
-%% simplifies the dynamic programming.
-Each insertion state~($I$) represents a potential mutation that 
-inserts an amino acid;
-each deletion state~($D$) represents a potental mutation that
-deletes an amino acid;
-and each match state~($M$) represents either
-an unmutated amino acid or a potential mutation that
-substitutes one amino acid for another.
-A~match state or an insertion state includes a table that assigns a
-nonzero probability to each amino acid.
-(The figure shows tables for the first two states only.)
-A~triplet of an insertion state, deletion state, 
-and match state formas a \emph{node}, which corresponds to a column in
-the alignment used to train the model.
-(Each node is shown in a dashed oval.)
-
-\caption{A hidden Markov model trained from a four-column alignment}
-
-\label{plan7} \end{figure}
 
@@
 
-\begin{eqnarray}    
-V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
+\begin{figure*}
+\def\goo{18pt}
+\def\gum{14pt}
+\[
+\begin{array}{l@{}c@{}l}
+V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
 \begin{array}{l l}
-V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
-V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
+V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}}\\
+V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}}\\
 V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
-\end{array} \right.\nonumber\\
-V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
+\end{array} \right.\\[\goo]
+V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
 \begin{array}{l l}
-V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
-V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
-\end{array} \right.\nonumber\\
-V_{j}^{D}(i) = max \left\{
+V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}}\\
+V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}}\\
+\end{array} \right.\\[\gum]
+V_{j}^{D}(i) &=& \max \left\{
 \begin{array}{l l}
-V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}},\\
+V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}}\\
 V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
-\end{array} \right.\nonumber\\
-\end{eqnarray}\label{viterbi_eqn}
-
-\begin{eqnarray}    
-V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
+\end{array} \right.\\
+\end{array}
+\qquad
+\begin{array}{l@{}c@{}l}
+V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
 \begin{array}{l l}
-V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
-V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
+V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}}\\
+V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}}\\
 V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
-\end{array} \right.\nonumber\\
-V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
+\end{array} \right.\\[\goo]
+V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
 \begin{array}{l l}
-V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
-V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
-\end{array} \right.\nonumber\\
-V_{j}^{\prime D}(i) = min \left\{
+V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}}\\
+V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}}\\
+\end{array} \right.\\[\gum]
+V_{j}^{\prime D}(i) &=& \min \left\{
 \begin{array}{l l}
-V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}},\\
+V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}}\\
 V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
-\end{array} \right.\nonumber\\
-\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
-\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
-\end{eqnarray}\label{viterbi_log_eqn}
+\end{array} \right.\\
+\end{array}
+\]
+
+\centerline{$a'_{s} = - \log a_{s} \nonumber$,
+$e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}$,
+and 
+$V_j^{\prime M}(i) = - V_j^{M}(i)$}
+\caption{Viterbi's equations, in original and negated forms}
+\label{viterbi}
+\end{figure*}
 
@@
 
-\subsection{MRFy Implementation}
+%\subsection{MRFy Implementation}
 
@@
 annealing, multistart simulated annealing, and a genetic algorithm.
+
+\subsection{Higher-order functions}
 
@@
 
-\subsection{Ugly code}
-
 Despite the beauty of some of the code we were able to write, we found several 
@@
 
+\subsection{Awkward Haskell}
+
 Another source of ugliness resulted from adding cost-center annotations when we 

### 2012-03-10 mrfy_experience_report.tex
Some wordsmithing and reorg of the MRF intro

@@
 
-\remark{Dropped the section on ``our background,'' but the information
-  should reappear when and where we discuss barriers to entry in more
-  detail. ---NR}
-
 \section{The Software}
@@
 $N$ routinely contain several hundred to a few thousand elements.
-
-%% break the viterbi types up and insert into the Plan7 explanation.
-%% go over this with NR; I can't make it flow.
-% 
-% \begin{figure}
-% \lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
-% \end{figure}
-%
 
@@
 
-\emph{[This might be a place for comparison with standard
-implementations of Viterbi.]}
+Standard implementations of the Viterbi algorithm, such as in HMMER, spend much of their code
+managing the dynamic programming tables, as even when memoization libraries are available,
+they are rarely used.
+Furthermore, standard implementations in imperative languages typically avoid the natural
+recursive representations because of the associated costs in languages such as C++.
+The efficient machine code that GHC generates for recursive functions allowed us to write
+a program that faithfully reflected the mathematics, without paying great costs in run-time
+efficiency.
 
@@
 HMMER.
-These implementations form is just one component of our new tool, MRFy.
+These implementations form just one component of our new tool, MRFy.
 
@@
 
-\emph{[Next up: how does this bonding phenomenon constrain sequences
-of homologous proteins?]}
-
-In particular, these approaches ``nail down'' 
+During the course of evolution, a mutation in an amino acid that participates
+in this beta-strand pairing may cause a protein to fail to fold into a useful
+structure, unless that mutation is accompanied by a \textit{compensatory}
+mutation in the amino acid with which it is paired. 
+Thus, homologous
+proteins---those that share structure and function---tend to \textit{conserve}
+such pairwise-conditional amino acids.
+
+In order to model these non-local interactions, computational biologists
+have extended hidden Markov models to include these non-local interactions,
+modeled as a \textit{dependency graph}, which transforms the model into a
+\textit{Markov random field}. These approaches ``nail down'' 
 certain subsequences of the model, corresponding to the non-local hydrogen-bond 
 interactions between beta strands.
-Because of their pairwise 
-interactions, they are highly conserved over the course of evolution.
-Moreover, 
-again because of these pairwise interactions, a change in one amino acid in a 
-beta strand may require a change in a corresponding, hydrogen-bonded residue in 
-another beta strand if the protein is to remain stable.
 The architecture of the 
@@
 
-These non-local interactions cause what had been a simple hidden Markov model 
-to become a Markov random field.
 SMURF~\cite{Menke:2010ti} and SMURFLite~\cite{Daniels:2012}, our previous work,
-attempted to compute the 
+compute the 
 optimal parse of a protein sequence onto the directed graph representing a 

### 2012-03-10 mrfy_experience_report.tex
wordsmithing, a few additions, and a steaming dump of an abstract. We need to lose half a page.

@@
 \begin{abstract}
-HELLO?  DRAFT OF AN ABSTRACT?
+In computational biology, the languages of choice for solving ad-hoc problems are high-level
+``scripting'' languages such as Python or domain-specific statistics languages such as R,
+while the language of choice for solving high-performance problems is C++.
+This dichotomy creates a difficulty when computational biologists wish to experiment with a
+variety of approaches to a problem, only some of which are likely to prove fruitful.
+This experience report details our use of Haskell in implementing a high-performance,
+computational biology application which is the first implementation of a particular
+novel algorithm for protein structure and function prediction. We discuss the highlights
+and difficulties of this project, as well as the potential improvements the Haskell
+community might make that would ease some of these difficulties.
 \end{abstract}
@@
 thus
-the problem of identifying proteins similar to a query sequence is called 
+the problem of identifying proteins similar to a \textit{query sequence} is called 
 \textit{homology detection}.
@@
 
-Homology-detection software most often used in one of two ways:
+Homology-detection software is most often used in one of two ways:
 to test a hypothesis about 
@@
 and the few proteins with gaps probably lost amino acids via
-\emph{deletions}. 
+\emph{deletions}.\remark{NMD: Missing is the fact that all these models
+are directionless; they don't care whether something was gained or lost
+over time. Perhaps this doesn't matter.}
 If~column~$j$ contains \emph{mostly} gaps, 
@@
 model.
-A parse maps each amino acid of the sequence under test onto a state
+A parse maps each amino acid of the query sequence onto a state
 of the model; equivalently, a parse is a \textit{path} through the model.
@@
 viterbi s n o
-  -- consume an observation AND a node
-  | s == mat = let trans t prev = 
-                   (score + tProb (n-1) t + (eProb n o), mat:path)
-                    where (score, path) = viterbi prev (n - 1) (o - 1)
+  | s == mat = 
+       let trans t prev = 
+        (scr + tProb (n-1) t + (eProb n o), mat:p)
+        where (scr, p) = viterbi prev (n-1) (o-1)
        in minimum $ [

### 2012-03-10 mrfy_experience_report.tex
wordsmithing.

@@
 productivity are more important;
-this software is usually written in a ``scripting'' language like
-Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
+this software is usually written in a high-level, dynamically-typed
+``scripting'' or domain-specific language like
+Perl, Python, Ruby, SPSS, or~R.\remark{Don't forget audience.}
 In~this Experience Report, we show that computational biologists can

### 2012-03-10 mrfy_experience_report.tex
minor tweaks
sketch of an alternate intro, to be considered later

@@
 newly discovered proteins.
+
+\emph{[Alternate introduction, trying to tie the points more closely
+to the content of the paper:\\
+In~this paper, we report on an attempt to replace both these families
+of languages with Haskell: 
+\begin{itemize}
+\item
+Although had to write a new implementation of a standard
+algorithm, the cost was low, and the connection between code and
+algorithm is unusually strong (\secref{viterbi}).
+The performance of our new implementation is comparable to the
+performance of~C++.
+\item
+As expected, higher-order functions made it unusually easy to
+implement and experiment with a family of related algorithms
+(\secref{hofs}).
+\item
+Haskell is surrounded by a penumbra of libraries and tools that
+promise powerful abstractions.
+Some deliver very high leverage, but others are not ready for prime
+time (\secref{penumbra}).
+We~found it difficult to tell which is which in advance.
+\item
+It is possible to use Haskell successfully for research even with a
+very modest background in functional programming. 
+\item
+The only significant impediment that Haskell pas presented to our
+research has been the difficulty of understanding and tuning the
+performance of Haskell programs.
+\end{itemize}
+]}
 

### 2012-03-10 mrfy_experience_report.tex
start of section 3 edits

@@
 \label{plan7} \end{figure}
-
-\subsection{Perspicuous Haskell}
-
-Computation of probabilities from hidden Markov models is well
-understood and is implemented in standard software packages.
-To~be able to use Haskell, we had to reimplement this standard
-algorithm.
-Haskell made it possible for us to write code that looks like the
-math,
-which made the code easy to write and gives us confidence that it is
-correct.
-
-We~begin with our representations.
-In~idiomatic Haskell, a query sequence is an immutable vector, and
-we might represent an individual amino acid~$x$
-using a value of algebraic data type, thus:
-\begin{smallverbatim}
-data Amino = Ala | Cys | Asp | Glu | Gly | ...
-\end{smallverbatim}
-But our primary goal was to get something working quickly with a
-legacy file format that represents each amino acid as a small integer.
-Our representations are therefore
-\begin{smallverbatim}
-type AA       = Int -- amino acid using HMMER coding
-type QuerySeq = Vector AA
-\end{smallverbatim}
-
-A~transition probability is simply a number.
-For accuracy in computation, we add log probabilities instead of
-multiplying probabilities.
-And to distinguish a transition probability from other numbers, we wrap it in
-the value constructor \texttt{TProb}:
-\begin{smallverbatim}
-type LogProb = Double
-data TProb  = TProb { logProbability :: LogProb }
-\end{smallverbatim}
-
-We group transition probabilities into 7-tuples.
-Each tuple corresponds to a ``node'' in the hidden Markov model
-(a~dashed ovel in \figref{plan7}), and to a column of the original
-alignment.
-As~noted above, 
-a match state may transition to any state in the successor node;
-an insertion state may transition to itself or to its successor match
-state;
-and
-a~deletion may transition to its successor match
-state or deletion state.
-\begin{smallverbatim}
-data TProbs = 
-     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
-            , i_m :: TProb, i_i :: TProb
-            , d_m :: TProb, d_d :: TProb }
-\end{smallverbatim}
-
-The
-and a vector of emission probabilities for the match and insertion states.
-\begin{smallverbatim}
-type EProbs = V.Vector LogProb
-data HmmNode = HmmNode { nodeNum :: Int
-                       , matEmissions :: EProbs
-                       , insEmissions :: EProbs
-                       , transitions :: TProbs
-                       }
-\end{smallverbatim}
-
-A hidden Markov model like that shown in \figref{plan7} can be used to
-determine the likelihood that a new protein shares structure and
-evolutionary history with the proteins used to train the model.
-(The new protein's amino-acid sequence is known, but its structure and
-function are unknown.)
-Such a likelihood can be computed by
-a~widely used software package called HMMER (``hammer'').
-HMMER computes the optimal \textit{parse}, or 
-alignment, of a protein sequence onto the directed graph of the hidden Markov 
-model.
-A parse maps each amino acid of the query sequence onto a state
-of the model; equivalently, a parse is a \textit{path} through the model.
-Each edge is weighted with a probability, and each emitting 
-state (match or insertion) maps each amino acid to a probability,
-and the 
-probability of a given path is the product of these individual probabilities.
-The most likely path, and its associated probability, are computed 
-using an algorithm developed by~\citet{Viterbi:1967hq}.
-
-The Viterbi algorithm comprises a set of recurrence 
-relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
-probability of an observed sequence (in this case, of amino acid ``letters'') 
-being emitted by a particular finite-state automaton (the hidden Markov model).
-
-To be efficient, the Viterbi 
-algorithm relies on dynamic programming, or memoization, and the resulting 
-asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
-distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
-letters in the protein sequence.
-In our particular problem domain, both $M$ and 
-$N$ routinely contain several hundred to a few thousand elements.
-
-The textbook Viterbi recurrence relations refer to logs of probabilities, so 
-that the terms may be added rather than multiplying raw probabilities.
-For 
-reasons including floating point underflow, the HMMER software (with which we 
-maintain file format compatibility) stores all probabilities in a trained HMM 
-file as negative natural logs.
-Thus, the Viterbi recurrence relations are 
-simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
-and because they are \textit{negative} logs, the problem transforms from 
-maximization to minimization.
 
@@
 
+\subsection{Perspicuous Haskell}
+
+Given a hidden Markov model, 
+an~established software package called HMMER (``hammer'') 
+can compute the probability
+that a new protein shares structure and
+evolutionary history with the proteins used to train the model.
+The computation finds the most likely path through the hidden Markov model.
+To~make best use of floating-point arithmetic, the software computes
+the \emph{logarithm} of the probability of each path, by summing
+the logs of the 
+probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
+(The logarithm function is monotonic, so the path that maximizes the
+log of the probability is the most likely path.)
+The~specification is shown on the left-hand side of \figref{viterbi}.
+The probability $V_j^M(i)$ \emph{[REPRESENTS WHAT EXACTLY?]}.
+
+To~be able to use Haskell, we had to reimplement this standard
+algorithm.
+Haskell made it possible for us to write code that looks like the
+math,
+which made the code easy to write and gives us confidence that it is
+correct.
+We~begin with our representations.
+
+In~idiomatic Haskell, 
+we might represent an individual amino acid~$x$
+using a value of algebraic data type, thus:
+\begin{smallverbatim}
+data Amino = Ala | Cys | Asp | Glu | Gly | ...
+\end{smallverbatim}
+But our primary goal was to get something working quickly with a
+legacy file format that represents each amino acid as a small integer.
+Our representation is therefore
+\begin{smallverbatim}
+type AA       = Int -- amino acid using HMMER coding
+\end{smallverbatim}
+Either way, a query sequence is (and should be) an immutable vector of
+amino acids.
+%type QuerySeq = Vector AA
+%\end{smallverbatim}
+
+A~transition probability is simply a number.
+For accuracy in computation, we add log probabilities instead of
+multiplying probabilities.
+And to distinguish a transition probability from other numbers, we wrap it in
+the value constructor \texttt{TProb}:
+\begin{smallverbatim}
+type LogProb = Double
+data TProb  = TProb { logProbability :: LogProb }
+\end{smallverbatim}
+
+We group transition probabilities into 7-tuples.
+Each tuple corresponds to a ``node'' in the hidden Markov model
+(a~dashed ovel in \figref{plan7}), and to a column of the original
+alignment.
+As~noted above, 
+a match state may transition to any state in the successor node;
+an insertion state may transition to itself or to its successor match
+state;
+and
+a~deletion may transition to its successor match
+state or deletion state.
+\begin{smallverbatim}
+data TProbs = 
+     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
+            , i_m :: TProb, i_i :: TProb
+            , d_m :: TProb, d_d :: TProb }
+\end{smallverbatim}
+
+The
+and a vector of emission probabilities for the match and insertion states.
+\begin{smallverbatim}
+type EProbs = V.Vector LogProb
+data HmmNode = HmmNode { nodeNum :: Int
+                       , matEmissions :: EProbs
+                       , insEmissions :: EProbs
+                       , transitions :: TProbs
+                       }
+\end{smallverbatim}
+
+The Viterbi algorithm comprises a set of recurrence 
+relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
+probability of an observed sequence (in this case, of amino acid ``letters'') 
+being emitted by a particular finite-state automaton (the hidden Markov model).
+
+To be efficient, the Viterbi 
+algorithm relies on dynamic programming, or memoization, and the resulting 
+asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
+distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
+letters in the protein sequence.
+In our particular problem domain, both $M$ and 
+$N$ routinely contain several hundred to a few thousand elements.
+
+The textbook Viterbi recurrence relations refer to logs of probabilities, so 
+that the terms may be added rather than multiplying raw probabilities.
+For 
+reasons including floating point underflow, the HMMER software (with which we 
+maintain file format compatibility) stores all probabilities in a trained HMM 
+file as negative natural logs.
+Thus, the Viterbi recurrence relations are 
+simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
+and because they are \textit{negative} logs, the problem transforms from 
+maximization to minimization.
+
 We implemented the Viterbi algorithm in two distinct ways.
@@
 The match equation is thus:
-\begin{verbatim}
+\begin{smallverbatim}
 viterbi s n o
@@
            ]
-\end{verbatim}
+\end{smallverbatim}
 % Similarly, the insertion equation is:
@@
 the Viterbi function, we wrote a higher-order function type 
-\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type.
+\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type:
+\begin{verbatim}
+consPath :: ScorePathCons a
+consPath x xs = x:xs
+
+consNoPath :: ScorePathCons a
+consNoPath _ _ = []
+\end{verbatim}
 
@@
 having two distinct functions.
-
-\lstinputlisting[label=scorepathcons,caption=ScorePathCons]{scorepathcons.hs}
 

### 2012-03-10 mrfy_experience_report.tex
add code for memo combinators, use smallverbatim

@@
 (again, a three-dimensional array).
+This combinator approach was easy to apply to the code after the fact:
+\begin{smallverbatim}
+viterbi state node obs = 
+  Memo.memo3 (Memo.arrayRange (mat, end)) 
+             (Memo.arrayRange (0, numNodes))
+             (Memo.arrayRange (0, seqlen)) 
+             viterbi' state node obs
+\end{smallverbatim}
 We determined that the run-time 
@@
 The central logic of this search function is simply:
-\begin{verbatim}
+\begin{smallverbatim}
 search' (s1:s2:seeds) oldPop hist age =
@@
            search' seeds oldPop hist (age + 1)
-\end{verbatim}
+\end{smallverbatim}
 As one of the parameters to the top-level \texttt{search} function,

### 2012-03-10 mrfy_experience_report.tex
another small verbatim instance
coming dangerously near collision with noah

--- 2012-03-10 mrfy_experience_report.tex
@@
 \newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
+\newcommand\alignwidth{\ensuremath R} % number of columns in an alignment
 
@@
 an amino acid.
+$N$~need not be the same as the number of columns in the alignment,
+which we write~\alignwidth.
 
@@
 
-\caption{A~structural alignment of four proteins}
+\caption{A~structural alignment of four proteins $(\alignwidth = 4)$}
 \label{alignment} 
@@
 For each column~$j$ of the alignment, the hidden Markov model has a
-\emph{match state}.
+\emph{match state}~$M_j$.
 The match state contains a table $e_{M_j}(x)$ which gives the
@@
 For each column~$j$ of the alignment, the hidden Markov model has a
-\emph{deletion state}.
+\emph{deletion state}~$D_j$.
 The deletion state determines the probability that the query sequence
@@
 For each column~$j$ of the alignment, the hidden Markov model has an
-\emph{insertion state}.
+\emph{insertion state}~$I_j$.
 The insertion state contains a table $e_{I_j}(x)$ which gives the
@@
 
-This model, trained from a four-column alignment,
+This model
 has begin and end states $B$~and~$E$,
@@
 
-\caption{A hidden Markov model}
+\caption{A hidden Markov model $(\alignwidth = 4)$}
 
@@
 V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
-\end{array} \right.\\
+\end{array} \right.\\[\gum]
+\multicolumn3{l}{%
+a'_{s} = - \log a_{s} 
+\qquad
+e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}
+\qquad
+V_j^{\prime M}(i) = - V_j^{M}(i)
+}
+\\
 \end{array}
@@
 
-\centerline{$a'_{s} = - \log a_{s} \nonumber$,
-$e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}$,
-and 
-$V_j^{\prime M}(i) = - V_j^{M}(i)$}
 \caption{Viterbi's equations, in original and negated forms}
@@
 The~specification is shown on the left-hand side of \figref{viterbi}.
-The probability $V_j^M(i)$ \emph{[REPRESENTS WHAT EXACTLY?]}.
-
-To~be able to use Haskell, we had to reimplement this standard
-algorithm.
+The probability $V_j^M(i)$ represents the probability of the most
+likely path of the first~$i$ amino acids in the query sequence,
+terminating with placement of amino acid $x_i$ in state~$M_j$.
+The probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
+
+To~be able to use Haskell, we had to reimplement the standard
+algorithm for solving Viterbi's equations.
 Haskell made it possible for us to write code that looks like the
@@
 
-A~transition probability is simply a number.
-For accuracy in computation, we add log probabilities instead of
-multiplying probabilities.
-And to distinguish a transition probability from other numbers, we wrap it in
+The log of a probability is simply a number.
+To distinguish a transition probability from other numbers, we wrap it in
 the value constructor \texttt{TProb}:
@@
 \end{smallverbatim}
-
 We group transition probabilities into 7-tuples.
@@
 
-The
-and a vector of emission probabilities for the match and insertion states.
+The representation of a node includes the probabilities of transitions
+into the three states of that node, plus the tables
+of emission probabilities for the match and insertion states.
 \begin{smallverbatim}
-type EProbs = V.Vector LogProb
-data HmmNode = HmmNode { nodeNum :: Int
+type EProbs  = Vector LogProb
+data HmmNode = HmmNode { transitions  :: TProbs
                        , matEmissions :: EProbs
                        , insEmissions :: EProbs
-                       , transitions :: TProbs
                        }
@@
 
-The Viterbi algorithm comprises a set of recurrence 
-relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
-probability of an observed sequence (in this case, of amino acid ``letters'') 
-being emitted by a particular finite-state automaton (the hidden Markov model).
-
-To be efficient, the Viterbi 
-algorithm relies on dynamic programming, or memoization, and the resulting 
-asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
-distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
-letters in the protein sequence.
-In our particular problem domain, both $M$ and 
-$N$ routinely contain several hundred to a few thousand elements.
-
-The textbook Viterbi recurrence relations refer to logs of probabilities, so 
-that the terms may be added rather than multiplying raw probabilities.
-For 
-reasons including floating point underflow, the HMMER software (with which we 
-maintain file format compatibility) stores all probabilities in a trained HMM 
-file as negative natural logs.
-Thus, the Viterbi recurrence relations are 
-simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
-and because they are \textit{negative} logs, the problem transforms from 
-maximization to minimization.
-
-We implemented the Viterbi algorithm in two distinct ways.
-We first wrote a 
-bottom-up dynamic-programming version, which began with the first node of the 
-hidden Markov model and first observation and built up the three-dimensional 
-array of possible paths through the model.
-The array containing the paths 
-served as the memoization table.
+In~Viterbi's equations,
+the probability in each state is a simple function of the probability
+in its predecessor states, 
+and all probabilities can be computed by a classic dynamic-programming
+algorithm
+which starts at the begin state,
+computes probabilites in nodes $1$~through~$\alignwidth$ in
+succession, and terminates at the end state.
+We~implemented this algorithm, storing the probabilities in an array.
+\remark{Why is the number of dimensions 3 and not 2 (indices $i$, $j$,
+and what??) and why do we care?}
+The cost is
+$O(|N|\times|\alignwidth|)$;
+in our work, $\alignwidth$~and~$N$ range from several hundred to a few
+thousand.
+
+%%  For 
+%%  reasons including floating point underflow, the HMMER software (with which we 
+%%  maintain file format compatibility) stores all probabilities in a trained HMM 
+%%  file as negative natural logs.
+%%  Thus, the Viterbi recurrence relations are 
+%%  simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
+%%  and because they are \textit{negative} logs, the problem transforms from 
+%%  maximization to minimization.
+
+\emph{[WHAT WAS UNSATISFACTORY ABOUT THE FIRST IMPLEMENTATION]}
+
 We next wrote a top-down dynamic-programming 

### 2012-03-10 mrfy_experience_report.tex

@@
 \lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type:
-\begin{verbatim}
+\begin{smallverbatim}
 consPath :: ScorePathCons a
@@
 consNoPath _ _ = []
-\end{verbatim}
+\end{smallverbatim}
 

### 2012-03-10 mrfy_experience_report.tex
refactoring noah code

@@
 
+\newcommand\naive{na\"\i ve}
 \usepackage{amsmath}
@@
 succession, and terminates at the end state.
-We~implemented this algorithm, storing the probabilities in an array.
+One of us implemented this algorithm, storing the probabilities in an array.
 \remark{Why is the number of dimensions 3 and not 2 (indices $i$, $j$,
 and what??) and why do we care?}
-The cost is
+The cost was
 $O(|N|\times|\alignwidth|)$;
@@
 
-\emph{[WHAT WAS UNSATISFACTORY ABOUT THE FIRST IMPLEMENTATION]}
-
-We next wrote a top-down dynamic-programming 
-version, which began with the last node of the model and the last observation, 
-and applied structural recursion down to the base cases involving the first 
-node or first observation (or both).
-The general case of this top-down, structurally recursive implementation closely
-resembles the Viterbi equations above.
-The match equation is thus:
+Another of us was curious to try to code Viterbi's equations
+directly as recursive functions.
+But like the classic recursive Fibonacci function, Viterbi's functions,
+when implemented \naive ly,
+perform extraordinarily badly.
+But just like the Fibonacci function, Viterbi's functions can be
+\emph{memoized}.
+To~show the memoized code, we have to confess to a small
+transformation of Viterbi's equations.
+The logarithms on the left of \figref{viterbi} are computed when the
+hidden Markov model is constructed, and the resulting numbers are
+stored in a file.
+Moreover, to make the file more pleasant to read, the logarithms are
+negated.
+(The logarithm of a probability is typically negative, and negating
+the logarithm eliminates hordes of minus signs from the file.)
+The negated log probability is called a \texttt{Score}.
+We~therefore implement a transformed version of Viterbi's equations,
+shown on the right-hand side of \figref{viterbi}, which minimize the
+negated log probability for each combination of node~$j$, amino
+acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{smallverbatim}
+viterbi :: State -> Int -> Int -> (Score, StatePath)
+viterbi Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
+                                          , trans i_m Ins
+                                          , trans d_m Del 
+                                          ]
+                  , Mat : path )
+  where (prevV, path) = viterbi prev (j-1) (i-1)
+        trans t prev = prevV + tProb hmm (j-1) t
+\end{smallverbatim}
+ALTERNATE VERSION FOLLOWS
+\begin{smallverbatim}
+viterbi :: WHAT IS MY TYPE????
 viterbi s n o

### 2012-03-10 mrfy_experience_report.tex
hammering on the HMM section complete

@@
 
-\subsection{Perspicuous Haskell}
+\subsection{Computing probabilities using perspicuous Haskell}
 
@@
 The probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
+Equations like these can be found in more recent literature
+\cite{Durbin:1998wz,Eddy:1998ut}.
 
@@
 the logarithm eliminates hordes of minus signs from the file.)
-The negated log probability is called a \texttt{Score}.
 We~therefore implement a transformed version of Viterbi's equations,
@@
 acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
+Our implementation computes not only the minimum
+negated log probability, called a \texttt{Score},
+but also the path of states which leads to that most likely outcome. 
+Because the cases for all three states are so similar, 
+we~show only the code for the match state:
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{smallverbatim}
-viterbi :: State -> Int -> Int -> (Score, StatePath)
-viterbi Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
+viterbi' :: State -> Int -> Int -> (Score, [State])
+viterbi' Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
                                           , trans i_m Ins
@@
         trans t prev = prevV + tProb hmm (j-1) t
-\end{smallverbatim}
-ALTERNATE VERSION FOLLOWS
-\begin{smallverbatim}
-viterbi :: WHAT IS MY TYPE????
-viterbi s n o
-  | s == mat = 
-       let trans t prev = 
-        (scr + tProb (n-1) t + (eProb n o), mat:p)
-        where (scr, p) = viterbi prev (n-1) (o-1)
-       in minimum $ [
-           trans m_m mat,
-           trans i_m ins,
-           trans d_m del
-           ]
 \end{smallverbatim}
@@
 % \end{verbatim}
-
-In this second approach, we used the
-\texttt{Data.Memocombinators} library to provide the memoization container 
-(again, a three-dimensional array).
-This combinator approach was easy to apply to the code after the fact:
+Programming the algorithm in this form was not only intellectually
+satisfying; keeping the code close to the math also helped us find
+bugs.
+The most memorable bug was that in one of our base cases, 
+we~mistakenly returned an empty path of states;
+the correct result was a singleton path containing the initial state.
+Very quickly after observing a missing state in the output, we were
+able to find the faulty case in the code.
+
+Whatever the attractions of the code above, in our work we also
+require high performance.
+To make the simple recursive code efficient, we used Luke Palmer's
+\texttt{Data.Memocombinators} package.
+The function \verb+viterbi'+ is unmemoized,
+and it makes recursive calls to \verb+viterbi+, which is memoized:
 \begin{smallverbatim}
-viterbi state node obs = 
-  Memo.memo3 (Memo.arrayRange (mat, end)) 
+viterbi state j i = 
+  Memo.memo3 (Memo.arrayRange (Mat, End)) 
              (Memo.arrayRange (0, numNodes))
              (Memo.arrayRange (0, seqlen)) 
-             viterbi' state node obs
+             viterbi' state j i
 \end{smallverbatim}
-We determined that the run-time 
-performances of both approaches were virtually indistinguishable, and we also 
-noted that the top-down version quite faithfully resembled the recurrence 
-relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
-(see Equation \ref{viterbi_eqn}).
-We chose to keep the top-down version, which 
-later proved to be wise when we had to track down a bug in one of our base 
-cases.
-This bug was observed as a missing state in the alignment our program 
-displayed, and we quickly discovered that one of our base cases was missing 
-this state: it was returning an empty list as the base of the path, rather than 
-a list containing a single node.
-In this, we were grateful for the resemblance 
-between the mathematical description of the algorithm and the top-down 
-dynamic-programming approach in Haskell, which resulted in perspicuous code.
-
-Standard implementations of the Viterbi algorithm, such as in HMMER, spend much of their code
-managing the dynamic programming tables, as even when memoization libraries are available,
-they are rarely used.
-Furthermore, standard implementations in imperative languages typically avoid the natural
-recursive representations because of the associated costs in languages such as C++.
-The efficient machine code that GHC generates for recursive functions allowed us to write
-a program that faithfully reflected the mathematics, without paying great costs in run-time
-efficiency.
+The value
+\texttt{numNodes} represents~$\alignwidth$,
+and \texttt{seqlen} represents~$N$.\remark{I hope this is right. ---NR}
+
+This memoized recursive code performs just as well as the classic
+dynamic-programming code.
+And~the short fragment above is the \emph{only} part of the code
+devoted to improving performance.
+By contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
+spend much of their code 
+managing dynamic-programming tables.
+Haskell enabled us write simple, performant code with little effort.
+%
+Because the memoized version so faithfully resembles the equations in
+\figref{viterbi}, we~retired the classic dynamic-programming version.
+
+\remark{NR doesn't think the bug story is that interesting.}
+
+%%  
+%%  
+%%  In this, we were grateful for the resemblance 
+%%  between the mathematical description of the algorithm and the top-down 
+%%  dynamic-programming approach in Haskell, which resulted in perspicuous code.
+%%  
+%%
+
+\subsection{Solving new problems using perspicuous Haskell}
 
@@
 Our implementations of hidden Markov models and Viterbi's algorithm
-simply duplicate 
-functionality that is already implemented in the C++~package
-HMMER.
-These implementations form just one component of our new tool, MRFy.
+duplicate existing functionality.
+They form just one component of our new tool, MRFy.
 

### 2012-03-10 mrfy_experience_report.tex
motivating MRfy

@@
 \newcommand\alignwidth{\ensuremath R} % number of columns in an alignment
+\newcommand\pairedwith[1]{{\pi(#1)}}
 
@@
 
-\subsection{Solving new problems using perspicuous Haskell}
+\subsection{Exploring new algorithms using higher-order functions}
 
@@
 
-A hidden Markov model effectively uses local information about
+A hidden Markov model uses local information about
 amino-acid sequences.
-But real proteins are three-dimensional structures,
-and their amino-acid sequences are subject to \emph{nonlocal}
-constraints. 
-And when a real protein folds, groups of amino acids
+But real proteins are three-dimensional,
+and when a real protein folds, groups of amino acids
 that are far away in the one-dimensional sequence can be
@@
 
-During the course of evolution, a mutation in an amino acid that participates
-in this beta-strand pairing may cause a protein to fail to fold into a useful
-structure, unless that mutation is accompanied by a \textit{compensatory}
-mutation in the amino acid with which it is paired. 
-Thus, homologous
-proteins---those that share structure and function---tend to \textit{conserve}
-such pairwise-conditional amino acids.
-
-In order to model these non-local interactions, computational biologists
-have extended hidden Markov models to include these non-local interactions,
-modeled as a \textit{dependency graph}, which transforms the model into a
-\textit{Markov random field}. These approaches ``nail down'' 
-certain subsequences of the model, corresponding to the non-local hydrogen-bond 
-interactions between beta strands.
-The architecture of the 
-resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
-illustrated in Figure~\ref{mrf}.
-
-SMURF~\cite{Menke:2010ti} and SMURFLite~\cite{Daniels:2012}, our previous work,
-compute the 
-optimal parse of a protein sequence onto the directed graph representing a 
-Markov random field using multidimensional dynamic programming.
-Mathematically, we can represent the probability of a query sequence 
-$x_{1}..x_{n}$ given a Markov random field with an additional term in the
-Viterbi relations.
-Specifically, each term $V_{j}^{\prime M}(i)$ becomes
-$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log Pr(x_{i} | x_{\pi(j)})$, where
-$x_{\pi(j)}$ is the amino acid in the query sequence that will be placed
-in the beta-strand position paired with position $j$.
-This nonlocal dependency prevents the problem from exhibiting optimal substructure,
-preventing dynamic programming from solving the problem quickly.
-SMURF's 
-computational complexity quickly exploded with complex protein structures, 
-while SMURFLite limited its computational complexity by also limiting the 
-complexity of protein structure it would understand.
-The algorithm underlying 
-MRFy grew out of an understanding of these limitations.
-
-Rather than attempt to compute exactly the optimal parse, MRFy treats the beta 
-strands as ``beads'' which can be slid along a string.
-In this analogy, the 
-string is the sequence of amino acids of the query protein.
-A given placement 
-of these beads corresponds to an assignment of particular amino acid residues 
-being hydrogen bonded to other particular amino acid residues, and thus MRFy 
-computes a probability based on observed frequencies of residues hydrogen 
-bonding in beta strands~\cite{Cowen:2002p588}.
-In between these beta strands, the 
-Markov random field is in fact a traditional hidden Markov model, whose optimal 
-solution can be easily computed via the Viterbi algorithm.
-MRFy performs a 
-stochastic search to place these beta strands, based on an objective function 
-that combines the probability computed by the Viterbi algorithm with that 
-computed from the paired, hydrogen-bonded residues in the beta strands.
-If a 
-structure has many beta strands, then it also has many miniature hidden Markov 
-models which can be solved via the Viterbi algorithm.
-
-\begin{figure}[h!] 
+A~mutation in a paired beta strand may break the pairing, in which
+ case the mutated protein will not fold properly and is unlikely to
+ survive natural selection.
+But a mutation in a paired beta strand \emph{can} survive if it is
+ accompanied by a compensatory mutation in the strand with which it is
+ paired. 
+The probability of finding amino acid $x_i$ in column~$j$ of an alignment may
+ therefore depend on the amino acid in the paired column
+ $\pairedwith j$.
+This nonlocal interaction changes
+Viterbi's equations:
+each term $V_{j}^{\prime M}(i)$ becomes
+$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
+The distance between columns~$j$ and $\pairedwith j$ can be as small
+ as a few columns or as large as a few hundreds of columns.
+
+The approach currently taken in computational biology
+is~to model each pair of beta strands not with two hidden Markov
+models, 
+but with a pair of sequences of match states.
+Each of the subsequences \emph{between} paired beta strands is then
+modeled with an ordinary hidden Markov model.
+The~combined model, as shown in \figref{mrf}, is called a
+\textit{Markov random field}. 
+Because of nonlocal interactions between paired beta strands, 
+the likelihood-maximization for a Markov random field does not
+ exhibit optimal substructure, and dynamic
+ programming cannot solve it quickly \cite{Menke:2010ti,Daniels:2012}.
+The~purpose of MRFy is to explore new methods of \emph{approximating}
+maximum likelihood.
+
+\begin{figure}
 \ifpdfmadness
@@
 \fi
-\caption{The directed graph representing the states of a MRFy Markov random 
-field finite state automaton.
-Shaded nodes represent beta-strand positions which have been fixed into match states, 
-while the dashed edges represent the long-range dependencies due to hydrogen 
-bonding.
-Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}
-
-MRFy contains two distinct algorithms: the classical Viterbi 
-dynamic-programming algorithm for hidden Markov models, and a stochastic search 
-algorithm for placing beta-strand positions.
-This stochastic search algorithm 
-implements several search heuristics: random-mutation hill climbing, simulated 
+In this Markov random field,
+each shaded node represents a beta-strand position.
+Nodes connected by dashed edges are hydrogen-bonded.
+
+\caption{A Markov random field with two beta-stand pairs}
+\label{mrf} 
+\end{figure}
+
+MRFy treats the beta 
+strands, which are build into the model,
+ as ``beads'' which can be slid along the query sequence to determine
+ a \emph{placement} of the beta strands.
+A~placement's probability is computed
+based on observed frequencies of paired amino acids in 
+hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
+Given a placement, the maximum likelihood of any subsequence that appears
+\emph{between} two beta strands can be computed exactly and
+efficiently using Viterbi's algorithm.
+(The result of the exact computation is a probability that is
+\emph{conditioned} on the placement.)
+MRFy searches for placements stochastically,
+trying to maximize the product of these probabilities.
+We~are exploring several
+heuristics: random-mutation hill climbing, simulated 
 annealing, multistart simulated annealing, and a genetic algorithm.
-
-\subsection{Higher-order functions}
-
-Just as the Viterbi algorithm is one major component of MRFy, stochastic search 
-is the other.
-In designing MRFy, we did not know in advance what sort of 
-stochastic search technique would converge the fastest, or give us the best 
-alignments with the fastest performance.
-We decided to implement several 
-different search strategies so that we could experiment with all of them.
-We 
-found this to be easy and pleasant thanks to Haskell's support of higher order 
-functions.
+Our~exploration of these heuristics relies on higher-order functions.
 
@@
 
+\vfill
+
+\begingroup
+\advance\leftskip by 0pt plus 10em
+\emph{The Haskell wave passes, and if you slide off the crest, you
+drown.}
+\endgroup
+
 \end{document}

### 2012-03-10 mrfy_experience_report.tex
Attempt at story & code for search.

@@
 
-We wrote a basic search function in only twenty-two lines of code.
-The central logic of this search function is simply:
+These various search heuristics differ in how they approach four subproblems:
+how they \textit{initialize} the search with a starting guess or guesses;
+how they \textit{mutate} a solution into a new one; how they decide whether
+to \textit{accept} a solution over a previous one; how they decide when to
+\textit{terminate} the search.
+For example, random-mutation hill climbing only accepts better guesses,
+while simulated annealing sometimes accepts worse guesses; a genetic algorithm
+combines ``parent'' guesses in its mutate function, while random-mutation
+hill climbing simply permutes a guess into a new one.
+
+We represent these various search heuristics as a \texttt{SearchStrategy}:
+\begin{verbatim}
+data SearchStrategy = 
+      SS { accept :: SearchParameters -> Seed -> History -> Age -> Bool
+         , terminate :: SearchParameters -> History -> Age -> Bool
+         , mutate :: SearchParameters -> Seed -> QuerySequence ->
+            Scorer -> [BetaStrand] -> [SearchSolution] ->
+            [SearchSolution]
+         , initialize :: SearchParameters -> Seed ->
+            QuerySequence -> [BetaStrand]-> [SearchGuess]
+         }
+type Scorer = QuerySequence -> [BetaStrand] -> SearchGuess -> SearchSolution
+\end{verbatim}
+This \texttt{SearchStrategy} is passed (as part of \texttt{SearchParameters})
+to our generic search function:
 \begin{smallverbatim}
+data SearchParameters = SP { strategy :: SearchStrategy
+                           , generations :: Int
+                           }
+
+search :: QuerySequence -> HMM -> [BetaStrand] -> SearchParameters -> [Seed] -> (SearchSolution, History)
+search query hmm betas searchP seeds = search' (tail seeds) initialGuessScore [] 0
+  where initialGuessScore = map (score hmm query betas) initialGuess
+
+        initialGuess = (initialize strat) hmm searchP (head seeds) query betas
+
+        strat = strategy searchP
+        search' :: [Seed] -> [SearchSolution] -> History -> Age -> (SearchSolution, History)
         search' (s1:s2:seeds) oldPop hist age =
-  let newPop = mutate oldPop
+          let newPop = mutate' oldPop
               score = fst $ minimum newPop
-      newHist = score : hist
-    in if accept newHist age then
-          if terminate newHist age then
+              newHist = (score, age) : hist
+            in if accept' newHist age then
+                  if terminate' newHist age then
                     (minimum newPop, newHist)
@@
                else
-         if terminate hist age then
+                  if terminate' hist age then
                     (minimum oldPop, hist)
@@
                     search' seeds oldPop hist (age + 1)
+            where mutate' = mutate strat searchP s1 query (score hmm) betas
+                  terminate' = terminate strat searchP
+                  accept' = accept strat searchP s2
 \end{smallverbatim}

### 2012-03-10 mrfy_experience_report.tex
add files derived from mrfy sources, and use them

--- 2012-03-10 mrfy_experience_report.tex
@@
 \newenvironment{smallverbatim}{\par\small\verbatim}{\endverbatim}
+\newcommand\smallverbatiminput[1]{{\par\unskip\small\verbatiminput{#1}}}
+\newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}
 
@@
 
-We wrote a basic search function in only twenty-two lines of code.
-The central logic of this search function is simply:
-\begin{smallverbatim}
-search' (s1:s2:seeds) oldPop hist age =
-  let newPop = mutate oldPop
-      score = fst $ minimum newPop
-      newHist = score : hist
-    in if accept newHist age then
-          if terminate newHist age then
-            (minimum newPop, newHist)
-          else
-            search' seeds newPop newHist (age + 1)
-       else
-         if terminate hist age then
-           (minimum oldPop, hist)
-         else
-           search' seeds oldPop hist (age + 1)
-\end{smallverbatim}
+These various search heuristics differ in how they approach four subproblems:
+how they \textit{initialize} the search with a starting guess or guesses;
+how they \textit{mutate} a solution into a new one; how they decide whether
+to \textit{accept} a solution over a previous one; how they decide when to
+\textit{terminate} the search.
+For example, random-mutation hill climbing only accepts better guesses,
+while simulated annealing sometimes accepts worse guesses; a genetic algorithm
+combines ``parent'' guesses in its mutate function, while random-mutation
+hill climbing simply permutes a guess into a new one.
+
+\emph{[In scope: Parameters, [BetaStrand], QuerySequence]}
+
+1. Start with [PLacement]
+2. At significant expense, any Placement can be assigned a Score
+
+ terminate determines when to finally stop the search. Examples might
+ be running for a predetermined number of generations, or stopping
+ when we've gone N generations without achieving more than E
+ improvement in score
+
+We represent these various search heuristics as a \texttt{SearchStrategy}:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\smallverbatiminput{strategy}
+
+This \texttt{SearchStrategy} is passed (as part of \texttt{SearchParameters})
+to our generic search function:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\smallverbatiminput{search}
+
 As one of the parameters to the top-level \texttt{search} function,

### 2012-03-10 search.tex
add files derived from mrfy sources, and use them

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Scored Age])
search strat score (s:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s
  runFrom :: [Seed] -> [Scored Placement] -> [Scored Age]
          -> Age -> (Scored Placement, [Scored Age])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist =
          Scored age (scoreOf $ minimum newPop) : oldHist
        (newPop, newHist) =
          if accept strat s2 newHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)

### 2012-03-10 mrfy_experience_report.tex
fix viterbi equations

--- 2012-03-10 mrfy_experience_report.tex
@@
 \begin{figure*}
+\newcommand\vsum[2]{#2&{}+{}& #1}
+
 \def\goo{18pt}
@@
 V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
-\begin{array}{l l}
-V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}}\\
-V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}}\\
-V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
+\begin{array}{l@{}c@{}l}
+\vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
+\vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
+\vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
 \end{array} \right.\\[\goo]
 V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
-\begin{array}{l l}
-V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}}\\
-V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}}\\
+\begin{array}{l@{}c@{}l}
+\vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
+\vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
 \end{array} \right.\\[\gum]
 V_{j}^{D}(i) &=& \max \left\{
-\begin{array}{l l}
-V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}}\\
-V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
+\begin{array}{l@{}c@{}l}
+\vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
+\vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
 \end{array} \right.\\
@@
 V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
-\begin{array}{l l}
-V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}}\\
-V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}}\\
-V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
+\begin{array}{l@{}c@{}l}
+\vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
+\vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
+\vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
 \end{array} \right.\\[\goo]
 V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
-\begin{array}{l l}
-V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}}\\
-V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}}\\
+\begin{array}{l@{}c@{}l}
+\vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
+\vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
 \end{array} \right.\\[\gum]
 V_{j}^{\prime D}(i) &=& \min \left\{
-\begin{array}{l l}
-V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}}\\
-V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
+\begin{array}{l@{}c@{}l}
+\vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
+\vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
 \end{array} \right.\\[\gum]
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\smallverbatiminput{viterbi}
 \begin{smallverbatim}
@@
 \emph{conditioned} on the placement.)
+The likelihood is computed by a function \texttt{score},
+which given a model and a {placement} returns the {placement}
+paired with a \emph{score}.
+A~score is a negated log probability, so lower scores are better.
+
 MRFy searches for placements stochastically,
-trying to maximize the product of these probabilities.
-We~are exploring several
-heuristics: random-mutation hill climbing, simulated 
+trying to minimize scores.
+We~are exploring several stochastic algorithms: random-mutation hill
+climbing, simulated  
 annealing, multistart simulated annealing, and a genetic algorithm.
-Our~exploration of these heuristics relies on higher-order functions.
-
-We wrote a basic search function in only twenty-two lines of code.
-The central logic of this search function is simply:
-\begin{smallverbatim}
-search' (s1:s2:seeds) oldPop hist age =
-  let newPop = mutate oldPop
-      score = fst $ minimum newPop
-      newHist = score : hist
-    in if accept newHist age then
-          if terminate newHist age then
-            (minimum newPop, newHist)
-          else
-            search' seeds newPop newHist (age + 1)
-       else
-         if terminate hist age then
-           (minimum oldPop, hist)
-         else
-           search' seeds oldPop hist (age + 1)
-\end{smallverbatim}
-As one of the parameters to the top-level \texttt{search} function,
-we provide a \texttt{SearchStrategy} data type, which we use to 
-define specific search techniques.
-We require a \texttt{SearchStrategy} to 
-define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
-and \texttt{initialize}; we use the Haskell module system to allow sharing of 
-some of these functions between different search strategies.
-For example, both 
-\textit{random mutation hill climb} and \textit{simulated annealing} use the 
-same initialize and mutate functions, but very different accept functions.
-In 
-all, we implemented four search strategies: \textit{random mutation hill 
-climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
-and \textit{genetic algorithm}.
-We found the implementation of the search 
-function and search strategies to be pleasant, and the resulting code to be 
-simple and elegant.
-In an imperative language with object orientation like Ruby 
-or Python, we likely would have used a class hierarchy to accomplish this.
-We 
-feel that this would have required more code, because object orientation is 
-meant for encapsulating data as well as functions on those data; here we only 
-wanted to encapsulate functions, as the data did not change.
-
-We note that \texttt{seed} in \texttt{search} comes from a monadic random 
-number generator in Main, but we wished for search itself to be pure at this 
-level.
-One advantage is that this allows for testing of the \texttt{search} 
-function with no randomness at all: if we pass in a known number as the seed, 
-behavior is entirely deterministic.
+\remark{It would be pleasant to connect these to the functions below,
+but we probably lack room.}
+Our~exploration relies on higher-order functions.
+
+MRFy's search works with a sequence of \emph{generations}.
+A~generation is a collection of scored \emph{placements}.
+We~pass the
+search function a \texttt{Scorer}, which is the result of partially
+applying \texttt{score} to the model.
+\smallverbatiminput{scoredecl}
+A~search algorithm is characterized by four functions, which are
+bundled into a record called \texttt{SearchStrategy}.
+\smallverbatiminput{strategy}
+In this record,
+\begin{itemize}
+\item
+Function \texttt{gen0} produces the first generation of placements.
+\item
+Function \texttt{nextGen} produces a new generation of placements from
+an existing generation.
+Because scoring a placement is expensive, 
+and some placements may persist from the old generation to the new
+generation, \texttt{nextGen} works with \emph{scored} placements, not
+bare placements.
+\item
+A~new generation produced by \texttt{nextGen} may not be any better
+than the generation it replaces.
+When it is first produced, the new generation is ``on trial.''
+The \texttt{accept} function decides whether to keep a new generation
+or to try \texttt{nextGen} again with a different seed.
+The \texttt{accept} considers the score of the best
+placement of the trial generation, along with the history of previous
+best scores.
+It~also considers the \texttt{Age} of the trial generation, which is
+the total number of 
+generations tried.
+\item
+Finally, the \texttt{quit} function decides when to stop the search.
+It~too considers the history of scores and the age of the current
+generation. 
+\end{itemize}
+This interface allows for considerable variation.
+For example, random-mutation hill climbing accepts only generations
+with better scores than their predecessors,
+while simulated annealing may accept a generation with a worse score.
+A~genetic algorithm uses \texttt{nextGen} to
+combine placements from the previous generation, while
+random-mutation
+hill climbing simply permutes a placement into a new one.
+The interface \texttt{quit} makes it easy
+to implement such termination criteria as a fixed number of generations
+or a limit of going $k$~generations without achieving more than $E$
+improvement in score.
+
+We hope you find the code simple and straightforward:
+\smallverbatiminput{search}
 

### 2012-03-10 search.tex
fix viterbi equations

@@
 search :: SearchStrategy -> Scorer -> [Seed]
-       -> (Scored Placement, [Scored Age])
+       -> (Scored Placement, [Score])
 search strat score (s:seeds) = runFrom seeds firstGen [] 0
@@
   firstGen = map score $ gen0 strat s
-  runFrom :: [Seed] -> [Scored Placement] -> [Scored Age]
-          -> Age -> (Scored Placement, [Scored Age])
+  runFrom :: [Seed] -> [Scored Placement] -> [Score]
+          -> Age -> (Scored Placement, [Score])
   runFrom (s1:s2:seeds) oldPop oldHist age =
     let trialPop  = nextGen strat s1 score oldPop
-        trialHist =
-          Scored age (scoreOf $ minimum newPop) : oldHist
+        trialHist = (scoreOf $ minimum newPop) : oldHist
         (newPop, newHist) =

### 2012-03-11 mrfy_experience_report.tex
finally finished an edit of all of section 3 (with a few holes)

@@
 The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
+We~represent a state using an algebraic data type:
+\begin{smallverbatim}
+data HMMState = Mat | Ins | Del | Beg | End
+\end{smallverbatim}
 
@@
 directly as recursive functions.
-But like the classic recursive Fibonacci function, Viterbi's functions,
+Like the classic recursive Fibonacci function, Viterbi's functions,
 when implemented \naive ly,
@@
 acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
+
 Our implementation computes not only the minimum
@@
 but also the path of states which leads to that most likely outcome. 
-Because the cases for all three states are so similar, 
-we~show only the code for the match state:
+Elsewhere, MRFy pairs not only paths but also other types of values
+with scores, so we have defined a small abstraction:
+\verbatiminput{vscore}
+We~made \texttt{Scored} an instance of~\texttt{Ord} and of
+\texttt{Functor}.
+
+Given the \texttt{Scored} abstraction, we can easily implement
+Viterbi's equations for a match state: 
+we~choose the best solution
+from among the three predecessor states; 
+we~use~\texttt{/+/} to add transition and emission scores,
+and we use \texttt{fmap} to add \texttt{Mat} to the list of states
+encountered up to this point:
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\smallverbatiminput{viterbi}
-\begin{smallverbatim}
-viterbi' :: State -> Int -> Int -> (Score, [State])
-viterbi' Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
-                                          , trans i_m Ins
-                                          , trans d_m Del 
-                                          ]
-                  , Mat : path )
-  where (prevV, path) = viterbi prev (j-1) (i-1)
-        trans t prev = prevV + tProb hmm (j-1) t
-\end{smallverbatim}
+\smallverbatiminput{vfix}
+Functions \texttt{tProb} and \texttt{edge} combine to select the
+appropriate transition score (one of the $a'$~values).
 % Similarly, the insertion equation is:
@@
 \end{itemize}
-This interface allows for considerable variation.
-For example, random-mutation hill climbing accepts only generations
+This interface provides considerable leeway for experiments.
+For~example, random-mutation hill climbing accepts only generations
 with better scores than their predecessors,
@@
 
-We hope you find the code simple and straightforward:
+\seclabel{search}
+
+We hope you find the search code simple and straightforward:
 \smallverbatiminput{search}
-
-In MRFy, the Viterbi function is called several times (once for each 
-non-beta-strand region) for each search generation.
-Furthermore, in a 
-population-based search such as \textit{multistart simulated annealing}, the 
-scoring function (which itself calls Viterbi) is called many times per 
-generation.
-Beyond traditional profiling, we found parallelization to be an 
-obvious way to improve runtime performance.
-We found this to be trivial thanks 
-to the \texttt{Control.Parallel} library.
-We pass the Viterbi function to a map over all 
-non-beta-strand regions of the model and query sequence, and these regions are 
-independent of one another.
-Similarly, we pass the scoring function to a map 
-over the population of potential solutions, which are also independent.
-We 
-merely substituted \texttt{parmap rseq} for map.
-MRFy now efficiently uses all six 
-cores on an AMD workstation for a non-population-based simulated annealing 
-search, and all forty-eight cores on a compute server for population-based 
-searches.
-
-Despite the beauty of some of the code we were able to write, we found several 
-instances where the easiest path seemed to be to write ugly code.
-The first of 
-these relates to the otherwise elegant Viterbi function.
-Our stochastic search 
-approach calls the Viterbi function a tremendous number of times, and most of 
-the program's runtime is spent in this function.
-However, the Viterbi algorithm 
-retains not just a score (the probability of the sequence given a state path) 
-but also the state path itself.
-We implemented this state path as a list of 
-\texttt{Int}, and every step in the recursion conses a new element onto the 
-list.
-However, at each step of the stochastic search, we discard the state 
-path; we are searching to optimize score only.
-We only need the state path at 
-the termination of the search, when we wish the program to output an alignment 
-of the query sequence to the Markov random field model.
-We found that 
-optimizing the Viterbi function by removing this state path entirely improved 
-runtime performance by nearly 50\%.
-However, we still needed the Viterbi 
-function to provide a state path when we called it after search finally 
-terminated.
-Thus, we wondered how to implement the Viterbi function so that a 
-state path was optional.
-
-In order to determine if this optimization was even worthwhile, we first wrote 
-a quick and dirty approach that is shameful in any language, but easy: we 
-duplicated code.
-We copied \texttt{viterbi} to produce a \texttt{viterbiF} 
-which did not retain state path.
-Next, however, we wondered how to avoid code 
-duplication.
-Clearly, littering the code with conditionals would produce code 
-that was ugly as well as less efficient.
-We knew how we would solve this 
-problem in other languages; we would have used macros in Lisp, or 
-metaprogramming via \texttt{eval} in Ruby, to produce two versions of the 
-function from one prototype.
-We struggled, however, to solve this problem 
-idiomatically in Haskell.
-
-Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
-the Viterbi function, we wrote a higher-order function type 
-\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type:
-\begin{smallverbatim}
-consPath :: ScorePathCons a
-consPath x xs = x:xs
-
-consNoPath :: ScorePathCons a
-consNoPath _ _ = []
-\end{smallverbatim}
-
-Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
-or \texttt{consNoPath} as a parameter, the path would or would not be 
-allocated.
-We found this higher-order function approach to be simple in code 
-and in concept, and it imposes negligible run-time overhead as compared to 
-having two distinct functions.
+\emph{[MENTION AND QUANTIFY NOAH'S SUCCESSES IN ADDING NEW HEURISTICS.]}
+
+\subsection{Performance}
+
+In MRFy, the \texttt{score} function calls \texttt{viterbi} several
+ times
+for each placement in each search generation.
+Calls on query sequences between beta strands are independent,
+as are calls on different placements.
+We~found them easy to parallelize
+using
+\texttt{Control.Parallel},
+simply substituting
+\texttt{parmap rseq} for \texttt{map}.
+Even when there is just one placement per generation,
+MRFy can keep six cores busy,
+and when our search algorithm uses multiple placements per generation,
+it~can use all forty-eight cores on a compute server.
+
+In~\secref{viterbi} above, the example case for the \texttt{viterbi}
+function computes not just a score but also an optimal path.
+MRFy's \texttt{search} function, in \secref{search}, uses only the
+score.
+Even though Haskell evaluation is lazy, the \texttt{viterbi} function
+is still allocating thunks.
+We~tried cloning and modifying the \texttt{viterbi} code to compute
+only a score, with no path, and this change improved 
+run-time performance by nearly 50\%.
+But when \texttt{search} finally decides on a placement, we need a
+version of \texttt{viterbi} that \emph{does} compute path, so we can
+extract the optimal parse for that placement.
+
+How could we keep the performance improvement without having to
+maintain two versions of \texttt{viterbi}?
+In~Lisp or Ruby we would have used macros or metaprogramming,
+but we were not confident of our ability to metaprogram a solution
+using Template Haskell.
+Instead, we replaced the primitive cons in \texttt{viterbi} with a
+call to a function of the same type, which we passed~in.
+Passing the primitive~``\texttt{:}'' gives us a path,
+and passing \verb+\_ _ -> []+ gives us the same
+speedup as the cloned and modified~code.
+But~although
+this trick is simple and easy to implement, 
+we~have an uneasy feeling that it may work only because of undocumented
+properties of GHC's inliner, which may be subject to change without
+notice.
 
@@
 
-Another source of ugliness resulted from adding cost-center annotations when we 
-wished to profile our code.
-GHC will happily add cost centers to top-level 
-functions, but much of our program relies on sub-functions.
-We found that 
-manually adding \texttt{\{-SCC -\}} annotations to dozens of guard clauses and 
-sub-functions harmed the readability of our program, such that we felt 
-compelled to remove all such annotations as soon as we could.
-In our prior 
-experience with profiling tools such as kcachegrind, simply compiling with 
-debug symbols -- rather than manually annotating many lines of code -- allowed 
-the profiling tools to provide us enough information.
-Without these 
-annotations, both call-site and allocation profiles are relatively useless in 
-GHC.
-Perhaps a compile-time flag that would auto-annotate every equation -- 
-even if the resulting cost center is simply named with a line number -- would 
-make this task easier.
-
-The final source of ugliness in our code came from attempts at debugging 
+In two major areas, our lack of skills or knowledge resulted in ugly
+Haskell.
+The first is profiling.
+GHC's profiling is based on cost centers \cite{sansome-peyton-jones},
+and although GHC 
+will happily add cost centers to top-level 
+functions, much of our code is in nested functions.
+We~know of no better technique than to add cost-center annotations to
+each guard clause and nested function,
+but these annotations made our code so ugly that we felt compelled to
+remove them as soon as possible.
+And without the annotations, we found GHC's call-site and allocation
+profiles did not provide enough information to be useful.
+We~miss valgrind, whose callgrind profiler gives us enough information
+without \emph{any} annotations in the source.
+
+Our other ugly code came from  attempts to debug
 run-time errors.
-MRFy relies on a source of input data -- the paired beta 
-strands in the Markov random field -- which is represented in an awkward manner 
-by the original SMURF program, and which is outside our control.
-Rather than 
-represent beta strands directly, SMURF represents pairings of residues, leaving 
-us to infer and reconstruct beta strands.
-This task was difficult due to 
-unclear invariants around overlapping, doubly-paired beta strands (and it would 
-have been difficult in any programming language; the difficulty was one of 
-representation).
-However, the debugging tools in GHC caused us some 
-difficulties in determining the sources of run-time errors.
-We used 
-\texttt{trace} extensively, but this style of ``\texttt{printf} debugging'' 
-littered our code.
-Being initially unaware of the backtrace feature of the GHC 
-profiler, we wrote wrappers around \texttt{Vector.slice} and \texttt{Vector.!} 
-which themselves called trace, which at least kept our ``\texttt{printf} 
-debugging'' less cluttered.
-Nonetheless, we struggled with these runtime errors 
-due to an awkward representation, and wished for debugging tools such as we 
-might find in gdb, which would allow us to examine the stack when a backtrace 
-occurs.
+MRFy uses a Markov random field which is represented in a legacy file
+format that uses a poorly documented representation---when beta
+strands overlap and are doubly paired, the invariants of the
+representation are not clear.
+Dealing with this representation is difficult no matter what
+programming language is used, and faults are to be expected.
+Using GHC, we had a hard time diagnosing run-time errors.
+Extensive use of 
+\texttt{trace} 
+littered our code,
+even when moved from our main code to wrapper functions.
+We~weren't aware of the backtrace feature of GHC's profiler,
+and even after we discovered that feature, the results weren't good enough.
+\emph{[WHY WASN'T THE BACKTRACE FEATURE GOOD ENOUGH?]}
+In~August 2011, 
+Lennart Augustsson said that the biggest advantage of Strict
+Haskell is getting a stack trace on error,
+and Simon Marlow said that he may have figured out how to track call
+stacks properly in a lazy functional language.
+We~can't wait.
 

### 2012-03-11 search.tex
finally finished an edit of all of section 3 (with a few holes)

@@
        -> (Scored Placement, [Score])
-search strat score (s:seeds) = runFrom seeds firstGen [] 0
+search strat score (s0:seeds) = runFrom seeds firstGen [] 0
  where
-  firstGen = map score $ gen0 strat s
+  firstGen = map score $ gen0 strat s0
   runFrom :: [Seed] -> [Scored Placement] -> [Score]

### 2012-03-11 mrfy_experience_report.tex
make the abstract shorter

@@
 \begin{abstract}
-In computational biology, the languages of choice for solving ad-hoc problems are high-level
-``scripting'' languages such as Python or domain-specific statistics languages such as R,
-while the language of choice for solving high-performance problems is C++.
-This dichotomy creates a difficulty when computational biologists wish to experiment with a
-variety of approaches to a problem, only some of which are likely to prove fruitful.
-This experience report details our use of Haskell in implementing a high-performance,
-computational biology application which is the first implementation of a particular
-novel algorithm for protein structure and function prediction. We discuss the highlights
-and difficulties of this project, as well as the potential improvements the Haskell
-community might make that would ease some of these difficulties.
+Haskell combines the flexibility and rapid prototyping of a scripting
+language with performance comparable to~C++.
+The Haskell language and libraries provide powerful abstractions, 
+but we couldn't always get them to work!
+Despite some speed bumps around profiling, debugging, and the
+uncertain status of any given package, we~found Haskell's ecosystem
+provided fine support for developing new algorithms in computational
+biology.
 \end{abstract}

### 2012-03-11 mrfy_experience_report.tex
completed one editing pass through the MS, and got below the page limit

@@
+\newif\ifnotcutting \notcuttingfalse % set if not cutting down to size
+
  \newif\ifpdfmadness
@@
 
+\seclabel{awkward-profiling}
+
 In two major areas, our lack of skills or knowledge resulted in ugly
@@
 The first is profiling.
-GHC's profiling is based on cost centers \cite{sansome-peyton-jones},
+GHC's profiling is based on cost centers \cite{sansom-pj},
 and although GHC 
@@
 
-\section{Our previous code base compared}
-
-Our previous computational biology projects include an enhancement to a protein 
-structural aligner that previously came from our group, Matt~\cite{Menke:2008wu}.
-
-Matt's code base contains approximately 12,000 lines of C++ code, and relies on 
-data structures such as mutable oct-trees, vectors, and arrays, with 
-significant amounts of clever pointer arithmetic.
-Our enhancement involved 
-modifying Matt to incorporate sequence information along with the structural 
-aligner.
-While we initially estimated this would be a three-month project, it 
-took most of a year because the mutable data structures were difficult to 
-repurpose, and the pointer arithmetic was too clever for its own good: since 
-invariants were not clear, nearly every change resulted in new segfaults.
-One 
-enhancement, partial alignments (essentially a cosmetic manipulation of the 
-output, but requiring deep information about some of the data structures) was 
-deemed infeasible.
-We believe we could have completed a ground-up rewrite in 
-Haskell, with most of the run-time performance, in fewer than nine months.
-
-In general, the failure modes we encounter most in C++ programming are 
-segmentation faults, memory errors (due to pointer arithmetic and uninitialized 
-memory) and heap exhaustion.
-In contrast, the failure modes we most commonly 
-see in Haskell code are type-check failures, or at worst, runtime errors around 
-bounds checking of arrays and vectors.
-While bounds checking failures have 
-proved difficult to debug, they are still less so than C++ memory errors.
-
-Haskell makes it difficult to write code in haste, because Haskell militates 
-towards careful type-checking and clearly thought-out function contracts.
-
-Unlike dynamic languages such as Ruby, and unlike C++, Haskell punishes the 
-programmer who attempts to write code quickly with the goal of later making it 
-work correctly.
-We have found MRFy to be easier to enhance and maintain than 
-our existing C++ code bases.
-Despite the learning curve of Haskell, we 
-implemented most of MRFy in roughly three months of part-time work.
-
-\section{State of the practice}
-
-\subsection{Our group then}
-Programmers generally avoid reimplementing algorithms, but there may be good reasons for doing so.
-Although the HMMER software contains an mature C++ implementation of the Viterbi algorithm,
-a previous member of our group implemented \textit{multidimensional} dynamic programming for SMURF
-in haste, resulting in 12,000 lines of undocumented code (including one function of 700 lines).
-Though we are using the same Viterbi algorithm HMMER implements, we chose to reimplement it because
-we wished to explore the design space of stochastic search techniques using higher order functions
-in Haskell.
-We found that reimplementing the Viterbi algorithm in Haskell was easy and pleasant.
-In contrast, the programming effort to transform SMURF into SMURFLite, a minor evolution, required
-several months of research to simply understand the SMURF codebase.
-
-A set of enhancements to another large C++ implementation, the Matt~\cite{Menke:2008wu} structural
-aligner, required nearly a year of programming due to complex mutable data structures being 
-hard to repurpose.
-In fact, we deemed one enhancement infeasible.
-Our experience working with these C++ codebases largely involved the failure modes of
-segmentation faults, memory errors, and heap exhaustion.
-
-\subsection{Our group now}
-Implementing MRFy in Haskell allowed us great flexibility to explore the space of stochastic search
-techniques, initial guess heuristics, and search termination.
-Higher order functions militate towards more flexibility, so that adding several planned features 
-to MRFy should be easy.
-The first functioning version of MRFy took roughly three months of part-time work.
-The Haskell failure modes we have experienced have largely been typecheck errors; at the worst,
-we have seen runtime errors around bounds checking of arrays and vectors, though we wanted
-better debugging tools for these runtime errors.
-
-\subsection{Computational biology at large}
-There exists some excellent software in the computational biology community, though little of it
-has been written in Haskell.
-The training component of MRFy was modified from that of HMMER, and working with the HMMER
-codebase was relatively pleasant, as data structures and their mutators are well documented.
-BioHaskell exists (and we use \texttt{Bio.Sequence.Fasta} as a parser) but is not nearly as
+\section{Our previous experience compared}
+
+When performance has mattered, the members of our group, like other
+computational biologists, have used~C++.
+Our~prior experience therefore includes building,
+maintaining, and extending research tools written in~C++.
+One~such tool is~Matt,
+a~structural protein aligner~\cite{Menke:2008wu}.
+Matt comprises about 12,000 lines of~C++.
+It uses such
+data structures such as mutable oct-trees, vectors, and arrays.
+It~uses clever pointer arithmetic.
+
+We~modified Matt to use information about sequences as well as structure.
+We~thought the modification would take three months, 
+but it took most of a~year.
+The mutable data structures were difficult to 
+repurpose, and the pointer arithmetic was too clever for its own good: 
+nearly every change resulted in new segfaults.
+We~had hoped to add support for partial alignments,
+which we~expected to require only 
+a cosmetic manipulation of the 
+output, but
+this feature wound up
+ requiring deep information about some data structures,
+and we had to give~up.
+We~believe we could write an equivalent tool in~Haskell,
+with most of Matt's performance, in at most nine months.
+
+Like many researchers, we~and our colleagues often write code hastily.
+Hastily written C++ code can get a job done, 
+but we have found it expensive in the long run.
+Our most painful experience was with an implementation of
+\textit{multidimensional}
+dynamic programming which was hastily added to the SMURF homology detector
+\cite{Menke:2010ti}.
+The implementation took 12,000~lines, including a single function of
+over 700 lines.
+When we added simulated evolution to this code, a relatively minor
+enhancement \cite{Daniels:2012}, 
+just understanding the existing code took several months.
+For comparison, 
+building the first functioning version of MRFy took three months of part-time work.
+And we expect that higher-order functions will make extending MRFy
+easy.
+\remark{I'm thinking this \P\ is not so strong. ---NR}
+
+Haskell encourages hasty programmers to slow down.
+We~have to get the types right,
+which makes it hard to write very large functions.
+To~get the code to typecheck, we~have to write type signatures, which
+also serve as documentation.
+And once the types are accepted by the compiler,
+it~is not much more work to write contracts for important functions.
+MRFy is still hasty work.
+Many types could be simplified;
+we're~sure we've missed opportunities for abstraction;
+and we know that MRFy's decomposition into modules could be improved.
+But despite being hasty programmers, we produced code 
+that is easy to understand and easy to extend.
+Our hastily written Haskell beats
+our hastily written Ruby and~C++.
+
+When we look beyond our own group to computational biology more
+broadly, our experiences with other software is better.
+Little of it is written in functional languages, 
+but much of the software shared by the community is excellent.
+MRFy's training component was derived from that of~HMMER,
+and
+working with HMMER 
+codebase was pleasant;
+data structures and their
+mutators are well documented. 
+There is a 
+BioHaskell library, part of which we use,
+but it is not nearly as 
 complete as BioPython or BioRuby, which are heavily used in the community.
-We hope that BioHaskell and other tools for computational biology in Haskell continue to mature.
-\section{What we think about it}
-
-\subsection{Our Background}
-
-\remark{I moved this section from the Intro as per NR's suggestion. ---AG}
-All three authors are members of the Tufts University Department of Computer
-Science.
-Noah Daniels has taken a graduate seminar in advanced functional
-programming, which included some Haskell, and spent ten years in industry as a
-professional programmer in languages such as Ruby, C, and C++.
-Andrew Gallant has taken a programming languages course heavily focused on 
-functional programming, but with no Haskell.
-Norman Ramsey is the local expert, but he wrote no code for this project.
+We hope that tools for computational biology in
+Haskell continue to mature.
+
+\section{What can you learn from our experience?}
+
+If~you are a computational biologist, 
+you can be productive in Haskell without extensive
+experience.
+Two~of us (Daniels and Gallant) are graduate students.
+Daniels has taken a graduate seminar in functional programming, which
+included some Haskell;
+Gallant has taken a programming-languages course which included
+significant functional programming but no Haskell.
+Ramsey is a professor who has used Haskell and other functional
+languages for several medium-sized projects,
+but his contribution was limited to minor ranting and refactoring.
 
 \subsection{Barriers}
-\remark{I feel like a lot of this is a summary of some of the problems we've
-  discussed previously in the paper. Is that okay? A summary may be worth it. 
----AG}
-
-As inexperienced functional programmers developing a complex piece of
-software in Haskell, we encountered some barriers.
-The first of these was profiling with GHC.
-We previously mentioned that profiling with GHC required annotating
-cost centers by hand, producing ugly code.
-Beyond that, we also found it very difficult to profile code at all.
-In particular, profiling required that all installed libraries be re-compiled
-with profiling support,which was a process that we could not replicate
-consistently.
-
-Another barrier we faced was how to test our code using QuickCheck.
-Indeed, the barrier was so formidable in our case that we still have not been 
-able to use it.
-The problems we faced specifically were two-fold: How do we start testing
-a complex module? How can we use QuickCheck to generate random inputs with 
-complex constraints? 
-We wished to test invariants around beta-strand placement 
-(potential solutions in our stochastic search), but found the QuickCheck generator
-to be difficult to write.
-
-Finally, one last barrier we faced was debugging runtime errors.
-Runtime errors in the form of out-of-bounds indexing were particularly 
-difficult to track down because they were not accompanied by a stack trace.
-(While profiling with GHC could help remedy this, profiling was itself a 
-barrier, as previously mentioned.)
-We had to resort to polluting our code with calls to the \texttt{trace} function 
-in order to track down where the out-of-bounds errors were occurring.
-
-% - profiling 
-% - debugging 
- % [ talk about profiling and debugging problems where?] 
-% [ something about quickcheck: how do we know when we have enough QC props? How do we start testing a complex module? ]
-
-\subsection{What you must know to succeed}
-If others would like to follow in our footsteps, there are a few
-things we believe one must know in order to succeed.
-Firstly, some experience with functional programming concepts is a necessity. 
-Specifically, coming from an imperative background, we found transforming input data 
-from the HMMER file format without the ability to mutate state was difficult.
-In contrast, writing ``pure math'' like the Viterbi algorithm felt easier
-and more natural than it would have been in C++.
-
-Secondly, identifying existing tools and libraries from Hackage that are mature 
-is also important.
-We found that some packages that could have been immensely useful---like one 
-that provides backtraces---were either not maintained or contained bugs.
-Some other packages like Pads were functional but immature, which resulted in 
-additional code to work around the parts of Pads that had not been implemented 
-yet. 
-In contrast, other packages such as \texttt{Data.Memocombinators} were
-useful and mature.
-
-Finally, a ``local expert'' or someone available to talk to with functional 
-programming expertise, can make the transition to a language like Haskell 
-easier.
-We don't believe a local expert is absolutely necessary, but answers 
-specifically related to functional programming idioms were easier to find when 
-a local expert was available.
-
-\subsection{Evaluation}
-\remark{I think we need more in this section. ---AG}
-In hindsight, we feel that there are several key facets of our project in 
-Haskell that deserve evaluation.
-Firstly, we found that existing tools and libraries were plentiful and 
-typically helped us greatly.
-In particular, \texttt{Data.Memocombinators} made dynamic 
-programming with Viterbi exquisitely simple and elegant.
-\texttt{Control.Parallel.Strategies} also made the parallelization of Viterbi trivial.
-Most of the tools we used came straight from Hackage, and although 
-documentation of some form was almost always present, it was sometimes 
-difficult to tell if a particular tool was right for us.
-In particular, parts of the documentation were inaccessible to us as beginning 
-functional programmers.
-Documentation that was accessible to us typically contained straight-forward 
-examples showing how to use that particular tool or library.
-
-Another component worthy of evaluation is the debugging process.
-Debugging in Haskell often takes the form of fixing type errors; which we think 
-is a great thing.
-However, in the case where debugging extends into runtime behavior, we couldn't 
-find any clearly defined approach to effective debugging.
-Stack traces are available via profiling, but this was difficult to find and 
-as we mentioned, profiling had its own barriers.
-
-Finally, we believe that we could have achieved a better division of modules in 
-our project, but that conflicted with one our goals: to get publishable results
-quickly. Despite being hasty programmers, we were still able to produce code 
-that is easy to understand and easily extensible. Namely, in our experience, 
-hasty Haskell code beats hasty code in a language like C or C++.
-
-In summary, with Haskell, we didn't have to master a completely different 
-design and development style but we did have to wrap our heads around 
-functional programming.\remark{Andrew, look at the comments below this
-part of the manuscript. See what you can do with them.}
-
-% - What does the local expert do? Do you have to have one? 
-%  
-% - To the FP community: Those who want us to abandon our old tools for debugging imperative code have not provided a clear path to follow (both intellectually, and via software tools). 
-%  
+
+In~\secref{awkward-profiling} above, we~mention the awkwardness of the
+cost-center annotations required for profiling.
+But~we also found it difficult just to get the profiler working.
+In~particular, profiling requires that all installed libraries be recompiled
+with profiling support.
+We~were not able to achieve such recompilations consistently.
+
+We~faced another barrier in testing.
+One of us (Ramsey) kept ranting that we should test our code using 
+QuickCheck \cite{claessen:quickcheck},
+but we could never figure out~how.
+We~were stumped by two problems:
+how to break down a complex module into parts that can usefully be
+unit tested,
+and how to implement QuickCheck generators for representations of complex
+data like beta-strand placements.
+We~know that QuickCheck is a powerful and proven tool,
+yet we were not smart enough to figure out how to use~it.
+
+Like other functional programmers, we~have found that often enough,
+once we have the types right, the code is right.
+But~much of our work involves computing with arrays and array indices,
+and in that domain, types don't help much.
+And we haven't been able to identify any systematic way of debugging
+run-time errors in Haskell programs.
+GHC's profiler can provide stack traces, but we~found this information
+difficult to discover, and as noted above,
+there are barriers to profiling.
+We're~aware that debugging lazy functional programs has been a topic
+of some research,\remark{Do we cite Hood, Hat, or Freya?}
+but one of the biggest barriers we encountered to using Haskell is
+that we have had to abandon our old approaches to debugging,
+and we~see no clear path to follow.
+
+\subsection{Information that will help you succeed}
+
+If~you want to use Haskell in your research, we~believe that you must
+have enough experience with functional programming that you can build
+\emph{all} the code you need, not only the code that is easy to write
+in a functional language.
+Implementing Viterbi's equations in Haskell was pure joy.
+Writing an iterative search in purely functional style was not
+difficult.
+When we had to transform data in the HMMER file format, \emph{without}
+using mutable state the way the C++ code does, we~struggled.
+
+You~also need to know that while the Haskell community offers many
+enticing tools, libraries, and packages,
+not all of them are worth using.
+Some are not ready for prime time, and some were once great but are
+no~longer maintained.
+The great packages, like \texttt{Data.Memocombinators} and Parallel
+Strategies, are truly great.
+But~for amateurs, it's not always easy to tell them from the wannabes
+and the has-beens.
+
+Most of the software we used came from Hackage, which is the Haskell
+community's software repository.
+We~sometimes found it hard to tell
+if a particular Hackage package was right for us.
+In particular, as beginning 
+functional programmers, we~found some documentation inaccessible.
+The documentation that we found most accessible typically contained
+examples showing how to use a particular tool or library.
+\remark{So what's our message here? What are we telling people to
+think or do differently? ---NR}
+
+As~in any endeavor, it's helpful to have access to experts.
+We~would have been better off if our in-house expert had been an
+enthusiastic student and not a busy professor.
+But we have been surprised and pleased by the help available from
+faraway experts on Stack Overflow and on Haskell mailing lists.
+Although a local expert makes things easier, one is not
+ absolutely necessary.
+
+\section{Conclusion}
+
+A~little knowledge of and a lot of love for functional programming was
+enough for us to carry out a successful research project in a language
+that is rarely used in computational biology.
+If~you \emph{want} to use Haskell---or one of your graduate students
+wants to use Haskell---you can
+succeed.
+
 % - No obvious body of knowledge on code improvement. 
 %
-% - The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies. 
-% How to tell what's good? How did we discover what's good? 
-% - Profiler can produce backtraces, but this was hard to discover.
-
-% \appendix
-% \section{Appendix Title}
-% 
-% This is the text of the appendix, if you need one.
-
+
+\ifnotcutting
 \acks
 
-We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford Larsen, and Nathan Ricci.
+We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford
+Larsen, and Nathan Ricci. 
+\fi
 
@@
 \begingroup
-\advance\leftskip by 0pt plus 10em
-\emph{The Haskell wave passes, and if you slide off the crest, you
+\parfillskip=0pt
+\advance\leftskip by 0pt plus 30em
+\emph{We live to surf the Haskell wave, but~if you slide off the crest, you
 drown.}
+\par
 \endgroup
 
+\break
+
+\appendix
+
+\section{Orphaned text}
+
+In~our C++ work, the faults we encounter most frequently are
+segmentation faults, 
+memory errors caused by pointer arithmetic or uninitialized 
+memory,
+and heap exhaustion.
+In~our Haskell work, the faults we encounter most frequently are
+type errors and bounds-checking errors in array and vector accesses.
+While we have found the bounds-checking errors difficult to debug,
+they are less difficult than memory errors.
+
 \end{document}

### 2012-03-11 mrfy_experience_report.tex
tex file scrubbed clean of all latex errors

@@
            {Department of Computer Science, Tufts University}
-           {\{ndaniels, agallant, nr\}@cs.tufts.edu}
+           {\char`\{ndaniels, agallant, nr\char`\}@cs.tufts.edu}
 
@@
 
+\seclabel{hofs}
+\seclabel{mrfy}
+
 %\subsection{MRFy Implementation}
@@
 bundled into a record called \texttt{SearchStrategy}.
-\smallverbatiminput{strategy}
+\smallfuzzverbatiminput{2.3pt}{strategy}
 In this record,
@@
 \end{itemize}
-This interface provides considerable leeway for experiments.
+This interface provides ample leeway for experiments.
 For~example, random-mutation hill climbing accepts only generations
@@
 We hope you find the search code simple and straightforward:
-\smallverbatiminput{search}
+\smallfuzzverbatiminput{11pt}{search}
 \emph{[MENTION AND QUANTIFY NOAH'S SUCCESSES IN ADDING NEW HEURISTICS.]}
@@
 
+\seclabel{penumbra}
+
 You~also need to know that while the Haskell community offers many
@@
 \advance\leftskip by 0pt plus 30em
-\emph{We live to surf the Haskell wave, but~if you slide off the crest, you
+\emph{You can live to surf the Haskell wave, but~if you slide off the crest, you
 drown.}

### 2012-03-11 mrfy_experience_report.tex
now those damn font warnings are dead!

@@
            {Department of Computer Science, Tufts University}
-           {\char`\{ndaniels, agallant, nr\char`\}@cs.tufts.edu}
+           {{\rmfamily\{}ndaniels, agallant, nr{\rmfamily\}}@cs.tufts.edu}
 

### 2012-03-11 mrfy_experience_report.tex
some love for verbs

@@
 enough for us to carry out a successful research project in a language
-that is rarely used in computational biology.
+that computational biologists seldom use.
 If~you \emph{want} to use Haskell---or one of your graduate students

### 2012-03-11 mrfy_experience_report.tex
consistent font sizes
remind what an Experience Report is

@@
 \begin{abstract}
+The Experience Report series is
+a body of published, refereed, citable \emph{evidence} that functional
+programming really works---or descriptions of obstacles that prevent
+it from working.   
+In~this Experience Report, we show that for computational biologists,
 Haskell combines the flexibility and rapid prototyping of a scripting
@@
 with scores, so we have defined a small abstraction:
-\verbatiminput{vscore}
+\smallverbatiminput{vscore}
 We~made \texttt{Scored} an instance of~\texttt{Ord} and of

### 2012-03-11 mrfy_experience_report.tex
remind what an Experience Report is
explain Scored in the paper

@@
 
+%
+% It can't hurt to remind everyone involved exactly what an Experience
+% Report is.
+%
+%
 \begin{abstract}
@@
 \texttt{Functor}.
+Think of a value of type \texttt{Scored~a} as a container holding
+ an~``\texttt a'' with a number written on the side.
+The \texttt{/+/} function adds to the number without touching the
+ container.
+Function 
+\begin{smallverbatim}
+fmap :: (a -> b) -> (Scored a -> Scored b)
+\end{smallverbatim}
+applies a function to what's inside the container, without touching
+the number.
+And containers are ordered purely by number, so applying
+\texttt{minimum} to a list of scored things chooses the thing with the
+smallest (and therefore best) score.
 
 Given the \texttt{Scored} abstraction, we can easily implement
-Viterbi's equations for a match state: 
-we~choose the best solution
+Viterbi's equations for a match state.
+We~choose the best solution
 from among the three predecessor states;

### 2012-03-11 mrfy_experience_report.tex
Fixed NR's botch of the search strategies and their types

@@
 or to try \texttt{nextGen} again with a different seed.
-The \texttt{accept} considers the score of the best
-placement of the trial generation, along with the history of previous
-best scores.
-It~also considers the \texttt{Age} of the trial generation, which is
-the total number of 
-generations tried.
+The \texttt{accept} function considers a list of type
+\texttt{[Scored~Age]}.
+At~the head of this list is the score of the best
+placement of the trial generation, together with the \texttt{Age} of the trial
+generation, which is the total number of generations tried.
+The remaining elements of the list hold the history of scores and ages for
+previously \emph{accepted} generations.
+(The list omits generations that didn't survive the \texttt{accept} trial.)
+For convenience, the \texttt{accept} function also receives
+the \texttt{Age} of the trial generation as an additional parameter.
 \item
 Finally, the \texttt{quit} function decides when to stop the search.
-It~too considers the history of scores and the age of the current
+It~too considers the list of accepted generations and the age of the current
 generation. 

### 2012-03-11 search.tex
Fixed NR's botch of the search strategies and their types

--- 2012-03-10 search.tex
@@
 search :: SearchStrategy -> Scorer -> [Seed]
-       -> (Scored Placement, [Score])
-search strat score (s:seeds) = runFrom seeds firstGen [] 0
+       -> (Scored Placement, [Scored Age])
+search strat score (s0:seeds) = runFrom seeds firstGen [] 0
  where
-  firstGen = map score $ gen0 strat s
-  runFrom :: [Seed] -> [Scored Placement] -> [Score]
-          -> Age -> (Scored Placement, [Score])
+  firstGen = map score $ gen0 strat s0
+  runFrom :: [Seed] -> [Scored Placement] -> [Scored Age]
+          -> Age -> (Scored Placement, [Scored Age])
   runFrom (s1:s2:seeds) oldPop oldHist age =
     let trialPop  = nextGen strat s1 score oldPop
-        trialHist = (scoreOf $ minimum newPop) : oldHist
+        trialHist = (fmap (const age) $ minimum trialPop)
+                  : oldHist
         (newPop, newHist) =
-          if accept strat s2 newHist age then
+          if accept strat s2 trialHist age then
             (trialPop, trialHist)

### 2012-03-11 mrfy_experience_report.tex
two questinos about the code
filled two 'holes' and added to the comparison section

@@
 encountered up to this point:
+\remark{Two questions: (1)~Why isn't the function called \texttt{vee}?
+(2)~Do we prefer \texttt{minimum (map from [Mat, Ins, Del])}?
+}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \smallfuzzverbatiminput{11pt}{search}
-\emph{[MENTION AND QUANTIFY NOAH'S SUCCESSES IN ADDING NEW HEURISTICS.]}
+
+We found it easy to write several stochastic search heuristics thanks to
+this framework.
+We first implemented random-mutation hill climbing in one day and roughly
+fifty lines of code.
+We next implemented simulated annealing, and thanks to Haskell's module system,
+we imported the \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} functions
+from the random-mutation hill climbing module.
+Indeed, we had only to write a new \texttt{accept} function, comprising only ten
+lines of code and half an hour of our time.
+We then implemented a genetic algorithm approach; this was still able to import
+existing \texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
+function was only a few lines of code.
+Understandably, though, the complexities of representing ``crossover'' of parent
+placements required forty lines of code and much of a day of thought.
 
@@
 and even after we discovered that feature, the results weren't good enough.
-\emph{[WHY WASN'T THE BACKTRACE FEATURE GOOD ENOUGH?]}
+Specifically, because the backtrace feature required profiling mode,
+it could only be used in very small test cases that didn't always trigger
+run-time errors, and one of us was unable to overcome difficulties in building
+various libraries in profiling mode.
 In~August 2011, 
@@
 It~uses clever pointer arithmetic.
+It relies on no external libraries beyond \texttt{zlib},
+but contains 1,000 lines of protein-structure code that ought
+to be in a library.
 
@@
 but it took most of a~year.
+This modification added 2,000 lines of code, and relies on
+system calls to external programs (sequence aligners) that we did not write.
 The mutable data structures were difficult to 
@@
 \cite{Menke:2010ti}.
-The implementation took 12,000~lines, including a single function of
+The implementation took 9,000~lines, including a single function of
 over 700 lines.
@@
 just understanding the existing code took several months.
+Part of SMURF's code includes the 1,000 lines of protein-structure code
+copied from the previous Matt software.
+
 For comparison, 
@@
 easy.
-\remark{I'm thinking this \P\ is not so strong. ---NR}
+MRFy currently contains 1,200 lines of Haskell code, though it relies heavily on
+a number of libraries beyond the standard Prelude.
+We found that adapting our software to explore a new search heuristic was typically
+less than a day's work.
+There are two major research ideas we wish to add to MRFy eventually, one of which
+\remark{how much should I say about profiles and alpha helices? ---NMD}
+involves looking at another type of nonlocal dependency (\textit{alpha helices} in contrast
+to the current beta strands).
+We expect that thanks to algebraic data types and typeclasses, these enhancements will be at
+most a few weeks of work; we do not believe these enhancements would even be feasible in SMURF.
+Indeed, Daniels tried to add another of these enhancements to SMURF two years ago and gave up due
+to the difficulty of the implementation.
+\remark{The above needs a lot of revision. It is a shitty first draft of a /P/. ---NMD}
 

### 2012-03-11 mrfy_experience_report.tex
questions are now TODO
edits to intro

@@
 The performance of our new implementation is comparable to the
-performance of~C++.
+performance of~C++ (\secref{perf}).
 \item
@@
 It is possible to use Haskell successfully for research even with a
-very modest background in functional programming. 
-\item
-The only significant impediment that Haskell pas presented to our
+very modest background in functional programming. We found our Haskell
+implementation to be more flexible about adding new research ideas
+than our past C++ code (\secref{comparo}).
+\item
+The only significant impediment that Haskell has presented to our
 research has been the difficulty of understanding and tuning the
@@
 encountered up to this point:
-\remark{Two questions: (1)~Why isn't the function called \texttt{vee}?
-(2)~Do we prefer \texttt{minimum (map from [Mat, Ins, Del])}?
+\remark{TODO: (1)~Call the inner function \texttt{vee}.
+(2)~Use \texttt{minimum (map from [Mat, Ins, Del])}
 }
@@
 \subsection{Performance}
+\seclabel{perf}
 
@@
 \section{Our previous experience compared}
+\seclabel{comparo}
 
@@
 computational biologists, have used~C++.
-Our~prior experience therefore includes building,
-maintaining, and extending research tools written in~C++.
-One~such tool is~Matt,
-a~structural protein aligner~\cite{Menke:2008wu}.
-Matt comprises about 12,000 lines of~C++.
+We~compare our experience with three tools:
+\begin{itemize}
+\item
+Matt~\cite{Menke:2008wu} is used to create alignments like that shown
+in \figref{alignment}.
+It~comprises about 12,000 lines of~C++.
+The only external tool or library it uses is \texttt{zlib}.
+Its initial development took \emph{[HOW LONG?]}.
+\item
+SMURF
+\cite{Menke:2010ti} is used \emph{[FOR WHAT?]}
+It~comprises about \emph{[HOW MANY?]} lines of~C++.
+Its initial development took \emph{[HOW LONG?]}.
+\item
+MRFy is used to detect homologous proteins in the presence of paired
+beta strands.
+It~comprises about \emph{[HOW MANY?]} lines of~Haskell
+Its initial development took about three months of part-time work.
+\remark{Loss of parallel structure---do we distinguish part time from
+full time?}
+\end{itemize}
+
 It uses such
@@
 dynamic programming which was hastily added to the SMURF homology detector
-\cite{Menke:2010ti}.
+
 The implementation took 9,000~lines, including a single function of
@@
 to the current beta strands).
-We expect that thanks to algebraic data types and typeclasses, these enhancements will be at
-most a few weeks of work; we do not believe these enhancements would even be feasible in SMURF.
-Indeed, Daniels tried to add another of these enhancements to SMURF two years ago and gave up due
+We expect that thanks to algebraic data types and typeclasses, these
+enhancements will be at 
+most a few weeks of work; we do not believe these enhancements would
+even be feasible in SMURF. 
+Indeed, Daniels tried to add another of these enhancements to SMURF
+two years ago and gave up due 
 to the difficulty of the implementation.

### 2012-03-11 mrfy_experience_report.tex
sketch of what NR is looking for in comparison section

@@
 The performance of our new implementation is comparable to the
-performance of~C++ (\secref{perf}).
+performance of~C++.
 \item
@@
 It is possible to use Haskell successfully for research even with a
-very modest background in functional programming. We found our Haskell
-implementation to be more flexible about adding new research ideas
-than our past C++ code (\secref{comparo}).
-\item
-The only significant impediment that Haskell has presented to our
+very modest background in functional programming. 
+\item
+The only significant impediment that Haskell pas presented to our
 research has been the difficulty of understanding and tuning the
@@
 \subsection{Performance}
-\seclabel{perf}
 
@@
 \section{Our previous experience compared}
-\seclabel{comparo}
 

### 2012-03-11 mrfy_experience_report.tex
holes filled, pushing back

--- 2012-03-11 mrfy_experience_report.tex
@@
 The only external tool or library it uses is \texttt{zlib}.
-Its initial development took \emph{[HOW LONG?]}.
+Its initial development took \emph{two years}.
 \item
 SMURF
-\cite{Menke:2010ti} is used \emph{[FOR WHAT?]}
-It~comprises about \emph{[HOW MANY?]} lines of~C++.
-Its initial development took \emph{[HOW LONG?]}.
+\cite{Menke:2010ti} is used to detect homologous proteins in the presence
+of paired beta strands.
+It~comprises about 9,000 lines of~C++.
+Its initial development took a year and a half.
 \item
@@
 beta strands.
-It~comprises about \emph{[HOW MANY?]} lines of~Haskell
-Its initial development took about three months of part-time work.
-\remark{Loss of parallel structure---do we distinguish part time from
-full time?}
+It~comprises about 1,200 lines of~Haskell
+Its initial development took about three months.
+\remark{Others were mostly full-time work, but no need to distinguish.}
 \end{itemize}
 
-It uses such
+Matt uses such
 data structures such as mutable oct-trees, vectors, and arrays.

### 2012-03-11 mrfy_experience_report.tex
massage the comparison section

@@
 \newcommand\naive{na\"\i ve}
+\usepackage{url}
 \usepackage{amsmath}
@@
 The only external tool or library it uses is \texttt{zlib}.
-Its initial development took \emph{two years}.
+Its initial development took two years.
 \item
@@
 of paired beta strands.
-It~comprises about 9,000 lines of~C++.
+It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
+shared with~Matt.
+It~uses no external tools or libraries.
 Its initial development took a year and a half.
@@
 It~comprises about 1,200 lines of~Haskell
+It~uses several external tools and libraries, of which the most
+notable are PADS-Haskell, the \path{Data.MemoCombinators} library,
+the BioHaskell bioinformatics library, and the \path{Data.Vector} library.
 Its initial development took about three months.
@@
 \end{itemize}
-
+Like much research software, all three of these tools were written
+ in~haste.
+We~have experience modifying the older tools.
+
+We~modified Matt to use information about sequences as well as structure.
+The modification added 2,000 lines of code, and it calls
+external sequence aligners that we did not write.
+We~thought the modification would take three months, 
+but it took most of a~year.
 Matt uses such
@@
 It~uses clever pointer arithmetic.
-It relies on no external libraries beyond \texttt{zlib},
-but contains 1,000 lines of protein-structure code that ought
-to be in a library.
-
-We~modified Matt to use information about sequences as well as structure.
-We~thought the modification would take three months, 
-but it took most of a~year.
-This modification added 2,000 lines of code, and relies on
-system calls to external programs (sequence aligners) that we did not write.
 The mutable data structures were difficult to 
@@
 nearly every change resulted in new segfaults.
-We~had hoped to add support for partial alignments,
+
+We~had hoped to extend Matt further, with support for partial alignments,
 which we~expected to require only 
@@
 
-Like many researchers, we~and our colleagues often write code hastily.
-Hastily written C++ code can get a job done, 
-but we have found it expensive in the long run.
-Our most painful experience was with an implementation of
-\textit{multidimensional}
-dynamic programming which was hastily added to the SMURF homology detector
-
-The implementation took 9,000~lines, including a single function of
-over 700 lines.
-When we added simulated evolution to this code, a relatively minor
-enhancement \cite{Daniels:2012}, 
+Our most painful experience was adding ``simulated evolution''
+to~SMURF \cite{Daniels:2012}. 
+Although simulated evolution represents a relatively minor
+enhancement,
 just understanding the existing code took several months.
-Part of SMURF's code includes the 1,000 lines of protein-structure code
-copied from the previous Matt software.
-
-For comparison, 
-building the first functioning version of MRFy took three months of part-time work.
-And we expect that higher-order functions will make extending MRFy
-easy.
-MRFy currently contains 1,200 lines of Haskell code, though it relies heavily on
-a number of libraries beyond the standard Prelude.
-We found that adapting our software to explore a new search heuristic was typically
-less than a day's work.
-There are two major research ideas we wish to add to MRFy eventually, one of which
-\remark{how much should I say about profiles and alpha helices? ---NMD}
-involves looking at another type of nonlocal dependency (\textit{alpha helices} in contrast
-to the current beta strands).
-We expect that thanks to algebraic data types and typeclasses, these
-enhancements will be at 
-most a few weeks of work; we do not believe these enhancements would
-even be feasible in SMURF. 
-Indeed, Daniels tried to add another of these enhancements to SMURF
-two years ago and gave up due 
-to the difficulty of the implementation.
-\remark{The above needs a lot of revision. It is a shitty first draft of a /P/. ---NMD}
+
+We~built MRFy quickly, and we expect that
+higher-order functions will MRFy easy to extend.
+Each new addition to MRFy's stochastic search has taken at most a day
+to implement.
+
+%
+% An Experience Report is about what you actually did, not what you
+% wish to do :-)
+%
+%%  There are two major research ideas we wish to add to MRFy eventually, one of which
+%%  \remark{how much should I say about profiles and alpha helices? ---NMD}
+%%  involves looking at another type of nonlocal dependency (\textit{alpha
+%%  helices} in contrast 
+%%  to the current beta strands).
+%%  We expect that thanks to algebraic data types and typeclasses, these
+%%  enhancements will be at 
+%%  most a few weeks of work; we do not believe these enhancements would
+%%  even be feasible in SMURF. 
+%%  Indeed, Daniels tried to add another of these enhancements to SMURF
+%%  two years ago and gave up due 
+%%  to the difficulty of the implementation.
+%%  \remark{The above needs a lot of revision. It is a shitty first
+%%  draft of a /P/. ---NMD}
 
@@
 
-When we look beyond our own group to computational biology more
-broadly, our experiences with other software is better.
+Looking beyond our research group to computational biology more
+broadly, our experience with other software is better.
 Little of it is written in functional languages, 

### 2012-03-11 mrfy_experience_report.tex
deploy the new intro

@@
 ``scripting'' or domain-specific language like
-Perl, Python, Ruby, SPSS, or~R.\remark{Don't forget audience.}
-In~this Experience Report, we show that computational biologists can
-successfully replace both these families of languages with Haskell:
-\begin{itemize}
-\item
-Our experimental Haskell programs are as easy to understand and change
- as the programs our group writes using scripting languages.
-\item
-The performance of our Haskell code is comparable to the performance
-of~C++.
-\item
-Although we have had to write a new implementation of a standard
-algorithm, the cost was low, and the connection between code and
-algorithm is unusually strong (\secref{viterbi}).
-\item
-We have experienced barriers to entry, of which the most salient has been
-the difficulty of understanding and tuning  the performance of Haskell
-programs. 
-We found the other barriers 
-relatively easy to overcome.
-\end{itemize}
-Our experience is based on MRFy (pronounced ``Murphy''), which
-implements a novel family of algorithms for finding remote homologs to 
-newly discovered proteins.
-
-\emph{[Alternate introduction, trying to tie the points more closely
-to the content of the paper:\\
+Perl, Python, Ruby, SPSS, or~R.
 In~this paper, we report on an attempt to replace both these families
@@
 The performance of our new implementation is comparable to the
-performance of~C++ (\secref{perf}).
+performance of~C++, and parallel performance was easy to achieve
+(\secref{perf}). 
 \item
@@
 Some deliver very high leverage, but others are not ready for prime
-time (\secref{penumbra}).
-We~found it difficult to tell which is which in advance.
+time.
+We~found it difficult to tell which is which in advance (\secref{penumbra}).
 \item
 It is possible to use Haskell successfully for research even with a
-very modest background in functional programming. We found our Haskell
-implementation to be more flexible about adding new research ideas
-than our past C++ code (\secref{comparo}).
+very modest background in functional programming. 
+Modifying our group's C++ code to try out new research ideas has been
+difficult;
+trying new ideas in Haskell has been easy
+(\secref{comparo}).
 \item
 The only significant impediment that Haskell has presented to our
-research has been the difficulty of understanding and tuning the
-performance of Haskell programs.
+research has been the difficulty of understanding and improving the
+performance of Haskell programs (\secref{awkward-profiling}).
 \end{itemize}
-]}
 
@@
 %% \subsection{The Viterbi Algorithm}
-
-\seclabel{viterbi}
 
@@
 
+\seclabel{viterbi}
+
 Given a hidden Markov model, 
@@
 \subsection{Performance}
+
 \seclabel{perf}

### 2012-03-11 mrfy_experience_report.tex
added a bit about runtime performance.

@@
 
+On the protein structures for which it was originally designed, SMURF can 
+compute the parse of a query sequence to a MRF based on those structures in 
+less than 10 seconds on a typical workstation.
+However, on somewhat more 
+complex protein structures, SMURF may take two to three minutes on the same 
+hardware. 
+On the most complex protein structures, SMURF exhausts all memory on 
+any system we have available, and must be killed. 
+In contrast, MRFy's runtime 
+for any of these protein structures is typically less than 30 seconds. 
+This 
+meets the performance goals we had, and means that MRFy is fast enough to scale 
+to whole-genome scans.
+
 Haskell encourages hasty programmers to slow down.

### 2012-03-11 mrfy_experience_report.tex
intro edited

@@
 \begin{abstract}
-The Experience Report series is
-a body of published, refereed, citable \emph{evidence} that functional
-programming really works---or descriptions of obstacles that prevent
-it from working.   
-In~this Experience Report, we show that for computational biologists,
+An Experience Report provides {evidence} that functional
+programming really works, or it describes obstacles that prevent
+functional programming from working.   
+Here, we show that for computational biologists,
 Haskell combines the flexibility and rapid prototyping of a scripting
 language with performance comparable to~C++.
-The Haskell language and libraries provide powerful abstractions, 
-but we couldn't always get them to work!
+The~Haskell language and libraries provide powerful abstractions, 
+but we couldn't always get them to work.
 Despite some speed bumps around profiling, debugging, and the
-uncertain status of any given package, we~found Haskell's ecosystem
-provided fine support for developing new algorithms in computational
+uncertain status of any given package, we~found that Haskell's ecosystem
+made it easy for us to develop new algorithms in computational
 biology.
@@
 productivity are more important;
-this software is usually written in a high-level, dynamically-typed
-``scripting'' or domain-specific language like
+this software is usually written in a high-level
+or domain-specific language like
 Perl, Python, Ruby, SPSS, or~R.
@@
 algorithm is unusually strong (\secref{viterbi}).
-The performance of our new implementation is comparable to the
-performance of~C++, and parallel performance was easy to achieve
+Our new Haskell code performs competitively with~C++,
+and parallel performance was easy to achieve
 (\secref{perf}). 
 \item
-As expected, higher-order functions made it unusually easy to
+Higher-order functions made it unusually easy to
 implement and experiment with a family of related algorithms
@@
 \item
-Haskell is surrounded by a penumbra of libraries and tools that
+The only significant impediment that Haskell presented to our
+research was the difficulty of understanding and improving the
+performance of Haskell programs (\secref{awkward-profiling}).
+\item
+We~used Haskell successfully even though we are computational
+biologists with little background in functional programming.
+Modifying our group's C++ code to try new research ideas has been
+difficult;
+trying new ideas in Haskell has been easy
+(\secref{comparo}).
+\item
+The Haskell language is surrounded by a penumbra of libraries and tools that
 promise powerful abstractions.
@@
 time.
-We~found it difficult to tell which is which in advance (\secref{penumbra}).
-\item
-It is possible to use Haskell successfully for research even with a
-very modest background in functional programming. 
-Modifying our group's C++ code to try out new research ideas has been
-difficult;
-trying new ideas in Haskell has been easy
-(\secref{comparo}).
-\item
-The only significant impediment that Haskell has presented to our
-research has been the difficulty of understanding and improving the
-performance of Haskell programs (\secref{awkward-profiling}).
+We~found the two families difficult to distinguish in advance (\secref{penumbra}).
 \end{itemize}
@@
 molecules to carry out the functions of living cells: metabolism, regulation, 
-signalling, and more.
+signalling, and so on.
 A~protein's function is determined by its structure, 
@@
 form the protein.
-\remark{Lost: there are only 20-odd amino acids}
 The amino-acid sequence is ultimately determined by a sequence of
@@
 Computational biologists detect homologies by building 
-algorithms which, given a \emph{query sequence} of amino acids,
+algorithms which, given a {query sequence} of amino acids,
 find known proteins of similar structure.
@@
 different amino-acid sequences, which are not close in edit distance.
-Our~research contribution---the MRFy algorithm---can detect homologies
+Our~research contribution---a software tool called MRFy (pronounced
+``Murphy'')---can detect homologies 
 in amino-acid sequences that are only distantly related.
@@
 %
-To~enable you to understand our experience with Haskell,
-we~sketch not only the MRFy algorithm (\secref{mrfy})
-but also one of the older hidden-Markov algorithms (\secref{viterbi}),
-which is incorporated into~MRFy.
-
-\remark{LOST: We conclude that implementations of mathematically-intensive 
-algorithms such as these lend themselves to the Haskell language, particularly 
-when the programmer needs a flexible implementation that simplifies 
-experimenting with variants of the algorithm.}
-
-% As a brief refresher in the standard dogma of genetics, recall that genes 
-% (strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
-% adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
-% and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
-% into the 20 naturally-occuring amino acids, which form peptide chains -- 
-% proteins -- which fold into complex structures in the cell. These proteins are 
-% cellular machinery, performing the varied functions an organism needs to 
-% survive.
-
-% However, determining what newly sequenced genes actually do -- when they become 
-% proteins, what structures those proteins fold into, and what functions they 
-% perform -- has not kept pace. Determining the atomic coordinates of a newly 
-% discovered protein may require as much as two graduate-student-years in a lab! 
-% Fortunately, determining the full structure of a protein is not always 
-% necessary. Often, we can identify well-known proteins of similar sequence or 
-% structure to a new protein, and thus make educated guesses as to the structure 
-% and function of the new protein.
-% 
-% The problem of taking a newly found protein sequence about which only the 
-% sequence is known, and determining what existing proteins of known structure 
-% and function most closely resemble that new protein, is known as homology 
-% detection. When proteins of similar sequence can be found, this problem is 
-% largely solved. To solve the homology detection problem, computational 
-% biologists have developed fast, approximate methods for determining the 
-% structure and function of proteins based on their sequence. The most popular 
-% software for homology detection is called HMMER, which uses a hidden Markov 
-% model to capture evolutionary change.
-
-% However, when there are no known proteins of similar sequence, only 
-% evolutionarily distant proteins that may share similar structure, this problem 
-% is called remote homology detection. Hidden Markov model approaches begin to 
-% fail as sequence similarity falls off. Recently, Markov random field 
-% approaches, which capture non-local interactions in the protein structure, have 
-% been shown to perform well. However, these approaches face the challenge of 
-% increased computational complexity; the SMURF program exhibits exponential time 
-% complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
-% simplifying the Markov random field's dependency graph. MRFy is a 
-% stochastic search approach to Markov random fields.
+
+%
+% I hope the new intro renders these two sentences unnecessary. ---NR
+%
+%
+%%  
+%%  To~enable you to understand our experience with Haskell,
+%%  we~sketch not only the MRFy algorithm (\secref{mrfy})
+%%  but also one of the older hidden-Markov algorithms (\secref{viterbi}),
+%%  which is incorporated into~MRFy.
 
 \section{The Software}
-
-%% \subsection{The Viterbi Algorithm}
 

### 2012-03-11 mrfy_experience_report.tex
edites through section 3.1

@@
+\newif\ifpagetuning
+
+\pagetuningtrue % trying to get page breaks in the best spots
+
 \newif\ifnotcutting \notcuttingfalse % set if not cutting down to size
@@
 \newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
-\newcommand\alignwidth{\ensuremath R} % number of columns in an alignment
+\newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
 \newcommand\pairedwith[1]{{\pi(#1)}}
+
+\ifpagetuning
+  \newcommand\tuningbox{\vbox}
+\else
+  \newcomming\tuningbox{\relax}
+\fi
 
@@
 \emph{alignment}. 
-This alignment represents structural
-correspondence among individual amino acids in a set of homologous proteins.
+This alignment relates individual amino acids in a set of homologous proteins.
 An~alignment may be represented as a matrix
-(Figure~\ref{alignment}) 
 in which each row corresponds to the amino-acid sequence of a protein,
 and each column groups amino acids that play similar roles in
-different proteins.
-
-An alignment may contain \emph{gaps} (notated as dashes in
-\figref{alignment}) for some proteins.
+different proteins (Figure~\ref{alignment}).
+
+An alignment may contain \emph{gaps}, which in 
+\figref{alignment} are shown as dashes.
 A~gap in row~2, column~$j$ indicates that as proteins evolved, either 
@@
 and the few proteins with gaps probably lost amino acids via
-\emph{deletions}.\remark{NMD: Missing is the fact that all these models
-are directionless; they don't care whether something was gained or lost
-over time. Perhaps this doesn't matter.}
+\emph{deletions}.
+%%  \remark{NMD: Missing is the fact that all these models
+%%  are directionless; they don't care whether something was gained or lost
+%%  over time. But this doesn't matter.}
 If~column~$j$ contains \emph{mostly} gaps, 
@@
 
-\caption{A~structural alignment of four proteins $(\alignwidth = 4)$}
+\caption{A~structural alignment of four proteins $(\alignwidth = 12)$}
 \label{alignment} 
@@
 
-A~hidden Markov model carries probabilities on its states and on its
+A~hidden Markov model carries probabilities on some states and on all
 state transitions.
@@
 deletions.
-\remark{citation needed?}
 In~other words, insertions and deletions cannot be mixed without an
 intervening match.
-This restriction means that between two adjacent columns, six
-transitions are possible.
-An~insertion state may also transition to itself.
+This restriction means that there are only seven possible transitions
+into the states of any one column \cite{Eddy:1998ut}. 
 \begin{itemize} 
@@
 Depending on the predecessor state, 
-the probability of such a transition is notated
+the probability of such a transition is 
 $\txprobj M M$, $\txprobj I M$, or $\txprobj D M$.
@@
  column~$j$ is more likely when column~$j$ is a non-consensus column.
-The probability of such a transition is notated
+The probability of such a transition is 
 $\txprobj M D$ or $\txprobj D D$.
@@
 A~transition into an insertion state in
- column~$j$ is relatively unlikely,
-but it is more likely when column~$j$ is a non-consensus column.
-The probability of such a transition is notated
+ column~$j$ is more likely when column~$j$ is a non-consensus column.
+The probability of such a transition is 
 $\txprobjj M I$ or $\txprobjj I I$.
@@
 Given a hidden Markov model, 
-an~established software package called HMMER (``hammer'') 
+an~established software package called HMMER (pronounced ``hammer'') 
 can compute the probability
@@
 probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
-(The logarithm function is monotonic, so the path that maximizes the
-log of the probability is the most likely path.)
-The~specification is shown on the left-hand side of \figref{viterbi}.
-The probability $V_j^M(i)$ represents the probability of the most
+The logarithm function is monotonic, so the path that maximizes the
+log of the probability is the most likely path.
+
+The computation is specified on the left-hand side of \figref{viterbi}.
+A~probability $V_j^M(i)$ represents the probability of the most
 likely path of the first~$i$ amino acids in the query sequence,
 terminating with placement of amino acid $x_i$ in state~$M_j$.
-The probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
-Equations like these can be found in more recent literature
-\cite{Durbin:1998wz,Eddy:1998ut}.
+Probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
+These
+equations are very clearly explained by
+\citet[Chapter~5]{Durbin:1998wz}
 
@@
 correct.
-We~begin with our representations.
+To~present some code,
+we~begin with representations.
 
@@
 we might represent an individual amino acid~$x$
-using a value of algebraic data type, thus:
+using a value of algebraic data type:
 \begin{smallverbatim}
@@
 \end{smallverbatim}
-But our primary goal was to get something working quickly with a
+But our primary goal was to get something working quickly,
+and our hidden Markov models are stored using a
 legacy file format that represents each amino acid as a small integer.
@@
 
-The log of a probability is simply a number.
+The log of a probability is just a number.
 To distinguish a transition probability from other numbers, we wrap it in
-the value constructor \texttt{TProb}:
+a value constructor:
 \begin{smallverbatim}
@@
 Each tuple corresponds to a ``node'' in the hidden Markov model
-(a~dashed ovel in \figref{plan7}), and to a column of the original
+(a~dashed oval in \figref{plan7}) and to a column of the original
 alignment.
@@
 \end{smallverbatim}
+In~our code, we select from this 7-tuple using the \texttt{edge}
+function:
+\smallverbatiminput{edge}
+In MRFy, the first two arguments to \texttt{edge} are known at compile
+time, and we expect \texttt{edge} to be inlined.
 
 The representation of a node includes the probabilities of transitions
-into the three states of that node, plus the tables
+into the states of that node, plus the tables
 of emission probabilities for the match and insertion states.
@@
 In~Viterbi's equations,
-the probability in each state is a simple function of the probability
+the probability in each state is a function of the probabilities
 in its predecessor states, 
 and all probabilities can be computed by a classic dynamic-programming
-algorithm
-which starts at the begin state,
+algorithm.
+This algorithm starts at the begin state,
 computes probabilites in nodes $1$~through~$\alignwidth$ in
@@
 One of us implemented this algorithm, storing the probabilities in an array.
-\remark{Why is the number of dimensions 3 and not 2 (indices $i$, $j$,
-and what??) and why do we care?}
 The cost was
 $O(|N|\times|\alignwidth|)$;
-in our work, $\alignwidth$~and~$N$ range from several hundred to a few
+in~MRFy, $\alignwidth$~and~$N$ range from several hundred to a few
 thousand.
@@
 
-Another of us was curious to try to code Viterbi's equations
+Another of us was curious to try coding Viterbi's equations
 directly as recursive functions.
@@
 \emph{memoized}.
-To~show the memoized code, we have to confess to a small
+To~show our code, we have to confess to a 
 transformation of Viterbi's equations.
-The logarithms on the left of \figref{viterbi} are computed when the
+The~logarithms on the left of \figref{viterbi} are computed when the
 hidden Markov model is constructed, and the resulting numbers are
 stored in a file.
-Moreover, to make the file more pleasant to read, the logarithms are
+And~to~make this file more pleasant to read, the logarithms are
 negated.
 (The logarithm of a probability is typically negative, and negating
-the logarithm eliminates hordes of minus signs from the file.)
+the logarithm eliminates hordes of minus signs.)
 We~therefore implement a transformed version of Viterbi's equations,
-shown on the right-hand side of \figref{viterbi}, which minimize the
+shown on the right-hand side of \figref{viterbi}.
+The transformed equations minimize the
 negated log probability for each combination of node~$j$, amino
 acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
-
-Our implementation computes not only the minimum
-negated log probability, called a \texttt{Score},
-but also the path of states which leads to that most likely outcome. 
-Elsewhere, MRFy pairs not only paths but also other types of values
-with scores, so we have defined a small abstraction:
+A~negated log probability is called a~\emph{score}.
+
+MRFy computes not only the minimum score but
+also the path of states which produces that score.
+Scores can be usefully paired with many types of values,
+so we have defined a small abstraction:
 \smallverbatiminput{vscore}
-We~made \texttt{Scored} an instance of~\texttt{Ord} and of
-\texttt{Functor}.
 Think of a value of type \texttt{Scored~a} as a container holding
- an~``\texttt a'' with a number written on the side.
-The \texttt{/+/} function adds to the number without touching the
+ an~``\texttt a'' with a score written on the side.
+The \texttt{/+/} function adds to the score without touching the
  container.
-Function 
+
+\tuningbox{
+We~also made \texttt{Scored} an instance of~\texttt{Ord} and of
+\texttt{Functor}, so
 \begin{smallverbatim}
@@
 applies a function to what's inside the container, without touching
-the number.
-And containers are ordered purely by number, so applying
+the score.
+And containers are ordered by score alone, so applying
 \texttt{minimum} to a list of scored things chooses the thing with the
 smallest (and therefore best) score.
+}
 
@@
 Viterbi's equations for a match state.
-We~choose the best solution
+To~implement the equation at the top right of \figref{viterbi},
+we~choose the best solution
 from among the three predecessor states;
@@
 encountered up to this point:
-\remark{TODO: (1)~Call the inner function \texttt{vee}.
+\remark{TODO: 
+(1)~Call the inner function \texttt{vee}
 (2)~Use \texttt{minimum (map from [Mat, Ins, Del])}
+(3)~Explain the recursive call
 }
@@
 appropriate transition score (one of the $a'$~values).
-% Similarly, the insertion equation is:
-% \begin{verbatim}                    
-%   -- consume an observation but not a node
-%   | s == ins = let trans t prev =
-%                    (score + tProb n t + (eProb n o), ins:path)
-%                     where (score, path) = viterbi prev n (o - 1)
-%                 in minimum [
-%                     trans m_i mat,
-%                     trans i_i ins
-%                     ]
-% \end{verbatim}
-% Finally, the deletion equation is:
-% \begin{verbatim}                    
-%   -- consume a node but not an observation
-%   | s == del = let trans t prev = 
-%                       (score + tProb (n-1) t, del:path)
-%                       where (score, path) = viterbi prev (n - 1) o
-%                 in minimum [
-%                     trans m_d mat,
-%                     trans d_d del
-%                     ]
-% \end{verbatim}
-Programming the algorithm in this form was not only intellectually
-satisfying; keeping the code close to the math also helped us find
-bugs.
+Implementing the equation in this form was not only intellectually
+satisfying; it~also helped us find
+bugs.\remark{NR doesn't think the bug story is that interesting.}
 The most memorable bug was that in one of our base cases, 
@@
 
-Whatever the attractions of the code above, in our work we also
-require high performance.
+Whatever the attractions of the code above, our work also
+requires high performance.
 To make the simple recursive code efficient, we used Luke Palmer's
@@
 and it makes recursive calls to \verb+viterbi+, which is memoized:
-\begin{smallverbatim}
-viterbi state j i = 
-  Memo.memo3 (Memo.arrayRange (Mat, End)) 
-             (Memo.arrayRange (0, numNodes))
-             (Memo.arrayRange (0, seqlen)) 
-             viterbi' state j i
-\end{smallverbatim}
+\smallverbatiminput{memo}
 The value
@@
 dynamic-programming code.
-And~the short fragment above is the \emph{only} part of the code
+And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
 devoted to improving performance.
-By contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
+By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
 spend much of their code 
@@
 \figref{viterbi}, we~retired the classic dynamic-programming version.
-
-\remark{NR doesn't think the bug story is that interesting.}
 

### 2012-03-11 mrfy_experience_report.tex
edit in progress (memoization)

@@
 
-%\subsection{MRFy Implementation}
-
 Our implementations of hidden Markov models and Viterbi's algorithm
 duplicate existing functionality.
-They form just one component of our new tool, MRFy.
+In~MRFy, they are used to implement new functionality.
 
@@
 amino-acid sequences.
-But real proteins are three-dimensional,
-and when a real protein folds, groups of amino acids
+But when a real protein folds in three dimensions, 
+groups of amino acids 
 that are far away in the one-dimensional sequence can be
@@
 But a mutation in a paired beta strand \emph{can} survive if it is
- accompanied by a compensatory mutation in the strand with which it is
+ accompanied by a \emph{compensatory} mutation in the strand with which it is
  paired. 
-The probability of finding amino acid $x_i$ in column~$j$ of an alignment may
+The probability of finding amino acid~$x_i$ in column~$j$ of an alignment may
  therefore depend on the amino acid in the paired column
@@
 modeled with an ordinary hidden Markov model.
-The~combined model, as shown in \figref{mrf}, is called a
+The~combined model, an~example of which is shown in \figref{mrf}, is called a
 \textit{Markov random field}. 
 Because of nonlocal interactions between paired beta strands, 
-the likelihood-maximization for a Markov random field does not
+the likelihood-maximization problem for a Markov random field does not
  exhibit optimal substructure, and dynamic
  programming cannot solve it quickly \cite{Menke:2010ti,Daniels:2012}.
-The~purpose of MRFy is to explore new methods of \emph{approximating}
+MRFy explores new methods of \emph{approximating}
 maximum likelihood.
@@
 MRFy treats the beta 
-strands, which are build into the model,
- as ``beads'' which can be slid along the query sequence to determine
- a \emph{placement} of the beta strands.
+strands, which are built into the model,
+ as ``beads'' which can be slid along the query sequence.
+A~set of positions of beta strands is called
+ a \emph{placement}.
 A~placement's probability is computed
@@
 \emph{conditioned} on the placement.)
-The likelihood is computed by a function \texttt{score},
-which given a model and a {placement} returns the {placement}
-paired with a \emph{score}.
-A~score is a negated log probability, so lower scores are better.
 
@@
 annealing, multistart simulated annealing, and a genetic algorithm.
-\remark{It would be pleasant to connect these to the functions below,
-but we probably lack room.}
 Our~exploration relies on higher-order functions.
 
-MRFy's search works with a sequence of \emph{generations}.
-A~generation is a collection of scored \emph{placements}.
+MRFy's search computes a sequence of \emph{generations}.
+A~generation is a collection of scored {placements}.
 We~pass the
-search function a \texttt{Scorer}, which is the result of partially
-applying \texttt{score} to the model.
+search function a \texttt{Scorer}, which runs Viterbi's algorithm.
 \smallverbatiminput{scoredecl}
-A~search algorithm is characterized by four functions, which are
+A~search heuristic is characterized by four functions, which are
 bundled into a record called \texttt{SearchStrategy}.
@@
 \item
-A~new generation produced by \texttt{nextGen} may not be any better
+A~new generation produced by \texttt{nextGen} may be no better
 than the generation it replaces.
@@
 For convenience, the \texttt{accept} function also receives
-the \texttt{Age} of the trial generation as an additional parameter.
+the \texttt{Age} of the trial generation.
 \item
@@
 hill climbing simply permutes a placement into a new one.
-The interface \texttt{quit} makes it easy
+The \texttt{quit} interface makes it easy
 to implement such termination criteria as a fixed number of generations

### 2012-03-11 mrfy_experience_report.tex
new story about memoization

@@
 encountered up to this point:
-\remark{TODO: 
-(1)~Call the inner function \texttt{vee}
-(2)~Use \texttt{minimum (map from [Mat, Ins, Del])}
-(3)~Explain the recursive call
-}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 appropriate transition score (one of the $a'$~values).
-Implementing the equation in this form was not only intellectually
-satisfying; it~also helped us find
-bugs.\remark{NR doesn't think the bug story is that interesting.}
-The most memorable bug was that in one of our base cases, 
-we~mistakenly returned an empty path of states;
-the correct result was a singleton path containing the initial state.
-Very quickly after observing a missing state in the output, we were
-able to find the faulty case in the code.
-
-Whatever the attractions of the code above, our work also
-requires high performance.
-To make the simple recursive code efficient, we used Luke Palmer's
-\texttt{Data.Memocombinators} package.
-The function \verb+viterbi'+ is unmemoized,
-and it makes recursive calls to \verb+viterbi+, which is memoized:
+Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
+so
+the call to~\verb+vee''+ produces the same result as a recursive call
+to~\verb+vee'+, but faster:
 \smallverbatiminput{memo}
+Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
+Luke Palmer's
+\path{Data.Memocombinators} package.
 The value
 \texttt{numNodes} represents~$\alignwidth$,
-and \texttt{seqlen} represents~$N$.\remark{I hope this is right. ---NR}
-
-This memoized recursive code performs just as well as the classic
+and \texttt{seqlen} represents~$N$.
+
+Memoization makes \verb+vee'+ perform just as well as the classic
 dynamic-programming code.
@@
 
-We hope you find the search code simple and straightforward:
+We hope you find the search simple and straightforward:
 \smallfuzzverbatiminput{11pt}{search}
@@
 
+Implementing the equation in this form was not only intellectually
+satisfying; it~also helped us find
+bugs.
+The most memorable bug was that in one of our base cases, 
+we~mistakenly returned an empty path of states;
+the correct result was a singleton path containing the initial state.
+Very quickly after observing a missing state in the output, we were
+able to find the faulty case in the code.
+
 \end{document}

### 2012-03-11 mrfy_experience_report.tex
finished an editing pass over section 3.2

@@
 \end{itemize}
-This interface provides ample leeway for experiments.
-For~example, random-mutation hill climbing accepts only generations
-with better scores than their predecessors,
-while simulated annealing may accept a generation with a worse score.
-A~genetic algorithm uses \texttt{nextGen} to
-combine placements from the previous generation, while
-random-mutation
-hill climbing simply permutes a placement into a new one.
-The \texttt{quit} interface makes it easy
-to implement such termination criteria as a fixed number of generations
-or a limit of going $k$~generations without achieving more than $E$
-improvement in score.
-
-\seclabel{search}
-
+\seclabel{search}%
 We hope you find the search simple and straightforward:
@@
 
-We found it easy to write several stochastic search heuristics thanks to
-this framework.
-We first implemented random-mutation hill climbing in one day and roughly
-fifty lines of code.
-We next implemented simulated annealing, and thanks to Haskell's module system,
-we imported the \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} functions
-from the random-mutation hill climbing module.
-Indeed, we had only to write a new \texttt{accept} function, comprising only ten
-lines of code and half an hour of our time.
-We then implemented a genetic algorithm approach; this was still able to import
-existing \texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
+The \texttt{SearchStrategy} interface provides ample scope for
+experiments. 
+Random-mutation hill took fifty lines of code and one~day to
+implement.
+Simulated annealing reuses
+ \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} 
+from random-mutation hill climbing.
+Simulated annealing's  new \texttt{accept} function took ten
+lines of code and half an hour to implement.
+(Hill climbing accepts a generation only if it scores better than its
+predecessor;
+simulated annealing may accept a generation with a worse score.)
+Our genetic algorithm reuses the same
+\texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
 function was only a few lines of code.
-Understandably, though, the complexities of representing ``crossover'' of parent
-placements required forty lines of code and much of a day of thought.
+But~the recombination of parent
+placements in a new \texttt{nextGen} function took forty lines of code
+and a full day to implement.
+Our~current \texttt{quit} function simply runs for a fixed number of
+generations, but its interface will enable us to try other criteria,
+such as stopping once $k$~generations have passed without more than $E$
+improvement in score.
 

### 2012-03-11 mrfy_experience_report.tex
some figure movement and spelling correction
some latex tightening

@@
 
-\begin{figure}
-\ifpdfmadness
-\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
-\else
-\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
-\fi
-In this Markov random field,
-each shaded node represents a beta-strand position.
-Nodes connected by dashed edges are hydrogen-bonded.
-
-\caption{A Markov random field with two beta-stand pairs}
-\label{mrf} 
-\end{figure}
-
 MRFy treats the beta 
@@
 Our~exploration relies on higher-order functions.
+
+\begin{figure}
+\ifpdfmadness
+\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
+\else
+\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
+\fi
+In this Markov random field,
+each shaded node represents a beta-strand position.
+Nodes connected by dashed edges are hydrogen-bonded.
+
+\caption{A Markov random field with two beta-strand pairs}
+\label{mrf} 
+\end{figure}
 
@@
 Finally, the \texttt{quit} function decides when to stop the search.
+\ifnotcutting
 It~too considers the list of accepted generations and the age of the current
 generation. 
+\fi
 \end{itemize}
@@
 experiments. 
-Random-mutation hill took fifty lines of code and one~day to
+Random-mutation hill climbing took fifty lines of code and one~day to
 implement.
-Simulated annealing reuses
- \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} 
-from random-mutation hill climbing.
+Simulated annealing reuses hill climbing's
+ \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} functions.
 Simulated annealing's  new \texttt{accept} function took ten
@@
 predecessor;
-simulated annealing may accept a generation with a worse score.)
+simulated annealing may accept a generation that scores worse.)
 Our genetic algorithm reuses the same
 \texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
-function was only a few lines of code.
+function took only a few lines of code.
 But~the recombination of parent
-placements in a new \texttt{nextGen} function took forty lines of code
+placements in its \texttt{nextGen} function took forty lines of code
 and a full day to implement.
+\ifnotcutting
 Our~current \texttt{quit} function simply runs for a fixed number of
@@
 improvement in score. 
+\fi
 

### 2012-03-11 mrfy_experience_report.tex
cut things we know will not appear

@@
 \newif\ifnotcutting \notcuttingfalse % set if not cutting down to size
+
+\newif\iffinaldraft \finaldraftfalse % true if it's the final version
 
@@
 
+\iffinaldraft
+
 \vfill
@@
 
-\break
-
-\appendix
-
-\section{Orphaned text}
-
-In~our C++ work, the faults we encounter most frequently are
-segmentation faults, 
-memory errors caused by pointer arithmetic or uninitialized 
-memory,
-and heap exhaustion.
-In~our Haskell work, the faults we encounter most frequently are
-type errors and bounds-checking errors in array and vector accesses.
-While we have found the bounds-checking errors difficult to debug,
-they are less difficult than memory errors.
-
-Implementing the equation in this form was not only intellectually
-satisfying; it~also helped us find
-bugs.
-The most memorable bug was that in one of our base cases, 
-we~mistakenly returned an empty path of states;
-the correct result was a singleton path containing the initial state.
-Very quickly after observing a missing state in the output, we were
-able to find the faulty case in the code.
+\fi
+
+%%%%    \break
+%%%%    
+%%%%    
+%%%%    \appendix
+%%%%    
+%%%%    
+%%%%    \section{Orphaned text}
+%%%%    
+%%%%    
+%%%%    In~our C++ work, the faults we encounter most frequently are
+%%%%    segmentation faults, 
+%%%%    memory errors caused by pointer arithmetic or uninitialized 
+%%%%    memory,
+%%%%    and heap exhaustion.
+%%%%    In~our Haskell work, the faults we encounter most frequently are
+%%%%    type errors and bounds-checking errors in array and vector accesses.
+%%%%    While we have found the bounds-checking errors difficult to debug,
+%%%%    they are less difficult than memory errors.
+%%%%    
+%%%%    
+%%%%    Implementing the equation in this form was not only intellectually
+%%%%    satisfying; it~also helped us find
+%%%%    bugs.
+%%%%    The most memorable bug was that in one of our base cases, 
+%%%%    we~mistakenly returned an empty path of states;
+%%%%    the correct result was a singleton path containing the initial state.
+%%%%    Very quickly after observing a missing state in the output, we were
+%%%%    able to find the faulty case in the code.
+%%%%    
+%%%%    
+%%%%
 

### 2012-03-12 mrfy_experience_report.tex
complete edited draft

@@
 
-In MRFy, the \texttt{score} function calls \texttt{viterbi} several
+MRFy calls \verb+vee'+ several
  times
@@
 using
-\texttt{Control.Parallel},
-simply substituting
+\texttt{Control.Parallel}:
+we substituted
 \texttt{parmap rseq} for \texttt{map}.
 Even when there is just one placement per generation,
-MRFy can keep six cores busy,
-and when our search algorithm uses multiple placements per generation,
-it~can use all forty-eight cores on a compute server.
-
-In~\secref{viterbi} above, the example case for the \texttt{viterbi}
+MRFy can keep six cores busy.
+Given multiple placements per generation,
+it~can use forty-eight cores.
+
+In~\secref{viterbi} above, the example case for the \verb+vee'+
 function computes not just a score but also an optimal path.
@@
 score.
-Even though Haskell evaluation is lazy, the \texttt{viterbi} function
-is still allocating thunks.
-We~tried cloning and modifying the \texttt{viterbi} code to compute
+Even though Haskell evaluation is lazy, the \verb+vee'+ 
+still allocates thunks that could compute paths.
+We~cloned \verb+vee'+ and modified it to compute
 only a score, with no path, and this change improved 
-run-time performance by nearly 50\%.
+run-time performance by nearly~50\%.
 But when \texttt{search} finally decides on a placement, we need a
-version of \texttt{viterbi} that \emph{does} compute path, so we can
+version of \verb+vee'+ that \emph{does} compute a path, so we can
 extract the optimal parse for that placement.
@@
 How could we keep the performance improvement without having to
-maintain two versions of \texttt{viterbi}?
+maintain two versions of \verb+vee'+?
 In~Lisp or Ruby we would have used macros or metaprogramming,
@@
 using Template Haskell.
-Instead, we replaced the primitive cons in \texttt{viterbi} with a
+Instead, we replaced the primitive cons in \verb+vee'+ with a
 call to a function of the same type, which we passed~in.
@@
 
-In two major areas, our lack of skills or knowledge resulted in ugly
-Haskell.
+We~think \verb+vee'+ is beautiful, but we wrote some ugly code, too,
+because we lack skills and knowledge in two areas.
 The first is profiling.
@@
 remove them as soon as possible.
-And without the annotations, we found GHC's call-site and allocation
-profiles did not provide enough information to be useful.
-We~miss valgrind, whose callgrind profiler gives us enough information
+And without the annotations, GHC's call-site and allocation
+profiles did not give us enough information to be useful.
+We~miss valgrind, whose callgrind profiler gives useful information
 without \emph{any} annotations in the source.
@@
 run-time errors.
-MRFy uses a Markov random field which is represented in a legacy file
-format that uses a poorly documented representation---when beta
+Our~group's legacy file
+format is poorly documented: when beta
 strands overlap and are doubly paired, the invariants of the
 representation are not clear.
-Dealing with this representation is difficult no matter what
+These files are hard to deal with no matter what
 programming language is used, and faults are to be expected.
-Using GHC, we had a hard time diagnosing run-time errors.
-Extensive use of 
+But using Haskell, we had a hard time diagnosing run-time errors.
+Calls to
 \texttt{trace} 
 littered our code,
-even when moved from our main code to wrapper functions.
+even when relegated to wrapper functions.
 We~weren't aware of the backtrace feature of GHC's profiler,
-and even after we discovered that feature, the results weren't good enough.
-Specifically, because the backtrace feature required profiling mode,
-it could only be used in very small test cases that didn't always trigger
-run-time errors, and one of us was unable to overcome difficulties in building
-various libraries in profiling mode.
+and even after we discovered that feature, the results weren't good enough:
+the profiler can be used only on very small test cases that didn't always trigger
+run-time errors.
 In~August 2011, 
@@
 
-When performance has mattered, the members of our group, like other
+When performance has mattered, members of our group, like other
 computational biologists, have used~C++.
@@
 It~comprises about 12,000 lines of~C++.
-The only external tool or library it uses is \texttt{zlib}.
-Its initial development took two years.
+The only external tool or library it uses is \texttt{zlib}, and
+its initial development took two years.
 \item
@@
 shared with~Matt.
-It~uses no external tools or libraries.
-Its initial development took a year and a half.
+It~uses no external tools or libraries, and
+its initial development took a year and a half.
 \item
@@
 Its initial development took about three months.
-\remark{Others were mostly full-time work, but no need to distinguish.}
+%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
 \end{itemize}
@@
 The mutable data structures were difficult to 
-repurpose, and the pointer arithmetic was too clever for its own good: 
+repurpose, and the pointer arithmetic was \emph{too} clever: 
 nearly every change resulted in new segfaults.
@@
 a cosmetic manipulation of the 
-output, but
+output.
+But
 this feature wound up
- requiring deep information about some data structures,
+requiring deep information about Matt's data structures,
 and we had to give~up.
@@
 %%  draft of a /P/. ---NMD}
-
-On the protein structures for which it was originally designed, SMURF can 
-compute the parse of a query sequence to a MRF based on those structures in 
-less than 10 seconds on a typical workstation.
-However, on somewhat more 
-complex protein structures, SMURF may take two to three minutes on the same 
-hardware. 
-On the most complex protein structures, SMURF exhausts all memory on 
-any system we have available, and must be killed. 
-In contrast, MRFy's runtime 
-for any of these protein structures is typically less than 30 seconds. 
-This 
-meets the performance goals we had, and means that MRFy is fast enough to scale 
-to whole-genome scans.
 
@@
 our hastily written Ruby and~C++.
+
+We~can also make a preliminary comparison of performance.
+SMURF can parse a simple protein structure, optimally, in under 10~seconds.
+It~can parse some more complex structures in 2--3~minutes.
+But when asked to parse the most complex structures, it runs out of
+memory and has to be killed, even on a system with 128GB of~RAM.
+By~contrast, MRFy can parse any of these structures, approximately,
+in 30~seconds.
+This performance
+meets our goals, and it means that MRFy can be used to scan whole genomes.
 
@@
 If~you are a computational biologist, 
-you can be productive in Haskell without extensive
+you can be productive in Haskell, and you don't need extensive
 experience.
@@
 significant functional programming but no Haskell.
-Ramsey is a professor who has used Haskell and other functional
-languages for several medium-sized projects,
-but his contribution was limited to minor ranting and refactoring.
-
-\subsection{Barriers}
+Ramsey is a professor who has used Haskell 
+for medium-sized projects,
+but his contribution to MRFy was limited to minor ranting and refactoring.
+
+\subsection{Barriers to be overcome}
 
@@
 We~were stumped by two problems:
-how to break down a complex module into parts that can usefully be
-unit tested,
+how to break down MRFy into parts that could be
+unit-tested,
 and how to implement QuickCheck generators for representations of complex
@@
 We~know that QuickCheck is a powerful and proven tool,
-yet we were not smart enough to figure out how to use~it.
-
-Like other functional programmers, we~have found that often enough,
-once we have the types right, the code is right.
-But~much of our work involves computing with arrays and array indices,
+but we could figure out how to exploit~it.
+
+Like other functional programmers, we~have found that 
+once we have our types right, our code is often right.
+But~MRFy computes with arrays and array indices,
 and in that domain, types don't help much.
-And we haven't been able to identify any systematic way of debugging
-run-time errors in Haskell programs.
+Bounds violations lead to run-time errors, which we have not been able
+to identify any systematic way to~debug.
 GHC's profiler can provide stack traces, but we~found this information
@@
 Implementing Viterbi's equations in Haskell was pure joy.
-Writing an iterative search in purely functional style was not
-difficult.
-When we had to transform data in the HMMER file format, \emph{without}
-using mutable state the way the C++ code does, we~struggled.
+Writing an iterative search in purely functional style was~easy.
+Transforming data in the HMMER file format, \emph{without}
+using mutable state the way the C++ code does, was difficult.
 
@@
 
-A~little knowledge of and a lot of love for functional programming was
-enough for us to carry out a successful research project in a language
+A~little knowledge of and a lot of love for functional programming
+enabled us to carry out a successful research project in a language
 that computational biologists seldom use.

### 2012-03-12 mrfy_experience_report.tex
delay marginal note about citing hood, hat, freya

@@
       \fi}
+
+\iffinaldraft
+  \newcommand\finalremark{\remark}
+\else
+  \newcommand\finalremark[1]{\relax}
+\fi
 
@@
 We're~aware that debugging lazy functional programs has been a topic
-of some research,\remark{Do we cite Hood, Hat, or Freya?}
+of some research,\finalremark{Do we cite Hood, Hat, or Freya?}
 but one of the biggest barriers we encountered to using Haskell is

### 2012-03-12 mrfy_experience_report.tex
turned doco paragraph into one sentence about examples

@@
 Strategies, are truly great.
-But~for amateurs, it's not always easy to tell them from the wannabes
+But~for amateurs, it's not always easy to tell the great packages from the wannabes
 and the has-beens.
-
-Most of the software we used came from Hackage, which is the Haskell
-community's software repository.
-We~sometimes found it hard to tell
-if a particular Hackage package was right for us.
-In particular, as beginning 
-functional programmers, we~found some documentation inaccessible.
-The documentation that we found most accessible typically contained
-examples showing how to use a particular tool or library.
-\remark{So what's our message here? What are we telling people to
-think or do differently? ---NR}
+And even some of the great packages could be better documented, with
+more examples.
+
+%%  Most of the software we used came from Hackage, which is the Haskell
+%%  community's software repository.
+%%  We~sometimes found it hard to tell
+%%  if a particular Hackage package was right for us.
+%%  In particular, as beginning 
+%%  functional programmers, we~found some documentation inaccessible.
+%%  The documentation that we found most accessible typically contained
+%%  examples showing how to use a particular tool or library.
+%%  \remark{So what's our message here? What are we telling people to
+%%  think or do differently? ---NR}
 

### 2012-03-12 mrfy_experience_report.tex
small fixes

--- 2012-03-12 mrfy_experience_report.tex
@@
 find known proteins of similar structure.
-When the similar proteins are formed from amino-acid sequences that
+When similar proteins are formed from amino-acid sequences that
 are not too different from the query sequence, the homology can be
@@
 proteins with similar structure and function may be formed from significantly 
-different amino-acid sequences, which are not close in edit distance.
+different amino-acid sequences, which are not close in sequence distance.
 Our~research contribution---a software tool called MRFy (pronounced

### 2012-03-12 mrfy_experience_report.tex
minor edits, plus addressing GHCi debugging

@@
 \item
-Although had to write a new implementation of a standard
+Although we had to write a new implementation of a standard
 algorithm, the cost was low, and the connection between code and
@@
 molecules to carry out the functions of living cells: metabolism, regulation, 
-signalling, and so on.
+signaling, and so on.
 A~protein's function is determined by its structure, 
@@
 
-\section{The Software}
+\section{The software}
 
@@
 state transitions.
-Both the probabilities and the
-the states are determined by the alignment:
+Both the probabilities and the states are determined by the alignment:
 \begin{itemize}
@@
 \end{smallverbatim}
-applies a function to what's inside the container, without touching
+applies a function to the thing inside the container, without touching
 the score.
@@
 run-time errors.
+This same limitation affected the GHCi debugger; our \texttt{vee'} function
+shown in \secref{viterbi}, when it is not compiled with
+optimizations, requires too much run time to be practical on non-trivial
+inputs.
+Moreover, the GHCi debugger can only set breakpoints at top-level functions
+or at specific column/line positions, which made debugging the memoized
+\texttt{vee''} function impractical.
 In~August 2011, 
@@
 We~built MRFy quickly, and we expect that
-higher-order functions will MRFy easy to extend.
+higher-order functions will make MRFy easy to extend.
 Each new addition to MRFy's stochastic search has taken at most a day
@@
 We~know that QuickCheck is a powerful and proven tool,
-but we could figure out how to exploit~it.
+but we could not figure out how to exploit~it.
 

### 2012-03-12 mrfy_experience_report.tex
revert one change

@@
 proteins with similar structure and function may be formed from significantly 
-different amino-acid sequences, which are not close in sequence distance.
+different amino-acid sequences, which are not close in edit distance.
 Our~research contribution---a software tool called MRFy (pronounced

### 2012-03-12 mrfy_experience_report.tex
Fix missing period

@@
 beta strands.
-It~comprises about 1,200 lines of~Haskell
+It~comprises about 1,200 lines of~Haskell.
 It~uses several external tools and libraries, of which the most

### 2012-03-12 mrfy_experience_report.tex
corrected a mistake: Matt also uses OpenMP.
minor wordsmithing. resubmitting.

@@
 It~comprises about 12,000 lines of~C++.
-The only external tool or library it uses is \texttt{zlib}, and
-its initial development took two years.
+The only external tools or libraries it uses are \texttt{zlib}
+and \texttt{OpenMP}, and its initial development took two years.
 \item
@@
 It~uses several external tools and libraries, of which the most
-notable are PADS-Haskell, the \path{Data.MemoCombinators} library,
-the BioHaskell bioinformatics library, and the \path{Data.Vector} library.
+notable are PADS-Haskell, the BioHaskell bioinformatics library, and
+the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
+and \path{Data.Vector}.
 Its initial development took about three months.

### 2012-05-08 icfp_reviewer_response.txt
reviewer response updated

We thank the reviewers for their insigful comments, and if our manuscript is 
accepted, we will incorporate changes that reflect these comments. We detail 
our responses to the reviewers below:

Reviewer A writes:

-- More discussion of the comparison between previous C++ software in 
-- that authors' research group and the new Haskell software would be 
-- much more interesting. In particular, the C++ tools described in 
-- section 4 seem significantly more complex than MRFy. Would Haskell 
-- scale to these problems just as well?

The C++ tools described are actually no more complex in terms of the algorithms 
they implement, though they comprise many more lines of code. We will clarify 
this language, and better contrast them.

-- Additionally, the performance discussion is so brief as to be almost 
-- uniterpretable. For example, it seems that MRFy parses protein 
-- structure (in what format?) in constant time, whereas SMURF seems to 
-- take longer on more complex proteins. Are the two tools implementing 
-- the same algorithm? Why is Haskell slower by a factor of 3 on the 
-- small structures? The discussion on parallelism is even briefer -- 
-- it's great to keep 48 cores warm, but what kind of parallel speedup is 
-- achieved?

We agree, and will provide more comprehensive quantitative results. In point of 
fact, MRFy is not implementing the same algorithm as SMURF, and we will better 
clarify this. We will also correct our wording to indicate that it's not 
*parsing* protein structures that we're talking about, but *computing* the 
optimal parse of a protein sequence onto a Markov random field. The complexity 
of the protein structures used to build the Markov random field is at issue 
here, and we will make this clearer. We will provide a better analysis of 
parallel performance, as well. In addition, while SMURF and MRFy differ 
algorithmically, we can and will provide a quantitative comparison of the 
run-time performance of the algorithm they share in common: Viterbi's algorithm.

-- The discussion of Viterbi's equations is extremely hard to follow for 
-- the uninitiated. This is especially true since the level of detail 
-- presented doesn't seem necessary for the larger points of the paper.

We agree, and given the request for more quantitative results, we will try to 
clarify the Viterbi discussion while at the same time going into less detail.

-- Is MRFy available? Can we try it out?

We plan to make MRFy publicly available by the time ICFP opens (and by which 
time we will have also published scientific results based on MRFy). We can make 
MRFy available for review if desired, but the primary investigator (Daniels' 
PhD advisor) is reluctant to release source code prior to publication of 
results.

-- I'm not sure that other structural biologists would agree that 
-- searching for homologous proteins is the "best known method of 
-- discovering" protein function.

We will clarify this to indicate that searching for homologous proteins is a 
well known and cost-effective approach that works in the absence of genetic 
interaction networks, given that protein sequence information is widely 
available. We will cite Pandey et al. 2007, "Computational Approaches for 
Protein Function Prediction: A Survey"

-- Why does the definition of HMMState on page 2 contain no data, just 
-- tags?

We realize that our use of "state" is misleading; our representation uses this 
HMMState data type as a tag, which along with node index and sequence residue 
index, allows us to index into our memoization table. We will change the data 
type to "StateLabel" and clarify this in the manuscript.

Reviewer B writes:

-- There is a report of positive experience using parallelism in Haskell, which is 
-- striking in its brevity. You could easily miss it! Apparently, at least for 
-- trivially parallel problems such as this, parallelism with GHC just works and 
-- there was nothing more to say. I'd like to see a bit more made of this result, 
-- since as a community we're often asked for evidence to support our hypothesis 
-- that functional programming is good for parallelism. Perhaps give parallelism a 
-- separate subsection and include a few numbers?

We plan to break parallelsm out into its own subsection, and include a 
quantitative analysis of scaling on several multi-core systems.

-- 3.1 
--    "data TProb" - why not use a newtype?

This is a historical artifact; TProb used to contain multiple fields. We will 
fix it.

-- 3.2 
--    - the code for "search" is a bit compressed - consider adding some line
--      breaks and blank lines to make it more readable

We will give "search" a bit more space.

-- 3.3 
--    - it sounds like your solution for vee' might be relying on lazy 
--      evaluation rather than properties of GHC's inliner, but it's hard 
--      to tell without more details.

We don't believe this to be the case, but it seems like a worthwhile and simple 
experiment. We will attempt to add strictness to vee' and report whether or not 
there is any impact.

-- 3.4 
--    - since version 7.4.1, GHC's profiler now annotates nested 
--      functions automatically.

This is terrific news! We plan to try GHC 7.4.1 once we resolve some library 
dependencies that are currently problematic. We will update the text to reflect 
this new support.

--   - it's not clear why the profiler could only be used on very small 
--     test cases: is it speed, memory use, or something else? If speed, 
--     does it help to use fewer annotations?

It was definitely runtime speed that was the problem here, for any test case 
large enough to be interesting, even if few annotations. We will state this 
more clearly and provide a more thorough evaluation of our profiling 
experiences.

-- 4. 
--    "working with HMMER codebase" -> "working with the HMMER codebase"

We will correct this.

Reviewer C writes:

-- The main thing I'm missing is more discussion of the effect of laziness -- in 
-- general and on the specific points mentioned. Was laziness useful for the 
-- algorithms described? Was it perhaps even instrumental? Or was it rather a 
-- hurdle (in particular with respect to the issues of debugging, profiling and 
-- performance)? Perhaps the authors can try to make some informed guess whether, 
-- with everything else being equal, a strict language might or might not have 
-- worked better for their problem domain.

We will include more discussion of laziness. We used laziness in two essential 
places, though one of them was not preserved: our original bottom-up 
dynamic-programming approach to Viterbi, and our use of a lazy infinite list of 
random numbers in our search implementation. We will discuss this in more 
detail.

In general, though, we don't particularly feel that laziness had a significant
bearing on our experience; other than one bug which we thought was due to
laziness, but actually turned out to be a mis-implemented typeclass derivation,
laziness was neither an impediment nor instrumental to our success.

-- Sec 3.1: In the type definition for Scored, what is type Score?

We will add this to section 3.1; as it happens it is "newtype Score = Score 
Double"

-- Does the code for vee' omit some cases, or what's happening for other 
-- HMMStates? Also, what is eProb?

Yes, this is only the Match case; we omitted the other recursive and base cases 
in the interests of brevity. We will enumerate the cases we've omitted, and 
refer the reader to Figure 3, but it would require too much space to include 
all of the code.

-- Sec 3.2: "Our implementation of hidden Markov models and Viterbi's algorithm 
-- duplicate existing functionality." I didn't get this. What functionality?

We will clarify this; we meant that we were reimplmenting some algorithms 
(namely, Viterbi) that have already been implemented (in C++) in SMURF and 
HMMER. The point of this contrast is to acknowledge that had we simply extended 
the SMURF code base to implement our new algorithm, we would not have had to 
reimplement Viterbi, for better or worse.

-- In the type definition for Scorer, what is type Placement?

In this case, [Int], but our real code is polymorphic. We ask the reviewers and 
committee: should we show the polymorphic version instead, though it may be 
less clear how it relates to this particular problem domain?

-- You say the use of higher-order functions was useful, e.g. for abstracting over 
-- search strategies. But isn't your type SearchStrategy effectively defining an 
-- object type? That is, wouldn't this have been as convenient with conventional 
-- object-oriented techniques?

We don't believe this would have been as convenient with conventional 
object-oriented techniques, as we are closing over a great deal of environment 
information (command-line parameters, some of which are search-strategy 
specific) which would have required instantiating various types of 
search-strategy-specific objects.

-- Sec 4: Typo: "Matt uses such data structures such as..." - double "such"

Indeed; we will fix this.

Reviewer D writes:

-- Perhaps the most useful feature of Haskell that is used here is function 
-- memoization which comes naturally to dynamic programming and is cumbersome in 
-- other programming languages save for the efficiency of the implementation of 
-- memoization in Haskell itself. It would have been useful to compare this to 
-- implementing table lookups in other programming languages.

We can provide a contrast to our experience implementing dynamic programming in 
C++ (unsurprisingly, there is a great deal more bookkeeping to be done). We do 
not feel we have the experience to provide a comparison over a wide range of 
programming languages.

-- While in the introduction the authors state that "considerations of performance 
-- are paramount" no numerical comparison is presented in the paper about 
-- performance of their implementations compared to others written in C or C++. 
-- Also, no viable way of comparing "convenience, readability, and productivity" 
-- to other implementations is presented in a quantitative manner that can help 
-- attract would-be Haskell programmers except for section on the use of 
-- memoization.

We promise a quantitative comparison should this manuscript be accepted. This 
comparison will largely have to be limited to code size and development time, 
but we can and will provide a quantitative comparison of the Viterbi algorithm 
run-time performance between SMURF and MRFy. However, the overall MRFy 
algorithm differs from SMURF, so overall run-time differences cannot be 
attributed to language choices as opposed to algorithmic differences.

-- Can the Authors identify other problems in computational biology, other than 
-- protein amino acid sequence comparison, where their experience presented in the 
-- paper would apply and be useful? I doubt it. For example in Section 5, "What 
-- can you learn from our experience?" The Authors state "If you are a 
-- computational biologist, you can be productive in Haskell, and you don’t need 
-- extensive experience." I would ask them to name one more problem in 
-- computational biology that would make use of their experience.

In fact, we can name several other problems we can (and may) address based on 
this experience. For example, we have applied graph-theoretic algorithms to the 
problems of gene functional annotation given a genetic interaction graph, as 
well as the problem of identifying compensatory genetic pathways 
("Between-pathway models") in the Yeast genome. We have written some Haskell 
code for the former problem (very much a work in progress), and while we 
approached the latter problem using Python, we feel that Haskell would have 
suited nicely. We will expand upon this discussion in the paper.

-- As stated earlier, identifying homologous proteins is a very important problem, 
-- but it is misleading to state that there are no other problems in computational 
-- biology that are important, the filed so very vast and problems so numerous 
-- which makes the title of the paper misleading. If the Conference Committee 
-- decides to accept the paper, I would urge them to ask the Authors to change the 
-- title of the paper.

If accepted, we will choose a more appropriately narrow title. It was not our 
intent to suggest that protein remote homology detection is the only field 
within computational biology, but rather that many computational biologists are 
unfamiliar with Haskell, and most computational biology software is written, as 
we suggest in the introduction, in C++, Java, or Python.

-- Also, In making a decision to accept or reject the paper, the Conference 
-- Committee should compare their decision with that of accepting or rejecting a 
-- paper that describes implementation of a dynamic programming algorithm in 
-- Hasekell. The phrase "computational biology" should not be a factor in their 
-- decision; the same methods presented in the paper could have been used in 
-- implementing an algorithm for comparing sequences of strings over a finite 
-- alphabet without change, regardless of the fact that in this case the alphabet 
-- encodes names of amino acids.

Clearly, we have not adequately explained the difference between our 
implementation (which combines dynamic programming, stochastic search, and a 
relatively new graphical model for Markov random fields, which are not the same 
as hidden Markov models) and past work from our group. We intend our experience 
report to be an exemplar of what computational biologists may experience should 
they choose to implement novel, performance-intensive algorithms in Haskell 
(or, perhaps, other functional languages). We will endeavor to make this goal 
clearer should our paper be accepted.

We would also like to better explain that MRFy represents a novel algorithm for 
solving Markov random fields, but whose components include implementations of 
the standard techniques of dynamic programming and stochastic search. We hope 
that some readers not yet familiar with Haskell will find this higher-order 
stochastic search implementation useful as well.

### 2012-05-08 icfp_reviewer_response.txt
cutting down response to reviewers

@@
-We thank the reviewers for their insigful comments, and if our manuscript is 
-accepted, we will incorporate changes that reflect these comments. We detail 
-our responses to the reviewers below:
+Thanks for the help!  We're especially grateful for detailed questions
+about such things as type definitions, which highlight issues we
+overlooked.  We believe we can revise the paper to address most
+concerns.  There are a couple of areas where reviewers may have
+misunderstood our intent and we are not sure how best to revise the
+paper.
 
-Reviewer A writes:
+As several reviewers recommend, we will present more comparison (and
+more quantitative comparison) with previous C++ software.  We have few
+quantitative measures that bear on "convenience, readability, and
+productivity", but we will present the measures we have and will make
+them easy to compare.  We will also discuss how Haskell might scale to
+problems previously solved using C++.
 
--- More discussion of the comparison between previous C++ software in 
--- that authors' research group and the new Haskell software would be 
--- much more interesting. In particular, the C++ tools described in 
--- section 4 seem significantly more complex than MRFy. Would Haskell 
--- scale to these problems just as well?
+We will provide a more thorough and more quantitative discussion of
+performance, including measurements of parallel speedup and discussion
+of parallelism.  Where possible we will present apples-to-apples
+comparisons of Haskell and C++ codes.  Elsewhere we can compare
+performance of Haskell code with the performance of C++ code on
+different but related problems.
 
-The C++ tools described are actually no more complex in terms of the algorithms 
-they implement, though they comprise many more lines of code. We will clarify 
-this language, and better contrast them.
+Reviewer A asks if MRFy is available.  The project PI (Daniels's PhD
+supervisor) has agreed to release the source code before ICFP opens.
+If reviewers want access now, we can provide a private copy.
 
--- Additionally, the performance discussion is so brief as to be almost 
--- uniterpretable. For example, it seems that MRFy parses protein 
--- structure (in what format?) in constant time, whereas SMURF seems to 
--- take longer on more complex proteins. Are the two tools implementing 
--- the same algorithm? Why is Haskell slower by a factor of 3 on the 
--- small structures? The discussion on parallelism is even briefer -- 
--- it's great to keep 48 cores warm, but what kind of parallel speedup is 
--- achieved?
+Reviewer A asks why HMMState contains only tags and not data.  On
+reflection, we think this type should not have been called a 'state'.
+We plan to change the name, improve the presentation, and explain the
+data type better.
 
-We agree, and will provide more comprehensive quantitative results. In point of 
-fact, MRFy is not implementing the same algorithm as SMURF, and we will better 
-clarify this. We will also correct our wording to indicate that it's not 
-*parsing* protein structures that we're talking about, but *computing* the 
-optimal parse of a protein sequence onto a Markov random field. The complexity 
-of the protein structures used to build the Markov random field is at issue 
-here, and we will make this clearer. We will provide a better analysis of 
-parallel performance, as well. In addition, while SMURF and MRFy differ 
-algorithmically, we can and will provide a quantitative comparison of the 
-run-time performance of the algorithm they share in common: Viterbi's algorithm.
+Two reviewers ask about laziness.  We believe it was useful but not
+instrumental.  To confirm, we will do and report on some experiments
+with strictness annotations.
 
--- The discussion of Viterbi's equations is extremely hard to follow for 
--- the uninitiated. This is especially true since the level of detail 
--- presented doesn't seem necessary for the larger points of the paper.
+We will report in more detail that the profile was not useful because
+it was too slow, even when using few annotations.
 
-We agree, and given the request for more quantitative results, we will try to 
-clarify the Viterbi discussion while at the same time going into less detail.
+We will do our best to clear up the presentation of the Viterbi
+algorithm and to characterize the cases we have omitted.
 
--- Is MRFy available? Can we try it out?
+Reviewer C asks what functionality was duplicated in Section 3.2.
+We will clarify: we meant that our Haskell code includes new
+implementations of algorithms already implemented in C++ in both SMURF
+and HMMER.  The revised paper will say more explicitly that this
+reimplementation is a cost of using Haskell.  We suspect other
+researchers may pay similar costs.
 
-We plan to make MRFy publicly available by the time ICFP opens (and by which 
-time we will have also published scientific results based on MRFy). We can make 
-MRFy available for review if desired, but the primary investigator (Daniels' 
-PhD advisor) is reluctant to release source code prior to publication of 
-results.
+Reviewer C asks if SearchStrategy would have been as convenient using
+conventional object-oriented techniques.  We will add the question to
+the paper, with this answer: it is *easier* to create a SearchStrategy
+in Haskell than using conventional object-oriented techniques, because
+all the functions in a SearchStrategy are partial applications of
+other functions.  In Haskell, capturing the partially applied
+arguments is effortless; using conventional object-oriented
+techniques, we would have to declare subclasses with instance
+variables and would have to write coded to initialize objects of those
+subclasses.
 
--- I'm not sure that other structural biologists would agree that 
--- searching for homologous proteins is the "best known method of 
--- discovering" protein function.
+Reviewer D suggests comparing memoization using Data.Memocombinators
+with table lookups in other languages.  We will contrast our Haskell
+experience with our prior work implementing dynamic programming in C++.  
+C++ requires much more bookkeeping.
 
-We will clarify this to indicate that searching for homologous proteins is a 
-well known and cost-effective approach that works in the absence of genetic 
-interaction networks, given that protein sequence information is widely 
-available. We will cite Pandey et al. 2007, "Computational Approaches for 
-Protein Function Prediction: A Survey"
+Referee D asks for other problems in computational biology that would
+make use of our experience.  We have applied Haskell to problems of
+gene functional annotation given a genetic-interaction graph, using
+standard graph algorithms.  We are also prototyping code to identify
+compensatory genetic pathways ("Between-pathway models") in the yeast
+genome.   Both of these efforts are much less mature than MRFy, and we
+are not sure what, if anything, to say about them in the paper.
 
--- Why does the definition of HMMState on page 2 contain no data, just 
--- tags?
+Referee D suggests the paper's title is too broad.  We certainly want
+the title to reflect our experience accurately; if the paper does not
+deliver on the promise implicit in the title, that's bad for us and
+for ICFP.  But we would also like the paper to reach an audience
+beyond the ICFP community.  In particular, if a graduate student in
+computational biology wants to try Haskell, we would like this paper
+to be usable as evidence.  We had hoped that a computational biologist
+faced with a new problem would not take a short paper as
+representative of all of computational biology, but instead would
+evaluate the experience we describe and would decide whether the
+evidence we present supports the use of Haskell for the new problem.
 
-We realize that our use of "state" is misleading; our representation uses this 
-HMMState data type as a tag, which along with node index and sequence residue 
-index, allows us to index into our memoization table. We will change the data 
-type to "StateLabel" and clarify this in the manuscript.
-
-Reviewer B writes:
-
--- There is a report of positive experience using parallelism in Haskell, which is 
--- striking in its brevity. You could easily miss it! Apparently, at least for 
--- trivially parallel problems such as this, parallelism with GHC just works and 
--- there was nothing more to say. I'd like to see a bit more made of this result, 
--- since as a community we're often asked for evidence to support our hypothesis 
--- that functional programming is good for parallelism. Perhaps give parallelism a 
--- separate subsection and include a few numbers?
-
-We plan to break parallelsm out into its own subsection, and include a 
-quantitative analysis of scaling on several multi-core systems.
-
--- 3.1 
---    "data TProb" - why not use a newtype?
-
-This is a historical artifact; TProb used to contain multiple fields. We will 
-fix it.
-
--- 3.2 
---    - the code for "search" is a bit compressed - consider adding some line
---      breaks and blank lines to make it more readable
-
-We will give "search" a bit more space.
-
--- 3.3 
---    - it sounds like your solution for vee' might be relying on lazy 
---      evaluation rather than properties of GHC's inliner, but it's hard 
---      to tell without more details.
-
-We don't believe this to be the case, but it seems like a worthwhile and simple 
-experiment. We will attempt to add strictness to vee' and report whether or not 
-there is any impact.
-
--- 3.4 
---    - since version 7.4.1, GHC's profiler now annotates nested 
---      functions automatically.
-
-This is terrific news! We plan to try GHC 7.4.1 once we resolve some library 
-dependencies that are currently problematic. We will update the text to reflect 
-this new support.
-
---   - it's not clear why the profiler could only be used on very small 
---     test cases: is it speed, memory use, or something else? If speed, 
---     does it help to use fewer annotations?
-
-It was definitely runtime speed that was the problem here, for any test case 
-large enough to be interesting, even if few annotations. We will state this 
-more clearly and provide a more thorough evaluation of our profiling 
-experiences.
-
--- 4. 
---    "working with HMMER codebase" -> "working with the HMMER codebase"
-
-We will correct this.
-
-Reviewer C writes:
-
--- The main thing I'm missing is more discussion of the effect of laziness -- in 
--- general and on the specific points mentioned. Was laziness useful for the 
--- algorithms described? Was it perhaps even instrumental? Or was it rather a 
--- hurdle (in particular with respect to the issues of debugging, profiling and 
--- performance)? Perhaps the authors can try to make some informed guess whether, 
--- with everything else being equal, a strict language might or might not have 
--- worked better for their problem domain.
-
-We will include more discussion of laziness. We used laziness in two essential 
-places, though one of them was not preserved: our original bottom-up 
-dynamic-programming approach to Viterbi, and our use of a lazy infinite list of 
-random numbers in our search implementation. We will discuss this in more 
-detail.
-
-In general, though, we don't particularly feel that laziness had a significant
-bearing on our experience; other than one bug which we thought was due to
-laziness, but actually turned out to be a mis-implemented typeclass derivation,
-laziness was neither an impediment nor instrumental to our success.
-
--- Sec 3.1: In the type definition for Scored, what is type Score?
-
-We will add this to section 3.1; as it happens it is "newtype Score = Score 
-Double"
-
--- Does the code for vee' omit some cases, or what's happening for other 
--- HMMStates? Also, what is eProb?
-
-Yes, this is only the Match case; we omitted the other recursive and base cases 
-in the interests of brevity. We will enumerate the cases we've omitted, and 
-refer the reader to Figure 3, but it would require too much space to include 
-all of the code.
-
--- Sec 3.2: "Our implementation of hidden Markov models and Viterbi's algorithm 
--- duplicate existing functionality." I didn't get this. What functionality?
-
-We will clarify this; we meant that we were reimplmenting some algorithms 
-(namely, Viterbi) that have already been implemented (in C++) in SMURF and 
-HMMER. The point of this contrast is to acknowledge that had we simply extended 
-the SMURF code base to implement our new algorithm, we would not have had to 
-reimplement Viterbi, for better or worse.
-
--- In the type definition for Scorer, what is type Placement?
-
-In this case, [Int], but our real code is polymorphic. We ask the reviewers and 
-committee: should we show the polymorphic version instead, though it may be 
-less clear how it relates to this particular problem domain?
-
--- You say the use of higher-order functions was useful, e.g. for abstracting over 
--- search strategies. But isn't your type SearchStrategy effectively defining an 
--- object type? That is, wouldn't this have been as convenient with conventional 
--- object-oriented techniques?
-
-We don't believe this would have been as convenient with conventional 
-object-oriented techniques, as we are closing over a great deal of environment 
-information (command-line parameters, some of which are search-strategy 
-specific) which would have required instantiating various types of 
-search-strategy-specific objects.
-
--- Sec 4: Typo: "Matt uses such data structures such as..." - double "such"
-
-Indeed; we will fix this.
-
-Reviewer D writes:
-
--- Perhaps the most useful feature of Haskell that is used here is function 
--- memoization which comes naturally to dynamic programming and is cumbersome in 
--- other programming languages save for the efficiency of the implementation of 
--- memoization in Haskell itself. It would have been useful to compare this to 
--- implementing table lookups in other programming languages.
-
-We can provide a contrast to our experience implementing dynamic programming in 
-C++ (unsurprisingly, there is a great deal more bookkeeping to be done). We do 
-not feel we have the experience to provide a comparison over a wide range of 
-programming languages.
-
--- While in the introduction the authors state that "considerations of performance 
--- are paramount" no numerical comparison is presented in the paper about 
--- performance of their implementations compared to others written in C or C++. 
--- Also, no viable way of comparing "convenience, readability, and productivity" 
--- to other implementations is presented in a quantitative manner that can help 
--- attract would-be Haskell programmers except for section on the use of 
--- memoization.
-
-We promise a quantitative comparison should this manuscript be accepted. This 
-comparison will largely have to be limited to code size and development time, 
-but we can and will provide a quantitative comparison of the Viterbi algorithm 
-run-time performance between SMURF and MRFy. However, the overall MRFy 
-algorithm differs from SMURF, so overall run-time differences cannot be 
-attributed to language choices as opposed to algorithmic differences.
-
--- Can the Authors identify other problems in computational biology, other than 
--- protein amino acid sequence comparison, where their experience presented in the 
--- paper would apply and be useful? I doubt it. For example in Section 5, "What 
--- can you learn from our experience?" The Authors state "If you are a 
--- computational biologist, you can be productive in Haskell, and you don’t need 
--- extensive experience." I would ask them to name one more problem in 
--- computational biology that would make use of their experience.
-
-In fact, we can name several other problems we can (and may) address based on 
-this experience. For example, we have applied graph-theoretic algorithms to the 
-problems of gene functional annotation given a genetic interaction graph, as 
-well as the problem of identifying compensatory genetic pathways 
-("Between-pathway models") in the Yeast genome. We have written some Haskell 
-code for the former problem (very much a work in progress), and while we 
-approached the latter problem using Python, we feel that Haskell would have 
-suited nicely. We will expand upon this discussion in the paper.
-
--- As stated earlier, identifying homologous proteins is a very important problem, 
--- but it is misleading to state that there are no other problems in computational 
--- biology that are important, the filed so very vast and problems so numerous 
--- which makes the title of the paper misleading. If the Conference Committee 
--- decides to accept the paper, I would urge them to ask the Authors to change the 
--- title of the paper.
-
-If accepted, we will choose a more appropriately narrow title. It was not our 
-intent to suggest that protein remote homology detection is the only field 
-within computational biology, but rather that many computational biologists are 
-unfamiliar with Haskell, and most computational biology software is written, as 
-we suggest in the introduction, in C++, Java, or Python.
-
--- Also, In making a decision to accept or reject the paper, the Conference 
--- Committee should compare their decision with that of accepting or rejecting a 
--- paper that describes implementation of a dynamic programming algorithm in 
--- Hasekell. The phrase "computational biology" should not be a factor in their 
--- decision; the same methods presented in the paper could have been used in 
--- implementing an algorithm for comparing sequences of strings over a finite 
--- alphabet without change, regardless of the fact that in this case the alphabet 
--- encodes names of amino acids.
-
-Clearly, we have not adequately explained the difference between our 
-implementation (which combines dynamic programming, stochastic search, and a 
-relatively new graphical model for Markov random fields, which are not the same 
-as hidden Markov models) and past work from our group. We intend our experience 
-report to be an exemplar of what computational biologists may experience should 
-they choose to implement novel, performance-intensive algorithms in Haskell 
-(or, perhaps, other functional languages). We will endeavor to make this goal 
-clearer should our paper be accepted.
-
-We would also like to better explain that MRFy represents a novel algorithm for 
-solving Markov random fields, but whose components include implementations of 
-the standard techniques of dynamic programming and stochastic search. We hope 
-that some readers not yet familiar with Haskell will find this higher-order 
-stochastic search implementation useful as well.
+Perhaps a member of the program committee can suggest a title that is
+accurate but is not so narrow as to rule out many problems in
+computational biology?

### 2012-05-08 icfp_reviewer_response.txt
Fixed one typo, and added back two concluding paragraphs that I feel are important. Norman, what do you think? It's now 103 lines.

@@
 techniques, we would have to declare subclasses with instance
-variables and would have to write coded to initialize objects of those
+variables and would have to write code to initialize objects of those
 subclasses.
@@
 computational biology?
+
+Reviewer D also implies that we have merely reimplemented a standard
+dynamic programming algorithm.
+Clearly, we have not adequately explained the difference between our 
+implementation (which combines dynamic programming, stochastic search, and a 
+relatively new graphical model for Markov random fields, which are not the same 
+as hidden Markov models) and past work from our group. We intend our experience 
+report to be an exemplar of what computational biologists may experience should 
+they choose to implement novel, performance-intensive algorithms in Haskell 
+(or, perhaps, other functional languages). We will endeavor to make this goal 
+clearer should our paper be accepted.
+
+We would also like to better explain that MRFy represents a novel algorithm for 
+solving Markov random fields, but whose components include implementations of 
+the standard techniques of dynamic programming and stochastic search. We hope 
+that some readers not yet familiar with Haskell will find this higher-order 
+stochastic search implementation useful as well.

### 2012-05-30 mrfy_experience_report.tex
fix caps
start of a note about quickcheck

@@
 Luke Palmer's
-\path{Data.Memocombinators} package.
+\path{Data.MemoCombinators} package.
 The value
@@
 
+\remark{The following \P\ will need to be connected up to the last \S\
+on ``our old approach to debugging.''}
+
+Debugging led also not to awkward code, but to awkward interactions on
+our team.
+The two junior members of the team wanted to apply the debugging
+skills they had honed through years of imperative programming.
+The senior partner kept nattering on about QuickCheck properties.
+No~progress was made until the first draft of this paper was prepared.
+At~that point, we were quickly about to write QuickCheck
+specifications for several properties of the Viterbi algorithm:
+\begin{itemize}
+\item
+The total number of \verb+Mat+ and \verb+Ins+ states returned equals
+the number of amino acids in the query sequence.
+\item
+The total number of matches, insertions, and deletions returned equals the
+number of nodes in the hidden Markov model,
+\emph{provided} that a maximal, nonempty, contiguous subsequence
+of~\verb+Ins+ states counts as a single node,
+and \verb+Mat+ and \verb+Del+ states count as one node each.
+\item
+The sequence returned satisfies the Plan7 invariant, that is, states
+\verb+Ins+ and \verb+Del+ are never adjacent.
+\end{itemize}
+\textbf{These properties need to be followed by more ambitious
+properties},
+such as proving that nearby sequences are less likely than the
+sequence returned by \verb+vee'+.
+
 \section{Our previous experience compared}
@@
 no~longer maintained.
-The great packages, like \texttt{Data.Memocombinators} and Parallel
+The great packages, like \texttt{Data.MemoCombinators} and Parallel
 Strategies, are truly great.

### 2012-06-03 mrfy_experience_report.tex
sketch of what is wrong with Norman's code

@@
 \documentclass[preprint,nonatbib,blockstyle,nocopyrightspace,times]{sigplanconf}
+
+\newcommand\mrfy{MRFy} % damn you Noah
 
@@
 A~generation is a collection of scored {placements}.
+\begingroup\em\relax
+Ideally, MRFy's search would use now-classic lazy, modular
+technique advocated by \citet{hughes:why}, in which one function
+computes an infinite sequence of generations, and another function
+uses a finite prefix to decide on an approximation.
+But making MRFy's search lazy and modular search is not so easy.
+ because MRFy's search is stochastic.
+
+In~a stochastic search, \emph{movement} between points in the search
+space is not the same as \emph{progress}.
+For example, suppose you are searching for the top of a hill
+and you take a small jump in a random direction.
+If~you arrive at a higher point than your point of origin, you have
+made progress.
+Otherwise you have just moved.
+\smallverbatiminput{movequality}
+We~hoped to implement \mrfy's search by composing functions:
+\begin{itemize}
+\item
+From a given point, a \emph{generator} produces an infinite sequence
+of moves.
+\item
+From such a sequence, a \emph{classifier} chooses the first move that
+represents progress.
+What sort of move constitutes progress is not only different for
+different kinds of searches, but may also be different at different
+\emph{times} in a search.
+\end{itemize}
+A~generator and classifier can be composed to produce a sequence of
+progressing generations, and a test function can terminate the search
+after examining a finite
+prefix.
+But this sketch contains a fatal flaw:
+a generator may produce an infinite sequence of generations without any
+progress.
+And then, when the test function tries to examine a finite prefix of
+progressing generations, it will diverge.
+We~might conclude that
+\mrfy's search algorithm is inherently non-modular:
+production of progressing generations cannot be decoupled from the
+test for termination.
+
+N.B.~The following transition no longer makes any sense:
+\endgroup
 We~pass the
@@
 \smallfuzzverbatiminput{11pt}{search}
+
+\remark{We need to explain why we don't just generate an infinite lazy
+list of thingies.  And perhaps we can consider an infinite tree!}
 

### 2012-06-20 everygen.tex
damnfino what's going on, but these are my changes

everyPt :: SearchGen pt r -> Age -> Scored pt
         -> Rand r [Aged (Utility (Scored pt))]
everyPt ss age startPt = do
  successors <- mapM (nextPt ss) (repeat startPt)
  tagged <- zipWithM agedUtility successors [succ age..]
  let (useless, Aged (Useful newPt) newAge : _) =
                        span (isUseless . unAged) tagged
  nextPts <- everyPt ss newAge newPt
  return $ Aged (Useful startPt) age : useless ++ nextPts

  where agedUtility pt age =
           utility ss move >>= \u -> return $ Aged u age
         where move = Move { older = startPt, younger = pt
                           , youngerAge = age }

### 2012-06-20 mrfy_experience_report.tex
damnfino what's going on, but these are my changes

--- 2012-05-30 mrfy_experience_report.tex
@@
 
-\newif\ifnotcutting \notcuttingfalse % set if not cutting down to size
+\newif\ifnotcutting \notcuttingtrue % set if not cutting down to size
 
@@
 \newcommand\smallverbatiminput[1]{{\par\unskip\small\verbatiminput{#1}}}
+\newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
 \newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}
@@
 annealing, multistart simulated annealing, and a genetic algorithm.
-Our~exploration relies on higher-order functions.
+Our~exploration exploits higher-order functions, existentially
+quantified types, and (to a degree) lazy evaluation.
 
@@
 
-MRFy's search computes a sequence of \emph{generations}.
-A~generation is a collection of scored {placements}.
+We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
+of \emph{points}. 
+Each point contains
+a collection of {placements}; different searches represent
+collections differently.
+Each point also has a \texttt{Score}; \mrfy\ looks for points with
+good scores.
+
+Ideally, MRFy's search would use now-classic lazy, modular
+technique advocated by \citet{hughes:why}, in which one function
+computes an infinite sequence of points, and another function
+uses a finite prefix to decide on an approximation.
+But because MRFy's search is stochastic, 
+making MRFy's search lazy and modular is not so easy.
+
+To illustrate the difficulties, we~use our simplest search:
+random hill climbing.
+The space of beta-strand placements has no notion of slope or
+gradient,
+so at any given point in the search space, this search moves a random
+distance in a random 
+direction.
+If~it leads to a better point, the move
+is \emph{useful};
+otherwise it is \emph{useless}.
+\smallverbatiminput{utility}
+(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
+With luck, an~infinite sequence of useful moves converges
+at a local optimum.
+
+\mrfy's search path follows only useful moves;
+if a move proves useless, 
+\mrfy\ abandons it and tries again (with a new random seed) at the
+previous point.
+Ideally, \mrfy\ would search by composing a \emph{generator}
+that produces an infinite sequence of moves, 
+a \emph{filter} that selects the useful moves, and 
+a \emph{test} function that enumerates finitely many useful moves
+and returns the final destination.
+But the generator may produce an infinite sequence of useless moves.
+(For example, if \mrfy\ should find a global optimum, every move from
+that point is guaranteed to be useless.)
+Given an infinite sequence of useless inputs, a filter would not be
+productive, and a test function would diverge.
+
+We~address this problem by combining ``generate and filter'' into a
+single abstraction, which has type \texttt{SearchGen pt r}.
+Type variable~\texttt{pt} is a point in the search
+space, and type variable~\texttt{r} is a
+random-number generator:
+\smallverbatiminput{gen.tex}
+The monadic computation \texttt{pt0} randomly selects a starting point
+for search;
+\texttt{nextPt} produces a new point from
+an existing point.
+Because scoring may be expensive, both \texttt{pt0}
+and \texttt{nextPt} use \emph{scored} points, and they can
+reuse scores from previous points.
+
+Why do we have \texttt{pt0} and \texttt{nextPt}
+instead 
+of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
+which could produce an infinite list?
+Because after the first move
+the starting point for the next move depends on whether the first move
+was useful.\remark{I hate this ``explanation.'' ---NR}
+That decision is made by the \texttt{utility} function,
+which scrutinizes a move represented as follows:
+\smallverbatiminput{move}
+The decision about utility uses not only a source of randomness but
+also the \emph{age} of the \texttt{younger} point which is
+moved~to.\remark{This word ``age'' is \emph{entirely} bogus!}
+For example, in~simulated annealing, as the search progresses,
+the \texttt{utility} function becomes less likely to express a move
+that results in a worse score.\remark{And I hate this explanation
+too.}
+
+\noindent
+\vbox{
+Using these pieces, function \texttt{everyGen} produces an infinite
+sequence containing a mix of useful and useless moves:
+\smallverbatiminput{everygen}
+}
+\noindent
+Because both \texttt{nextPt} and \texttt{utility} consume randomness,
+the computation is very sequential, but it does use a little bit of
+laziness:
+from its starting point, \texttt{everyPt} provides an infinite list of
+randomly chosen successor points, then calls \texttt{agedUtility} to
+tag each one with its age and utility.\remark{There's that bogus age
+again.}
+The Prelude function \texttt{span} finds the first useful successor,
+but in case there \emph{is} no successor, the code must also produce
+all the useless successors.
+If~we do find a useful successor, we start searching anew from that
+point, with a recursive call to \texttt{everyPt}.
+
+The most telling line of code in \texttt{everyPt} is the last line of
+the body, which shows that the result begins with a useful point,
+is followed by a (possibly infinite, possibly empty) list of useless
+points, and then continues recursively with another useful point.
+When the computation returned by \texttt{everyPt} is evaluated against
+a random-number generator, it returns an infinite list containing a
+mix of useful and useless points.
+This mix can be passed to a test function of type \texttt{SearchStop~pt}:\\
+\vbox{
+\smallverbatiminput{strat}
+\smallverbatiminputx{stop}
+\smallverbatiminputx{search}
+}
+\noindent
+We have implemented two kinds of stop functions...
+
+N.B. WE HAVE TWO KINDS OF NON-MODULARITY.  FIRST, TERMINATION HAS TO
+LOOK AT \emph{ALL} STATES, EVEN THOUGH WE WOULD PREFER THAT IT LOOK
+ONLY AT USEFUL STATES.  SECOND, AGE IS EVERYWHERE.  IN PARTICULAR, WE
+CANNOT SEPARATE THE AGE OF THE CURRENT STATE FROM THE UTILITY
+DECISION.
+
+\remark{We need to explain why we don't just generate an infinite lazy
+list of thingies.  And perhaps we can consider an infinite tree!}
+
+The \texttt{SearchStrategy} interface provides ample scope for
+experiments.\remark{\P\ needs rewriting}
+Random-mutation hill climbing took fifty lines of code and one~day to
+implement.
+Simulated annealing reuses hill climbing's
+ \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
+Simulated annealing's  new \texttt{accept} function took ten
+lines of code and half an hour to implement.
+(Hill climbing accepts a point only if it scores better than its
+predecessor;
+simulated annealing may accept a point that scores worse.)
+Our genetic algorithm reuses the same
+\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
+function took only a few lines of code.
+But~the recombination of parent
+placements in its \texttt{nextPt} function took forty lines of code
+and a full day to implement.
+\ifnotcutting
+Our~current \texttt{quit} function simply runs for a fixed number of
+generations, but its interface will enable us to try other criteria,
+such as stopping once $k$~generations have passed without more than $E$
+improvement in score. 
+\fi
+
+\subsection*{The old search algorithm}
+
 We~pass the
 search function a \texttt{Scorer}, which runs Viterbi's algorithm.
-\smallverbatiminput{scoredecl}
+\smallverbatiminput{orig-scoredecl}
 A~search heuristic is characterized by four functions, which are
 bundled into a record called \texttt{SearchStrategy}.
-\smallfuzzverbatiminput{2.3pt}{strategy}
+\smallfuzzverbatiminput{2.3pt}{orig-strategy}
 In this record,
@@
 \item
-Function \texttt{gen0} produces the first generation of placements.
+Function \texttt{pt0} produces the first generation of placements.
 \item
@@
 We hope you find the search simple and straightforward:
-\smallfuzzverbatiminput{11pt}{search}
-
-The \texttt{SearchStrategy} interface provides ample scope for
-experiments. 
-Random-mutation hill climbing took fifty lines of code and one~day to
-implement.
-Simulated annealing reuses hill climbing's
- \texttt{nextGen}, \texttt{gen0}, and \texttt{quit} functions.
-Simulated annealing's  new \texttt{accept} function took ten
-lines of code and half an hour to implement.
-(Hill climbing accepts a generation only if it scores better than its
-predecessor;
-simulated annealing may accept a generation that scores worse.)
-Our genetic algorithm reuses the same
-\texttt{gen0} and \texttt{quit} functions, and its \texttt{accept}
-function took only a few lines of code.
-But~the recombination of parent
-placements in its \texttt{nextGen} function took forty lines of code
-and a full day to implement.
-\ifnotcutting
-Our~current \texttt{quit} function simply runs for a fixed number of
-generations, but its interface will enable us to try other criteria,
-such as stopping once $k$~generations have passed without more than $E$
-improvement in score. 
-\fi
+\smallfuzzverbatiminput{11pt}{orig-search}
 
@@
 
-We thank Lenore Cowen, Kathleen Fisher, Benjamin Hescott, Bradford
+An~anonymous referee\remark{Just one referee?} spurred us to think
+more deeply about laziness.
+Josef Svenningsson suggested a correspondence between \mrfy's search
+and stream fusion, which led us eventually to the \texttt{Utility} type.
+We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
 Larsen, and Nathan Ricci. 

### 2012-06-20 orig-search.tex
damnfino what's going on, but these are my changes

search :: forall placement 
        . SearchStrategy placement 
       -> Scorer placement 
       -> [Seed]
       -> (Scored placement, History placement)
search strat scorer (s0:seeds) = runFrom seeds firstGen (History []) 0
 where
  firstGen = map scorer $ gen0 strat s0
  runFrom :: [Seed] -> [Scored placement] -> History placement
          -> Age -> (Scored placement, History placement)
  runFrom (s1:s2:seeeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 scorer oldPop
        trialHist = (winner, age) `hcons` oldHist
        winner = minimum trialPop
        (newPop, newHist) =
          if accept strat s2 trialHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then -- TODO quit must change: consider best-ever, convergence
          (fst $ minimum (unHistory newHist), newHist) 
        else
          runFrom seeds newPop newHist (age + 1)

### 2012-06-20 orig-strategy.tex
damnfino what's going on, but these are my changes

type Age  = Int -- number of generations explored
type Seed = Int -- source of stochastic variation
data History placement = History { unHistory :: [(Scored placement, Age)] }
  deriving (Show, Eq, Ord)
hcons :: (Scored placement, Age) -> History placement -> History placement
hcons a (History as) = History (a:as)
emptyHistory :: History a
emptyHistory = History []

-- is there a better name for seed?
data SearchStrategy placement = 
 SS { gen0    :: Seed -> [placement]
    , nextGen :: Seed -> Scorer placement
              -> [Scored placement] -> [Scored placement]
    , accept  :: Seed -> History placement -> Age -> Bool
    , quit    ::         History placement -> Age -> Bool
    }

### 2012-06-20 search.tex
damnfino what's going on, but these are my changes

--- 2012-03-10 search.tex
@@
-search :: SearchStrategy -> Scorer -> [Seed]
-       -> (Scored Placement, [Score])
-search strat score (s:seeds) = runFrom seeds firstGen [] 0
- where
-  firstGen = map score $ gen0 strat s
-  runFrom :: [Seed] -> [Scored Placement] -> [Score]
-          -> Age -> (Scored Placement, [Score])
-  runFrom (s1:s2:seeds) oldPop oldHist age =
-    let trialPop  = nextGen strat s1 score oldPop
-        trialHist = (scoreOf $ minimum newPop) : oldHist
-        (newPop, newHist) =
-          if accept strat s2 newHist age then
-            (trialPop, trialHist)
-          else
-            (oldPop, oldHist)
-    in  if quit strat newHist age then
-          (minimum newPop, newHist)
-        else
-          runFrom seeds newPop newHist (age + 1)
+search :: SearchStrategy pt r -> Rand r (History pt)
+search (SS strat test) =
+  fmap test . everyPt strat 0 =<< pt0 strat

### 2012-06-21 mrfy_experience_report.tex
A few crude changes, and many breadcrumbs (in remarks) that need to be swept up and dealt with.

@@
 We~represent a state using an algebraic data type:
+\remark{we promised a better name for HMMState. Thoughts?}
 \begin{smallverbatim}
@@
 and we use \texttt{fmap} to add \texttt{Mat} to the list of states
-encountered up to this point:
+encountered up to this point:\remark{we promised to characterize the missing cases}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Our implementations of hidden Markov models and Viterbi's algorithm
-duplicate existing functionality.
-In~MRFy, they are used to implement new functionality.
+duplicate existing functionality; these same algorithms were
+implemented in C++ for HMMER and SMURF.
+This reimplementation is a cost of using Haskell, and one that
+other researchers may have to pay as well.
+However, the authors of SMURF and HMMER implemented these algorithms
+anew, even though existing implementations existed in C++.
+In~MRFy, we use these algorithms as building blocks to implement
+new functionality.
 
@@
 \end{itemize}
+This \texttt{SearchStrategy} interface somewhat resembles an object-oriented
+style of programming.
+However, it is \textit{easier} to create a SearchStrategy
+in Haskell than using conventional object-oriented techniques, because
+all the functions in a SearchStrategy are partial applications of
+other functions.
+If we were using conventional object-oriented
+techniques, we would have to declare subclasses with instance
+variables and would have to write code to initialize objects of those
+subclasses.
+
 \seclabel{search}%
 We hope you find the search simple and straightforward:
+
 \smallfuzzverbatiminput{11pt}{orig-search}
@@
 it~can use forty-eight cores.
+However, we have found that MRFy performs best, at least on our
+hardware, with between 12 and 16 cores. We have access to only
+one 48-core system, and while MRFy will keep all 48 cores busy, it
+performs better when given no more than 16 cores.\remark{give specific
+numbers here}
+
+MRFy's predecessor, SMURF, does not perform stochastic search,
+and thus its overall run-time performance cannot be compared directly to MRFy.
+However, both programs implement the Viterbi algorithm.
+On an exceedingly large, contrived example, comprising 56,000 amino acids
+(larger than any known protein) and 343 nodes, and ignoring all beta-strand
+information, SMURF's Viterbi implementation completed in 0.497 seconds on a
+2.6GHz Core i7.
+On the same benchmark, MRFy's Viterbi implementation required 1 minute 57
+seconds.
+Despite the inferior runtime performance, MRFy still meets our performance
+goals.
+This is because the stochastic search algorithm relies on the Viterbi algorithm
+applied to many, significantly smaller hidden Markov models than SMURF's
+multidimensional programming must tackle.
+MRFy computes an alignment for a 12-stranded ``beta sandwich'' structure in
+generally less than one minute, while SMURF consumes over 16GB of memory
+and does not terminate even after eight hours.
+Had we implemented the SMURF algorithm in Haskell, we believe that performance
+would be unacceptable.
+However, for the algorithm underlying MRFy, we are satisfied with the runtime
+performance of Haskell and GHC.
 
@@
 GHC's profiling is based on cost centers \cite{sansom-pj},
-and although GHC 
+and for most of MRFy's development, we used GHC 7.0, which
 will happily add cost centers to top-level 
-functions, much of our code is in nested functions.
+functions only, while much of our code is in nested functions.
 We~know of no better technique than to add cost-center annotations to
@@
 without \emph{any} annotations in the source.
+Late in MRFy's development, we began using GHC 7.4, which provides much
+better profiling tools.
+We are thrilled by these tool developments.
 
@@
 in \figref{alignment}.
-It~comprises about 12,000 lines of~C++.
+It~comprises about 12,000 lines of~C++.\remark{reviewers seemed to want a
+table of this information, but that seems like a huge waste of space.}
 The only external tools or libraries it uses are \texttt{zlib}
@@
 It~uses several external tools and libraries, of which the most
-notable are PADS-Haskell, the BioHaskell bioinformatics library, and
+notable are Parsec, the BioHaskell bioinformatics library, and
 the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
@@
 to implement.
-
-%
-% An Experience Report is about what you actually did, not what you
-% wish to do :-)
-%
-%%  There are two major research ideas we wish to add to MRFy eventually, one of which
-%%  \remark{how much should I say about profiles and alpha helices? ---NMD}
-%%  involves looking at another type of nonlocal dependency (\textit{alpha
-%%  helices} in contrast 
-%%  to the current beta strands).
-%%  We expect that thanks to algebraic data types and typeclasses, these
-%%  enhancements will be at 
-%%  most a few weeks of work; we do not believe these enhancements would
-%%  even be feasible in SMURF. 
-%%  Indeed, Daniels tried to add another of these enhancements to SMURF
-%%  two years ago and gave up due 
-%%  to the difficulty of the implementation.
-%%  \remark{The above needs a lot of revision. It is a shitty first
-%%  draft of a /P/. ---NMD}
 
@@
 By~contrast, MRFy can parse any of these structures, approximately,
-in 30~seconds.
+in 30~seconds.\remark{reconcile this vague paragraph with performance section above}
 This performance

### 2012-06-27 mrfy_experience_report.tex
added a little sentence
Incorporated reviewer comments. Still need to cut some space.

@@
 function of the protein that gene codes for.
-The best known method of discovering such function is
+One of the best known methods of discovering such function is
 to find other proteins of 
@@
 The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
-We~represent a state using an algebraic data type:
+We~represent the state chosen for each node using an algebraic data type:
 \remark{we promised a better name for HMMState. Thoughts?}
@@
 
-The log of a probability is just a number.
-To distinguish a transition probability from other numbers, we wrap it in
+The log of a probability is just a number, which we call a \texttt{Score}.
+A transition probability connects two \texttt{HMMState}s, and so we wrap it in
 a value constructor:
 \begin{smallverbatim}
-type LogProb = Double
-data TProb   = TProb { logProbability :: LogProb }
+newtype Score = Score Double
+
+data TProb =
+     TProb { logProbability :: Score
+           , fromState :: HMMState
+           , toState :: HMMState
+           }
+
 \end{smallverbatim}
@@
 and we use \texttt{fmap} to add \texttt{Mat} to the list of states
-encountered up to this point:\remark{we promised to characterize the missing cases}
+encountered up to this point:
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \smallverbatiminput{vfix}
+The other cases (\texttt{Ins} and \texttt{Del}) look similar.
 Functions \texttt{tProb} and \texttt{edge} combine to select the
@@
 This reimplementation is a cost of using Haskell, and one that
-other researchers may have to pay as well.
+other researchers may have to pay as well, as fewer computational biology
+libraries exist for Haskell. For example, the National Center for
+Biotechnology Information publishes its own library of algorithms
+for computational biology.
 However, the authors of SMURF and HMMER implemented these algorithms
@@
 the starting point for the next move depends on whether the first move
-was useful.\remark{I hate this ``explanation.'' ---NR}
+was useful, whereas the first move is always useful.
 That decision is made by the \texttt{utility} function,
@@
 The decision about utility uses not only a source of randomness but
-also the \emph{age} of the \texttt{younger} point which is
-moved~to.\remark{This word ``age'' is \emph{entirely} bogus!}
+also the \emph{age} of the search, which corresponds to the
+computational cost to reach a point in the search.
 For example, in~simulated annealing, as the search progresses,
 the \texttt{utility} function becomes less likely to express a move
-that results in a worse score.\remark{And I hate this explanation
-too.}
+that results in a worse score.
 
@@
 randomly chosen successor points, then calls \texttt{agedUtility} to
-tag each one with its age and utility.\remark{There's that bogus age
-again.}
+tag each one with the search's age and the point's utility.
 The Prelude function \texttt{span} finds the first useful successor,
@@
 \noindent
-We have implemented two kinds of stop functions...
-
-N.B. WE HAVE TWO KINDS OF NON-MODULARITY.  FIRST, TERMINATION HAS TO
-LOOK AT \emph{ALL} STATES, EVEN THOUGH WE WOULD PREFER THAT IT LOOK
-ONLY AT USEFUL STATES.  SECOND, AGE IS EVERYWHERE.  IN PARTICULAR, WE
-CANNOT SEPARATE THE AGE OF THE CURRENT STATE FROM THE UTILITY
-DECISION.
-
-\remark{We need to explain why we don't just generate an infinite lazy
-list of thingies.  And perhaps we can consider an infinite tree!}
+
+We have two kinds of non-modularity.
+First, termination has to
+look at \emph{all} states, even though we would prefer that it look
+only at useful states.
+This is because otherwise we risk non-termination:
+we are not guaranteed a useful move after any finite number of steps.
+Second, age of the search shows up in multiple places.
+In particular, we cannot separate the age of the search at the
+current state from the utility decision.
 
@@
 lines of code and half an hour to implement.
-(Hill climbing accepts a point only if it scores better than its
+(Hill climbing accepts a point if and only if it scores better than its
 predecessor;
@@
 
-\subsection*{The old search algorithm}
-
-We~pass the
-search function a \texttt{Scorer}, which runs Viterbi's algorithm.
-\smallverbatiminput{orig-scoredecl}
-A~search heuristic is characterized by four functions, which are
-bundled into a record called \texttt{SearchStrategy}.
-\smallfuzzverbatiminput{2.3pt}{orig-strategy}
-In this record,
-\begin{itemize}
-\item
-Function \texttt{pt0} produces the first generation of placements.
-\item
-Function \texttt{nextGen} produces a new generation of placements from
-an existing generation.
-Because scoring a placement is expensive, 
-and some placements may persist from the old generation to the new
-generation, \texttt{nextGen} works with \emph{scored} placements, not
-bare placements.
-\item
-A~new generation produced by \texttt{nextGen} may be no better
-than the generation it replaces.
-When it is first produced, the new generation is ``on trial.''
-The \texttt{accept} function decides whether to keep a new generation
-or to try \texttt{nextGen} again with a different seed.
-The \texttt{accept} function considers a list of type
-\texttt{[Scored~Age]}.
-At~the head of this list is the score of the best
-placement of the trial generation, together with the \texttt{Age} of the trial
-generation, which is the total number of generations tried.
-The remaining elements of the list hold the history of scores and ages for
-previously \emph{accepted} generations.
-(The list omits generations that didn't survive the \texttt{accept} trial.)
-For convenience, the \texttt{accept} function also receives
-the \texttt{Age} of the trial generation.
-\item
-Finally, the \texttt{quit} function decides when to stop the search.
-\ifnotcutting
-It~too considers the list of accepted generations and the age of the current
-generation. 
-\fi
-\end{itemize}
-This \texttt{SearchStrategy} interface somewhat resembles an object-oriented
+Our \texttt{SearchStrategy} interface somewhat resembles an object-oriented
 style of programming.
@@
 subclasses.
-
-\seclabel{search}%
-We hope you find the search simple and straightforward:
-
-\smallfuzzverbatiminput{11pt}{orig-search}
 
@@
 However, we have found that MRFy performs best, at least on our
-hardware, with between 12 and 16 cores. We have access to only
+hardware, with between 12 and 16 cores.
+We have access to only
 one 48-core system, and while MRFy will keep all 48 cores busy, it
-performs better when given no more than 16 cores.\remark{give specific
-numbers here}
+performs only slightly better, at an average of 35 seconds to find an alignment
+on a large protein structure, than when given 16 cores, which takes an average
+of 37 seconds.
 
@@
 performance of Haskell and GHC.
+Nonetheless, we expect that we can still improve MRFy's run-time performance
+with further profiling and tuning.
 
@@
 function computes not just a score but also an optimal path.
-MRFy's \texttt{search} function, in \secref{search}, uses only the
+MRFy's \texttt{search} function, in \secref{mrfy}, uses only the
 score.
@@
 \item
-The total number of matches, insertions, and deletions returned equals the
+The total number of matches, and deletions returned equals the
 number of nodes in the hidden Markov model,
-\emph{provided} that a maximal, nonempty, contiguous subsequence
+\emph{except} that an initial sequence of
 of~\verb+Ins+ states counts as a single node,
-and \verb+Mat+ and \verb+Del+ states count as one node each.
+due to an odd property of the HMMER file format (namely, the zeroth
+node has no delete state and a non-emitting match state which
+is always occupied, but a normal insert state).
 \item
@@
 sequence returned by \verb+vee'+.
+One practice we recommend for those following in our footsteps is to
+write instances of \texttt{Arbitrary} for data types as those data types
+are written, when relevant invariants are fresh in memory.
 
@@
 Matt uses such
-data structures such as mutable oct-trees, vectors, and arrays.
+data structures as mutable oct-trees, vectors, and arrays.
 It~uses clever pointer arithmetic.
@@
 By~contrast, MRFy can parse any of these structures, approximately,
-in 30~seconds.\remark{reconcile this vague paragraph with performance section above}
+in 30~seconds.
 This performance
@@
 and
-working with HMMER 
+working with the HMMER 
 codebase was pleasant;
@@
 
-An~anonymous referee\remark{Just one referee?} spurred us to think
+An~anonymous referee spurred us to think
 more deeply about laziness.

### 2012-06-27 mrfy_experience_report.tex
formatting tweaks and stuff. In final style. Need to cut. Please look at all remarks.

@@
 
-\documentclass[preprint,nonatbib,blockstyle,nocopyrightspace,times]{sigplanconf}
+\documentclass[nonatbib]{sigplanconf}
 
@@
 
-\conferenceinfo{ICFP '12}{September 9-15, Copenhagen.} 
-\copyrightyear{2012} 
-\copyrightdata{[to be supplied]}
+\conferenceinfo{ICFP'12,} {September 9--15, 2012, Copenhagen, Denmark.}
+\CopyrightYear{2012}
+\copyrightdata{978-1-4503-1054-3/12/09}
 
@@
 in \figref{alignment}.
-It~comprises about 12,000 lines of~C++.\remark{reviewers seemed to want a
-table of this information, but that seems like a huge waste of space.}
+It~comprises about 12,000 lines of~C++.
 The only external tools or libraries it uses are \texttt{zlib}
@@
 its initial development took a year and a half.
+It uses multidimensional dynamic programming to exactly compute the alignments
+for which MRFy relies on stochastic search.
+As a result, in the presence of complex beta-strand topologies, SMURF is
+computationally intractable.
 \item
 MRFy is used to detect homologous proteins in the presence of paired
-beta strands.
+beta strands; it effectively supplants SMURF.
 It~comprises about 1,200 lines of~Haskell.
@@
  absolutely necessary.
+
+MRFy is available at \url{mrfy.cs.tufts.edu}
 

### 2012-06-27 everygen.tex
an abstract that more accurately reflects the essential information in the paper

@@
-everyPt :: SearchGen pt r -> Age -> Scored pt
+everyPt :: RandomGen r
+        => SearchGen pt r -> Age -> Scored pt
         -> Rand r [Aged (Utility (Scored pt))]
-everyPt ss age startPt = do
-  successors <- mapM (nextPt ss) (repeat startPt)
+everyPt sg age startPt = do
+  successors <- mapM (nextPt sg) (repeat startPt)
   tagged <- zipWithM agedUtility successors [succ age..]
@@
                         span (isUseless . unAged) tagged
-  nextPts <- everyPt ss newAge newPt
+  nextPts <- everyPt sg newAge newPt
   return $ Aged (Useful startPt) age : useless ++ nextPts
@@
   where agedUtility pt age =
-           utility ss move >>= \u -> return $ Aged u age
+           utility sg move >>= \u -> return $ Aged u age
          where move = Move { older = startPt, younger = pt

### 2012-06-27 mrfy_experience_report.tex
an abstract that more accurately reflects the essential information in the paper

@@
 
-\newif\iffinaldraft \finaldraftfalse % true if it's the final version
+\newif\iffinaldraft \finaldrafttrue % true if it's the final version
 
@@
 \begin{abstract}
-An Experience Report provides {evidence} that functional
-programming really works, or it describes obstacles that prevent
-functional programming from working.   
-Here, we show that for computational biologists,
-Haskell combines the flexibility and rapid prototyping of a scripting
-language with performance comparable to~C++.
-The~Haskell language and libraries provide powerful abstractions, 
-but we couldn't always get them to work.
-Despite some speed bumps around profiling, debugging, and the
-uncertain status of any given package, we~found that Haskell's ecosystem
+%An Experience Report provides {evidence} that functional
+%programming really works, or it describes obstacles that prevent
+%functional programming from working.   
+Haskell gives computational biologists
+the flexibility and rapid prototyping of a scripting
+language along with performance comparable to~C++.
+In~our experience, higher-order functions, lazy evaluation, and
+monads really worked, but
+profiling and debugging presented obstacles.
+Haskell libraries varied greatly:
+libraries providing memoization combinators and parallel-evaluation
+strategies gave us tremendous leverage,
+but other, nameless libraries mostly got in our way.
+Despite the obstacles and the uncertain quality of some libraries,
+Haskell's ecosystem
 made it easy for us to develop new algorithms in computational

### 2012-06-27 mrfy_experience_report.tex
edit the intro (first column) for verbs and for parallel structure

@@
 libraries providing memoization combinators and parallel-evaluation
-strategies gave us tremendous leverage,
+strategies helped us a lot,
 but other, nameless libraries mostly got in our way.
@@
 acids (proteomic data). 
-For some software, considerations of performance are paramount; this
+When performance is paramount,
 software is usually written in C~or~C++. 
-For other software, considerations of convenience, readability, and
-productivity are more important;
-this software is usually written in a high-level
+When convenience, readability, and
+productivity are more important,
+software is usually written in a dynamically typed
 or domain-specific language like
 Perl, Python, Ruby, SPSS, or~R.
-In~this paper, we report on an attempt to replace both these families
-of languages with Haskell: 
+In~this paper, we report on experience using a third kind of language,
+Haskell: 
 \begin{itemize}
 \item
-Although we had to write a new implementation of a standard
-algorithm, the cost was low, and the connection between code and
-algorithm is unusually strong (\secref{viterbi}).
+We had to reimplement an
+algorithm already implemented in~C++, 
+but the implementation required little effort,
+and the connection between code and
+algorithm is unusually clear (\secref{viterbi}).
 Our new Haskell code performs competitively with~C++,
-and parallel performance was easy to achieve
+and it was easy to parallelize
 (\secref{perf}). 
+\remark{Can we summarize the performance results?  Speedup?}
 \item
 Higher-order functions made it unusually easy to
-implement and experiment with a family of related algorithms
+create and experiment with new stochastic-search algorithms
 (\secref{hofs}).
 \item
-The only significant impediment that Haskell presented to our
-research was the difficulty of understanding and improving the
-performance of Haskell programs (\secref{awkward-profiling}).
-\item
-We~used Haskell successfully even though we are computational
-biologists with little background in functional programming.
-Modifying our group's C++ code to try new research ideas has been
-difficult;
-trying new ideas in Haskell has been easy
+Haskell slowed us down in only one way: it's hard to understand and
+improve the
+performance of a Haskell program (\secref{awkward-profiling}).
+%%  \item
+%%  The only significant impediment that Haskell presented to our
+%%  research was the difficulty of understanding and improving the
+%%  performance of Haskell programs (\secref{awkward-profiling}).
+\item
+Although authors Daniels and Gallant are computational
+biologists with little background in functional programming,
+we~used Haskell successfully.
+And Haskell has made it easy to explore new research ideas;
+by~contrast,
+our group's C++ code has not been a good vehicle for exploring new ideas
 (\secref{comparo}).
 \item
-The Haskell language is surrounded by a penumbra of libraries and tools that
+The Haskell community offers libraries and tools that
 promise powerful abstractions.
-Some deliver very high leverage, but others are not ready for prime
-time.
-We~found the two families difficult to distinguish in advance (\secref{penumbra}).
+Some worked as advertised, saved us lots of time, and were a pleasure
+to~use.
+Others, not so much.
+And~we could not tell ahead of time which experience we were likely to have
+ (\secref{penumbra}).
 \end{itemize}

### 2012-06-27 mrfy_experience_report.tex
good libraries save "effort" not "time
edit section 2 and start section 3

@@
 promise powerful abstractions.
-Some worked as advertised, saved us lots of time, and were a pleasure
+Some worked as advertised, saved us lots of effort, and were a pleasure
 to~use.
@@
 
-Proteins are cellular machinery. They interact with one another and with other 
+Proteins %%% are cellular machinery. They
+interact with one another and with other 
 molecules to carry out the functions of living cells: metabolism, regulation, 
@@
 Computational biologists detect homologies by building 
-algorithms which, given a {query sequence} of amino acids,
+algorithms which, given a {query sequence}, %% of amino acids,
 find known proteins of similar structure.
@@
 different amino-acid sequences, which are not close in edit distance.
-Our~research contribution---a software tool called MRFy (pronounced
-``Murphy'')---can detect homologies 
+Our~research software, MRFy (pronounced
+``Murphy''), can detect homologies 
 in amino-acid sequences that are only distantly related.
@@
 These proteins are identified by a biologist, who puts
-their amino-acid sequences into a correspondence, also called an
+their amino-acid sequences into an
 \emph{alignment}. 
@@
 Both the probabilities and the states are determined by the alignment:
+\remark{These bullets need proper parallel structure}
 \begin{itemize}
@@
 The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
-We~represent the state chosen for each node using an algebraic data type:
-\remark{we promised a better name for HMMState. Thoughts?}
-\begin{smallverbatim}
-data HMMState = Mat | Ins | Del | Beg | End
-\end{smallverbatim}
-
-The particular hidden Markov model we use requires that every nonempty
-sequence of insertions be followed by a match, and similarly for
-deletions.
-In~other words, insertions and deletions cannot be mixed without an
-intervening match.
-This restriction means that there are only seven possible transitions
-into the states of any one column \cite{Eddy:1998ut}. 
+In~our representation, each state contains a probability or a table of
+probabilities, and it is also labelled with one of these labels:
+\smallverbatiminput{statelabel}
+
+We~use the ``Plan7'' hidden Markov model, which forbids direct
+transitions between insertion 
+states and deletion states \cite{Eddy:1998ut}. 
+The model gets its name because there are exactly
+seven possible transitions
+into the states of any column~$j$.
+Each transition has its own probability:
 \begin{itemize} 
 \item
-A~transition into a match state in
-column~$j$ is more likely when column~$j$ is a consensus column.
+A~transition into a match state 
+is more likely when column~$j$ is a consensus column.
 Depending on the predecessor state, 
 the probability of such a transition is 
-$\txprobj M M$, $\txprobj I M$, or $\txprobj D M$.
-\item
-A~transition into a deletion state in
- column~$j$ is more likely when column~$j$ is a non-consensus column.
+$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
+\item
+A~transition into a deletion state 
+is more likely when column~$j$ is a non-consensus column.
 The probability of such a transition is 
-$\txprobj M D$ or $\txprobj D D$.
-\item
-A~transition into an insertion state in
- column~$j$ is more likely when column~$j$ is a non-consensus column.
+$\txprobj M D$~or~$\txprobj D D$.
+\item
+A~transition into an insertion state 
+is more likely when column~$j$ is a non-consensus column.
 The probability of such a transition is 
-$\txprobjj M I$ or $\txprobjj I I$.
+$\txprobjj M I$~or~$\txprobjj I I$.
 \end{itemize}
@@
 The log of a probability is just a number, which we call a \texttt{Score}.
-A transition probability connects two \texttt{HMMState}s, and so we wrap it in
+A transition probability connects two \texttt{StateLabel}s, and so we wrap it in
 a value constructor:
@@
      TProb { logProbability :: Score
-           , fromState :: HMMState
-           , toState :: HMMState
+           , fromState :: StateLabel
+           , toState :: StateLabel
            }

### 2012-06-27 mrfy_experience_report.tex
import hmmnode definition direct from the code
reflect new AA type
Be more honest about Viterbi: it's still slower than C++ by a lot (but the C++ version may not even be correct!)

@@
 algorithm is unusually clear (\secref{viterbi}).
-Our new Haskell code performs competitively with~C++,
+Our new Haskell code performs slower than the~C++
+implementation of the same algorithm,
+but still meets our performance goals,
 and it was easy to parallelize
 (\secref{perf}).
-\remark{Can we summarize the performance results?  Speedup?}
+However, this algorithm is only one part of a larger
+algorithm, which on the whole performs significantly
+better than the~C++ software that precedes it.
 \item
@@
 \begin{smallverbatim}
-type AA = Int -- amino acid using HMMER coding
+newtype AA = AA Int -- amino acid using HMMER coding
 \end{smallverbatim}
@@
 of emission probabilities for the match and insertion states.
-\begin{smallverbatim}
-type EProbs  = Vector LogProb
-data HmmNode = HmmNode { transitions  :: TProbs
-                       , matEmissions :: EProbs
-                       , insEmissions :: EProbs
-                       }
-\end{smallverbatim}
+\remark{Why strictness annotations?}
+\smallverbatiminput{hmmnode}
 

### 2012-06-27 mrfy_experience_report.tex
I've restored block style.  It's not enough for our paper to appear in the proceedings; for it to have an impact, people have to read it. Well worth the additional two column inches.
A paragraph about The QuickCheck Experience (tm).

@@
 
-\documentclass[nonatbib]{sigplanconf}
+\documentclass[nonatbib,blockstyle,times]{sigplanconf}
+  % at 9pt anything but block style is unreadable
 
@@
 
+This was a practice that the junior members of our team did not follow, and
+they sorely regret it.
+We speculate that one reason why we didn't follow this practice was a poor 
+understanding of the \texttt{Arbitrary} type class.
+Namely, a lot of the examples we had seen using QuickCheck focused 
+predominantly on writing and testing properties with data types that already 
+implemented \texttt{Arbitrary}.
+We recall a smattering about \texttt{Arbitrary}, but we were still too baffled 
+by the \texttt{Gen} monad to truly grasp its importance in using QuickCheck.
+When it finally came time to use QuickCheck, we were overwhelmed by the amount 
+of work that had to be done to write and test implementations of 
+\texttt{Arbitrary} for each of our data types.
+
 \section{Our previous experience compared}

### 2012-06-28 mrfy_experience_report.tex
intro crammed back into  one column,  even in block style

@@
 
-\documentclass[nonatbib,blockstyle,times]{sigplanconf}
+\documentclass[nonatbib,preprint,blockstyle,times]{sigplanconf}
   % at 9pt anything but block style is unreadable
@@
 
-\newenvironment{smallverbatim}{\par\small\verbatim}{\endverbatim}
-\newcommand\smallverbatiminput[1]{{\par\unskip\small\verbatiminput{#1}}}
+\newif\ifverbatimsmall
+\verbatimsmallfalse
+\makeatletter
+\def\verbatim@font{%
+   \normalfont\ttfamily
+   \ifverbatimsmall\small\fi
+   \hyphenchar\font\m@ne
+   \@noligs}
+\makeatother
+
+\newenvironment{smallverbatim}{\par\small\verbatimsmalltrue\verbatim}{\endverbatim}
+\newcommand\smallverbatiminput[1]{%
+  \everypar=\expandafter{\the\everypar
+       \verbatimsmallfalse
+       \belowdisplayskip=\abovedisplayskip
+       \belowdisplayshortskip=\abovedisplayshortskip
+       \topsep=\standardvspace}%
+  \topsep=0.78\topsep
+  \belowdisplayskip=0.78\belowdisplayskip
+  \belowdisplayshortskip=0.78\belowdisplayshortskip
+  \verbatimsmalltrue\verbatiminput{#1}}
 \newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
@@
 the flexibility and rapid prototyping of a scripting
-language along with performance comparable to~C++.
+language but performance comparable to~C++.
 In~our experience, higher-order functions, lazy evaluation, and
@@
 profiling and debugging presented obstacles.
-Haskell libraries varied greatly:
-libraries providing memoization combinators and parallel-evaluation
+Haskell libraries vary greatly:
+memoization combinators and parallel-evaluation
 strategies helped us a lot,
@@
 algorithm already implemented in~C++, 
-but the implementation required little effort,
-and the connection between code and
-algorithm is unusually clear (\secref{viterbi}).
-Our new Haskell code performs slower than the~C++
-implementation of the same algorithm,
-but still meets our performance goals,
-and it was easy to parallelize
+and the Haskell code is slower.
+But the implementation required little effort;
+the connection between code and
+algorithm is unusually clear (\secref{viterbi});
+the code was easy to parallelize; 
+and it meets our performance needs
 (\secref{perf}).
-However, this algorithm is only one part of a larger
-algorithm, which on the whole performs significantly
-better than the~C++ software that precedes it.
+Moreover, measured as a whole, the software that incorporates this new
+code 
+performs significantly
+better than the~C++ software that preceded~it.
 \item
@@
 \item
-Haskell slowed us down in only one way: it's hard to understand and
-improve the
-performance of a Haskell program (\secref{awkward-profiling}).
+Haskell slowed us down in only one area:  understanding and
+improving
+performance (\secref{awkward-profiling}).
 %%  \item
@@
 \item
-Although authors Daniels and Gallant are computational
+Although the first two authors are computational
 biologists with little background in functional programming,

### 2012-06-28 mrfy_experience_report.tex
work on layout, formatting, and sections 2 and 3

@@
 
-\newif\ifnotcutting \notcuttingtrue % set if not cutting down to size
+\newif\ifnotcutting \notcuttingfalse % set if not cutting down to size
 
@@
 \begin{document}
+\parskip=0.9\parskip % cheating
 
@@
 Each transition has its own probability:
+\remark{Problem: the subscripts are not so easy to read, and notation
+is not quite consistent with \figref{viterbi}, definition of $a'$ at
+bottom right.}
 \begin{itemize} 
@@
 can compute the probability
-that a new protein shares structure and
-evolutionary history with the proteins used to train the model.
+that a new protein shares structure 
+\ifnotcutting and evolutionary history \fi
+with the proteins used to train the model.
 The computation finds the most likely path through the hidden Markov model.
@@
 probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
-The logarithm function is monotonic, so the path that maximizes the
+The~path that maximizes the
 log of the probability is the most likely path.
@@
 correct.
-To~present some code,
-we~begin with representations.
-
+
+Our code represents a query sequence as an immutable vector of amino
+acids.
 In~idiomatic Haskell, 
-we might represent an individual amino acid~$x$
+we might represent an individual amino acid~$x_i$
 using a value of algebraic data type:
@@
 \end{smallverbatim}
-But our primary goal was to get something working quickly,
-and our hidden Markov models are stored using a
+But our models use a
 legacy file format that represents each amino acid as a small integer.
 Our representation is therefore
+\smallverbatiminput{aa}
+The log of a probability is a negative number.
+To~make these numbers more pleasant to read, we negate them,
+which eliminates hordes of minus signs from our file formats.
+The negated logarithm of a probability is called a \emph{score}.
+\smallverbatiminput{score}
+Type \texttt{Score} has a limited \texttt{Num} instance which permits
+scores to be added and subtracted but not multiplied.
+
+In a hidden Markov model, each transition is actually labeled with a
+score, not a probability.
+And we include the labels of the states that are
+connected:\remark{\emph{Why} are the labels there?
+And since they are apparently never used, why do they have strictness annotations???}
 \begin{smallverbatim}
-newtype AA = AA Int -- amino acid using HMMER coding
-\end{smallverbatim}
-Either way, a query sequence is (and should be) an immutable vector of
-amino acids.
-%type QuerySeq = Vector AA
-%\end{smallverbatim}
-
-The log of a probability is just a number, which we call a \texttt{Score}.
-A transition probability connects two \texttt{StateLabel}s, and so we wrap it in
-a value constructor:
-\begin{smallverbatim}
-newtype Score = Score Double
-
-data TProb =
+data TProb =   --- THIS DEFN IS A DAMN LIE
      TProb { logProbability :: Score
@@
            }
-
 \end{smallverbatim}
-We group transition probabilities into 7-tuples.
-Each tuple corresponds to a ``node'' in the hidden Markov model
-(a~dashed oval in \figref{plan7}) and to a column of the original
-alignment.
-As~noted above, 
-a match state may transition to any state in the successor node;
-an insertion state may transition to itself or to its successor match
-state;
-and
-a~deletion may transition to its successor match
-state or deletion state.
+In~a Plan7 hidden Markov model, as discussed above, 
+there are seven transitions into each node:
 \begin{smallverbatim}
-data TProbs = 
+data TProbs =  -- MORE DAMN LIES
      TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
@@
 In~our code, we select from this 7-tuple using the \texttt{edge}
-function:
+function:\remark{Broken!  This function no longer appears in our examples}
 \smallverbatiminput{edge}
@@
 
-The representation of a node includes the probabilities of transitions
-into the states of that node, plus the tables
-of emission probabilities for the match and insertion states.
-\remark{Why strictness annotations?}
+Nodes are numbered, and the representation of a node includes 
+the tables
+of emission probabilities for the match and insertion states,
+plus
+the
+probabilities of transitions 
+into the states of that node.
 \smallverbatiminput{hmmnode}
@@
 \emph{memoized}.
-To~show our code, we have to confess to a 
-transformation of Viterbi's equations.
-The~logarithms on the left of \figref{viterbi} are computed when the
-hidden Markov model is constructed, and the resulting numbers are
-stored in a file.
-And~to~make this file more pleasant to read, the logarithms are
-negated.
-(The logarithm of a probability is typically negative, and negating
-the logarithm eliminates hordes of minus signs.)
-We~therefore implement a transformed version of Viterbi's equations,
+Our code implements a transformed version of Viterbi's equations, 
+which operates on scores instead of on probabilities.
+This version is
 shown on the right-hand side of \figref{viterbi}.
+THIS PART IS BROKEN.  THE A ARRAY NEEDS TWO SUBSCRIPTS, AND THE E
+NOTATION IS INCONSISTENT (HAS MAGICALLY CHANGED FROM BRACKETS TO SUBSCRIPTS).
 The transformed equations minimize the
-negated log probability for each combination of node~$j$, amino
+score (i.e., the negated log probability) for each combination of
+node~$j$, amino 
 acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
-A~negated log probability is called a~\emph{score}.
-
-MRFy computes not only the minimum score but
-also the path of states which produces that score.
+
 Scores can be usefully paired with many types of values,
@@
 
-\tuningbox{
 We~also made \texttt{Scored} an instance of~\texttt{Ord} and of
@@
 applies a function to the thing inside the container, without touching
-the score.
+the score.\remark{Changes in the code appear to have
+made \texttt{fmap} irrelevant}
 And containers are ordered by score alone, so applying
@@
 smallest (and therefore best) score.
-}
-
-Given the \texttt{Scored} abstraction, we can easily implement
-Viterbi's equations for a match state.
-To~implement the equation at the top right of \figref{viterbi},
-we~choose the best solution
-from among the three predecessor states;
+
+The \texttt{Scored} abstraction makes it easy to implement
+Viterbi's equations.
+For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
+right of \figref{viterbi}, we call 
+\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}, 
+which chooses the best (\texttt{minimum}) solution
+from among the three predecessor states:
+\smallverbatiminput{vfix}
+In this code, EXPLAIN EXTEND, EXPLAIN ESCORE, EXPLAIN
+ASCORE. 
 we~use~\texttt{/+/} to add transition and emission scores,
 and we use \texttt{fmap} to add \texttt{Mat} to the list of states
-encountered up to this point:
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\smallverbatiminput{vfix}
-The other cases (\texttt{Ins} and \texttt{Del}) look similar.
-Functions \texttt{tProb} and \texttt{edge} combine to select the
-appropriate transition score (one of the $a'$~values).
-Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
-so
-the call to~\verb+vee''+ produces the same result as a recursive call
-to~\verb+vee'+, but faster:
+encountered up to this point:\remark{BROKEN! \texttt{fmap} is gone}
+
+Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+.
+Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,
+but faster: 
 \smallverbatiminput{memo}
@@
 And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
-devoted to improving performance.
+devoted to dynamic programming.
 By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,

### 2012-06-28 mrfy_experience_report.tex
Some cleanup of section 3. NR: please note the missing 'ashore' file in a remark.

@@
 \item 
-For each column~$j$ of the alignment, the hidden Markov model has a
-\emph{deletion state}~$D_j$.
-The deletion state determines the probability that the query sequence
-has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
-\item 
 For each column~$j$ of the alignment, the hidden Markov model has an
@@
 insertion at column~$j$.
+\item
+For each column~$j$ of the alignment, the hidden Markov model has a
+\emph{deletion state}~$D_j$.
+The deletion state determines the probability that the query sequence
+has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
 \end{itemize}
@@
 Each transition has its own probability:
-\remark{Problem: the subscripts are not so easy to read, and notation
-is not quite consistent with \figref{viterbi}, definition of $a'$ at
-bottom right.}
+\remark{Norman: how do you propose to make the subscripts more readable?}
 \begin{itemize} 
@@
 \multicolumn3{l}{%
-a'_{s} = - \log a_{s} 
+a'_{s_{j}s_{k}} = - \log a_{s_{j}s_{k}} 
 \qquad
-e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}
+e^{\prime}_{s}(x) = - \log\frac{e_{s}(x)}{q_{x}}
 \qquad
@@
 \begin{smallverbatim}
-data TProb =   --- THIS DEFN IS A DAMN LIE
+data TProb =
      TProb { logProbability :: Score
@@
 \begin{smallverbatim}
-data TProbs =  -- MORE DAMN LIES
+data TProbs =
      TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
@@
 \end{smallverbatim}
-In~our code, we select from this 7-tuple using the \texttt{edge}
-function:\remark{Broken!  This function no longer appears in our examples}
-\smallverbatiminput{edge}
-In MRFy, the first two arguments to \texttt{edge} are known at compile
-time, and we expect \texttt{edge} to be inlined.
+In~our code, we select from this 7-tuple using the \texttt{ascore}
+function:\remark{Norman: please put ascore into a file}
+\smallverbatiminput{ascore}
+In MRFy, the first two arguments to \texttt{ascore} are known at compile
+time, and we expect \texttt{ascore} to be inlined.
 
@@
 shown on the right-hand side of \figref{viterbi}.
-THIS PART IS BROKEN.  THE A ARRAY NEEDS TWO SUBSCRIPTS, AND THE E
-NOTATION IS INCONSISTENT (HAS MAGICALLY CHANGED FROM BRACKETS TO SUBSCRIPTS).
 The transformed equations minimize the
@@
 
-We~also made \texttt{Scored} an instance of~\texttt{Ord} and of
-\texttt{Functor}, so
-\begin{smallverbatim}
-fmap :: (a -> b) -> (Scored a -> Scored b)
-\end{smallverbatim}
-applies a function to the thing inside the container, without touching
-the score.\remark{Changes in the code appear to have
-made \texttt{fmap} irrelevant}
-And containers are ordered by score alone, so applying
+We~also made \texttt{Scored} an instance of~\texttt{Ord}.
+Containers are ordered by score alone, so applying
 \texttt{minimum} to a list of scored things chooses the thing with the
@@
 \smallverbatiminput{vfix}
-In this code, EXPLAIN EXTEND, EXPLAIN ESCORE, EXPLAIN
-ASCORE. 
+
+In this code, we look up emission scores for a given node and amino
+acid using \texttt{eScore},
 we~use~\texttt{/+/} to add transition and emission scores,
-and we use \texttt{fmap} to add \texttt{Mat} to the list of states
-encountered up to this point:\remark{BROKEN! \texttt{fmap} is gone}
+and we use \texttt{extend} to add \texttt{Mat} to the list of states
+encountered up to this point.
 

### 2012-06-28 mrfy_experience_report.tex
wordsmithing section 3.2

--- 2012-06-28 mrfy_experience_report.tex
@@
        \verbatimsmallfalse
-       \belowdisplayskip=\abovedisplayskip
-       \belowdisplayshortskip=\abovedisplayshortskip
+%       \belowdisplayskip=\abovedisplayskip
+%       \belowdisplayshortskip=\abovedisplayshortskip
        \topsep=\standardvspace}%
   \topsep=0.78\topsep
-  \belowdisplayskip=0.78\belowdisplayskip
-  \belowdisplayshortskip=0.78\belowdisplayshortskip
+%  \belowdisplayskip=0.78\belowdisplayskip
+%  \belowdisplayshortskip=0.78\belowdisplayshortskip
   \verbatimsmalltrue\verbatiminput{#1}}
@@
 
-Our implementations of hidden Markov models and Viterbi's algorithm
-duplicate existing functionality; these same algorithms were
-implemented in C++ for HMMER and SMURF.
-This reimplementation is a cost of using Haskell, and one that
-other researchers may have to pay as well, as fewer computational biology
-libraries exist for Haskell. For example, the National Center for
-Biotechnology Information publishes its own library of algorithms
-for computational biology.
-However, the authors of SMURF and HMMER implemented these algorithms
-anew, even though existing implementations existed in C++.
-In~MRFy, we use these algorithms as building blocks to implement
-new functionality.
-
-A hidden Markov model uses local information about
-amino-acid sequences.
-But when a real protein folds in three dimensions, 
-groups of amino acids 
+Our Haskell code for hidden Markov models and Viterbi's algorithm
+solves the same problems as existing C++~code.
+Other researchers using Haskell may also have to reimplement code,
+but researchers may also reimplement code out of choice, not
+necessity.
+For~example, both SMURF and HMMER contain new implementations of these
+same algorithms.
+In~MRFy, we are using the Haskell implementations to help solve new
+problems.
+
+When a real protein folds in three dimensions, 
+amino acids 
 that are far away in the one-dimensional sequence can be
 adjacent in three-dimensional space.
-We~are studying adjacent groups called \emph{beta strands}, which
+We~are studying groups of such acids called \emph{beta
+strands}.\remark{Fix this; we're not really studying beta strands}
+Beta strands
 can become hydrogen-bonded to each other,
-and are effectively ``stuck together.''
-
-A~mutation in a paired beta strand may break the pairing, in which
- case the mutated protein will not fold properly and is unlikely to
- survive natural selection.
-But a mutation in a paired beta strand \emph{can} survive if it is
- accompanied by a \emph{compensatory} mutation in the strand with which it is
- paired. 
-The probability of finding amino acid~$x_i$ in column~$j$ of an alignment may
- therefore depend on the amino acid in the paired column
- $\pairedwith j$.
-This nonlocal interaction changes
-Viterbi's equations:
-each term $V_{j}^{\prime M}(i)$ becomes
+making them ``stuck together.''
+MRFy detects homologous proteins that include hydrogen-bonded beta
+strands; using prior methods, many instances of this problem are
+intractable.
+
+To~account for beta strands, computational biologists must use both a
+richer model of protein structure and a revised version of Viterbi's
+equations.
+The~richer model recognizes that only certain pairs of amino acids are
+able to bond to each other.
+%%  The presence of beta strands changes both the natu
+%%   the probability that a given
+%%  query sequence fit
+%%  A~mutation in a paired beta strand may break the pairing, in which
+%%   case the mutated protein will not fold properly and is unlikely to
+%%   survive natural selection.
+%%  But a mutation in a paired beta strand \emph{can} survive if it is
+%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
+%%   paired. 
+When column~$j$ of an alignment is part of a beta strand and is paired
+with another column  $\pairedwith j$,
+the probability of finding amino acid~$x_i$ in column~$j$ 
+depends on~$x_{\pairedwith j}$.
+The~relevant term in
+Viterbi's equations 
+$V_{j}^{\prime M}(i)$ becomes
 $$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
-The distance between columns~$j$ and $\pairedwith j$ can be as small
- as a few columns or as large as a few hundreds of columns.
-
-The approach currently taken in computational biology
-is~to model each pair of beta strands not with two hidden Markov
-models, 
-but with a pair of sequences of match states.
-Each of the subsequences \emph{between} paired beta strands is then
-modeled with an ordinary hidden Markov model.
+Columns~$j$ and $\pairedwith j$ can be as close together as a few
+columns or as far apart as a few hundreds of columns.
+
+The presence of beta strands also changes the model.
+Within a beta strand, amino acids are not inserted or deleted;
+a~bonded pair of beta strands is modeled by
+ a pair of sequences of match states.
+Between the beta strands, the model is structured as before.
 The~combined model, an~example of which is shown in \figref{mrf}, is called a
 \textit{Markov random field}. 
-Because of nonlocal interactions between paired beta strands, 
-the likelihood-maximization problem for a Markov random field does not
- exhibit optimal substructure, and dynamic
- programming cannot solve it quickly \cite{Menke:2010ti,Daniels:2012}.
+Becayuse $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
+also on the nonlocal $x_{\pairedwith j}$, 
+dynamic programming cannot compute the maximum likelihood quickly
+\cite{Menke:2010ti,Daniels:2012}.
 MRFy explores new methods of \emph{approximating}
@@
 MRFy treats the beta 
-strands, which are built into the model,
+strands in the model
  as ``beads'' which can be slid along the query sequence.
-A~set of positions of beta strands is called
+A~positioning of the beta strands is called
  a \emph{placement}.
@@
 hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
-Given a placement, the maximum likelihood of any subsequence that appears
-\emph{between} two beta strands can be computed exactly and
-efficiently using Viterbi's algorithm.
-(The result of the exact computation is a probability that is
-\emph{conditioned} on the placement.)
+Given a placement, the maximum likelihood of the remaining parts of
+the query sequence, which appear between and around beta strands,
+can be computed quickly and exactly 
+using Viterbi's algorithm.
+The exact result is a probability that is
+\emph{conditioned} on the placement.
 

### 2012-06-28 mrfy_experience_report.tex
s hat.
splice more things into code

@@
 And since they are apparently never used, why do they have strictness annotations???}
-\begin{smallverbatim}
-data TProb =
-     TProb { logProbability :: Score
-           , fromState :: StateLabel
-           , toState :: StateLabel
-           }
-\end{smallverbatim}
+\smallverbatiminput{tprob}
 In~a Plan7 hidden Markov model, as discussed above, 
 there are seven transitions into each node:
-\begin{smallverbatim}
-data TProbs =
-     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
-            , i_m :: TProb, i_i :: TProb
-            , d_m :: TProb, d_d :: TProb }
-\end{smallverbatim}
+\smallverbatiminput{tprobs}
 In~our code, we select from this 7-tuple using the \texttt{ascore}
 function:\remark{Norman: please put ascore into a file}
-\smallverbatiminput{ascore}
+%\smallverbatiminput{ascore}
 In MRFy, the first two arguments to \texttt{ascore} are known at compile
@@
 columns or as far apart as a few hundreds of columns.
+Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
+also on the nonlocal $x_{\pairedwith j}$, 
+dynamic programming cannot compute the maximum likelihood quickly
+\cite{Menke:2010ti,Daniels:2012}.
+MRFy explores new methods of \emph{approximating}
+maximum likelihood.
 
@@
 \textit{Markov random field}.
-Becayuse $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
-also on the nonlocal $x_{\pairedwith j}$, 
-dynamic programming cannot compute the maximum likelihood quickly
-\cite{Menke:2010ti,Daniels:2012}.
-MRFy explores new methods of \emph{approximating}
-maximum likelihood.
 

### 2012-06-28 mrfy_experience_report.tex
ccs stuff added
new explanation of function vee-prime

@@
 \multicolumn3{l}{%
-a'_{s_{j}s_{k}} = - \log a_{s_{j}s_{k}} 
+a'_{s \hat{s}} = - \log a_{s \hat{s}} 
 \qquad
@@
 \smallverbatiminput{tprobs}
-In~our code, we select from this 7-tuple using the \texttt{ascore}
-function:\remark{Norman: please put ascore into a file}
-%\smallverbatiminput{ascore}
-In MRFy, the first two arguments to \texttt{ascore} are known at compile
-time, and we expect \texttt{ascore} to be inlined.
-
-Nodes are numbered, and the representation of a node includes 
-the tables
-of emission probabilities for the match and insertion states,
-plus
-the
-probabilities of transitions 
-into the states of that node.
-\smallverbatiminput{hmmnode}
+In~our code, we select from this 7-tuple using the \texttt{aScore}
+function;
+\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {{\hat s}}$},
+where $s$~and~$\hat s$ form one of the 7~pairs above.
+
+A~7-tuple of scores is included in the representation of each node in
+the model.
+This representation also includes tables of emission probabilities
+$e_{M_j}$ and $e_{I_j}$.
+These tables are read using function
+\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.
+
+%%  DROP THIS PARA??
+%%  Nodes are numbered, and the representation of a node includes 
+%%  the tables
+%%  of emission probabilities for the match and insertion states,
+%%  plus
+%%  the
+%%  probabilities of transitions 
+%%  into the states of that node.
+%%  \smallverbatiminput{hmmnode}
 
@@
 right of \figref{viterbi}, we call 
-\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}, 
-which chooses the best (\texttt{minimum}) solution
-from among the three predecessor states:
+\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
+The equation adds $e^{\prime}_{M_{j}}(x_{i})$, computed
+with \texttt{eScore}, to a minimum.
+The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
+is computed by function \texttt{avSum},
+the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
 \smallverbatiminput{vfix}
-
-In this code, we look up emission scores for a given node and amino
-acid using \texttt{eScore},
-we~use~\texttt{/+/} to add transition and emission scores,
-and we use \texttt{extend} to add \texttt{Mat} to the list of states
-encountered up to this point.
+The entire right-hand side is wrapped in \texttt{extend~Mat}.
+This call performs a computation \emph{not} shown in \figref{viterbi}:
+in~addition to computing the most likely BUZZWORD,
+\mrfy\ also computes most likely path (sequence of states).
+Function \texttt{extend} adds~$M$ to the head of that path.
 

### 2012-06-28 mrfy_experience_report.tex
Added first names to bib file. A few edits to paper.

@@
 
+\category{J.3}{Biology and genetics}{}
+\category{D.1.1}{Applicative (Functional) Programming}{}
 % \category{CR-number}{subcategory}{third-level}
@@
 Both the probabilities and the states are determined by the alignment:
-\remark{These bullets need proper parallel structure}
 \begin{itemize}
@@
  column~$j$.
-This probability is called an \emph{emission} probability.
 \item 
@@
 \emph{deletion state}~$D_j$.
-The deletion state determines the probability that the query sequence
-has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
+The deletion state determines the probability that a homologous protein
+has lost an amino acid  in column~$j$, i.e., that it has a gap.
 \end{itemize}
-The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
+The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are called emission probabilities.
+The hidden Markov model also has distinguished \emph{Begin} and \emph{End} states.
 In~our representation, each state contains a probability or a table of
@@
 Each transition has its own probability:
-\remark{Norman: how do you propose to make the subscripts more readable?}
 \begin{itemize} 
@@
 And we include the labels of the states that are
-connected:\remark{\emph{Why} are the labels there?
-And since they are apparently never used, why do they have strictness annotations???}
+connected:
 \smallverbatiminput{tprob}
@@
 This call performs a computation \emph{not} shown in \figref{viterbi}:
-in~addition to computing the most likely BUZZWORD,
-\mrfy\ also computes most likely path (sequence of states).
+in~addition to computing the probability of the most likely path,
+\mrfy\ also computes the most likely path (sequence of states).
 Function \texttt{extend} adds~$M$ to the head of that path.
@@
 adjacent in three-dimensional space.
-We~are studying groups of such acids called \emph{beta
-strands}.\remark{Fix this; we're not really studying beta strands}
+Some proteins contain groups of such acids called \emph{beta
+strands}.
 Beta strands
 can become hydrogen-bonded to each other,
-making them ``stuck together.''
+making them ``stuck together,''
+and these beta strands help distinguish groups of homologous
+proteins from one another.
 MRFy detects homologous proteins that include hydrogen-bonded beta

### 2012-06-28 everygen.tex
wordsmithing, death of age, revised search

--- 2012-06-20 everygen.tex
@@
-everyPt :: SearchGen pt r -> Age -> Scored pt
-         -> Rand r [Aged (Utility (Scored pt))]
-everyPt ss age startPt = do
-  successors <- mapM (nextPt ss) (repeat startPt)
+everyPt :: RandomGen r
+        => SearchGen pt r -> CCost -> Scored pt
+        -> Rand r [CCosted (Utility (Scored pt))]
+everyPt sg age startPt = do
+  successors <- mapM (nextPt sg) (repeat startPt)
   tagged <- zipWithM agedUtility successors [succ age..]
-  let (useless, Aged (Useful newPt) newAge : _) =
-                        span (isUseless . unAged) tagged
-  nextPts <- everyPt ss newAge newPt
-  return $ Aged (Useful startPt) age : useless ++ nextPts
+  let (useless, CCosted (Useful newPt) newCCost : _) =
+                        span (isUseless . unCCosted) tagged
+  nextPts <- everyPt sg newCCost newPt
+  return $ CCosted (Useful startPt) age : useless ++ nextPts
 
   where agedUtility pt age =
-           utility ss move >>= \u -> return $ Aged u age
+           utility sg move >>= \u -> return $ CCosted u age
          where move = Move { older = startPt, younger = pt
-                           , youngerAge = age }
+                           , youngerCCost = age }

### 2012-06-28 mrfy_experience_report.tex
wordsmithing, death of age, revised search

--- 2012-06-28 mrfy_experience_report.tex
@@
 
+\category{J.3}{Biology and genetics}{}
+\category{D.3.2}{Language Classifications}{Applicative (functional) languages}
 % \category{CR-number}{subcategory}{third-level}
@@
 \mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
-The equation adds $e^{\prime}_{M_{j}}(x_{i})$, computed
+The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
 with \texttt{eScore}, to a minimum.
@@
 
-To~account for beta strands, computational biologists must use both a
-richer model of protein structure and a revised version of Viterbi's
-equations.
-The~richer model recognizes that only certain pairs of amino acids are
-able to bond to each other.
+To~account for beta strands, computational biologists add
+a term to Viterbi's equations, and they build
+richer models of protein structure.
+%%  The~richer model recognizes that only certain pairs of amino acids are
+%%  able to bond to each other.
 %%  The presence of beta strands changes both the natu
@@
 \cite{Menke:2010ti,Daniels:2012}.
-MRFy explores new methods of \emph{approximating}
-maximum likelihood.
-
-The presence of beta strands also changes the model.
-Within a beta strand, amino acids are not inserted or deleted;
-a~bonded pair of beta strands is modeled by
+
+The new terms in Viterbi's equations are accompanied by a new model.
+Within a beta strand, amino acids are not inserted or deleted, so
+a bonded pair of beta strands is modeled by
  a pair of sequences of match states.
@@
 
-MRFy searches for placements stochastically,
-trying to minimize scores.
-We~are exploring several stochastic algorithms: random-mutation hill
+MRFy searches for good placements stochastically.
+\mrfy\ implements
+random hill
 climbing, simulated  
 annealing, multistart simulated annealing, and a genetic algorithm.
-Our~exploration exploits higher-order functions, existentially
-quantified types, and (to a degree) lazy evaluation.
+These algorithms share much code, and they use higher-order functions,
+existentially 
+quantified types, and lazy evaluation.
+
+We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
+of \emph{points} in a search space.
+A~point is typically a single placement or perhaps a collection of
+placements. 
+Each point also has a \texttt{Score}; \mrfy\ looks for points with
+good scores.
+
+Ideally, MRFy would use now-classic lazy, modular
+technique advocated by \citet{hughes:why}, in which one function
+computes an infinite sequence of points, and another function
+uses a finite prefix to decide on an approximation.
+But because MRFy's search is stochastic, 
+making MRFy's search lazy and modular is not so easy.
+
+To illustrate the difficulties, we~discuss our simplest search:
+random hill climbing.
+%The space of beta-strand placements has no notion of slope or
+%gradient,
+From any given point in the search space, this search moves a random
+distance in a random 
+direction.
+If~the move leads to a better point, it
+is \emph{useful};
+otherwise it is \emph{useless}.
+\smallverbatiminput{utility}
+(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
+With luck, an~infinite sequence of useful moves converges
+at a local optimum.
+
+\mrfy's search path follows only useful moves;
+if a move is useless, 
+\mrfy\ abandons it and moves again (in a new random direction) from the
+previous point.
+Ideally, \mrfy\ would search by composing a \emph{generator}
+that produces an infinite sequence of moves, 
+a \emph{filter} that selects the useful moves, and 
+a \emph{test} function that enumerates finitely many useful moves
+and returns the final destination.
+But a generator may produce an infinite sequence of useless moves.
+(For example, if \mrfy\ should find a global optimum, every move from
+that point is guaranteed to be useless.)
+Given an infinite sequence of useless inputs, a filter would not
+produce any values, and the search would diverge.
 
@@
 \fi
-In this Markov random field,
-each shaded node represents a beta-strand position.
+Each shaded node represents a beta-strand position.
 Nodes connected by dashed edges are hydrogen-bonded.
@@
 
-We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
-of \emph{points}. 
-Each point contains
-a collection of {placements}; different searches represent
-collections differently.
-Each point also has a \texttt{Score}; \mrfy\ looks for points with
-good scores.
-
-Ideally, MRFy's search would use now-classic lazy, modular
-technique advocated by \citet{hughes:why}, in which one function
-computes an infinite sequence of points, and another function
-uses a finite prefix to decide on an approximation.
-But because MRFy's search is stochastic, 
-making MRFy's search lazy and modular is not so easy.
-
-To illustrate the difficulties, we~use our simplest search:
-random hill climbing.
-The space of beta-strand placements has no notion of slope or
-gradient,
-so at any given point in the search space, this search moves a random
-distance in a random 
-direction.
-If~it leads to a better point, the move
-is \emph{useful};
-otherwise it is \emph{useless}.
-\smallverbatiminput{utility}
-(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
-With luck, an~infinite sequence of useful moves converges
-at a local optimum.
-
-\mrfy's search path follows only useful moves;
-if a move proves useless, 
-\mrfy\ abandons it and tries again (with a new random seed) at the
-previous point.
-Ideally, \mrfy\ would search by composing a \emph{generator}
-that produces an infinite sequence of moves, 
-a \emph{filter} that selects the useful moves, and 
-a \emph{test} function that enumerates finitely many useful moves
-and returns the final destination.
-But the generator may produce an infinite sequence of useless moves.
-(For example, if \mrfy\ should find a global optimum, every move from
-that point is guaranteed to be useless.)
-Given an infinite sequence of useless inputs, a filter would not be
-productive, and a test function would diverge.
-
 We~address this problem by combining ``generate and filter'' into a
@@
 Type variable~\texttt{pt} is a point in the search
-space, and type variable~\texttt{r} is a
-random-number generator:
+space, and \texttt{r} is a
+random-number generator.
+\texttt{Rand~r} is a monad of stochastic computations:
 \smallverbatiminput{gen.tex}
@@
 an existing point.
-Because scoring may be expensive, both \texttt{pt0}
+Because scoring can be expensive, both \texttt{pt0}
 and \texttt{nextPt} use \emph{scored} points, and they can
@@
 
-Why do we have \texttt{pt0} and \texttt{nextPt}
+Why have \texttt{pt0} and \texttt{nextPt}
 instead 
 of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
-which could produce an infinite list?
-Because after the first move
+which could produce an infinite list of points?
+Because if the infinite list contains a useless point,
+
 the starting point for the next move depends on whether the first move

### 2012-06-28 mrfy_experience_report.tex

@@
 \category{J.3}{Biology and genetics}{}
-\category{D.3.2}{Language Classifications}{Applicative (functional) languages}
+\category{D.1.1}{Applicative (Functional) Programming}{}
 % \category{CR-number}{subcategory}{third-level}

### 2012-06-28 mrfy_experience_report.tex
intro once again compressed into first column (this can't continue indefinitely)

@@
 
+\category{D.1.1}{Applicative (Functional) Programming}{}
 \category{J.3}{Biology and genetics}{}
-\category{D.1.1}{Applicative (Functional) Programming}{}
 % \category{CR-number}{subcategory}{third-level}
@@
 (\secref{perf}).
-Moreover, measured as a whole, the software that incorporates this new
+\ifpagetuning
+And overall,
+\else
+Moreover, measured as a whole, 
+\fi
+the software that incorporates this new
 code 
@@
 promise powerful abstractions.
-Some worked as advertised, saved us lots of effort, and were a pleasure
+Some kept the promise, saved us lots of effort, and were a pleasure
 to~use.
 Others, not so much.
-And~we could not tell ahead of time which experience we were likely to have
+And~we couldn't tell in advance which 
+\ifpagetuning
+would be which
+\else
+experience we were likely to have
+\fi
  (\secref{penumbra}).
+\looseness=-1
 \end{itemize}

### 2012-06-28 mrfy_experience_report.tex
more wordsmithing; new description of everyPt

@@
 in amino-acid sequences that are only distantly related.
+MRFy is available at \url{mrfy.cs.tufts.edu}.
 %
@@
 reuse scores from previous points.
-
-Why have \texttt{pt0} and \texttt{nextPt}
-instead 
-of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
-which could produce an infinite list of points?
-Because if the infinite list contains a useless point,
-
-the starting point for the next move depends on whether the first move
-was useful, whereas the first move is always useful.
-That decision is made by the \texttt{utility} function,
+\remark{Question ``why not an infinite list'' has gone missing.}
+
+%%Why have \texttt{pt0} and \texttt{nextPt}
+%%instead 
+%%of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
+%%which could produce an infinite list of points?
+%%Because if the infinite list contains a useless point,
+%%
+%%
+%%the starting point for the next move depends on whether the first move
+%%was useful, whereas the first move is always useful.
+To tell if a point returned by \texttt{nextPt} is useful, we call
+ the \texttt{utility} function,
 which scrutinizes a move represented as follows:
@@
 The decision about utility uses not only a source of randomness but
-also the \emph{age} of the search, which corresponds to the
-computational cost to reach a point in the search.
-For example, in~simulated annealing, as the search progresses,
-the \texttt{utility} function becomes less likely to express a move
+also the \emph{cumulative cost} of the point, which we define to be
+the number of points explored previously.
+The~cumulative cost of the current point is also the age of the
+search,
+and
+in~simulated annealing, for example, as the search ages,
+the \texttt{utility} function becomes less likely to accept a move
 that results in a worse score.
@@
 from its starting point, \texttt{everyPt} provides an infinite list of
-randomly chosen successor points, then calls \texttt{agedUtility} to
+randomly chosen successor points, then calls \texttt{costedUtility} to
 tag each one with the search's age and the point's utility.
-The Prelude function \texttt{span} finds the first useful successor,
-but in case there \emph{is} no successor, the code must also produce
+And the computation of \texttt{successors} should explain why we
+separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
+function that produces an infinite list:
+We~don't \emph{want} the infinite list that would result from
+applying \texttt{nextPt} many times in succession;
+we~want the infinite lists that results from applying \texttt{nextPt}
+to \texttt{startPt} many times in succession, each time with a
+different source of randomness.
+
+Once the successors have been computed and tagged,
+\texttt{span} finds the first useful successor.
+In~case there \emph{is} no successor, \texttt{everyPt} also returns
 all the useless successors.
+\remark{Story about the killer strict Rand monad? Where?}
 If~we do find a useful successor, we start searching anew from that
 point, with a recursive call to \texttt{everyPt}.
-
-The most telling line of code in \texttt{everyPt} is the last line of
-the body, which shows that the result begins with a useful point,
+Because \texttt{everyPt} is monadic, the points accumulated so far are
+appended to its result using the \verb+<$>+~operator.
+%
+The most informative part of \texttt{everyPt} is last expression of
+the \texttt{do} block,
+which shows that the result begins with a useful point,
 is followed by a (possibly infinite, possibly empty) list of useless
 points, and then continues recursively with another useful point.
+
 When the computation returned by \texttt{everyPt} is evaluated against
@@
 mix of useful and useless points.
+\remark{This part is gross and still needs work!}
 This mix can be passed to a test function of type \texttt{SearchStop~pt}:\\
@@
 We have two kinds of non-modularity.
+\remark{Still gross}
 First, termination has to
@@
 and a full day to implement.
-\ifnotcutting
-Our~current \texttt{quit} function simply runs for a fixed number of
-generations, but its interface will enable us to try other criteria,
-such as stopping once $k$~generations have passed without more than $E$
-improvement in score. 
-\fi
-
-Our \texttt{SearchStrategy} interface somewhat resembles an object-oriented
-style of programming.
-However, it is \textit{easier} to create a SearchStrategy
-in Haskell than using conventional object-oriented techniques, because
-all the functions in a SearchStrategy are partial applications of
-other functions.
-If we were using conventional object-oriented
-techniques, we would have to declare subclasses with instance
-variables and would have to write code to initialize objects of those
-subclasses.
+
+\remark{Stop functions not as composable as we would like.
+Wanted: combinator library.}
+
+%%  Our \texttt{SearchStrategy} interface somewhat resembles an object-oriented
+%%  style of programming.
+%%  However, it is \textit{easier} to create a SearchStrategy
+%%  in Haskell than using conventional object-oriented techniques, because
+%%  all the functions in a SearchStrategy are partial applications of
+%%  other functions.
+%%  If we were using conventional object-oriented
+%%  techniques, we would have to declare subclasses with instance
+%%  variables and would have to write code to initialize objects of those
+%%  subclasses.
 
@@
 
+At each point in its search,
 MRFy calls \verb+vee'+ several
- times
-for each placement in each search generation.
+ times.
 Calls on query sequences between beta strands are independent,
@@
 \texttt{parmap rseq} for \texttt{map}.
-Even when there is just one placement per generation,
+Even when there is just one placement per point,
 MRFy can keep six cores busy.
-Given multiple placements per generation,
+Given multiple placements per point,
 it~can use forty-eight cores.
@@
 performs only slightly better, at an average of 35 seconds to find an alignment
-on a large protein structure, than when given 16 cores, which takes an average
+on a large\remark{QUANTIFY!!} protein structure, than when given 16 cores, which takes an average
 of 37 seconds.
@@
 and thus its overall run-time performance cannot be compared directly to MRFy.
+\remark{This \P\ is stultifying}
 However, both programs implement the Viterbi algorithm.
@@
 SMURF can parse a simple protein structure, optimally, in under 10~seconds.
+\remark{At this stage ``parse'' is kind of a new word.}
 It~can parse some more complex structures in 2--3~minutes.
@@
 in 30~seconds.
-This performance
-meets our goals, and it means that MRFy can be used to scan whole genomes.
+That's fast enough
+%This performance
+%meets our goals, and it means that MRFy can be used
+to scan whole genomes.
 
@@
 for medium-sized projects,
-but his contribution to MRFy was limited to minor ranting and refactoring.
+but his primary contributions to MRFy have been limited to \emph{post hoc}
+refactoring and testing.
 
@@
 We~faced another barrier in testing.
+\remark{Where is Andrew's shitty paragraph?  It shoudl replace this one?}
 One of us (Ramsey) kept ranting that we should test our code using 
@@
 
-MRFy is available at \url{mrfy.cs.tufts.edu}
-
 \section{Conclusion}
@@
 
-\ifnotcutting
 \acks
@@
 and stream fusion, which led us eventually to the \texttt{Utility} type.
+Koen Claessen instantly diagnosed an inappropriately strict monad.
 We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
 Larsen, and Nathan Ricci.
-\fi
 

### 2012-06-28 mrfy_experience_report.tex
fix bogus para break

--- 2012-06-28 mrfy_experience_report.tex
@@
 Because if the infinite list contains a useless point,
-
 the starting point for the next move depends on whether the first move

### 2012-06-29 mrfy_experience_report.tex
another pass over the introduction

--- 2012-06-28 mrfy_experience_report.tex
@@
 and the Haskell code is slower.
-But the implementation required little effort;
-the connection between code and
-algorithm is unusually clear (\secref{viterbi});
-the code was easy to parallelize; 
-and it meets our performance needs
+But the Haskell code was easy to write,     % coherent subject!
+it is clearly connected to the underlying
+algorithm (\secref{viterbi}),
+it was easy to parallelize,
+and it performs well enough
 (\secref{perf}).
@@
 \fi
-the software that incorporates this new
+the software that incorporates our new
 code 
@@
 biologists with little background in functional programming,
-we~used Haskell successfully.
-And Haskell has made it easy to explore new research ideas;
-by~contrast,
-our group's C++ code has not been a good vehicle for exploring new ideas
+Haskell has made it easy for us to explore new research ideas.
+By~contrast,
+our group's C++ code has not
+made it easy to explore
+% been a good vehicle for exploring 
+new ideas
 (\secref{comparo}).
@@
 reuse scores from previous points.
-\remark{Question ``why not an infinite list'' has gone missing.}
-
-%%Why have \texttt{pt0} and \texttt{nextPt}
-%%instead 
-%%of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
-%%which could produce an infinite list of points?
-%%Because if the infinite list contains a useless point,
-%%
-%%
-%%the starting point for the next move depends on whether the first move
-%%was useful, whereas the first move is always useful.
+
 To tell if a point returned by \texttt{nextPt} is useful, we call

### 2012-06-29 mrfy_experience_report.tex
lots of rewriting, much still in flight

@@
 \newcommand\pairedwith[1]{{\pi(#1)}}
+
+\newcommand\primcons{\texttt{\small(:)}}
 
@@
 \newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
-\newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}
+\newcommand\smallfuzzverbatiminput[2]{%
+  \everypar\expandafter{\the\everypar\hfuzz=0pt\relax}%
+  \hfuzz=#1 \smallverbatiminput{#2}}
 
@@
 \parskip=0.9\parskip % cheating
+\advance \parskip by 0pt plus 0.1pt
 
@@
 so we have defined a small abstraction:
-\smallverbatiminput{vscore}
+\smallfuzzverbatiminput{10.8pt}{vscore}
 Think of a value of type \texttt{Scored~a} as a container holding
@@
 smallest (and therefore best) score.
+
+\seclabel{cons}
+\seclabel{vee-prime}
 
@@
 \smallverbatiminput{vfix}
-The entire right-hand side is wrapped in \texttt{extend~Mat}.
+The right-hand side is wrapped in a call to
+\texttt{\mbox{fmap (Mat `cons`)}}.
 This call performs a computation \emph{not} shown in \figref{viterbi}:
-in~addition to computing the probability of the most likely path,
-\mrfy\ also computes the most likely path (sequence of states).
-Function \texttt{extend} adds~$M$ to the head of that path.
+\mrfy\ not only computes the probability of the most likely path but 
+also computes the path itself.
+Function \texttt{cons} adds~$M$ to a path;
+we~avoid primitive~\primcons\ for reasons explained
+in \secref{performance} below.
 
@@
 sequence containing a mix of useful and useless moves:
-\smallverbatiminput{everygen}
+\smallfuzzverbatiminput{10.8pt}{everygen}
 }
@@
 \seclabel{perf}
+\seclabel{performance}
 
@@
  times.
+Our~\verb+vee'+
+function computes a \mbox{\small \texttt{Scored [StateLabel]}}, that~is, 
+an~optimal path and its score.
+But at intermediate points in \mrfy's search,
+\mrfy\ uses only the
+score.
+Even though Haskell evaluation is lazy, \verb+vee'+ 
+still allocates thunks that could compute paths.
+To~measure the relevant overhead,
+we~cloned \verb+vee'+ and modified it to compute
+only a score, with no path.
+This change improved 
+run time by nearly~50\%.
+%%But when \texttt{search} finally decides on a placement, we need a
+%%version of \verb+vee'+ that \emph{does} compute a path, so we can
+%%extract the optimal parse for that placement.
+
+How could we keep the improvement without 
+maintaining two versions of \verb+vee'+?
+In~Lisp or Ruby we would have used macros or metaprogramming,
+but we were not confident of our ability to metaprogram a solution
+using Template Haskell.
+Instead, we use higher-order functions.
+As~shown in \secref{vee-prime},
+\verb+vee'+ does not use primitive~\primcons\ but instead
+uses an unknown function \texttt{cons}, which is
+passed~in.
+To~get a path, we pass in primitive~\primcons;
+to~get just a score, we pass in
+\verb+\_ _ -> []+.
+This trick is simple and easy to implement, and it provides the same
+speedup as the cloned and modified~code.
+But~we have an uneasy feeling that it may work only because of
+undocumented properties of GHC's inliner, which may 
+change without warning.
+
+We~found another opportunity to improve performance through parallelism.
 Calls on query sequences between beta strands are independent,
@@
 we substituted
-\texttt{parmap rseq} for \texttt{map}.
+\texttt{parmap rseq} for \texttt{map}.\remark{Did NR break this?}
 Even when there is just one placement per point,
-MRFy can keep six cores busy.
+MRFy can keep 6~cores busy.
+\remark{This paragraph needs to talk about \textbf{speedups}.}
 Given multiple placements per point,
-it~can use forty-eight cores.
+it~can use 48~cores.
 However, we have found that MRFy performs best, at least on our
@@
 performs only slightly better, at an average of 35 seconds to find an alignment
-on a large\remark{QUANTIFY!!} protein structure, than when given 16 cores, which takes an average
+on a large\remark{QUANTIFY!!} protein structure, than when given 16
+cores, which takes an average 
 of 37 seconds.
 
-MRFy's predecessor, SMURF, does not perform stochastic search,
-and thus its overall run-time performance cannot be compared directly to MRFy.
-\remark{This \P\ is stultifying}
-However, both programs implement the Viterbi algorithm.
-On an exceedingly large, contrived example, comprising 56,000 amino acids
-(larger than any known protein) and 343 nodes, and ignoring all beta-strand
-information, SMURF's Viterbi implementation completed in 0.497 seconds on a
-2.6GHz Core i7.
-On the same benchmark, MRFy's Viterbi implementation required 1 minute 57
-seconds.
-Despite the inferior runtime performance, MRFy still meets our performance
-goals.
-This is because the stochastic search algorithm relies on the Viterbi algorithm
-applied to many, significantly smaller hidden Markov models than SMURF's
-multidimensional programming must tackle.
-MRFy computes an alignment for a 12-stranded ``beta sandwich'' structure in
-generally less than one minute, while SMURF consumes over 16GB of memory
+MRFy's implementation of Viterbi's algorithm appears dramatically
+slower than the C++ version implemented in \mrfy's predecessor,
+SMURF.
+On a 2.6GHz Intel Core i7,
+given a contrived query sequence of 56,000 amino acids,
+which is larger than any known protein,
+with a hidden Markov model of 343 nodes and no beta strands,
+\mrfy\ takes almost two minutes to compute an optimal parse.
+SMURF parses the same sequence in half a second.
+\remark{This result looks really bad. What is this contrived benchmark supposed to
+mean, anyway??  Can we get a more realistic benchmark?
+What fraction of time is spent in Viterbi?}
+But unlike SMURF, \mrfy\ doesn't have to run on large models.
+{\large{\textbf{QUANTIFY!}}}
+\mrfy\ repeatedly runs Viterbi's algorithm on 
+a~great many smaller models.
+{\large{\textbf{QUANTIFY!}}}
+Here, performance is good enough that \mrfy\ can do its job.
+Given a Markov random field with a 12-stranded ``beta sandwich''
+structure,
+\mrfy\ typically computes an alignment within a minute;\remark{with or
+without parallelism?}
+SMURF allocates over 16GB of memory
 and does not terminate even after eight hours.
-Had we implemented the SMURF algorithm in Haskell, we believe that performance
-would be unacceptable.
-However, for the algorithm underlying MRFy, we are satisfied with the runtime
-performance of Haskell and GHC.
-Nonetheless, we expect that we can still improve MRFy's run-time performance
-with further profiling and tuning.
-
-In~\secref{viterbi} above, the example case for the \verb+vee'+
-function computes not just a score but also an optimal path.
-MRFy's \texttt{search} function, in \secref{mrfy}, uses only the
-score.
-Even though Haskell evaluation is lazy, the \verb+vee'+ 
-still allocates thunks that could compute paths.
-We~cloned \verb+vee'+ and modified it to compute
-only a score, with no path, and this change improved 
-run-time performance by nearly~50\%.
-But when \texttt{search} finally decides on a placement, we need a
-version of \verb+vee'+ that \emph{does} compute a path, so we can
-extract the optimal parse for that placement.
-
-How could we keep the performance improvement without having to
-maintain two versions of \verb+vee'+?
-In~Lisp or Ruby we would have used macros or metaprogramming,
-but we were not confident of our ability to metaprogram a solution
-using Template Haskell.
-Instead, we replaced the primitive cons in \verb+vee'+ with a
-call to a function of the same type, which we passed~in.
-Passing the primitive~``\texttt{:}'' gives us a path,
-and passing \verb+\_ _ -> []+ gives us the same
-speedup as the cloned and modified~code.
-But~although
-this trick is simple and easy to implement, 
-we~have an uneasy feeling that it may work only because of undocumented
-properties of GHC's inliner, which may be subject to change without
-notice.
-
-\subsection{Awkward Haskell}
+%%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
+%%% would be unacceptable.
+%%% However, for the algorithm underlying MRFy, we are satisfied with the runtime
+%%% performance of Haskell and GHC.
+%
+% Nonetheless, we expect that we can still improve MRFy's run-time performance
+% with further profiling and tuning.  
+% And we expect that if we cross the street, we will get to the other side.
+
+\subsection{Awkward profiling and debugging}
 
@@
 
-We~think \verb+vee'+ is beautiful, but we wrote some ugly code, too,
-because we lack skills and knowledge in two areas.
-The first is profiling.
-GHC's profiling is based on cost centers \cite{sansom-pj},
-and for most of MRFy's development, we used GHC 7.0, which
-will happily add cost centers to top-level 
-functions only, while much of our code is in nested functions.
-We~know of no better technique than to add cost-center annotations to
-each guard clause and nested function,
-but these annotations made our code so ugly that we felt compelled to
+We've shown you some of the code we like, but we've also written 
+code we dislike.
+The~code we like least is what we've written to work around
+obstacles we 
+encountered in profiling and debugging.
+
+GHC's profiler assigns costs not to functions but to ``cost
+centers'' \cite{sansom-pj}.
+In GHC~7.0, which we used 
+for most of MRFy's development, 
+cost-center annotations can be added automatically only
+to top-level 
+functions.
+But most of our code is in nested functions.
+We~annotated such functions manually,
+but the annotations
+made our code so ugly that we felt compelled to
 remove them as soon as possible.
-And without the annotations, GHC's call-site and allocation
+Without the annotations, however, GHC's call-site and allocation
 profiles did not give us enough information to be useful.
-We~miss valgrind, whose callgrind profiler gives useful information
-without \emph{any} annotations in the source.
-Late in MRFy's development, we began using GHC 7.4, which provides much
+We~miss valgrind, whose profiler gives useful information
+without \emph{any} annotations.
+(GHC~7.4 provides much
 better profiling tools.
-We are thrilled by these tool developments.
-
-Our other ugly code came from  attempts to debug
+We~look forward to using them.)
+
+We hate the code we've written to help debug
 run-time errors.
@@
 strands overlap and are doubly paired, the invariants of the
-representation are not clear.
-These files are hard to deal with no matter what
+format are not clear.
+This format is hard to deal with no matter what
 programming language is used, and faults are to be expected.
@@
 We~weren't aware of the backtrace feature of GHC's profiler,
-and even after we discovered that feature, the results weren't good enough:
+and even after we discovered that feature, it didn't help:
 the profiler can be used only on very small test cases that didn't always trigger
 run-time errors.
-This same limitation affected the GHCi debugger; our \texttt{vee'} function
-shown in \secref{viterbi}, when it is not compiled with
-optimizations, requires too much run time to be practical on non-trivial
+This same limitation affected the GHCi debugger; 
+in~GHCi, our \texttt{vee'} function
+is too slow to be runnable on nontrivial
 inputs.
@@
 
-\remark{The following \P\ will need to be connected up to the last \S\
-on ``our old approach to debugging.''}
-
-Debugging led also not to awkward code, but to awkward interactions on
+Our difficulties with debugging led to significant disagreements on
 our team.
-The two junior members of the team wanted to apply the debugging
+The two junior members wanted to apply the debugging
 skills they had honed through years of imperative programming.
-The senior partner kept nattering on about QuickCheck properties.
-No~progress was made until the first draft of this paper was prepared.
-At~that point, we were quickly about to write QuickCheck
-specifications for several properties of the Viterbi algorithm:
+These skills did not transfer well to Haskell.
+The senior member of the team kept saying that any proper approach to
+debugging Haskell code should involve QuickCheck.
+This advice was not helpful.
+
+Our experience was that only the senior member of our team was able to use
+QuickCheck easily.
+In~retrospect, we~have identified some obstacles to using
+QuickCheck.
 \begin{itemize}
 \item
-The total number of \verb+Mat+ and \verb+Ins+ states returned equals
-the number of amino acids in the query sequence.
-\item
-The total number of matches, and deletions returned equals the
-number of nodes in the hidden Markov model,
-\emph{except} that an initial sequence of
-of~\verb+Ins+ states counts as a single node,
-due to an odd property of the HMMER file format (namely, the zeroth
-node has no delete state and a non-emitting match state which
-is always occupied, but a normal insert state).
-\item
-The sequence returned satisfies the Plan7 invariant, that is, states
-\verb+Ins+ and \verb+Del+ are never adjacent.
+The examples and tutorials we found focused 
+predominantly on writing and testing properties using data types that
+already implemented class \texttt{Arbitrary}.
+We~didn't understand the \texttt{Arbitrary} class very well, 
+perhaps because the overloaded \texttt{arbitrary} value is not a
+function.
+For programmers who have seen lot of object-oriented dispatch
+on \emph{arguments}, it is hard to grasp how the appropriate version
+of \texttt{arbitrary} is chosen by Haskell's type-class system.
+\item
+Our difficulties were compounded by a weak understanding of monads.
+We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
+true importance.
+\item
+We~continually overlooked the critical role of \emph{shrinking}.
+As~a result, on the one or two occasions we did uncover a problem
+using QuickCheck, we got witnesses that were much too large to be
+informative. 
 \end{itemize}
-\textbf{These properties need to be followed by more ambitious
-properties},
-such as proving that nearby sequences are less likely than the
-sequence returned by \verb+vee'+.
-One practice we recommend for those following in our footsteps is to
-write instances of \texttt{Arbitrary} for data types as those data types
-are written, when relevant invariants are fresh in memory.
-
-This was a practice that the junior members of our team did not follow, and
-they sorely regret it.
-We speculate that one reason why we didn't follow this practice was a poor 
-understanding of the \texttt{Arbitrary} type class.
-Namely, a lot of the examples we had seen using QuickCheck focused 
-predominantly on writing and testing properties with data types that already 
-implemented \texttt{Arbitrary}.
-We recall a smattering about \texttt{Arbitrary}, but we were still too baffled 
-by the \texttt{Gen} monad to truly grasp its importance in using QuickCheck.
-When it finally came time to use QuickCheck, we were overwhelmed by the amount 
-of work that had to be done to write and test implementations of 
-\texttt{Arbitrary} for each of our data types.
+Because of these obstacles, 
+we wrote thousands of lines of code without ever defining an instance
+of \texttt{Arbitrary} and therefore
+without looking hard at QuickCheck.
+After the fact, we were overwhelmed by the work
+involved in writing and testing instances
+of
+\texttt{Arbitrary}.
+The work got done only because the whole team got involved in order to
+meet the deadlines for this paper.
+
+SUMMARY OF WHAT LITTLE WE HAVE LEARNED FROM QUICKCHECK
+Have we tested that \verb+vee'+ returns a most likely sequence?
+And didn't we find an inconsistency?
+
+%%  At~that point, we were quickly about to write QuickCheck
+%%  specifications for several properties of the Viterbi algorithm:
+%%  \begin{itemize}
+%%  \item
+%%  The total number of \verb+Mat+ and \verb+Ins+ states returned equals
+%%  the number of amino acids in the query sequence.
+%%  \item
+%%  The total number of matches, and deletions returned equals the
+%%  number of nodes in the hidden Markov model,
+%%  \emph{except} that an initial sequence of
+%%  of~\verb+Ins+ states counts as a single node,
+%%  due to an odd property of the HMMER file format (namely, the zeroth
+%%  node has no delete state and a non-emitting match state which
+%%  is always occupied, but a normal insert state).
+%%  \item
+%%  The sequence returned satisfies the Plan7 invariant, that is, states
+%%  \verb+Ins+ and \verb+Del+ are never adjacent.
+%%  \end{itemize}
+%%  \textbf{These properties need to be followed by more ambitious
+%%  properties},
 
@@
 computational biologists, have used~C++.
-We~compare our experience with three tools:
+To~compare our Haskell experience with our C++ experience,
+we discuss three tools:
 \begin{itemize}
@@
 with profiling support.
-We~were not able to achieve such recompilations consistently.
+Using Haskell's Cabal tool, which manages installed libraries,
+we~were not able to achieve such recompilations consistently.
+
+REWRITE THE ADVICE ON USING QUICKCHECK
+
+One practice we recommend for those following in our footsteps is to
+write instances of \texttt{Arbitrary} for data types as those data types
+are written, when relevant invariants are fresh in memory.
+
+This was a practice that the junior members of our team did not follow, and
+they sorely regret it.
 

### 2012-06-29 mrfy_experience_report.tex
edits and more edits
also some work on latex warning messages

@@
 \advance \parskip by 0pt plus 0.1pt
+
+\overfullrule=10pt
 
@@
 \[
-\begin{array}{l@{}c@{}l}
+\def\maxiquad{\hskip 1.2em\relax}
+%\mskip -5mu
+\begin{array}{@{}l@{}c@{}l}
 V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
@@
 \end{array}
-\qquad
-\begin{array}{l@{}c@{}l}
+%
+\hskip 1.5em\relax % \qquad too big
+%
+\begin{array}{l@{}c@{}l@{}}
 V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
@@
   \end{array} \right.\\[\gum]
-\multicolumn3{l}{%
+%
+\multicolumn3{l@{}}{%
   a'_{s \hat{s}} = - \log a_{s \hat{s}} 
@@
 \end{array}
+%\mskip -5mu
 \]
@@
 The right-hand side is wrapped in a call to
-\texttt{\mbox{fmap (Mat `cons`)}}.
+\mono{\mbox{fmap (Mat `cons`)}}.
 This call performs a computation \emph{not} shown in \figref{viterbi}:
@@
 We~address this problem by combining ``generate and filter'' into a
-single abstraction, which has type \texttt{SearchGen pt r}.
+single abstraction, which has type \mono{SearchGen pt r}.
 Type variable~\texttt{pt} is a point in the search
@@
 random-number generator.
-\texttt{Rand~r} is a monad of stochastic computations:
+\mono{Rand~r} is a monad of stochastic computations:
 \smallverbatiminput{gen.tex}
@@
 
+\seclabel{awkward-quickcheck}
+
 Our experience was that only the senior member of our team was able to use
@@
 If~you are a computational biologist, 
-you can be productive in Haskell, and you don't need extensive
-experience.
+you can be productive in Haskell without extensive preparation.
 Two~of us (Daniels and Gallant) are graduate students.
@@
 for medium-sized projects,
-but his primary contributions to MRFy have been limited to \emph{post hoc}
+but his contributions to MRFy have been limited, mostly to \emph{post hoc}
 refactoring and testing.
 
-\subsection{Barriers to be overcome}
+\subsection{Obstacles to be overcome}
 
@@
 we~were not able to achieve such recompilations consistently.
-
-REWRITE THE ADVICE ON USING QUICKCHECK
-
-One practice we recommend for those following in our footsteps is to
-write instances of \texttt{Arbitrary} for data types as those data types
-are written, when relevant invariants are fresh in memory.
-
-This was a practice that the junior members of our team did not follow, and
-they sorely regret it.
-
-We~faced another barrier in testing.
-\remark{Where is Andrew's shitty paragraph?  It shoudl replace this one?}
-One of us (Ramsey) kept ranting that we should test our code using 
-QuickCheck \cite{claessen:quickcheck},
-but we could never figure out~how.
-We~were stumped by two problems:
-how to break down MRFy into parts that could be
-unit-tested,
-and how to implement QuickCheck generators for representations of complex
-data like beta-strand placements.
-We~know that QuickCheck is a powerful and proven tool,
-but we could not figure out how to exploit~it.
 
@@
 difficult to discover, and as noted above,
-there are barriers to profiling.
+there are obstacles to profiling.
 We're~aware that debugging lazy functional programs has been a topic
-of some research,\finalremark{Do we cite Hood, Hat, or Freya?}
-but one of the biggest barriers we encountered to using Haskell is
-that we have had to abandon our old approaches to debugging,
-and we~see no clear path to follow.
+of some research,
+but one of the biggest obstacles we encountered to using Haskell is
+that we have had to abandon our old approaches to debugging.
+
+Ideally we would use QuickCheck to find bugs, but as we mention in
+\secref{awkward-quickcheck}, we found obstacles.
+We~have now overcome these obstacles, but
+we~sorely regret not doing so earlier.
+In~light of our experience, we have instituted a new programming practice:
+whenever we introduce a new data type,
+we~write the instance of
+\texttt{Arbitrary} right away, while
+relevant invariants are still fresh in memory.
+When invariants are not enforced by Haskell's static type system,
+we write them as Haskell predicates.
+We can then immediately run QuickCheck on each predicate, to verify
+that our Arbitrary instance agrees with the predicate.
+For each predicate~\texttt{p} we also check
+\mono{fmap (all p . shrink) arbitrary}.
 
@@
 
-An~anonymous referee spurred us to think
-more deeply about laziness.
+Anonymous referees spurred us to think
+more deeply about laziness and to write more carefully about performance.
 Josef Svenningsson suggested a correspondence between \mrfy's search
 and stream fusion, which led us eventually to the \texttt{Utility} type.
-Koen Claessen instantly diagnosed an inappropriately strict monad.
+Koen Claessen instantly diagnosed an inappropriately strict
+implementation of the \texttt{Rand} monad.
 We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad

### 2012-06-29 mrfy_experience_report.tex
more on search

@@
 point, with a recursive call to \texttt{everyPt}.
-Because \texttt{everyPt} is monadic, the points accumulated so far are
-appended to its result using the \verb+<$>+~operator.
+(Because \texttt{everyPt} is monadic, the points accumulated so far are
+appended to its result using the \verb+<$>+~operator.)
 %
@@
 is followed by a (possibly infinite, possibly empty) list of useless
-points, and then continues recursively with another useful point.
-
-When the computation returned by \texttt{everyPt} is evaluated against
-a random-number generator, it returns an infinite list containing a
-mix of useful and useless points.
-\remark{This part is gross and still needs work!}
-This mix can be passed to a test function of type \texttt{SearchStop~pt}:\\
-\vbox{
-\smallverbatiminput{strat}
+points, and then continues recursively with the results of \texttt{everyPt}.
+
+The rest of the search uses Hughes's
+classic composition of generator  and test function.
+Because our code is monadic, we~use the monadic composition
+operator~\texttt{=<<}, which is the bind operator with its arguments
+swapped: 
+\smallverbatiminput{search}
+The \texttt{test} function has type \mono{SearchStop~pt}:\\
 \smallverbatiminputx{stop}
-\smallverbatiminputx{search}
-}
-\noindent
-
-We have two kinds of non-modularity.
-\remark{Still gross}
-First, termination has to
-look at \emph{all} states, even though we would prefer that it look
-only at useful states.
-This is because otherwise we risk non-termination:
-we are not guaranteed a useful move after any finite number of steps.
-Second, age of the search shows up in multiple places.
-In particular, we cannot separate the age of the search at the
-current state from the utility decision.
-
-The \texttt{SearchStrategy} interface provides ample scope for
-experiments.\remark{\P\ needs rewriting}
-Random-mutation hill climbing took fifty lines of code and one~day to
+Type \mono{History~pt} retains only the useful points.
+(Internally, \mrfy\ needs only the \emph{final} useful point,
+but because we want to study how different search algorithms behave,
+we~keep all the useful points.)
+
+The definition of \texttt{SearchStop} reveals the forms of
+non-modularity which are inherent in \mrfy's search algorithm.
+First, we need \texttt{Utility} because if we omit the useless states,
+search might not terminate.
+Second, we need \texttt{CCosted} because some of our test functions
+decide to terminate based either on the cumulative cost of the most
+recent point or on the difference between costs of successive points.
+
+Despite these non-modular aspects, the search
+interface provides ample scope for
+experiments.
+Random hill climbing took fifty lines of code and one~day to
 implement.

### 2012-06-29 mrfy_experience_report.tex
another note

@@
 experiments.
+\remark{This \P\ needs revising in light of the ``monad madness'' refactoring.}
 Random hill climbing took fifty lines of code and one~day to
@@
 that our Arbitrary instance agrees with the predicate.
-For each predicate~\texttt{p} we also check
+For each predicate~\texttt{p} we can also check
 \mono{fmap (all p . shrink) arbitrary}.

### 2012-06-29 mrfy_experience_report.tex
put timestamp on paper
a pass over most of the paper.

@@
 \newif\iffinaldraft \finaldrafttrue % true if it's the final version
+
+\newif\iftimestamp\timestamptrue  % show MD5 stamp of paper
+
+% \timestampfalse % it's post-submission time
+
+\IfFileExists{timestamp.tex}{}{\timestampfalse}
 
@@
 
-\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
-\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
+\newcommand\txprobj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
+\newcommand\txprobjj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
 \newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
@@
 
+\IfFileExists{timestamp.tex}{\input{timestamp}}{}
+\renewcommand{\floatpagefraction}{0.9} % must be less than \topfraction
+\renewcommand{\topfraction}{0.95}
+\renewcommand{\textfraction}{0.05}
+
 \begin{document}
@@
 \preprintfooter{Haskell in Computational Biology}   % 'preprint' option specified.
+\iftimestamp
+\preprintfooter{\mdfivestamp} 
+\fi
 
@@
 profiling and debugging presented obstacles.
-Haskell libraries vary greatly:
+Also, Haskell libraries vary greatly:
 memoization combinators and parallel-evaluation
@@
 The deletion state determines the probability that a homologous protein
-has lost an amino acid  in column~$j$, i.e., that it has a gap.
+has lost an amino acid by deletion from column~$j$.
 \end{itemize}
-The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are called emission probabilities.
-The hidden Markov model also has distinguished \emph{Begin} and \emph{End} states.
+The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.
+
+A hidden Markov model also has distinguished \emph{Begin} and \emph{End} states.
 In~our representation, each state contains a probability or a table of
-probabilities, and it is also labelled with one of these labels:
+probabilities, and it is also labeled with one of these labels:
 \smallverbatiminput{statelabel}
@@
 score, not a probability.
-And we include the labels of the states that are
-connected:
-\smallverbatiminput{tprob}
-In~a Plan7 hidden Markov model, as discussed above, 
+And in~a Plan7 model, as discussed above, 
 there are seven transitions into each node:
-\smallverbatiminput{tprobs}
-In~our code, we select from this 7-tuple using the \texttt{aScore}
-function;
-\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {{\hat s}}$},
+\smallverbatiminput{tprob-tprobs}
+%\smallverbatiminput{tprobs}
+In~our code, we select from this 7-tuple using \texttt{aScore}, a
+function which satisfies
+\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$},
 where $s$~and~$\hat s$ form one of the 7~pairs above.
-
-A~7-tuple of scores is included in the representation of each node in
+Such a 7-tuple of scores is stored in each node in
 the model.
-This representation also includes tables of emission probabilities
-$e_{M_j}$ and $e_{I_j}$.
+A~node also contains tables of emission probabilities
+$e_{M_j}$~and~$e_{I_j}$.
 These tables are read using function
@@
 directly as recursive functions.
-Like the classic recursive Fibonacci function, Viterbi's functions,
+Like a recursive Fibonacci function, Viterbi's functions,
 when implemented \naive ly,
-perform extraordinarily badly.
-But just like the Fibonacci function, Viterbi's functions can be
+take exponential time.
+But like the Fibonacci function, Viterbi's functions can be
 \emph{memoized}.
 Our code implements a transformed version of Viterbi's equations, 
-which operates on scores instead of on probabilities.
+which operates on scores, not on probabilities.
 This version is
@@
 The transformed equations minimize the
-score (i.e., the negated log probability) for each combination of
+score (the negated log probability) for each combination of
 node~$j$, amino 
@@
 \smallfuzzverbatiminput{10.8pt}{vscore}
-Think of a value of type \texttt{Scored~a} as a container holding
+Think of a value of type \mono{Scored~a} as a container holding
  an~``\texttt a'' with a score written on the side.
@@
 The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
-with \texttt{eScore}, to a minimum.
-The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
-is computed by function \texttt{avSum},
-the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
+with \texttt{eScore}, to a minimum of sums.
+The sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
+is computed by function \texttt{avSum}, in which
+the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
 \smallverbatiminput{vfix}
-The right-hand side is wrapped in a call to
-\mono{\mbox{fmap (Mat `cons`)}}.
+%%%%%%%%%% horrible line break
+% The right-hand side of \verb+vee'+ is wrapped in a call to
+What about the call to
+\mono{\mbox{fmap (Mat `cons`)}}?
 This call performs a computation \emph{not} shown in \figref{viterbi}:
-\mrfy\ not only computes the probability of the most likely path but 
-also computes the path itself.
-Function \texttt{cons} adds~$M$ to a path;
-we~avoid primitive~\primcons\ for reasons explained
+\mrfy\ computes not only the {probability} of the most likely path but 
+also the actual path itself.
+Function \mono{(Mat `cons`)} adds~$M$ to a path;
+we~avoid \mbox{\mono{(Mat :)}} for reasons explained
 in \secref{performance} below.
@@
 
-Memoization makes \verb+vee'+ perform just as well as the classic
+Memoization makes \verb+vee'+ perform as well as our classic
 dynamic-programming code.
@@
 devoted to dynamic programming.
-By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
+By~contrast, standard implementations of Viterbi's algorithm, such as in HMMER,
 spend much of their code 
@@
 Because the memoized version so faithfully resembles the equations in
-\figref{viterbi}, we~retired the classic dynamic-programming version.
+\figref{viterbi}, we~retired the classic version.
 
@@
 solves the same problems as existing C++~code.
+\remark{Drop this \P\ or move it}
 Other researchers using Haskell may also have to reimplement code,
@@
 necessity.
+IN COMPUTATIONAL BIOLOGY, REIMPLEMENTING EXISTING ALGORITHMS IS UNREMARKABLE.
 For~example, both SMURF and HMMER contain new implementations of these
@@
  a pair of sequences of match states.
-Between the beta strands, the model is structured as before.
+Between beta strands, the model is structured as before.
 The~combined model, an~example of which is shown in \figref{mrf}, is called a
@@
 strands in the model
- as ``beads'' which can be slid along the query sequence.
+ as ``beads'' which can slide along the query sequence.
 A~positioning of the beta strands is called
@@
 annealing, multistart simulated annealing, and a genetic algorithm.
-These algorithms share much code, and they use higher-order functions,
+These algorithms share much code, and \mrfy\ implements them using
+higher-order functions, 
 existentially 
@@
 of \emph{points} in a search space.
-A~point is typically a single placement or perhaps a collection of
-placements. 
+The type of point is existentially quantified, but it is typically
+a single placement or perhaps a population of placements. 
 Each point also has a \texttt{Score}; \mrfy\ looks for points with
@@
 
-Ideally, MRFy would use now-classic lazy, modular
+Ideally, MRFy would use now-classic, lazy, modular
 technique advocated by \citet{hughes:why}, in which one function
@@
 direction.
-If~the move leads to a better point, it
-is \emph{useful};
+If~the move leads to a better point, 
+we call it \emph{useful};
 otherwise it is \emph{useless}.
@@
 random-number generator.
-\mono{Rand~r} is a monad of stochastic computations:
+\mono{Rand~r} is a lazy monad of stochastic computations:
 \smallverbatiminput{gen.tex}
@@
 the \texttt{utility} function becomes less likely to accept a move
-that results in a worse score.
+that worsens the score.
 
@@
 \vbox{
-Using these pieces, function \texttt{everyGen} produces an infinite
+Using these pieces, function \texttt{everyPt} produces an infinite
 sequence containing a mix of useful and useless moves:
@@
 \noindent
-Because both \texttt{nextPt} and \texttt{utility} consume randomness,
-the computation is very sequential, but it does use a little bit of
+Both \texttt{nextPt} and \texttt{utility} are monadic, but 
+we can still use a little bit of
 laziness:
-from its starting point, \texttt{everyPt} provides an infinite list of
+from its starting point, \texttt{everyPt} produces an infinite list of
 randomly chosen successor points, then calls \texttt{costedUtility} to
-tag each one with the search's age and the point's utility.
-And the computation of \texttt{successors} should explain why we
+tag each one with a cumulative cost and a utility.
+If~you look carefully at how \texttt{successors} is computed,
+we~hope you will understand why we
 separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
@@
 We~don't \emph{want} the infinite list that would result from
-applying \texttt{nextPt} many times in succession;
-we~want the infinite lists that results from applying \texttt{nextPt}
+applying \texttt{nextPt} to many points in succession;
+we~want the infinite list that results from applying \texttt{nextPt}
 to \texttt{startPt} many times in succession, each time with a
@@
 all the useless successors.
-\remark{Story about the killer strict Rand monad? Where?}
 If~we do find a useful successor, we start searching anew from that
@@
 
-The definition of \texttt{SearchStop} reveals the forms of
+The definition of \texttt{SearchStop} reveals two forms of
 non-modularity which are inherent in \mrfy's search algorithm.
@@
 Our~\verb+vee'+
-function computes a \mbox{\small \texttt{Scored [StateLabel]}}, that~is, 
+function computes a \mbox{\mono{Scored [StateLabel]}}, that~is, 
 an~optimal path and its score.
@@
 
-We~found another opportunity to improve performance through parallelism.
+We~also found an opportunity to improve performance through parallelism.
 Calls on query sequences between beta strands are independent,
@@
 we substituted
-\texttt{parmap rseq} for \texttt{map}.\remark{Did NR break this?}
+\texttt{parmap rseq} for \texttt{map}.
 Even when there is just one placement per point,
@@
 inputs.
-Moreover, the GHCi debugger can only set breakpoints at top-level functions
+Moreover, the GHCi debugger can set breakpoints only at top-level functions
 or at specific column/line positions, which made debugging the memoized
@@
 
-Our difficulties with debugging led to significant disagreements on
-our team.
-The two junior members wanted to apply the debugging
+Our difficulties with debugging led to internal disagreements.
+The~junior members of our team wanted to apply the debugging
 skills they had honed through years of imperative programming.
@@
 debugging Haskell code should involve QuickCheck.
-This advice was not helpful.
+This exhortation was unhelpful.
 
@@
 
-Our experience was that only the senior member of our team was able to use
+Only the senior member of our team was able to use
 QuickCheck easily.
-In~retrospect, we~have identified some obstacles to using
+In~retrospect, we~have identified some obstacles that prevented the
+junior people from using
 QuickCheck.
@@
 function.
-For programmers who have seen lot of object-oriented dispatch
-on \emph{arguments}, it is hard to grasp how the appropriate version
-of \texttt{arbitrary} is chosen by Haskell's type-class system.
+For programmers accustomed to object-oriented dispatch
+on \emph{arguments}, it is hard to grasp how Haskell's type-class
+system can find the proper version
+of \texttt{arbitrary}.
 \item
@@
 We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
-true importance.
+importance.
 \item
 We~continually overlooked the critical role of \emph{shrinking}.
-As~a result, on the one or two occasions we did uncover a problem
-using QuickCheck, we got witnesses that were much too large to be
+As~a result, on the one or two occasions we did 
+use QuickCheck, the counterexamples were too large to be
 informative. 
@@
 It~comprises about 1,200 lines of~Haskell.
+\remark{HA HA HA.  It's over 3,500 lines now. Maybe around 800 lines
+are tests or quickcheck properties?}
 It~uses several external tools and libraries, of which the most
@@
 we~sorely regret not doing so earlier.
-In~light of our experience, we have instituted a new programming practice:
+In~light of our experience, we will institute a new programming practice:
 whenever we introduce a new data type,
-we~write the instance of
+we~will write the instance of
 \texttt{Arbitrary} right away, while
@@
 When invariants are not enforced by Haskell's static type system,
-we write them as Haskell predicates.
+we~will write them as Haskell predicates.
 We can then immediately run QuickCheck on each predicate, to verify
@@
 
-You~also need to know that while the Haskell community offers many
+While the Haskell community offers many
 enticing tools, libraries, and packages,
+be aware that
 not all of them are worth using.
@@
 
-As~in any endeavor, it's helpful to have access to experts.
+As~in any endeavor, access to experts helps.
 We~would have been better off if our in-house expert had been an

### 2012-06-30 mrfy_experience_report.tex
fix line count

@@
 beta strands; it effectively supplants SMURF.
-It~comprises about 1,200 lines of~Haskell.
-\remark{HA HA HA.  It's over 3,500 lines now. Maybe around 800 lines
-are tests or quickcheck properties?}
-It~uses several external tools and libraries, of which the most
+It~comprises about 2300 lines of~Haskell, 300 of which are devoted
+to tests and QuickCheck properties and generators.
+Neither C++ codebase includes any testing facilities.
+MRFy~uses several external tools and libraries, of which the most
 notable are Parsec, the BioHaskell bioinformatics library, and

### 2012-06-30 mrfy_experience_report.tex
Many changes. Ugly plot included, need help to fix.

@@
 \seclabel{mrfy}
-
-Our Haskell code for hidden Markov models and Viterbi's algorithm
-solves the same problems as existing C++~code.
-\remark{Drop this \P\ or move it}
-Other researchers using Haskell may also have to reimplement code,
-but researchers may also reimplement code out of choice, not
-necessity.
-IN COMPUTATIONAL BIOLOGY, REIMPLEMENTING EXISTING ALGORITHMS IS UNREMARKABLE.
-For~example, both SMURF and HMMER contain new implementations of these
-same algorithms.
-In~MRFy, we are using the Haskell implementations to help solve new
-problems.
 
@@
 experiments.
-\remark{This \P\ needs revising in light of the ``monad madness'' refactoring.}
 Random hill climbing took fifty lines of code and one~day to
@@
 Simulated annealing reuses hill climbing's
- \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
-Simulated annealing's  new \texttt{accept} function took ten
+ \texttt{nextPt} function.
+Simulated annealing's  new \texttt{utility} function took ten
 lines of code and half an hour to implement.
@@
 Our genetic algorithm reuses the same
-\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
+\texttt{pt0} and \texttt{SearchStop} functions, and its \texttt{utility}
 function took only a few lines of code.
@@
 
-\remark{Stop functions not as composable as we would like.
-Wanted: combinator library.}
+We note that the \texttt{SearchStop} functions are not as composable as we
+would like, and we would like to have a combinator library for searches.
 
@@
 \texttt{parmap rseq} for \texttt{map}.
-Even when there is just one placement per point,
-MRFy can keep 6~cores busy.
-\remark{This paragraph needs to talk about \textbf{speedups}.}
-Given multiple placements per point,
-it~can use 48~cores.
-However, we have found that MRFy performs best, at least on our
-hardware, with between 12 and 16 cores.
-We have access to only
-one 48-core system, and while MRFy will keep all 48 cores busy, it
-performs only slightly better, at an average of 35 seconds to find an alignment
-on a large\remark{QUANTIFY!!} protein structure, than when given 16
-cores, which takes an average 
-of 37 seconds.
+We have access to a
+one 48-core, 2.3GHz AMD Opteron 6176 system, on which we benchmarked \mrfy\ performing
+a search on a large protein structure (an ``8-bladed beta-propeller'')
+comprising 343 hidden Markov model nodes and 40 beta strands, and a query sequence
+592 amino acids long.
+Figure~\ref{multicore} illustrates the mean speedup \mrfy\ achieves as the number of
+processing cores increases.
+Optimal speedup appears to occur between 12 and 16 processors.
+
+\begin{figure}[h!]
+\centerline{\includegraphics[width=6cm]{mrfy_multicore_log}}
+\caption{MRFy run-time performance with varying
+numbers of processing cores on a large protein}\label{multicore}
+\end{figure}
 
 MRFy's implementation of Viterbi's algorithm appears dramatically
-slower than the C++ version implemented in \mrfy's predecessor,
+slower than the C++ version implemented in \mrfy\'s predecessor,
 SMURF.
+Both \mrfy\ and SMURF, when provided with a Markov random field containing no beta strands,
+simply compute the optimal solution using Viterbi's algorithm.
+For both \mrfy\ and SMURF, this task admits no parallelism.
 On a 2.6GHz Intel Core i7,
-given a contrived query sequence of 56,000 amino acids,
-which is larger than any known protein,
-with a hidden Markov model of 343 nodes and no beta strands,
-\mrfy\ takes almost two minutes to compute an optimal parse.
-SMURF parses the same sequence in half a second.
-\remark{This result looks really bad. What is this contrived benchmark supposed to
-mean, anyway??  Can we get a more realistic benchmark?
-What fraction of time is spent in Viterbi?}
+given a micro-benchmark consisting of a contrived query sequence
+of 2000 amino acids (which represents a large protein),
+and a hidden Markov model of 343 nodes and no beta strands,
+\mrfy\ takes 2.32 seconds to compute a Viterbi solution.
+SMURF performs the same computation in 0.29 seconds.
+
 But unlike SMURF, \mrfy\ doesn't have to run on large models.
-{\large{\textbf{QUANTIFY!}}}
-\mrfy\ repeatedly runs Viterbi's algorithm on 
-a~great many smaller models.
-{\large{\textbf{QUANTIFY!}}}
+For example, on the ``8-bladed propeller'' benchmark, the beta strands
+break the Markov random field into 41 pieces, each of which is much smaller
+than the original (typically 10 or fewer nodes rather than 343).
+And, each \texttt{placement} breaks the query sequence into 41 pieces, each of
+which is much smaller than the original (typically 20 or fewer amino acids
+rather than 592).
+Thus, \mrfy\ repeatedly runs Viterbi's algorithm on 
+a~great many smaller models, and the independence of these models
+means that \mrfy\ can execute them in parallel.
 Here, performance is good enough that \mrfy\ can do its job.
@@
 structure,
-\mrfy\ typically computes an alignment within a minute;\remark{with or
-without parallelism?}
+\mrfy\ typically computes an alignment within a minute;
 SMURF allocates over 16GB of memory
 and does not terminate even after eight hours.
+
+Clearly, micro-benchmarks do not tell the whole story.
+Our goal is to solve a biological problem, and though our implementation of
+Viterbi's algorithm is comparatively slow, we can still solve a biological problem
+quickly.
 %%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
@@
 
-\subsection{Awkward profiling and debugging}
+\subsection{Awkward debugging and testing}
 
@@
 obstacles we 
-encountered in profiling and debugging.
-
-GHC's profiler assigns costs not to functions but to ``cost
-centers'' \cite{sansom-pj}.
-In GHC~7.0, which we used 
-for most of MRFy's development, 
-cost-center annotations can be added automatically only
-to top-level 
-functions.
-But most of our code is in nested functions.
-We~annotated such functions manually,
-but the annotations
-made our code so ugly that we felt compelled to
-remove them as soon as possible.
-Without the annotations, however, GHC's call-site and allocation
-profiles did not give us enough information to be useful.
-We~miss valgrind, whose profiler gives useful information
-without \emph{any} annotations.
-(GHC~7.4 provides much
-better profiling tools.
-We~look forward to using them.)
+encountered in debugging.
 
@@
 
-SUMMARY OF WHAT LITTLE WE HAVE LEARNED FROM QUICKCHECK
-Have we tested that \verb+vee'+ returns a most likely sequence?
-And didn't we find an inconsistency?
+Based on the QuickCheck properties we did write, we have identified and
+easily fixed a bug in our implementation of Viterbi's algorithm; we had not
+been including the score for the transition from the final node of the hidden
+Markov model to the special ``end'' state.
+Without QuickCheck, it is unlikely that we would have found this bug,
+and if we had, determining the cause would have required much time.
 
@@
 
+Our Haskell code for hidden Markov models and Viterbi's algorithm
+solves the same problems as existing C++~code.
+Other researchers using Haskell may also have to reimplement code,
+but researchers may also reimplement code out of choice, not
+necessity.
+In computational biology, reimplementing existing algorithms is unremarkable.
+For~example, both SMURF and HMMER contain new implementations of these
+same algorithms.
+In~MRFy, we are using the Haskell implementations to help solve new
+problems.
+
 When performance has mattered, members of our group, like other
@@
 
-In~\secref{awkward-profiling} above, we~mention the awkwardness of the
-cost-center annotations required for profiling.
-But~we also found it difficult just to get the profiler working.
-In~particular, profiling requires that all installed libraries be recompiled
-with profiling support.
-Using Haskell's Cabal tool, which manages installed libraries,
-we~were not able to achieve such recompilations consistently.
+GHC's profiler assigns costs not to functions but to ``cost
+centers'' \cite{sansom-pj}.
+In GHC~7.0, which we used 
+for most of MRFy's development, 
+cost-center annotations can be added automatically only
+to top-level 
+functions.
+But most of our code is in nested functions.
+We~annotated such functions manually,
+but the annotations
+made our code so ugly that we felt compelled to
+remove them as soon as possible.
+Without the annotations, however, GHC's call-site and allocation
+profiles did not give us enough information to be useful.
+We~miss valgrind, whose profiler gives useful information
+without \emph{any} annotations.
+GHC~7.4 provides much
+better profiling tools, which eliminate most of these obstacles,
+rendering this experience no longer as relevant.
 

### 2012-07-01 mrfy_experience_report.tex
new graphs
more work on graphs
possible bug in exposition

@@
 \begin{document}
-\parskip=0.9\parskip % cheating
-\advance \parskip by 0pt plus 0.1pt
+%\parskip=0.9\parskip % cheating
+%\advance \parskip by 0pt plus 0.1pt
 
@@
 
+We're using our implementation of Viterbi's algorithm to help detect
+homologies in new kinds of proteins.
 When a real protein folds in three dimensions, 
@@
 depends on~$x_{\pairedwith j}$.
+\remark{This looks wrong.  $x$ should be subscripted with a
+query-sequence index, but $\pairedwith j$ is a model index.}
 The~relevant term in
@@
 
+%%  \begin{figure}[h!]
+%%  \centerline{\includegraphics[width=6cm]{mrfy_multicore_log}}
+%%  \caption{MRFy run-time performance with varying
+%%  numbers of processing cores on a large protein}
+%%  \label{multicore}
+%%  \end{figure}
+
 \begin{figure}[h!]
-\centerline{\includegraphics[width=6cm]{mrfy_multicore_log}}
-\caption{MRFy run-time performance with varying
-numbers of processing cores on a large protein}\label{multicore}
+\centerline{\includegraphics{speedup}}
+\bigskip
+\caption{\mrfy's parallel speedup}
+\label{fig:speedup}
+\end{figure}
+
+\begin{figure}[h!]
+\centerline{\includegraphics{efficiency}}
+\bigskip
+\caption{\mrfy's parallel efficiency}
+\label{fig:efficiency}
 \end{figure}

### 2012-07-01 mrfy_experience_report.tex
Performance *not* comparable to C++.
extra care in the description of search

@@
 the flexibility and rapid prototyping of a scripting
-language but performance comparable to~C++.
+language, plus the performance of native code.
 In~our experience, higher-order functions, lazy evaluation, and
@@
 tag each one with a cumulative cost and a utility.
-If~you look carefully at how \texttt{successors} is computed,
-we~hope you will understand why we
+We~hope that if you look carefully at how \texttt{successors} is computed,
+you will understand why we
 separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
@@
 decide to terminate based either on the cumulative cost of the most
-recent point or on the difference between costs of successive points.
+recent point or on the difference between costs of successive useful points.
 

### 2012-07-01 mrfy_experience_report.tex
Revised commentary on the code

@@
 experiments.
-Random hill climbing took fifty lines of code and one~day to
+Random hill climbing took 50~lines of code and one~day to
 implement.  
-Simulated annealing reuses hill climbing's
- \texttt{nextPt} function.
-Simulated annealing's  new \texttt{utility} function took ten
-lines of code and half an hour to implement.
+Simulated annealing requires only a new
+\texttt{utility} function, which took 15~lines of code and half an hour to implement.
 (Hill climbing accepts a point if and only if it scores better than its
@@
 function took only a few lines of code.
+\remark{This claim isn't supported by inspection of the code.}
 But~the recombination of parent
@@
 
-We note that the \texttt{SearchStop} functions are not as composable as we
-would like, and we would like to have a combinator library for searches.
+We're not entirely happy with the way we're writing individual
+functions.
+In~particular, \texttt{SearchStop} functions aren't composable;
+we~can't, for example, combine two functions to say that we'd like to
+stop if scores aren't improving or if we've tried a thousand points,
+whichever comes first.
+Eventually, we'd like to have
+combinator libraries for \texttt{SearchStop} and \texttt{nextGen}, at~least.
 

### 2012-07-01 mrfy_experience_report.tex
problems with parallelism

@@
 \texttt{parmap rseq} for \texttt{map}.
-We have access to a
-one 48-core, 2.3GHz AMD Opteron 6176 system, on which we benchmarked \mrfy\ performing
-a search on a large protein structure (an ``8-bladed beta-propeller'')
-comprising 343 hidden Markov model nodes and 40 beta strands, and a query sequence
-592 amino acids long.
-Figure~\ref{multicore} illustrates the mean speedup \mrfy\ achieves as the number of
-processing cores increases.
-Optimal speedup appears to occur between 12 and 16 processors.
+On a 48-core, 2.3GHz AMD Opteron 6176 system, 
+parallel performance is good up to about 12~cores.
+\figref{fig:efficiency} shows parallel efficiency when
+using a model of an ``8-bladed beta propeller;''
+the model has 343 ordinary nodes and 40~beta strands TOTALLING HOW
+MANY NODES?
+We~benchmarked \mrfy\ using a query sequence of
+592 amino acids.
+Figure~\ref{fig:speedup} SHOWS NOTHING OF VALUE BECAUSE RANDOMNESS
+BREAKS EVERYTHING.  AND IT SHOULD HAVE ERROR BARS.
+(Parallel efficiency is 1.0 when the speedup is equal to the number of cores
+used.
+Lower speedups mean lower efficiency.)
 
@@
 
-\begin{figure}[h!]
-\centerline{\includegraphics{speedup}}
-\bigskip
-\caption{\mrfy's parallel speedup}
-\label{fig:speedup}
-\end{figure}
-
-\begin{figure}[h!]
+\begin{figure}%%%[h!]
 \centerline{\includegraphics{efficiency}}
 \bigskip
-\caption{\mrfy's parallel efficiency}
+\caption{\mrfy's WRONG parallel efficiency}
 \label{fig:efficiency}
@@
 
+\break
+
+\begin{figure}%%%[h!]
+\centerline{\includegraphics{speedup}}
+\bigskip
+\caption{\mrfy's parallel speedup}
+\label{fig:speedup}
+\end{figure}
+
 \end{document}

### 2012-07-01 mrfy_experience_report.tex
performance section soundly whacked

@@
 \begin{document}
-%\parskip=0.9\parskip % cheating
-%\advance \parskip by 0pt plus 0.1pt
+\parskip=0.9\parskip % cheating
+\advance \parskip by 0pt plus 0.1pt
 
@@
 
-We~also found an opportunity to improve performance through parallelism.
+Even with this trick,
+MRFy's implementation of Viterbi's algorithm is much
+slower than the C++ version implemented in \mrfy\'s predecessor,
+SMURF.
+For example, on a micro-benchmark that searches for a
+structural motif of 343 nodes in a protein of 2000 amino acids,
+\mrfy\ takes 2.32~seconds and
+SMURF takes 0.29~seconds.
+
+But \mrfy's job is not to run Viterbi's algorithm on large models;
+\mrfy's job is to detect homologies to structures on which Viterbi's
+algorithm is intractable.
+We~benchmarked \mrfy\ 
+using a model of an ``8-bladed beta propeller,''
+which has 343 nodes, of which NNN appear in 40~beta strands.
+The models between beta strands typically have at most 10~nodes.
+We~used a query sequence of 592 amino acids, but each placement breaks
+the sequence into 41~pieces, each of which typically has at most 20 amino
+acids.
+CAN WE PLEASE REPLACE THE TEXT ABOVE WITH A DESCRIPTION OF WHAT'S
+GOING ON IN THE CODE WE ACTUALLY RAN???
+On~these smaller problems, 
+performance is good enough that \mrfy\ can do its job.
+%%
+%%
+%%
+SUDDEN CHANGE OF BENCHMARK CONSIDERED HARMFUL.
+%%
+%%
+Given a Markov random field with a 12-stranded ``beta sandwich''
+structure,
+\mrfy\ typically computes an alignment within a minute;
+SMURF allocates over 16GB of memory
+and does not terminate even after eight hours.
+
+Because \mrfy\ works on many subproblems independently, we~found it easy
+to improve performance through parallelism.
 Calls on query sequences between beta strands are independent,
 as are calls on different placements.
-We~found them easy to parallelize
-using
-\texttt{Control.Parallel}:
-we substituted
+Using
+\texttt{Control.Parallel}, parallelizing this computation was as easy
+as  substituting
 \texttt{parmap rseq} for \texttt{map}.
@@
 
-MRFy's implementation of Viterbi's algorithm appears dramatically
-slower than the C++ version implemented in \mrfy\'s predecessor,
-SMURF.
-Both \mrfy\ and SMURF, when provided with a Markov random field containing no beta strands,
-simply compute the optimal solution using Viterbi's algorithm.
-For both \mrfy\ and SMURF, this task admits no parallelism.
-On a 2.6GHz Intel Core i7,
-given a micro-benchmark consisting of a contrived query sequence
-of 2000 amino acids (which represents a large protein),
-and a hidden Markov model of 343 nodes and no beta strands,
-\mrfy\ takes 2.32 seconds to compute a Viterbi solution.
-SMURF performs the same computation in 0.29 seconds.
-
-But unlike SMURF, \mrfy\ doesn't have to run on large models.
-For example, on the ``8-bladed propeller'' benchmark, the beta strands
-break the Markov random field into 41 pieces, each of which is much smaller
-than the original (typically 10 or fewer nodes rather than 343).
-And, each \texttt{placement} breaks the query sequence into 41 pieces, each of
-which is much smaller than the original (typically 20 or fewer amino acids
-rather than 592).
-Thus, \mrfy\ repeatedly runs Viterbi's algorithm on 
-a~great many smaller models, and the independence of these models
-means that \mrfy\ can execute them in parallel.
-Here, performance is good enough that \mrfy\ can do its job.
-Given a Markov random field with a 12-stranded ``beta sandwich''
-structure,
-\mrfy\ typically computes an alignment within a minute;
-SMURF allocates over 16GB of memory
-and does not terminate even after eight hours.
-
-Clearly, micro-benchmarks do not tell the whole story.
-Our goal is to solve a biological problem, and though our implementation of
-Viterbi's algorithm is comparatively slow, we can still solve a biological problem
-quickly.
 %%% Had we implemented the SMURF algorithm in Haskell, we believe that performance

### 2012-07-01 everygen.tex
mk for noah

--- 2012-06-20 everygen.tex
@@
-everyPt :: SearchGen pt r -> Age -> Scored pt
-         -> Rand r [Aged (Utility (Scored pt))]
-everyPt ss age startPt = do
-  successors <- mapM (nextPt ss) (repeat startPt)
-  tagged <- zipWithM agedUtility successors [succ age..]
-  let (useless, Aged (Useful newPt) newAge : _) =
-                        span (isUseless . unAged) tagged
-  nextPts <- everyPt ss newAge newPt
-  return $ Aged (Useful startPt) age : useless ++ nextPts
-
-  where agedUtility pt age =
-           utility ss move >>= \u -> return $ Aged u age
+everyPt :: RandomGen r
+        => SearchGen pt r -> CCost -> Scored pt
+        -> Rand r [CCosted (Utility (Scored pt))]
+everyPt sg cost startPt = do
+  successors <- mapM (nextPt sg) (repeat startPt)
+  tagged <- zipWithM costedUtility successors [succ cost..]
+  let (useless, CCosted (Useful newPt) newCost : _) =
+                      span (isUseless . unCCosted) tagged
+  ((CCosted (Useful startPt) cost : useless) ++) <$>
+                                  everyPt sg newCost newPt
+ where costedUtility pt cost =
+         utility sg move >>= \u -> return $ CCosted u cost
         where move = Move { older = startPt, younger = pt
-                           , youngerAge = age }
+                          , youngerCCost = cost }

### 2012-07-01 everygen.tex
revised graph scales; fixed code bogons

@@
                       span (isUseless . unCCosted) tagged
-  ((CCosted (Useful startPt) cost : useless) ++) <$>
+  (++) (CCosted (Useful startPt) cost : useless) <$>
                                   everyPt sg newCost newPt

### 2012-07-01 mrfy_experience_report.tex

@@
 the probability of finding amino acid~$x_i$ in column~$j$ 
-depends on~$x_{\pairedwith j}$.
-\remark{This looks wrong.  $x$ should be subscripted with a
-query-sequence index, but $\pairedwith j$ is a model index.}
+depends on~$x_{\pairedwith i}$.
 The~relevant term in
@@
 $V_{j}^{\prime M}(i)$ becomes
-$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
+$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith i}).$$
 Columns~$j$ and $\pairedwith j$ can be as close together as a few
@@
 using a model of an ``8-bladed beta propeller,''
-which has 343 nodes, of which NNN appear in 40~beta strands.
+which has 343 nodes, of which 178 appear in 40~beta strands.
 The models between beta strands typically have at most 10~nodes.
@@
 using a model of an ``8-bladed beta propeller;''
-the model has 343 ordinary nodes and 40~beta strands TOTALLING HOW
-MANY NODES?
+the model has 165 ordinary nodes and 40~beta strands totalling
+178 nodes.
 We~benchmarked \mrfy\ using a query sequence of
@@
 
-We~can also make a preliminary comparison of performance.
-SMURF can parse a simple protein structure, optimally, in under 10~seconds.
-\remark{At this stage ``parse'' is kind of a new word.}
-It~can parse some more complex structures in 2--3~minutes.
-But when asked to parse the most complex structures, it runs out of
-memory and has to be killed, even on a system with 128GB of~RAM.
-By~contrast, MRFy can parse any of these structures, approximately,
-in 30~seconds.
-That's fast enough
-%This performance
-%meets our goals, and it means that MRFy can be used
-to scan whole genomes.
-
 Looking beyond our research group to computational biology more
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{abbrvnat}
 

### 2012-07-01 mrfy_experience_report.tex
plainnat bibstyle
correct graph
parallel dangling ref pulled

@@
 parallel performance is good up to about 12~cores.
-\figref{fig:efficiency} shows parallel efficiency when
-using a model of an ``8-bladed beta propeller;''
-the model has 165 ordinary nodes and 40~beta strands totalling
-178 nodes.
-We~benchmarked \mrfy\ using a query sequence of
+\figref{fig:speedup} shows speedup when
+using a model of an ``8-bladed beta propeller'' with 1~to~48 cores.
+The model has 165 ordinary nodes and 40~beta strands totalling
+178 nodes; the benchmark query sequence has
 592 amino acids.
-Figure~\ref{fig:speedup} SHOWS NOTHING OF VALUE BECAUSE RANDOMNESS
-BREAKS EVERYTHING.  AND IT SHOULD HAVE ERROR BARS.
-(Parallel efficiency is 1.0 when the speedup is equal to the number of cores
-used.
-Lower speedups mean lower efficiency.)
+AND WHAT DO WE CONCLUDE FROM THE FIGURE, JOHNNY?
 
@@
 \begin{figure}%%%[h!]
-\centerline{\includegraphics{efficiency}}
+\centerline{\includegraphics{speedup}}
 \bigskip
-\caption{\mrfy's WRONG parallel efficiency}
-\label{fig:efficiency}
+\caption{Parallel speedup}
+\label{fig:speedup}
 \end{figure}
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{plainnat}
 
@@
 
-\break
-
-\begin{figure}%%%[h!]
-\centerline{\includegraphics{speedup}}
-\bigskip
-\caption{\mrfy's parallel speedup}
-\label{fig:speedup}
-\end{figure}
+% \break
+% 
+% 
+% \begin{figure}%%%[h!]
+% \centerline{\includegraphics{efficiency}}
+% \bigskip
+% Beautiful $\frac 1 {\log p}$!
+% \caption{\mrfy's parallel efficiency}
+% \label{fig:speedup}
+% \end{figure}
 

### 2012-07-01 mrfy_experience_report.tex
fixed the nonlocal pairing for beta strands, sort of
this was a lot harder than I thought, and I couldn't get an equeation

@@
 the probability of finding amino acid~$x_i$ in column~$j$ 
-depends on~$x_{\pairedwith i}$.
-The~relevant term in
-Viterbi's equations 
-$V_{j}^{\prime M}(i)$ becomes
-$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith i}).$$
-Columns~$j$ and $\pairedwith j$ can be as close together as a few
-columns or as far apart as a few hundreds of columns.
-Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
-also on the nonlocal $x_{\pairedwith j}$, 
+depends on whatever amino acid~$x'$  is in column~${\pairedwith i}$.
+If~$x'$ is located in position~$i'$ in the query sequence, Viterbi's
+equations must be altered so that, for example
+$V_{j}^{\prime M}(i)$ depends not only on
+$V_{j-1}^{\prime M}(i-1)$ but also on
+$V_{\pairedwith j}^{\prime M}(i')$.\remark{I can't get an equation I
+believe in. ---NR}
+The distance between $j$~and~$\pairedwith j$ can be as small as a few
+columns or as large as a few hundreds of columns.
+Because $V_j^{\prime M}(i)$~depends not only on nearby values but also on
+$V_{\pairedwith j}^{\prime M}(i')$,
 dynamic programming cannot compute the maximum likelihood quickly

### 2012-07-01 mrfy_experience_report.tex
sensible pdf build without latexmk
A couple remarks.

--- 2012-07-01 mrfy_experience_report.tex
@@
 \smallverbatiminput{aa}
+
+\remark{I'm not sure if this is the whole story, since the reader might simply
+  ask, ``Why not map 0 to Ala, 1 to Cys, etc.?''. The real reason we keep the
+  ``Int'' here I think, is that it provides easy indexing into the match and
+  insertion emission vectors.}
+
 The log of a probability is a negative number.
@@
 
+\remark{The opening sentence here sounds awkward, but I'm not sure if it's
+  because I'm reading it wrong. It seems like ``technique'' should be
+  pluralized, or perhaps ``would use now-classic: a lazy and modular
+  techniqe''.}
 Ideally, MRFy would use now-classic, lazy, modular

### 2012-07-01 mrfy_experience_report.tex
correcter 1-core data, rewrote perf section

@@
 % 
-% \keywords
-% keyword1, keyword2
+\keywords
+memoization, stochastic search, QuickCheck, homology detection
 
@@
 simulated annealing may accept a point that scores worse.)
-Our genetic algorithm reuses the same
-\texttt{pt0} and \texttt{SearchStop} functions, and its \texttt{utility}
-function took only a few lines of code.
-\remark{This claim isn't supported by inspection of the code.}
-But~the recombination of parent
+Our genetic algorithm is very similar in three out of its four
+functions,
+but~the recombination of parent
 placements in its \texttt{nextPt} function took forty lines of code
@@
 \mrfy's job is to detect homologies to structures on which Viterbi's
-algorithm is intractable.
-We~benchmarked \mrfy\ 
-using a model of an ``8-bladed beta propeller,''
+algorithm, and SMURF's more complex algorithm, are unsuited.
+SMURF may be faster than \mrfy\ in some cases,
+but \mrfy\ can solve problems that SMURF cannot.
+\mrfy\ can compute an alignment to a complex, 12-stranded ``beta sandwich''
+structure within one minute, while on the same structure, SMURF allocates
+over 16GB of memory and does not terminate even after eight ours.
+The Markov random field model for this structure contains 252 nodes,
+97 of which appear in the 12 beta strands.
+The models between beta strands typically have at most 12~nodes.
+We used a query sequence of 237 amino acids, but each placement
+breaks the sequence into 13 pieces, each of which typically has at most
+11~amino acids.
+On these smaller problems, performance is good enough that \mrfy\ can
+do its job.
+
+We benchmarked \mrfy\ using a model of an ``8-bladed beta propeller,''
 which has 343 nodes, of which 178 appear in 40~beta strands.
@@
 acids.
-CAN WE PLEASE REPLACE THE TEXT ABOVE WITH A DESCRIPTION OF WHAT'S
-GOING ON IN THE CODE WE ACTUALLY RAN???
-On~these smaller problems, 
-performance is good enough that \mrfy\ can do its job.
-%%
-%%
-%%
-SUDDEN CHANGE OF BENCHMARK CONSIDERED HARMFUL.
-%%
-%%
-Given a Markov random field with a 12-stranded ``beta sandwich''
-structure,
-\mrfy\ typically computes an alignment within a minute;
-SMURF allocates over 16GB of memory
-and does not terminate even after eight hours.
-
-Because \mrfy\ works on many subproblems independently, we~found it easy
+Because \mrfy\ can solve the models between the beta strands independently,
+this benchmark has a very high degree of parallelism.
+And, in Haskell, we~found it easy
 to improve performance through parallelism.
@@
 using a model of an ``8-bladed beta propeller'' with 1~to~48 cores.
-The model has 165 ordinary nodes and 40~beta strands totalling
-178 nodes; the benchmark query sequence has
-592 amino acids.
-AND WHAT DO WE CONCLUDE FROM THE FIGURE, JOHNNY?
+We can conclude that, at least on this particular system,
+a single instance of \mrfy\ scales well to about 12~cores.
+However, typical usage of \mrfy\ includes running many searches
+in parallel, which gives us a way to use all 48 cores on this system.
 
@@
 Larsen, and Nathan Ricci.
+Noah Daniels and Andrew Gallant were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).
 

### 2012-07-01 mrfy_experience_report.tex
address an AG comment, respond to an NR comment.

@@
 $V_{\pairedwith j}^{\prime M}(i')$.\remark{I can't get an equation I
-believe in. ---NR}
+believe in. ---NR I think it's fine, unless we want to introduce
+notation to distinguish paired amino acids from paired
+model nodes --ND}
 The distance between $j$~and~$\pairedwith j$ can be as small as a few
@@
 
-\remark{The opening sentence here sounds awkward, but I'm not sure if it's
-  because I'm reading it wrong. It seems like ``technique'' should be
-  pluralized, or perhaps ``would use now-classic: a lazy and modular
-  techniqe''.}
-Ideally, MRFy would use now-classic, lazy, modular
+Ideally, MRFy would use the now-classic, lazy, modular
 technique advocated by \citet{hughes:why}, in which one function

### 2012-07-02 mrfy_experience_report.tex
clarify the viterbi-only benchmark
Tried to shorten a few things.

@@
 For example, on a micro-benchmark that searches for a
-structural motif of 343 nodes in a protein of 2000 amino acids,
+structural motif of 343 nodes in a protein of 2000 amino acids
+using only the Viterbi algorithm and no beta-strand information,
 \mrfy\ takes 2.32~seconds and
@@
 but \mrfy\ can solve problems that SMURF cannot.
-\mrfy\ can compute an alignment to a complex, 12-stranded ``beta sandwich''
-structure within one minute, while on the same structure, SMURF allocates
+We have tested \mrfy\ on a complex, 12-stranded ``beta sandwich''
+structure, which contains 252 nodes,
+97 of which appear in the 12 beta strands.
+% The models between beta strands typically have at most 12~nodes.
+% We used a query sequence of 237 amino acids, but each placement
+% breaks the sequence into 13 pieces, each of which typically has at most
+% 11~amino acids.
+\mrfy\ computes an alignment in this instance in under one minute,
+while on the same structure, SMURF allocates
 over 16GB of memory and does not terminate even after eight ours.
-The Markov random field model for this structure contains 252 nodes,
-97 of which appear in the 12 beta strands.
-The models between beta strands typically have at most 12~nodes.
-We used a query sequence of 237 amino acids, but each placement
-breaks the sequence into 13 pieces, each of which typically has at most
-11~amino acids.
-On these smaller problems, performance is good enough that \mrfy\ can
-do its job.
-
-We benchmarked \mrfy\ using a model of an ``8-bladed beta propeller,''
+
+We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller,''
 which has 343 nodes, of which 178 appear in 40~beta strands.
@@
 rendering this experience no longer as relevant.
+\remark{given that this experience is no longers as relevant,
+and we're tight on space, should we cut this entire paragraph
+about profiling? --ND}
 
@@
 Larsen, and Nathan Ricci.
-Noah Daniels and Andrew Gallant were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).
+ND and AG were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).
 

### 2012-07-02 mrfy_experience_report.tex
fixed first column

@@
 
-\documentclass[nonatbib,preprint,blockstyle,times]{sigplanconf}
+\documentclass[nonatbib,blockstyle,times]{sigplanconf}
   % at 9pt anything but block style is unreadable
@@
 
+\begingroup
+\ifpagetuning
+\makeatletter
+\@paragraphaboveskip = 0.5\@paragraphaboveskip
+\makeatother
+\fi
+
 \category{D.1.1}{Applicative (Functional) Programming}{}
@@
 \keywords
-memoization, stochastic search, QuickCheck, homology detection
+memoization, stochastic search, parallel strategies,  QuickCheck, remote homology detection
+
+\endgroup
 
@@
 But the Haskell code was easy to write,     % coherent subject!
-it is clearly connected to the underlying
-algorithm (\secref{viterbi}),
-it was easy to parallelize,
-and it performs well enough
+clearly implements the underlying mathematics
+ (\secref{viterbi}),
+ was easy to parallelize,
+and performs well enough
 (\secref{perf}).
+\ifpagetuning
+  And our new tool solves a problem that could not be solved by
+  the~C++ tool which preceded~it.
+\else
   \ifpagetuning
@@
   better than the~C++ software that preceded~it.
+\fi
 \item
@@
 Although the first two authors are computational
-biologists with little background in functional programming,
-Haskell has made it easy for us to explore new research ideas.
+biologists
+with little functional-programming experience,
+Haskell made it easy for us to explore new research ideas.
 By~contrast,
-our group's C++ code has not
-made it easy to explore
+our group's C++ code has 
+made it hard to explore
 % been a good vehicle for exploring 

### 2012-07-02 mrfy_experience_report.tex
carefully crafted page breaks on pages 1 to 3

@@
 
-\documentclass[nonatbib,blockstyle,times]{sigplanconf}
+\documentclass[preprint,nonatbib,blockstyle,times]{sigplanconf}
   % at 9pt anything but block style is unreadable
@@
 states and deletion states \cite{Eddy:1998ut}. 
-The model gets its name because there are exactly
-seven possible transitions
+\ifpagetuning
+``Plan7'' implies that
+\else
+The model gets its name because
+\fi
+ there are exactly~7 possible transitions
 into the states of any column~$j$.
@@
 And in~a Plan7 model, as discussed above, 
-there are seven transitions into each node:
+there are 7~transitions into each node,
+each labeled $a_{s\hat s}$,
+where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
+labels: 
 \smallverbatiminput{tprob-tprobs}
 %\smallverbatiminput{tprobs}
-In~our code, we select from this 7-tuple using \texttt{aScore}, a
-function which satisfies
-\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$},
-where $s$~and~$\hat s$ form one of the 7~pairs above.
-Such a 7-tuple of scores is stored in each node in
+To~choose a transition score from this 7-tuple, we use
+function \texttt{aScore}, 
+where
+\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.
+A~7-tuple is stored in each node in
 the model.
-A~node also contains tables of emission probabilities
-$e_{M_j}$~and~$e_{I_j}$.
-These tables are read using function
+Each node also contains tables of emission probabilities
+$e_{M_j}$~and~$e_{I_j}$, which are read using function
 \mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.
@@
 
-% The bibliography should be embedded for final submission.
-
+\iffalse
 \bibliography{mrfy_experience_report}
+\else
+\begin{thebibliography}{9}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Cowen et~al.(2002)Cowen, Bradley, Menke, King, and
+  Berger]{Cowen:2002p588}
+Lenore Cowen, Philip Bradley, Matt Menke, Jonathan King, and Bonnie Berger.
+\newblock {Predicting the beta-helix fold from protein sequence data}.
+\newblock \emph{Journal of Computational Biology}, 2002.
+
+\bibitem[Daniels et~al.(2012)Daniels, Hosur, Berger, and Cowen]{Daniels:2012}
+Noah Daniels, Raghavendra Hosur, Bonnie Berger, and Lenore Cowen.
+\newblock {SMURFLite: combining simplified Markov random fields with simulated
+  evolution improves remote homology detection for beta-structural proteins
+  into the twilight zone}.
+\newblock \emph{Bioinformatics (to appear)}, 2012.
+
+\bibitem[Durbin et~al.(1998)Durbin, Eddy, Krogh, and Mitchison]{Durbin:1998wz}
+Rirchard Durbin, Sean Eddy, Anders Krogh, and Graeme Mitchison.
+\newblock \emph{{Biological Sequence Analysis: Probabilistic Models of Proteins
+  and Nucleic Acids}}.
+\newblock Cambridge University Press, May 1998.
+
+\bibitem[Eddy(1998)]{Eddy:1998ut}
+Sean Eddy.
+\newblock {Profile hidden Markov models}.
+\newblock \emph{Bioinformatics}, 14:\penalty0 755--763, 1998.
+
+\bibitem[Hughes(1989)]{hughes:why}
+John Hughes.
+\newblock Why functional programming matters.
+\newblock \emph{The Computer Journal}, 32\penalty0 (2):\penalty0 98--107, April
+  1989.
+
+\bibitem[Menke et~al.(2008)Menke, Berger, and Cowen]{Menke:2008wu}
+Matthew Menke, Bonnie Berger, and Lenore Cowen.
+\newblock {Matt: local flexibility aids protein multiple structure alignment}.
+\newblock \emph{PLoS Computational Biology}, 2008.
+
+\bibitem[Menke et~al.(2010)Menke, Berger, and Cowen]{Menke:2010ti}
+Matthew Menke, Bonnie Berger, and Lenore Cowen.
+\newblock {Markov random fields reveal an N-terminal double beta-propeller
+  motif as part of a bacterial hybrid two-component sensor system}.
+\newblock \emph{Proceedings of the National Academy of Science}, 2010.
+
+\bibitem[Sansom and Peyton~Jones(1997)]{sansom-pj}
+Patrick Sansom and Simon~L Peyton~Jones.
+\newblock Formally based profiling for higher-order functional languages.
+\newblock \emph{ACM TOPLAS}, 19\penalty0 (2):\penalty0 334--385, 1997.
+
+\bibitem[Viterbi(1967)]{Viterbi:1967hq}
+Andrew Viterbi.
+\newblock {Error bounds for convolutional codes and an asymptotically optimum
+  decoding algorithm}.
+\newblock \emph{IEEE Transactions on Information Theory}, 13\penalty0
+  (2):\penalty0 260--269, April 1967.
+
+\end{thebibliography}
+\fi
 

### 2012-07-02 mrfy_experience_report.tex
good page breaks; now let's check content

@@
 \keywords
-memoization, stochastic search, parallel strategies,  QuickCheck, remote homology detection
+memoization, stochastic search, parallel strategies,  QuickCheck,
+remote homology detection 
+{\raggedright\par}
 
@@
 
-To~account for beta strands, computational biologists add
-a term to Viterbi's equations, and they build
+To~account for beta strands, computational biologists extend
+Viterbi's equations, and they build
 richer models of protein structure.
@@
 the probability of finding amino acid~$x_i$ in column~$j$ 
+\ifpagetuning
+depends on the amino acid~$x'$ in column~${\pairedwith i}$.
+\else
 depends on whatever amino acid~$x'$  is in column~${\pairedwith i}$.
-If~$x'$ is located in position~$i'$ in the query sequence, Viterbi's
-equations must be altered so that, for example
+\fi
+If~$x'$ is in position~$i'$ in the query sequence, Viterbi's
+equations are altered so that, for example
 $V_{j}^{\prime M}(i)$ depends not only on
@@
 
-The new terms in Viterbi's equations are accompanied by a new model.
+The change in Viterbi's equations is accompanied by a new model.
 Within a beta strand, amino acids are not inserted or deleted, so
@@
  a \emph{placement}.
-A~placement's probability is computed
-based on observed frequencies of paired amino acids in 
+A~placement's likelihood is computed
+based on frequencies of amino-acid pairs observed in 
 hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
-Given a placement, the maximum likelihood of the remaining parts of
-the query sequence, which appear between and around beta strands,
-can be computed quickly and exactly 
+Given a placement, the maximum likelihood of the rest of
+the query sequence, between and around beta strands,
+is computed quickly and exactly 
 using Viterbi's algorithm.
+\ifpagetuning
+This likelihood is \emph{conditioned} on the placement.
+\else
 The exact result is a probability that is
 \emph{conditioned} on the placement.
+\fi
 
@@
 quantified types, and lazy evaluation.
+
+\begin{figure}
+\ifpdfmadness
+\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.pdf}} 
+\else
+\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.eps}} 
+\fi
+Each shaded node represents a beta-strand position.
+Nodes connected by dashed edges are hydrogen-bonded.
+
+\caption{A Markov random field with two beta-strand pairs}
+\label{mrf} 
+\end{figure}
 
@@
 produce any values, and the search would diverge.
-
-\begin{figure}
-\ifpdfmadness
-\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
-\else
-\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
-\fi
-Each shaded node represents a beta-strand position.
-Nodes connected by dashed edges are hydrogen-bonded.
-
-\caption{A Markov random field with two beta-strand pairs}
-\label{mrf} 
-\end{figure}
 
@@
 \centerline{\includegraphics{speedup}}
-\bigskip
+\medskip
 \caption{Parallel speedup}

### 2012-07-02 mrfy_experience_report.tex
page 1, column 1

@@
 
+\newskip\presvtopsep
+
 \newenvironment{smallverbatim}{\par\small\verbatimsmalltrue\verbatim}{\endverbatim}
 \newcommand\smallverbatiminput[1]{%
-  \everypar=\expandafter{\the\everypar
+  \verbatimsmalltrue
+  \presvtopsep=\topsep
+  \topsep=0.78\topsep
+  \verbatiminput{#1}%
   \verbatimsmallfalse
-%       \belowdisplayskip=\abovedisplayskip
-%       \belowdisplayshortskip=\abovedisplayshortskip
-       \topsep=\standardvspace}%
-  \topsep=0.78\topsep
-%  \belowdisplayskip=0.78\belowdisplayskip
-%  \belowdisplayshortskip=0.78\belowdisplayshortskip
-  \verbatimsmalltrue\verbatiminput{#1}}
-\newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
+  \topsep=\presvtopsep
+}
 \newcommand\smallfuzzverbatiminput[2]{%
-  \everypar\expandafter{\the\everypar\hfuzz=0pt\relax}%
-  \hfuzz=#1 \smallverbatiminput{#2}}
+  \hfuzz=#1 \smallverbatiminput{#2}\hfuzz=0pt }
 
@@
 Others, not so much.
-And~we couldn't tell in advance which 
+We couldn't tell in advance which 
 \ifpagetuning
@@
 \smallverbatiminput{search}
-The \texttt{test} function has type \mono{SearchStop~pt}:\\
-\smallverbatiminputx{stop}
+The \texttt{test} function has type \mono{SearchStop~pt}:
+\smallverbatiminput{stop}
 Type \mono{History~pt} retains only the useful points.

### 2012-07-02 mrfy_experience_report.tex
page 1, column 2

@@
 
-Proteins %%% are cellular machinery. They
-interact with one another and with other 
-molecules to carry out the functions of living cells: metabolism, regulation, 
+Proteins, by
+interacting with one another and with other 
+molecules, carry out the functions of living cells: metabolism, regulation, 
 signaling, and so on.
@@
 nucleic acids in DNA, which we call a gene.
-Given a genetic sequence, biologists wish to know the cellular
-function of the protein that gene codes for.
+Given a gene, biologists wish to know the cellular
+function of the protein the gene codes for.
 One of the best known methods of discovering such function is
@@
 algorithms which, given a {query sequence}, %% of amino acids,
-find known proteins of similar structure.
-When similar proteins are formed from amino-acid sequences that
-are not too different from the query sequence, the homology can be
+compare it with known proteins.
+When the known proteins have amino-acid sequences that
+are not too different from the query sequence, homology can be
 detected by

### 2012-07-02 mrfy_experience_report.tex
major restructuring of section 3.1 (models and viterbi)

@@
 an amino acid.
-$N$~need not be the same as the number of columns in the alignment,
+The number of amino acids $N$ can differ from the number of columns in the alignment,
 which we write~\alignwidth.
@@
 
-A hidden Markov model also has distinguished \emph{Begin} and \emph{End} states.
+A hidden Markov model also has distinguished ``begin'' and ``end''
+states.
+%%
+%% for consistency with the text above 'begin' and 'end' are neither
+%% capitalized nor italicized.
+%%
 In~our representation, each state contains a probability or a table of
@@
 These
-equations are very clearly explained by
+equations are explained very clearly by
 \citet[Chapter~5]{Durbin:1998wz}
@@
 \begin{smallverbatim}
-data Amino = Ala | Cys | Asp | Glu | Gly | ...
+data Amino = Ala | Cys | Asp | Glu | ...   -- not used
 \end{smallverbatim}
 But our models use a
-legacy file format that represents each amino acid as a small integer.
-Our representation is therefore
+legacy file format in which each amino acid is a small integer
+used only to index arrays.
+%which we use only as an array index.
+We therefore chose 
 \smallverbatiminput{aa}
 
-\remark{I'm not sure if this is the whole story, since the reader might simply
-  ask, ``Why not map 0 to Ala, 1 to Cys, etc.?''. The real reason we keep the
-  ``Int'' here I think, is that it provides easy indexing into the match and
-  insertion emission vectors.}
-
-The log of a probability is a negative number.
-To~make these numbers more pleasant to read, we negate them,
-which eliminates hordes of minus signs from our file formats.
+%%  \remark{I'm not sure if this is the whole story, since the reader might simply
+%%    ask, ``Why not map 0 to Ala, 1 to Cys, etc.?''. The real reason we keep the
+%%    ``Int'' here I think, is that it provides easy indexing into the match and
+%%    insertion emission vectors.}
+
+Our legacy file format also represents
+the log of a probability \emph{negated}, which makes it a positive number.
+%%To~make these numbers more pleasant to read, we negate them,
+%%which eliminates hordes of minus signs from our file formats.
 The negated logarithm of a probability is called a \emph{score}.
@@
 
-In a hidden Markov model, each transition is actually labeled with a
-score, not a probability.
-And in~a Plan7 model, as discussed above, 
-there are 7~transitions into each node,
-each labeled $a_{s\hat s}$,
+In a real hidden Markov model, each probability is represented as a score.
+Our code implements a transformed version of Viterbi's equations, 
+which operates on scores, not on probabilities.
+This version is
+shown on the right-hand side of \figref{viterbi}.
+The transformed equations minimize the
+score (the negated log probability) for each combination of
+column~$j$, amino 
+acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
+
+A~model is represented as a sequence of \emph{nodes};
+node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
+as well as the probabilities of transitions into that node.
+Each node contains tables of emission scores
+$e'_{M_j}$~and~$e'_{I_j}$.
+These table are read by function
+\texttt{eScore}, which satisfies
+the law % parallel structure for line breaking below
+\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
+%
+We place the transition probabilities 
+into a record in which each field is labeled
+\texttt{$s$\_$\mskip 1mu\hat s$},
 where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
@@
 \smallverbatiminput{tprob-tprobs}
-%\smallverbatiminput{tprobs}
-To~choose a transition score from this 7-tuple, we use
+These scores are read by
 function \texttt{aScore}, 
-where
+which satisfies
+the law % line breaking
 \mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.
-A~7-tuple is stored in each node in
-the model.
-Each node also contains tables of emission probabilities
-$e_{M_j}$~and~$e_{I_j}$, which are read using function
-\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.
 
@@
 
-In~Viterbi's equations,
-the probability in each state is a function of the probabilities
+Scores can be usefully attached to many types of values,
+so we have defined a small abstraction:
+\smallfuzzverbatiminput{10.8pt}{vscore}
+Think of a value of type \mono{Scored~a} as a container holding
+ an~``\texttt a'' with a score written on the side.
+The \texttt{/+/} function adds to the score without touching the
+ container.
+%
+We~also made \texttt{Scored} an instance of~\texttt{Ord}.
+Containers are ordered by score alone, so applying
+\texttt{minimum} to a list of scored things chooses the thing with the
+smallest (and therefore best) score.
+
+Armed with our models and with the \texttt{Scored} abstraction, we
+attacked Viterbi's equations. 
+The probability in each state is a function of the probabilities
 in its predecessor states, 
@@
 algorithm.
-This algorithm starts at the begin state,
+This algorithm starts at the {begin} state,
 computes probabilites in nodes $1$~through~$\alignwidth$ in
-succession, and terminates at the end state.
+succession, and terminates at the {end} state.
 One of us implemented this algorithm, storing the probabilities in an array.
@@
 
+\seclabel{cons}
+\seclabel{vee-prime}
+
 Another of us was curious to try coding Viterbi's equations
@@
 \emph{memoized}.
-Our code implements a transformed version of Viterbi's equations, 
-which operates on scores, not on probabilities.
-This version is
-shown on the right-hand side of \figref{viterbi}.
-The transformed equations minimize the
-score (the negated log probability) for each combination of
-node~$j$, amino 
-acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
-
-Scores can be usefully paired with many types of values,
-so we have defined a small abstraction:
-\smallfuzzverbatiminput{10.8pt}{vscore}
-Think of a value of type \mono{Scored~a} as a container holding
- an~``\texttt a'' with a score written on the side.
-The \texttt{/+/} function adds to the score without touching the
- container.
-
-We~also made \texttt{Scored} an instance of~\texttt{Ord}.
-Containers are ordered by score alone, so applying
-\texttt{minimum} to a list of scored things chooses the thing with the
-smallest (and therefore best) score.
-
-\seclabel{cons}
-\seclabel{vee-prime}
-
-The \texttt{Scored} abstraction makes it easy to implement
-Viterbi's equations.
 For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
-right of \figref{viterbi}, we call 
+right of \figref{viterbi}, we define
 \mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
@@
 with \texttt{eScore}, to a minimum of sums.
-The sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
+The~sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
 is computed by function \texttt{avSum}, in which
 the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
-\smallverbatiminput{vfix}
+\smallfuzzverbatiminput{2.9pt}{vfix}
 %%%%%%%%%% horrible line break
@@
 \mrfy\ computes not only the {probability} of the most likely path but 
-also the actual path itself.
+also the path itself.
 Function \mono{(Mat `cons`)} adds~$M$ to a path;
@@
 
-Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+.
+Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+.
 Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,

### 2012-07-02 mrfy_experience_report.tex
fix typo
through page 3

@@
 node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
-as well as the probabilities of transitions into that node.
+as well as the probabilities of transitions out of that node.
 Each node contains tables of emission scores
@@
 
-We're using our implementation of Viterbi's algorithm to help detect
+We use Viterbi's algorithm to help detect
 homologies in new kinds of proteins.
@@
 adjacent in three-dimensional space.
-Some proteins contain groups of such acids called \emph{beta
-strands}.
+Some groups of such acids are called \emph{beta strands}.
 Beta strands
-can become hydrogen-bonded to each other,
-making them ``stuck together,''
-and these beta strands help distinguish groups of homologous
-proteins from one another.
+can be hydrogen-bonded to each other,
+making them ``stuck together.''
+These beta strands help identify groups of homologous
+proteins.
 MRFy detects homologous proteins that include hydrogen-bonded beta
@@
 
-To~account for beta strands, computational biologists extend
-Viterbi's equations, and they build
+Beta strands require new equations and
 richer models of protein structure.
@@
 If~$x'$ is in position~$i'$ in the query sequence, Viterbi's
-equations are altered so that, for example
+equations are altered; for example,
 $V_{j}^{\prime M}(i)$ depends not only on
 $V_{j-1}^{\prime M}(i-1)$ but also on
-$V_{\pairedwith j}^{\prime M}(i')$.\remark{I can't get an equation I
-believe in. ---NR I think it's fine, unless we want to introduce
-notation to distinguish paired amino acids from paired
-model nodes --ND}
+$V_{\pairedwith j}^{\prime M}(i')$.
+%%  \remark{I can't get an equation I
+%%  believe in. ---NR I think it's fine, unless we want to introduce
+%%  notation to distinguish paired amino acids from paired
+%%  model nodes --ND}
 The distance between $j$~and~$\pairedwith j$ can be as small as a few
@@
 
-The change in Viterbi's equations is accompanied by a new model.
-Within a beta strand, amino acids are not inserted or deleted, so
-a bonded pair of beta strands is modeled by
+The new equations are accompanied by a new model.
+Within a beta strand, amino acids are not inserted or deleted, so~a
+bonded pair of beta strands is modeled by 
  a pair of sequences of match states.

### 2012-07-02 mrfy_experience_report.tex
page 4

@@
 $e'_{M_j}$~and~$e'_{I_j}$.
-These table are read by function
+These tables are read by function
 \texttt{eScore}, which satisfies
@@
 
+%
+% N.B. very tight paragraph...
+%
 MRFy treats the beta 
@@
 
-MRFy searches for good placements stochastically.
+MRFy searches for likely placements stochastically.
 \mrfy\ implements
@@
 But because MRFy's search is stochastic, 
-making MRFy's search lazy and modular is not so easy.
+making MRFy's search modular is not so easy.
 
@@
 But a generator may produce an infinite sequence of useless moves.
-(For example, if \mrfy\ should find a global optimum, every move from
-that point is guaranteed to be useless.)
+(For example, if \mrfy\ should stumble upon a global optimum, every move from
+that point would be useless.)
 Given an infinite sequence of useless inputs, a filter would not
@@
 Both \texttt{nextPt} and \texttt{utility} are monadic, but 
-we can still use a little bit of
+we can still use 
 laziness:
@@
 is followed by a (possibly infinite, possibly empty) list of useless
-points, and then continues recursively with the results of \texttt{everyPt}.
+points, and then continues recursively with another call to \texttt{everyPt}.
 

### 2012-07-02 mrfy_experience_report.tex
pages 4 and 5, and minor fixes earlier

@@
 We use Viterbi's algorithm to help detect
-homologies in new kinds of proteins.
+homologies in proteins with specific kinds of structure.
 When a real protein folds in three dimensions, 
@@
 
+%
+%  placement at top left column is not ideal, but right column results
+%  in a bad page break elsewhere
+%
 \begin{figure}
@@
 Both \texttt{nextPt} and \texttt{utility} are monadic, but 
-we can still use 
-laziness:
+we can still exploit laziness:
 from its starting point, \texttt{everyPt} produces an infinite list of
@@
 
+%%\begingroup\parskip=0.9\parskip
+%% \makeatletter\@sectionaboveskip=0.9\@sectionaboveskip\makeatother
 The rest of the search uses Hughes's
@@
 non-modularity which are inherent in \mrfy's search algorithm.
-First, we need \texttt{Utility} because if we omit the useless states,
+First, we need \texttt{Utility}, because if we omit the useless states,
 search might not terminate.
-Second, we need \texttt{CCosted} because some of our test functions
+Second, we need \texttt{CCosted}, because some of our test functions
 decide to terminate based either on the cumulative cost of the most
@@
 implement.  
-Simulated annealing requires only a new
+Simulated annealing required only a new
 \texttt{utility} function, which took 15~lines of code and half an hour to implement.
@@
 simulated annealing may accept a point that scores worse.)
-Our genetic algorithm is very similar in three out of its four
-functions,
-but~the recombination of parent
-placements in its \texttt{nextPt} function took forty lines of code
+Our genetic algorithm uses very similar
+functions, except for \texttt{nextPt}:
+recombination of parent
+placements took forty lines of code
 and a full day to implement.
 
-We're not entirely happy with the way we're writing individual
+We're not entirely happy with the way we're writing all the individual
 functions.
@@
 Eventually, we'd like to have
-combinator libraries for \texttt{SearchStop} and \texttt{nextGen}, at~least.
+combinator libraries for \texttt{SearchStop} and \texttt{nextPt}, at~least.
 
@@
 
-How could we keep the improvement without 
+Could we keep the improvement without 
 maintaining two versions of \verb+vee'+?
 In~Lisp or Ruby we would have used macros or metaprogramming,
-but we were not confident of our ability to metaprogram a solution
-using Template Haskell.
+but we were not confident of our ability to use Template Haskell.
 Instead, we use higher-order functions.
@@
 speedup as the cloned and modified~code.
-But~we have an uneasy feeling that it may work only because of
+But~we worry that it may work only because of
 undocumented properties of GHC's inliner, which may 
-change without warning.
+change.
 
@@
 MRFy's implementation of Viterbi's algorithm is much
-slower than the C++ version implemented in \mrfy\'s predecessor,
+slower than the C++ version in \mrfy's predecessor,
 SMURF.
-For example, on a micro-benchmark that searches for a
-structural motif of 343 nodes in a protein of 2000 amino acids
-using only the Viterbi algorithm and no beta-strand information,
+For example, on a microbenchmark that searches for a
+structural motif of 343~nodes in a protein of 2000 amino acids
+using only Viterbi's algorithm and no beta-strand information,
 \mrfy\ takes 2.32~seconds and
@@
 But \mrfy's job is not to run Viterbi's algorithm on large models;
-\mrfy's job is to detect homologies to structures on which Viterbi's
-algorithm, and SMURF's more complex algorithm, are unsuited.
-SMURF may be faster than \mrfy\ in some cases,
-but \mrfy\ can solve problems that SMURF cannot.
-We have tested \mrfy\ on a complex, 12-stranded ``beta sandwich''
-structure, which contains 252 nodes,
-97 of which appear in the 12 beta strands.
+\mrfy's job is to detect homologies to structures for which both Viterbi's
+algorithm and SMURF's more complex algorithm are unsuited.
+\mrfy\ can solve problems that SMURF cannot.
+For~example, we tried both programs on a complex, 12-stranded ``beta
+sandwich'' 
+model.
+The~model contains 252~nodes,
+97~of which appear in the 12~beta strands.
 % The models between beta strands typically have at most 12~nodes.
@@
 % 11~amino acids.
-\mrfy\ computes an alignment in this instance in under one minute,
-while on the same structure, SMURF allocates
-over 16GB of memory and does not terminate even after eight ours.
-
-We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller,''
-which has 343 nodes, of which 178 appear in 40~beta strands.
-The models between beta strands typically have at most 10~nodes.
+\mrfy\ computes an alignment in under a~minute,
+but SMURF allocates
+over 16GB of memory and does not terminate even after eight hours.
+
+We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
+The model has 343~nodes, of which 178 appear in 40~beta strands.
+The segments between beta strands typically have at most 10~nodes.
 We~used a query sequence of 592 amino acids, but each placement breaks
@@
 Because \mrfy\ can solve the models between the beta strands independently,
-this benchmark has a very high degree of parallelism.
-And, in Haskell, we~found it easy
-to improve performance through parallelism.
-Calls on query sequences between beta strands are independent,
-as are calls on different placements.
+this benchmark has a lot of parallelism,
+which Haskell made it easy to exploit.
 Using
-\texttt{Control.Parallel}, parallelizing this computation was as easy
+\texttt{Control.Parallel}, parallelizing the computation was as easy
 as  substituting
 \texttt{parmap rseq} for \texttt{map}.
-On a 48-core, 2.3GHz AMD Opteron 6176 system, 
-parallel performance is good up to about 12~cores.
-\figref{fig:speedup} shows speedup when
-using a model of an ``8-bladed beta propeller'' with 1~to~48 cores.
-We can conclude that, at least on this particular system,
-a single instance of \mrfy\ scales well to about 12~cores.
-However, typical usage of \mrfy\ includes running many searches
-in parallel, which gives us a way to use all 48 cores on this system.
+\figref{fig:speedup} shows speedups when using from 
+ 1~to~48 of the cores 
+on a 48-core, 2.3GHz AMD Opteron 6176 system.
+After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
+code, speedup rolls off.
+By~running 4~instance of \mrfy\ in parallel on different searches,
+we~hope to be able to use all 48~cores with about 50\%~efficiency.
 
@@
 
-We've shown you some of the code we like, but we've also written 
-code we dislike.
-The~code we like least is what we've written to work around
-obstacles we 
-encountered in debugging.
-
-We hate the code we've written to help debug
-run-time errors.
-Our~group's legacy file
-format is poorly documented: when beta
-strands overlap and are doubly paired, the invariants of the
-format are not clear.
-This format is hard to deal with no matter what
-programming language is used, and faults are to be expected.
-But using Haskell, we had a hard time diagnosing run-time errors.
+Our experience writing code and trying new ideas was excellent, as~was
+the ease of parallelizing \mrfy.
+Higher-order functions, memoization, laziness, and parallel strategies
+really worked.
+But~we also encountered obstacles that prevented functional
+programming from serving us as well as we would like.
+The~most significant obstacles were in debugging and testing.
+
+We had a hard time diagnosing run-time errors.
+We~expected some run-time errors;
+our~group's legacy file
+format is poorly documented and hard to deal with.
+(When beta
+strands overlap and are doubly paired, even the invariants of the
+format are unclear.)
+But using Haskell, we~found the errors hard to diagnose.
 Calls to
@@
 even when relegated to wrapper functions.
-We~weren't aware of the backtrace feature of GHC's profiler,
-and even after we discovered that feature, it didn't help:
-the profiler can be used only on very small test cases that didn't always trigger
-run-time errors.
-This same limitation affected the GHCi debugger; 
+We~didn't know about the backtrace feature of GHC's profiler,
+and even after we learned about it, it didn't help:
+the profiler can be used only on very small test cases,
+which didn't always trigger the errors.
+This same limitation affected GHCi's debugger; 
 in~GHCi, our \texttt{vee'} function
@@
 inputs.
-Moreover, the GHCi debugger can set breakpoints only at top-level functions
+Moreover,  GHCi's debugger can set breakpoints only at top-level functions
 or at specific column/line positions, which made debugging the memoized
 \texttt{vee''} function impractical.
-In~August 2011, 
+In~August~2011, 
 Lennart Augustsson said that the biggest advantage of Strict

### 2012-07-02 mrfy_experience_report.tex
looseness for later page breaking; edits pp 5-7

@@
 and a full day to implement.
+{\ifpagetuning\looseness=1 \par\fi}
 
@@
 SMURF.
+\ifpagetuning\vadjust{\break}\fi
 For example, on a microbenchmark that searches for a
@@
 we~hope to be able to use all 48~cores with about 50\%~efficiency.
+{\ifpagetuning\looseness=1 \par\fi}
 
@@
 skills they had honed through years of imperative programming.
-These skills did not transfer well to Haskell.
-The senior member of the team kept saying that any proper approach to
+But these skills did not transfer well to Haskell.
+The senior member of the team kept repeating that a proper approach to
 debugging Haskell code should involve QuickCheck.
-This exhortation was unhelpful.
+But mere exhortation was unhelpful.
 
@@
 \texttt{Arbitrary}.
-The work got done only because the whole team got involved in order to
+The work got done only when the whole team pitched in to
 meet the deadlines for this paper.
 
-Based on the QuickCheck properties we did write, we have identified and
-easily fixed a bug in our implementation of Viterbi's algorithm; we had not
-been including the score for the transition from the final node of the hidden
+At the last minute, QuickCheck did find
+a bug in our implementation of Viterbi's algorithm: we had omitted
+the score for the transition from the final node of the hidden
 Markov model to the special ``end'' state.
-Without QuickCheck, it is unlikely that we would have found this bug,
-and if we had, determining the cause would have required much time.
+Without QuickCheck, we~probably wouldn't have known anything was wrong.
+%it is unlikely that we would have found this bug,
+%and if we had, determining the cause would have required much time.
 
@@
 Other researchers using Haskell may also have to reimplement code,
-but researchers may also reimplement code out of choice, not
-necessity.
-In computational biology, reimplementing existing algorithms is unremarkable.
-For~example, both SMURF and HMMER contain new implementations of these
-same algorithms.
-In~MRFy, we are using the Haskell implementations to help solve new
-problems.
+but
+in computational biology, reimplementing existing algorithms is unremarkable.
+For~example, both SMURF and HMMER contain new implementations of
+hidden Markov models and Viterbi's algorithm.
 
@@
 beta strands; it effectively supplants SMURF.
-It~comprises about 2300 lines of~Haskell, 300 of which are devoted
-to tests and QuickCheck properties and generators.
-Neither C++ codebase includes any testing facilities.
+It~comprises about 2,500 lines of~Haskell, about 500 of which are devoted
+to tests, QuickCheck properties, and generators.
+Neither C++ codebase includes any test code.
 MRFy~uses several external tools and libraries, of which the most
@@
 and \path{Data.Vector}.
-Its initial development took about three months.
+\mrfy's initial development took about three months.
 %%%% \remark{Others were mostly full-time work, but no need to distinguish.}
 \end{itemize}
-Like much research software, all three of these tools were written
+Like much research software, all three tools were written
  in~haste.
@@
 
-If~you are a computational biologist, 
-you can be productive in Haskell without extensive preparation.
+If~you are a computational biologist and you are interested in
+functional programming, you don't need extensive preparation
+to be productive in Haskell.
 Two~of us (Daniels and Gallant) are graduate students.
-Daniels has taken a graduate seminar in functional programming, which
+Daniels has taken a seminar in functional programming, which
 included some Haskell;
@@
 enticing tools, libraries, and packages,
-be aware that
 not all of them are worth using.
@@
 Larsen, and Nathan Ricci.
-ND and AG were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).
+
+This work was funded in part by NIH grant 1R01GM080330.
 

### 2012-07-02 mrfy_experience_report.tex
fix reference
rewrite of profiling para

@@
 for most of MRFy's development, 
-cost-center annotations can be added automatically only
-to top-level 
-functions.
-But most of our code is in nested functions.
-We~annotated such functions manually,
-but the annotations
+cost-center annotations must be added manually
+to nested functions.
+The annotations
 made our code so ugly that we felt compelled to
 remove them as soon as possible.
-Without the annotations, however, GHC's call-site and allocation
-profiles did not give us enough information to be useful.
+Nonetheless, GHC's profiler led us to the performance
+improvements discussed in \secref{perf}.
 We~miss valgrind, whose profiler gives useful information
@@
 better profiling tools, which eliminate most of these obstacles,
-rendering this experience no longer as relevant.
-\remark{given that this experience is no longers as relevant,
-and we're tight on space, should we cut this entire paragraph
-about profiling? --ND}
+and we look forward to further profiling MRFy.
 
@@
   into the twilight zone}.
-\newblock \emph{Bioinformatics (to appear)}, 2012.
+\newblock \emph{Bioinformatics}, March 2012.
 

### 2012-07-02 mrfy_experience_report.tex
small changes

@@
 an amino acid.
-The number of amino acids $N$ can differ from the number of columns in the alignment,
-which we write~\alignwidth.
+The number of amino acids, $N$, can differ from the number of columns
+in the alignment,~\alignwidth.
 
@@
 terminating with placement of amino acid $x_i$ in state~$M_j$.
-Probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
-These
+Probabilities $V_j^I(i)$ and $V_j^D(i)$ are similar.
+The
 equations are explained very clearly by
@@
 
-Our code represents a query sequence as an immutable vector of amino
+Our code represents a query sequence as an immutable array of amino
 acids.
@@
 
-Our legacy file format also represents
-the log of a probability \emph{negated}, which makes it a positive number.
+Our legacy file format also \emph{negates}
+the log of each probability, making it a positive number.
 %%To~make these numbers more pleasant to read, we negate them,
@@
 In a real hidden Markov model, each probability is represented as a score.
-Our code implements a transformed version of Viterbi's equations, 
-which operates on scores, not on probabilities.
-This version is
+Our code implements a transformed version of Viterbi's equations
+which operates on scores.
+The transformed equations are
 shown on the right-hand side of \figref{viterbi}.
-The transformed equations minimize the
+They minimize the
 score (the negated log probability) for each combination of
 column~$j$, amino 
-acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
+acid~$x_i$, and state $M_j$, $I_j$, or~$D_j$.
 
@@
 These tables are read by function
-\texttt{eScore}, which satisfies
-the law % parallel structure for line breaking below
+\texttt{eScore}, whose specification
+is % parallel structure for line breaking below
 \mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
@@
 function \texttt{aScore}, 
-which satisfies
-the law % line breaking
+whose specification is
 \mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.
@@
  container.
-%
-We~also made \texttt{Scored} an instance of~\texttt{Ord}.
+Function \texttt{fmap} is also defined; it~applies a function to a
+ container's contents. 
+%
+Finally, we~made \texttt{Scored} an instance of~\texttt{Ord}.
 Containers are ordered by score alone, so applying
@@
 but we were not confident of our ability to use Template Haskell.
-Instead, we use higher-order functions.
+Instead, we used higher-order functions.
 As~shown in \secref{vee-prime},
@@
 code, speedup rolls off.
-By~running 4~instance of \mrfy\ in parallel on different searches,
+By~running 4~instances of \mrfy\ in parallel on different searches,
 we~hope to be able to use all 48~cores with about 50\%~efficiency.
@@
 \medskip
-\caption{Parallel speedup}
+\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
 \label{fig:speedup}
@@
 But~we also encountered obstacles that prevented functional
-programming from serving us as well as we would like.
+programming from working as well as we would have liked.
 The~most significant obstacles were in debugging and testing.
@@
 function.
-For programmers accustomed to object-oriented dispatch
+(For programmers accustomed to object-oriented dispatch
 on \emph{arguments}, it is hard to grasp how Haskell's type-class
 system can find the proper version
-of \texttt{arbitrary}.
+of \texttt{arbitrary} using only the type of a \emph{value}.)
 \item
@@
 in computational biology, reimplementing existing algorithms is unremarkable.
-For~example, both SMURF and HMMER contain new implementations of
+For~example, both SMURF and HMMER also contain new implementations of
 hidden Markov models and Viterbi's algorithm.
@@
 to tests, QuickCheck properties, and generators.
-Neither C++ codebase includes any test code.
+Neither Matt nor SMURF includes test code.
 MRFy~uses several external tools and libraries, of which the most
@@
 %%%% \remark{Others were mostly full-time work, but no need to distinguish.}
+{\ifpagetuning\looseness=1 \par\fi}
 \end{itemize}
@@
 repurpose, and the pointer arithmetic was \emph{too} clever: 
-nearly every change resulted in new segfaults.
+nearly every change resulted in new~segfaults.
+{\ifpagetuning\looseness=1 \par\fi}
 
@@
 our hastily written Ruby and~C++.
+{\ifpagetuning\looseness=1 \par\fi}
 

### 2012-07-03 icfp29ex-daniels.tex
for reference, submitted files are included.

--- 2012-07-02 mrfy_experience_report.tex
@@
 
-\documentclass[preprint,nonatbib,blockstyle,times]{sigplanconf}
+\documentclass[nonatbib,blockstyle,times]{sigplanconf}
   % at 9pt anything but block style is unreadable
@@
 on a 48-core, 2.3GHz AMD Opteron 6176 system.
+Errors are estimated from 5~runs.
 After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
@@
 we~hope to be able to use all 48~cores with about 50\%~efficiency.
-{\ifpagetuning\looseness=1 \par\fi}
 
@@
 
-GHC's profiler assigns costs not to functions but to ``cost
-centers'' \cite{sansom-pj}.
-In GHC~7.0, which we used 
+We~had quite some difficulty profiling, but we hope that
+this difficulty may be mitigated by new profiling tools released early
+in~2012 with
+GHC~7.4.
+GHC assigns costs to ``cost centers'' \cite{sansom-pj},
+and
+in GHC~7.0, which we used 
 for most of MRFy's development, 
-cost-center annotations must be added manually
+cost-center annotations had to be added manually
 to nested functions.
-The annotations
+Although these annotations
 made our code so ugly that we felt compelled to
-remove them as soon as possible.
-Nonetheless, GHC's profiler led us to the performance
-improvements discussed in \secref{perf}.
-We~miss valgrind, whose profiler gives useful information
-without \emph{any} annotations.
-GHC~7.4 provides much
-better profiling tools, which eliminate most of these obstacles,
-and we look forward to further profiling MRFy.
+remove them, they did enable us to improve the performance of~\verb+vee'+ as
+discussed in \secref{perf}.
+%We~miss valgrind, whose profiler gives useful information
+%without \emph{any} annotations.
+GHC~7.4 provides 
+more sophisticated profiling tools, which we look forward to using.
+Difficulties using Cabal to enable profiling of installed libraries
+may remain.
 

### 2012-10-31 mrfy_experience_report.tex
error bars on graphs
make reprint version with the proper page number (and remove dependencies on the source code)

@@
 
-\documentclass[preprint,nonatbib,blockstyle,times]{sigplanconf}
+\documentclass[nonatbib,blockstyle,times,reprint]{sigplanconf}
   % at 9pt anything but block style is unreadable
@@
 %\titlebanner{banner above paper title}        % These are ignored unless
+\titlebanner{\textsf{Reprinted from \textrm{\emph{Proceedings of the
+2012 {ACM} {SIGPLAN} International
+Conference on Functional Programming}} (ICFP'12)}}
+\setcounter{page}{227}
+
 \preprintfooter{Haskell in Computational Biology}   % 'preprint' option specified.

### 2013-04-28 jfp.tex
jfp-ified draft

--- 2012-07-03 icfp29ex-daniels.tex
@@
 
-\pagetuningtrue % trying to get page breaks in the best spots
-
-\newif\ifnotcutting \notcuttingfalse % set if not cutting down to size
+\pagetuningfalse % trying to get page breaks in the best spots
+
+\newif\ifnotcutting \notcuttingtrue % set if not cutting down to size
 
@@
 
-\documentclass[nonatbib,blockstyle,times]{sigplanconf}
-  % at 9pt anything but block style is unreadable
+\documentclass[compmod]{jfp1}
 
@@
 \else
-  \newcomming\tuningbox{\relax}
+  \newcommand\tuningbox{\relax}
 \fi
@@
 \usepackage{url}
-\usepackage{amsmath}
+%\usepackage{amsmath}
 \usepackage{array}
@@
    \@noligs}
+\newcommand{\mono}[1]{%
+  {\@tempdima = \fontdimen2\font
+   \texttt{\spaceskip = 1.1\@tempdima #1}}}
 \makeatother
@@
 
+\title{Haskell in Computational Biology}
+
+\author[N. M. Daniels, A. Gallant, and N. Ramsey]
+        {Noah M. Daniels \and Andrew Gallant \and Norman Ramsey \\
+           {Department of Computer Science, Tufts University}\\
+           {\email{{\rmfamily\{}ndaniels, agallant, nr{\rmfamily\}}@cs.tufts.edu}}}
+
 \begin{document}
-%\parskip=0.9\parskip % cheating
-%\advance \parskip by 0pt plus 0.1pt
 
@@
 
-\conferenceinfo{ICFP'12,} {September 9--15, 2012, Copenhagen, Denmark.}
-\CopyrightYear{2012}
-\copyrightdata{978-1-4503-1054-3/12/09}
-
-%\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{Haskell in Computational Biology}   % 'preprint' option specified.
 \iftimestamp
-\preprintfooter{\mdfivestamp} 
+%\preprintfooter{\mdfivestamp} 
 \fi
-
-\title{Experience Report: Haskell in Computational Biology}
-% \subtitle{Subtitle Text, if any}
-
-\authorinfo{Noah M. Daniels \and Andrew Gallant \and Norman Ramsey}
-           {Department of Computer Science, Tufts University}
-           {{\rmfamily\{}ndaniels, agallant, nr{\rmfamily\}}@cs.tufts.edu}
 
@@
 \end{abstract}
-
-\begingroup
-\ifpagetuning
-\makeatletter
-\@paragraphaboveskip = 0.5\@paragraphaboveskip
-\makeatother
-\fi
-
-\category{D.1.1}{Applicative (Functional) Programming}{}
-\category{J.3}{Biology and genetics}{}
-% \category{CR-number}{subcategory}{third-level}
-% 
-% \terms
-% term1, term2
-% 
-\keywords
-memoization, stochastic search, parallel strategies,  QuickCheck,
-remote homology detection 
-{\raggedright\par}
-
-\endgroup
 
@@
 
-\acks
+\section*{Acknowledgments}
 
@@
 implementation of the \texttt{Rand} monad.
+John Tibell taught us the \texttt{UNPACK} pragma.
 We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
@@
 
-\bibliographystyle{plainnat}
-
-\iffalse
 \bibliography{mrfy_experience_report}
-\else
-\begin{thebibliography}{9}
-\providecommand{\natexlab}[1]{#1}
-\providecommand{\url}[1]{\texttt{#1}}
-\expandafter\ifx\csname urlstyle\endcsname\relax
-  \providecommand{\doi}[1]{doi: #1}\else
-  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
-
-\bibitem[Cowen et~al.(2002)Cowen, Bradley, Menke, King, and
-  Berger]{Cowen:2002p588}
-Lenore Cowen, Philip Bradley, Matt Menke, Jonathan King, and Bonnie Berger.
-\newblock {Predicting the beta-helix fold from protein sequence data}.
-\newblock \emph{Journal of Computational Biology}, 2002.
-
-\bibitem[Daniels et~al.(2012)Daniels, Hosur, Berger, and Cowen]{Daniels:2012}
-Noah Daniels, Raghavendra Hosur, Bonnie Berger, and Lenore Cowen.
-\newblock {SMURFLite: combining simplified Markov random fields with simulated
-  evolution improves remote homology detection for beta-structural proteins
-  into the twilight zone}.
-\newblock \emph{Bioinformatics}, March 2012.
-
-\bibitem[Durbin et~al.(1998)Durbin, Eddy, Krogh, and Mitchison]{Durbin:1998wz}
-Rirchard Durbin, Sean Eddy, Anders Krogh, and Graeme Mitchison.
-\newblock \emph{{Biological Sequence Analysis: Probabilistic Models of Proteins
-  and Nucleic Acids}}.
-\newblock Cambridge University Press, May 1998.
-
-\bibitem[Eddy(1998)]{Eddy:1998ut}
-Sean Eddy.
-\newblock {Profile hidden Markov models}.
-\newblock \emph{Bioinformatics}, 14:\penalty0 755--763, 1998.
-
-\bibitem[Hughes(1989)]{hughes:why}
-John Hughes.
-\newblock Why functional programming matters.
-\newblock \emph{The Computer Journal}, 32\penalty0 (2):\penalty0 98--107, April
-  1989.
-
-\bibitem[Menke et~al.(2008)Menke, Berger, and Cowen]{Menke:2008wu}
-Matthew Menke, Bonnie Berger, and Lenore Cowen.
-\newblock {Matt: local flexibility aids protein multiple structure alignment}.
-\newblock \emph{PLoS Computational Biology}, 2008.
-
-\bibitem[Menke et~al.(2010)Menke, Berger, and Cowen]{Menke:2010ti}
-Matthew Menke, Bonnie Berger, and Lenore Cowen.
-\newblock {Markov random fields reveal an N-terminal double beta-propeller
-  motif as part of a bacterial hybrid two-component sensor system}.
-\newblock \emph{Proceedings of the National Academy of Science}, 2010.
-
-\bibitem[Sansom and Peyton~Jones(1997)]{sansom-pj}
-Patrick Sansom and Simon~L Peyton~Jones.
-\newblock Formally based profiling for higher-order functional languages.
-\newblock \emph{ACM TOPLAS}, 19\penalty0 (2):\penalty0 334--385, 1997.
-
-\bibitem[Viterbi(1967)]{Viterbi:1967hq}
-Andrew Viterbi.
-\newblock {Error bounds for convolutional codes and an asymptotically optimum
-  decoding algorithm}.
-\newblock \emph{IEEE Transactions on Information Theory}, 13\penalty0
-  (2):\penalty0 260--269, April 1967.
-
-\end{thebibliography}
-\fi
-
-\iffinaldraft
-
-\vfill
-
-\begingroup
-\parfillskip=0pt
-\advance\leftskip by 0pt plus 30em
-\emph{You can live to surf the Haskell wave, but~if you slide off the crest, you
-drown.}
-\par
-\endgroup
-
-\fi
+\bibliographystyle{jfp}
+
+%%  \iffinaldraft
+%%  
+%%  
+%%  \vfill
+%%  
+%%  \begingroup
+%%  \parfillskip=0pt
+%%  \advance\leftskip by 0pt plus 30em
+%%  \emph{You can live to surf the Haskell wave, but~if you slide off the crest, you
+%%  drown.}
+%%  \par
+%%  \endgroup
+%%  
+%%  \fi
 

### 2013-04-28 jfp.tex
jfp draft with notes

@@
 
-\documentclass[compmod]{jfp1}
+\documentclass[]{jfp1}
 
@@
 \author[N. M. Daniels, A. Gallant, and N. Ramsey]
-        {Noah M. Daniels \and Andrew Gallant \and Norman Ramsey \\
+        {Noah M. Daniels, Andrew Gallant, and Norman Ramsey \\
            {Department of Computer Science, Tufts University}\\
@@
 \end{abstract}
+
+\section{Notes on the new paper}
+
+\input{jfpnotes}
 

### 2013-04-28 jfpnotes.tex
jfp draft with notes

\subsection{Introduction}

A preliminary version of this paper was presented as an \emph{Experience
Report} at ICFP 2012. In this revised paper, we add to our original work
the results of two different attempts to improve and deepen the
application of functional programming to the detection of remote
homologies in proteins:

\begin{itemize}
\item
  To detect the likelihood that a protein is drawn from a family
  represented by a given Hidden Markov Model, we implemented Viterbi's
  algorithm in Haskell. The algorithm uses dynamic programming, which we
  implemented by memoizing a recursive function.

  Our original Haskell code ran 7 to 8 times slower than C++ code built
  in the same research group for the same purpose. Because MRFy exploits
  parallelism well, we could afford to give up this factor of 7 or 8,
  but we felt this number was awfully high. In Section XXXX we report
  our efforts to speed up a native Haskell implementation.
\item
  Our code for Viterbi's algorithm is carefully crafted to reflect the
  equations usually used in textbooks to describe the algorithm. But an
  experienced functional programmer might find our original code
  distasteful:

  \begin{itemize}
  \item
    The central data structure is an array, and it is used in a way that
    is more typical of FORTRAN than of any functional language. (Lots of
    index arithmetic, no wholemeal programming.)
  \item
    There is a good deal of fiddling with indices at the boundaries.
  \item
    The elegant structure of the Plan 7 Hidden Markov Model is
    \emph{not} reflected in the static types used in the Haskell code.
  \item
    There are an alarming number of cases.
  \end{itemize}

  In Section YYYY, we report on our efforts to make the data structures
  and algorithms ``more functional.'' We refactor our original code and
  assess it on three dimensions:

  \begin{itemize}
  \item
    How much is it still manifestly an implementation of Viterbi's
    equations?
  \item
    How well do the static types model the structure of the problem?
  \item
    How do we feel about the code?
  \item
    How does making the code ``more functional'' affect performance?
  \end{itemize}
\end{itemize}

\subsection{XXXX: Experience trying to make Haskell competitive with
C++}

We asked, according to our relatively naive understanding of what
factors affect performance, how Haskell code might differ from C++ code.
We came up with these ideas:

\begin{enumerate}[1.]
\item
  In C and C++, record structures compose with no additional cost in
  allocations or indirections. That is, a record of records is allocated
  in a single block of memory, and a member of a member can be addressed
  with only one address-arithmetic operation---there is no
  ``pointer-chasing''.

  In Haskell, because the language is polymorphic, we believe that a
  typical value must be represented in a single machine word. If the
  value is too big to fit in a machine word, it must be allocated on the
  heap and represented by a pointer. Such a representation is called
  ``boxed''.

  Because Haskell is lazy, even individual members of a record may be
  either evaluated or unevaluated, and may therefore be boxed.

  We set out to make the \emph{machine-level} representation of our
  Haskell data structures more like the machine-level representation
  that a C programmer would use. We were able to discover two mechanism:

  \begin{itemize}
  \item
    A \emph{strictness annotation} can force the evaluation of a part of
    a record and therefore prevent it from being boxed in a thunk.
  \item
    The \texttt{UNPACK} pragma: where do we learn about it (aside from
    Johan Tibell), what do we cite, how is it known how to use it?
  \item
    Does GHC provide an option that enables us to look at some kind of
    GHC IR and know if we have succeeded?
  \end{itemize}
\item
  C and C++ use pointer arithmetic and unsafe array access.

  \begin{itemize}
  \item
    What happens if we change the record of transition probabilities to
    an array indexed by pair of states?

    What should the index type be?

    Can we look at GHC output (maybe \texttt{-ddump-simpl}) and figure
    out whether we are getting a flat two-dimensional array vs an array
    of arrays?

    Is there a one-dimensional array that works?

    Since we are indexing using every element of an enumeration type,
    what is a way we can introduce unsafe indexing that we believe is
    safe according to the type system? Can we convince the compiler?
  \end{itemize}
\end{enumerate}

\subsection{YYYY: Experience making things ``more functional''}

After the experience of presenting our work at ICFP and meeting other
functional programmers there, we became aware that our central data
structure is not what a functional programmer might expect to see. We
were curious to see what effect it might have to change this data
structure. {[}questions from above{]}?

Diagramatically, a Hidden Markov Model looks like a Begin node followed
by zero (one?) or more ordinary nodes followed by an End node. But in
our original code, all nodes are represented in the same way, and the
distinct properties of Begin and End nodes are present only in the code,
not in the types. (As new Haskell programmers, we tended to choose
representations similar to the C representations we had used before, and
because C does not provide discriminated-union type, the responsibility
for knowing which node we're looking at \emph{always} resides in the
code. Our groups C code does not even use C unions (\textbf{check
this}); there would be little benefit.)

\begin{itemize}
\item
  Because every node has to be capable of representing a Begin node,
  every node stores two probabilities used only for transitions out of
  Begin nodes, even though in all but one of the nodes, these
  probabilities go unused.
\item
  (Mumbo jumbo about chopping the node sequence into pieces, so that any
  node might become a Begin node after beta strands are placed. Again,
  the different interpretations of the ``same'' node are not explicit in
  the type system.)
\end{itemize}

Experiments to be carried out:

\begin{enumerate}[1.]
\item
  Refactor the basic representation to eliminate the unused transition
  probabilities.
\item
  Introduce an explicit Begin node with its own custom representation.
\item
  Create a \texttt{Model} type with the form
  \texttt{Begin \{Middle\} End}. Inspired by REPA-style indexing tricks
  (also seen elsewhere), represent the middle sequence as a function of
  type \texttt{Int -\textgreater{}     Middle}. Implement a slicing
  function
  \texttt{Array Node -\textgreater{} Interval -\textgreater{}     Model}.
\item
  Rewrite Viterbi with the new representation.
\end{enumerate}

N.B. The elimination of unused transition probabilities is
\emph{incompatible} with the change in representation to use a
two-dimensional array. These experiments will have to be handled
carefully; we may need to build all four variants from

\begin{verbatim}
{ one node type, three node types } * { probability record, probability array }
\end{verbatim}

A more radical change in the representation is to use the functional
programmer's mainstay: a list instead of an array. This means lists of
nodes as well as lists of residues. This change was motivated by two
observations:

\begin{itemize}
\item
  Our Viterbi code was doing an awful lot of array indexing and index
  arithmetic. This felt more like FORTRAN than like Haskell.
\item
  There were a \emph{lot} of special cases around indices like 0 and -1.
  {[}Some of these cases arose because we didn't have a distinct Begin
  node, so we'll have to see what the code looks like when a Begin node
  is introduced.{]}
\end{itemize}

Changing to lists had several effects:

\begin{itemize}
\item
  The algorithm suddenly came into focus as something familiar to
  functional programmers. Instead of being something blindly translated
  from equations in a textbook, it became (probabilistic) minimum edit
  distance. We thought this was pretty cool.
\item
  It's no longer so obvious how our code relates to Viterbi's algorithm
  as explained in textbooks. Instead, we are relying on a deeper
  understanding of the underlying problem. Perhaps good for us, but
  maybe not so good for communicating with, e.g., new graduate students
  in the group. The jury is still out here.
\item
  Memoization is \emph{much} hairier. The junior members of the team
  might not have been able to pull this off on their own. Lots of
  QuickCheck action, and significant uncertainty.
\item
  We've knowingly introduced new pointer indirections, with who knows
  what effect on locality or on memory traffic. How did this affect
  performance? {[}This problem militates toward testing on multiple
  hardware platforms, including NR's ancient AMD chip before it goes
  away.{]}
\end{itemize}

\subsection{Benchmarking}

We created several benchmarks without beta strands so that MRFy
\emph{only} executes a single pass of Viterbi without executing a
stochastic search. For every benchmark, we tested a change in either the
implementation of \texttt{viterbi} or a change in the representation. We
report figures relative to a baseline, which was established to account
for IO and parsing.

\subsection{Naming the parts of the design space, Take One}

\{Since there are many changes, my idea here is to give names to each of
the changes. Ideally, they'd correspond to a branch name. I have not
thought this through.\}

\begin{itemize}
\item
  \texttt{special-directives} - \{I don't know what this is. These were
  suggested by John Tibell and apparently already added?\}
\item
  \texttt{indirection-removal} - \{I don't know what this is. See
  \#11.\}
\item
  \texttt{state-arrays} - In \texttt{viterbi}, instead of using case
  analysis on HMM states, we use arrays indexed by HMM states.
\item
  \texttt{unsafe-arrays} - Use unsafe array ranges for memoization.
\item
  \texttt{separate-begin} - The ``begin'' HMM node requires special case
  analysis, so we therefore explicitly separate it from all other HMM
  nodes.
\item
  \texttt{static-trans} - Static encoding of HMM transition
  probabilities. \{??? See issue \#2.\}
\item
  \texttt{HMM-list} - Represent an HMM as a \emph{list} of nodes and
  memoize on its prefix.
\end{itemize}

The number of all possible tests given these changes is quite large, so
we tested each change individually and tested some interesting
combinations of changes: \{I don't think we know yet.\}

\{We talked about a commutative diagram in issue \#8. But it seems like
we don't need that for all tests---maybe only the \texttt{static-trans}
and \texttt{HMM-list} tests? What about \texttt{separate-begin}? With
\texttt{separate-begin}, what if we constructed the special node within
\texttt{viterbi}?\}

\subsection{Naming the parts of the design space, Take Two}

The design space is combinatorial. I suggest we name the individual
elements:

\begin{itemize}
\item
  Records of records: nested, flat

  \begin{itemize}
  \item
    relates to special directives and indirection removal?
  \end{itemize}
\item
  Transition tables: named fields, single array, array of arrays
\item
  Transition array indexing: safe, unsafe
\item
  Viterbi decreases: lists, array indices
\item
  Node array indexing: safe, unsafe, not used in inner loop
\item
  Residue array indexing: safe, unsafe, not used in inner loop
\end{itemize}

Some of these are independent and some are connected. For example, if
Viterbi decreases array indices, then that indexing must be safe or
unsafe. If Viterbi decreases lists, then array indexing is not used in
the inner loop.

### 2013-05-06 jfp.tex
added a note about the first delta in the software.  it sucks.

@@
 \input{jfpnotes}
+
+\subsection{Software deltas}
+
+\subsubsection{Reducing cases in Viterbi}
+
+Original Viterbi took a state and two indices as argument, and it
+dealt with impermissible transitions by assigning a very large score
+(which translates to a very small probability).
+Indices 1, 0, and $-1$ were all used to indicate special conditions,
+and 
+the function had 25~cases overall.
+
+\newcommand\pv{\textit{Predecessor Viterbi}}
+\newcommand\ov{{Original Viterbi}}
+
+Our first step in adapting the code was to create \pv.
+Like \ov, \pv\ takes a state and two indices as an argument.
+But at top level, there is no case analysis on the state, and the only
+indices considered are zero and nonzero, so there are only four cases.
+(The state is used in two ways: to compute transition probabilities,
+and to decide which states may permissibly precede the current state.
+The set of permissible preceding states is looked up in a data
+structure, not computed by case analysis.)
+
+Here is the general case in the new code:
+\smallverbatiminput{hoviterbi}
+THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
+WE SHOULD PROBABLY FIX THAT.
+In the old code, there are three corresponding cases: one each for
+match, insert, and delete states.  
+Here is the code for the match case:
+\smallverbatiminput{viterbi}
+The other case are similar; the second argument to \texttt{aScore} is
+always equal to the first (state) argument to \texttt{vee'},
+and the argument to \texttt{map from} is different in each case.
+(In the new code, this argument is identical to \texttt{preceders !
+stateRight}.)
+NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(
+
+The main benefit is that the Viterbi function is reduced from 90~lines
+of code to just 40~lines.
+This change laid the groundwork for more experiments.
+
+BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
 

### 2013-05-06 jfp.tex
A note trying to make the lazy search stuff more explicit.

@@
 tag each one with a cumulative cost and a utility.
-We~hope that if you look carefully at how \texttt{successors} is computed,
-you will understand why we
-separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
-function that produces an infinite list:
-We~don't \emph{want} the infinite list that would result from
-applying \texttt{nextPt} to many points in succession;
-we~want the infinite list that results from applying \texttt{nextPt}
-to \texttt{startPt} many times in succession, each time with a
-different source of randomness.
+
+In Hughes's \citeyearpar{hughes:why} classic model of lazy search,
+we~might expect to use an infinite sequence of points obtained by
+applying \texttt{nextPt} to many points in succession, as done by
+the Prelude function \texttt{iterate}.
+But~if you look carefully at how \texttt{successors} is computed,
+that's not what's happening:
+instead, we~compute an infinite list by applying \texttt{nextPt} to
+a~\emph{single} initial point using infinitely many distinct sources
+of random variation.
+Then, when we find a useful \texttt{newPt}, a~recursive call
+to \texttt{everyPt} computes a \emph{fresh} infinite sequence of
+single jumps starting at~\texttt{newPt}.
+\mrfy's search may explore an arbitrary number of infinite
+lists of points, and the representation we~found easiest to think
+about is the pair of elements
+\texttt{pt0}~and~\texttt{nextPt} in the \texttt{SearchGen} record.
 

### 2013-05-06 jfp.tex
note on where to link figure on search

@@
 of random variation.
+\remark{This is the place to bring in a figure, I~think. ---NR}
 Then, when we find a useful \texttt{newPt}, a~recursive call

### 2013-05-14 jfp.tex
line edits for jfp draft
closes #12

@@
    \fi
+\fi
+
+\ifpdfmadness
+  \def\loadpdf[#1]#2{\includegraphics[#1]{#2.pdf}}
+\else
+  \def\loadpdf[#1]#2{\includegraphics[#1]{#2.eps}}
 \fi
@@
 
-\newcommand\figref[1]{Figure~\ref{#1}}
+\newcommand\figref[1]{Figure~\ref{fig:#1}}
+\newcommand\figrefpair[3][and]{Figures \ref{fig:#2}~#1~\ref{fig:#3}}
+\newcommand\figlabel[1]{\label{fig:#1}}
 \newcommand\secref[1]{Section~\ref{sec:#1}}
@@
 
-\section{Notes on the new paper}
-
-\input{jfpnotes}
-
-\subsection{Software deltas}
-
-\subsubsection{Reducing cases in Viterbi}
-
-Original Viterbi took a state and two indices as argument, and it
-dealt with impermissible transitions by assigning a very large score
-(which translates to a very small probability).
-Indices 1, 0, and $-1$ were all used to indicate special conditions,
-and 
-the function had 25~cases overall.
-
-\newcommand\pv{\textit{Predecessor Viterbi}}
-\newcommand\ov{{Original Viterbi}}
-
-Our first step in adapting the code was to create \pv.
-Like \ov, \pv\ takes a state and two indices as an argument.
-But at top level, there is no case analysis on the state, and the only
-indices considered are zero and nonzero, so there are only four cases.
-(The state is used in two ways: to compute transition probabilities,
-and to decide which states may permissibly precede the current state.
-The set of permissible preceding states is looked up in a data
-structure, not computed by case analysis.)
-
-Here is the general case in the new code:
-\smallverbatiminput{hoviterbi}
-THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
-WE SHOULD PROBABLY FIX THAT.
-In the old code, there are three corresponding cases: one each for
-match, insert, and delete states.  
-Here is the code for the match case:
-\smallverbatiminput{viterbi}
-The other case are similar; the second argument to \texttt{aScore} is
-always equal to the first (state) argument to \texttt{vee'},
-and the argument to \texttt{map from} is different in each case.
-(In the new code, this argument is identical to \texttt{preceders !
-stateRight}.)
-NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(
-
-The main benefit is that the Viterbi function is reduced from 90~lines
-of code to just 40~lines.
-This change laid the groundwork for more experiments.
-
-BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
-
 \section{Introduction}
@@
 In~this paper, we report on experience using a third kind of language,
-Haskell: 
+Haskell.
+Although we encountered some obstacles that kept Haskell from working
+as well as we would have liked, our overall experience was positive:
+Haskell really worked for us, and we believe it will work equally well
+for other computational biologists.
 \begin{itemize}
@@
  (\secref{penumbra}).
-\looseness=-1
+\item
+\emph{[POTENTIAL BULLETS ABOUT SPEED AND ``MORE FUNCTIONAL'']}
 \end{itemize}
+A~preliminary version of this
+paper \citep{daniels:haskell-in-bio:2012} was presented as 
+an \emph{Experience Report} in ICFP~2012.
+This~paper adds to the original paper by describing our experience in
+trying to simplify and speed up the central algorithm.
 
@@
 \ifpdfmadness
-\centerline{\includegraphics[width=6cm]{alignment.pdf}} 
+\centerline{\includegraphics[width=8cm]{alignment.pdf}} 
 \else
-\centerline{\includegraphics[width=6cm]{alignment.eps}} 
+\centerline{\includegraphics[width=8cm]{alignment.eps}} 
 \fi
@@
 \caption{A~structural alignment of four proteins $(\alignwidth = 12)$}
-\label{alignment} 
+\figlabel{alignment} 
 \end{figure}
@@
 has lost an amino acid by deletion from column~$j$.
+\hfuzz=1.5pt % fix blot
 \end{itemize}
@@
 
+\begin{figure} 
+\ifpdfmadness
+\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
+\else
+\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
+\fi
+
+\caption{A hidden Markov model $(\alignwidth = 4)$}
+
+\figlabel{plan7}
+\end{figure}
+
 We~use the ``Plan7'' hidden Markov model, which forbids direct
@@
 states and deletion states \cite{Eddy:1998ut}. 
+\seclabel{plan7}%
+\remark{Reference to \figref{plan7}??}%
 \ifpagetuning
@@
 
-\begin{figure} 
-\ifpdfmadness
-\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
-\else
-\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
-\fi
-
 This model
@@
 
-\caption{A hidden Markov model $(\alignwidth = 4)$}
-
-\label{plan7} \end{figure}
-
-\begin{figure*}
-\newcommand\vsum[2]{#2&{}+{}& #1}
+\subsection{Computing probabilities using perspicuous Haskell}
+
+\seclabel{viterbi}
+
+Given a hidden Markov model, 
+an~established software package called HMMER (pronounced ``hammer'') 
+can compute the probability
+that a new protein shares structure 
+\ifnotcutting and evolutionary history \fi
+with the proteins used to train the model.
+The computation finds the most likely path through the hidden Markov model.
+To~make best use of floating-point arithmetic, the software computes
+the \emph{logarithm} of the probability of each path, by~summing
+the logs of the 
+probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
+The~path that maximizes the
+log of the probability is the most likely path.
 
@@
 \def\gum{14pt}
+\newcommand\vsum[2]{#2&{}+{}& #1}
+
+\begin{figure}
+
 \[
@@
 \end{array}
-%
-\hskip 1.5em\relax % \qquad too big
-%
+\]
+
+\caption{Viterbi's original equations}
+\figlabel{viterbi-original}
+
+\bigskip
+
+%\begin{figure}
+\[
 \begin{array}{l@{}c@{}l@{}}
@@
 \end{array}
-%\mskip -5mu
 \]
 
-\caption{Viterbi's equations, in original and negated forms}
-\label{viterbi}
-\end{figure*}
-
-\subsection{Computing probabilities using perspicuous Haskell}
-
-\seclabel{viterbi}
-
-Given a hidden Markov model, 
-an~established software package called HMMER (pronounced ``hammer'') 
-can compute the probability
-that a new protein shares structure 
-\ifnotcutting and evolutionary history \fi
-with the proteins used to train the model.
-The computation finds the most likely path through the hidden Markov model.
-To~make best use of floating-point arithmetic, the software computes
-the \emph{logarithm} of the probability of each path, by summing
-the logs of the 
-probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
-The~path that maximizes the
-log of the probability is the most likely path.
-
-The computation is specified on the left-hand side of \figref{viterbi}.
+\caption{Viterbi's equations transformed to work with scores}
+\figlabel{viterbi-transformed}
+\end{figure}
+
+The computation is specified in \figref{viterbi-original}.
 A~probability $V_j^M(i)$ represents the probability of the most
@@
 which operates on scores.
-The transformed equations are
-shown on the right-hand side of \figref{viterbi}.
+The~transformed equations are
+shown in \figref{viterbi-transformed}.
 They minimize the
@@
 \smallverbatiminput{tprob-tprobs}
-These scores are read by
+The fields of this record are read by
 function \texttt{aScore}, 
-whose specification is
+which is given states $s$~and~$\hat s$ and a node index; its %% fixes a bad line break
+specification is
 \mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.
@@
 For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
-right of \figref{viterbi}, we define
+of \figref{viterbi-transformed}, we define
 \mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
@@
 \mono{\mbox{fmap (Mat `cons`)}}?
-This call performs a computation \emph{not} shown in \figref{viterbi}:
+This call performs a computation \emph{not} shown
+in \figrefpair[or]{viterbi-original}{viterbi-transformed} :
 \mrfy\ computes not only the {probability} of the most likely path but 
@@
 Luke Palmer's
-\path{Data.MemoCombinators} package.
+package
+\path{Data.MemoCombinators}.
 The value
@@
 Because the memoized version so faithfully resembles the equations in
-\figref{viterbi}, we~retired the classic version.
+\figref{viterbi-transformed}, we~retired the classic version.
 
@@
 proteins.
-MRFy detects homologous proteins that include hydrogen-bonded beta
+MRFy's \emph{raison d'\^etre} is to detect homologous proteins that
+include hydrogen-bonded beta 
 strands; using prior methods, many instances of this problem are
-intractable.
+intractable \cite{daniels:thesis}.
 
@@
 
+\begin{figure}
+\ifpdfmadness
+\centerline{\includegraphics[width=11cm]{mrf_interleave_diagram.pdf}} 
+\else
+\centerline{\includegraphics[width=11cm]{mrf_interleave_diagram.eps}} 
+\fi
+Each shaded node represents a beta-strand position.\\
+Nodes connected by dashed edges are hydrogen-bonded.
+
+\smallskip
+
+\caption{A Markov random field with two beta-strand pairs}
+\figlabel{mrf} 
+\end{figure}
+
 The new equations are accompanied by a new model.
@@
 
-%
-%  placement at top left column is not ideal, but right column results
-%  in a bad page break elsewhere
-%
-\begin{figure}
-\ifpdfmadness
-\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.pdf}} 
-\else
-\centerline{\includegraphics[width=7.8cm]{mrf_interleave_diagram.eps}} 
-\fi
-Each shaded node represents a beta-strand position.
-Nodes connected by dashed edges are hydrogen-bonded.
-
-\caption{A Markov random field with two beta-strand pairs}
-\label{mrf} 
-\end{figure}
-
 We~describe \mrfy's search abstractly: \mrfy\ computes a sequence
@@
 good scores.
+
+\begin{figure}
+\tabcolsep=0.5cm
+\centerline{%
+\begin{tabular}{cc}
+\loadpdf[width=5cm]{search-figs/search-graph1}&
+\loadpdf[width=5cm]{search-figs/search-graph5}\\
+\begin{minipage}{4cm}
+\caption{Stochastic search with useful and useless steps}%
+\figlabel{search-steps}
+\end{minipage}%
+&
+\begin{minipage}{4cm}
+\caption{Filtering out useless steps could cause nontermination}%
+\figlabel{filtered-search-steps}
+\end{minipage}\\
+\end{tabular}%
+}
+\end{figure}
 
@@
 \caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
-\label{fig:speedup}
+\figlabel{speedup}
 \end{figure}
@@
 This work was funded in part by NIH grant 1R01GM080330.
+
+\section{Notes on the new paper}
+
+\input{jfpnotes}
+
+\subsection{Software deltas}
+
+\subsubsection{Reducing cases in Viterbi}
+
+Original Viterbi took a state and two indices as argument, and it
+dealt with impermissible transitions by assigning a very large score
+(which translates to a very small probability).
+Indices 1, 0, and $-1$ were all used to indicate special conditions,
+and 
+the function had 25~cases overall.
+
+\newcommand\pv{\textit{Predecessor Viterbi}}
+\newcommand\ov{{Original Viterbi}}
+
+Our first step in adapting the code was to create \pv.
+Like \ov, \pv\ takes a state and two indices as an argument.
+But at top level, there is no case analysis on the state, and the only
+indices considered are zero and nonzero, so there are only four cases.
+(The state is used in two ways: to compute transition probabilities,
+and to decide which states may permissibly precede the current state.
+The set of permissible preceding states is looked up in a data
+structure, not computed by case analysis.)
+
+Here is the general case in the new code:
+\smallverbatiminput{hoviterbi}
+THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
+WE SHOULD PROBABLY FIX THAT.
+In the old code, there are three corresponding cases: one each for
+match, insert, and delete states.  
+Here is the code for the match case:
+\smallverbatiminput{viterbi}
+The other case are similar; the second argument to \texttt{aScore} is
+always equal to the first (state) argument to \texttt{vee'},
+and the argument to \texttt{map from} is different in each case.
+(In the new code, this argument is identical to \texttt{preceders !
+stateRight}.)
+NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(
+
+The main benefit is that the Viterbi function is reduced from 90~lines
+of code to just 40~lines.
+This change laid the groundwork for more experiments.
+
+BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
 

### 2013-05-17 jfp.tex
work on draft

@@
 where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
-labels: 
+labels:\label{sec:tprobs}
 \smallverbatiminput{tprob-tprobs}
+\remark{Need to explain those last two fields. HMMer?}
 The fields of this record are read by
@@
 
+\subsection{Competing with C++; less FORTRAN; more Haskell}
+
+EXPLAIN SECTION TITLE.
+
+WHY IS THIS SECTION RELEVANT?  TO SHOW TRAPS AND PITFALLS FOR NEW
+HASKELL PROGRAMMERS.
+
+FIRST STEP WAS TO SIMPLIFY THE CODE.
+
+As shown in \figrefpair{viterbi-original}{viterbi-transformed} and in
+the code for \verb+vee'+ on page~\pageref{code:vee-prime}, Viterbi's
+$V'$ is indexed by a residue index~$i$, a~node index~$j$,
+and
+a state $M$, $I$, or~$D$.
+A~query sequence is simply an array of~\texttt{AA} indexed by~$i$,
+and a model is an array of nodes indexed by~$j$.
+The node array and its indices are handled as in our group's C++ code:
+all nodes have the same representation, and every node can be
+interpreted in one of the special states Begin and End.
+(This possible interpretation causes the introduction of the
+mysterious probabilities \verb+b_m+ and \verb+m_e+ into the transition
+table on page~\pageref{sec:tprobs}.)
+These conventions of representation lead to a full implementation of
+Viterbi's algorithm that is more reminiscent of FORTRAN than of
+Haskell:
+\begin{itemize}
+\item
+Recursive computations are done on smaller array indices, not smaller
+models.
+\item
+Array indices 1, 0, and $-1$ are all used to indicate special
+conditions.
+\item
+Senseless combinations of array indices and states are handled by
+assigning each such combination the largest representable score, which
+corresponds to the smallest representable probability.
+\end{itemize}
+This is the code in which \mrfy\ spends 98\%\ of its time, and that
+we~wanted to speed up to make it more competitive with~C++.
+But the central function \verb+vee'+ has 25~cases overall.
+Before thinking seriously about how to tune it, we~decided to simplify~it.
+
+Our first step in simplification was to start thinking like Haskell
+programmers and make more effective use of both the static type system:
+(AND THE PATTERN-MATCH COMPILER)
+\begin{itemize}
+\item
+We introduced a \texttt{Model} type that consists of a single Begin
+node, an~unbounded sequence of Middle nodes, and a single End node.
+\item
+We~observed that the main algorithm where \mrfy\ spends all its time
+uses only Match, Insert, and Delete states.
+We~redefined the \texttt{StateLabel} type to include only these
+states; the Begin and End labels are now part of
+type \texttt{FullStateLabel}. 
+\end{itemize}
+As a result of these changes, in the revised code,
+\begin{itemize}
+\item
+Only array index~0 is used to indicate a special
+conditions: the transition from the leftmost middle node to the unique
+begin node.
+\item
+All well-typed combinations of array indices and states make sense;
+senseless combinations are ruled out by the static type system.
+\end{itemize}
+
+\begin{figure}
+\def\vsum#1#2{#1+#2}
+\[
+V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
+  + \min\limits_s \left\{
+       \vsum{V_{j-1}^{\prime s}(i - 1)} {a^{\prime}_{s \,\hat
+       s}}, \mbox{ where $s$ can precede $\hat s$} \right \}
+\]
+
+\caption{Viterbi's transformed equations, abstracted over states $s$
+and $\hat s$, defining $e^{\prime}_D(x) = 0$}
+\figlabel{viterbi-abstracted}
+\end{figure}
+
+We also
+
+The right-hand sideOur 
+Indices 1, 0, and $-1$ were all used to indicate special conditions,
+and
+
+\subsubsection{Reducing cases in Viterbi}
+
+\newcommand\pv{\textit{Predecessor Viterbi}}
+\newcommand\ov{{Original Viterbi}}
+
+Our first step in adapting the code was to create \pv.
+Like \ov, \pv\ takes a state and two indices as an argument.
+But at top level, there is no case analysis on the state, and the only
+indices considered are zero and nonzero, so there are only four cases.
+(The state is used in two ways: to compute transition probabilities,
+and to decide which states may permissibly precede the current state.
+The set of permissible preceding states is looked up in a data
+structure, not computed by case analysis.)
+
+Here is the general case in the new code:
+\smallverbatiminput{hoviterbi}
+THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
+WE SHOULD PROBABLY FIX THAT.
+In the old code, there are three corresponding cases: one each for
+match, insert, and delete states.  
+Here is the code for the match case:
+\smallverbatiminput{viterbi}
+The other case are similar; the second argument to \texttt{aScore} is
+always equal to the first (state) argument to \texttt{vee'},
+and the argument to \texttt{map from} is different in each case.
+(In the new code, this argument is identical to \texttt{preceders !
+stateRight}.)
+NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(
+
+The main benefit is that the Viterbi function is reduced from 90~lines
+of code to just 40~lines.
+This change laid the groundwork for more experiments.
+
+BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
+
 \section{Our previous experience compared}
@@
 
+NEW PARAGRAPH: STEP SLOWLY AWAY FROM YOUR CODE, AND SEE IF YOU THINK
+IT RESEMBLES ANYTHING A HASKELL PROGRAMMER WOULD HAVE WRITTEN.
+
 \seclabel{penumbra}
@@
 \input{jfpnotes}
-
-\subsection{Software deltas}
-
-\subsubsection{Reducing cases in Viterbi}
-
-Original Viterbi took a state and two indices as argument, and it
-dealt with impermissible transitions by assigning a very large score
-(which translates to a very small probability).
-Indices 1, 0, and $-1$ were all used to indicate special conditions,
-and 
-the function had 25~cases overall.
-
-\newcommand\pv{\textit{Predecessor Viterbi}}
-\newcommand\ov{{Original Viterbi}}
-
-Our first step in adapting the code was to create \pv.
-Like \ov, \pv\ takes a state and two indices as an argument.
-But at top level, there is no case analysis on the state, and the only
-indices considered are zero and nonzero, so there are only four cases.
-(The state is used in two ways: to compute transition probabilities,
-and to decide which states may permissibly precede the current state.
-The set of permissible preceding states is looked up in a data
-structure, not computed by case analysis.)
-
-Here is the general case in the new code:
-\smallverbatiminput{hoviterbi}
-THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
-WE SHOULD PROBABLY FIX THAT.
-In the old code, there are three corresponding cases: one each for
-match, insert, and delete states.  
-Here is the code for the match case:
-\smallverbatiminput{viterbi}
-The other case are similar; the second argument to \texttt{aScore} is
-always equal to the first (state) argument to \texttt{vee'},
-and the argument to \texttt{map from} is different in each case.
-(In the new code, this argument is identical to \texttt{preceders !
-stateRight}.)
-NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(
-
-The main benefit is that the Viterbi function is reduced from 90~lines
-of code to just 40~lines.
-This change laid the groundwork for more experiments.
-
-BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
 

### 2013-05-25 jfp.tex
line edites sections 1-3

@@
 
-\newcommand\txprobj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
-\newcommand\txprobjj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
+\newcommand\txprobj[3][]{a#1_{{#2}_{j-1}{#3}_j}} % probability entering j
+\newcommand\txprobxj[3][]{a#1_{{#2}_{j}{#3}_{j+1}}}  % probability exiting j
+\newcommand\txprobjj[3][]{a#1_{{#2}_{j}{#3}_j}} % probability within j
 \newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
@@
 trying to simplify and speed up the central algorithm.
+AND FIX THE CODE AND...
 
@@
 has lost an amino acid by deletion from column~$j$.
+A~deletion state contains no tables of probabilities.
 \hfuzz=1.5pt % fix blot
@@
 The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.
-
-A hidden Markov model also has distinguished ``begin'' and ``end''
-states.
+The names of the states are easily confused;
+one~way to keep them straight is to imagine that the model of the
+alignment describes a prototype sequence of length~$\alignwidth$,
+and that each state describes an action used to build the query
+sequence from the prototype: 
+match a prototypical amino~acid,
+insert an unexpected amino~acid,
+or
+delete a prototypical amino~acid.
+
+A~full hidden Markov model also has distinguished ``begin'' and ``end''
+states; begin and end states are like match states, but they
+correspond to no amino acids and contain no emission probabilities.
 %%
@@
 %%
-In~our representation, each state contains a probability or a table of
-probabilities, and it is also labeled with one of these labels:
+In~our representation, each state is referred to by its position~$j$
+within the alignment and by one of these labels:
 \smallverbatiminput{statelabel}
-
-\begin{figure} 
-\ifpdfmadness
-\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
-\else
-\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
-\fi
-
-\caption{A hidden Markov model $(\alignwidth = 4)$}
-
-\figlabel{plan7}
-\end{figure}
-
-We~use the ``Plan7'' hidden Markov model, which forbids direct
-transitions between insertion 
-states and deletion states \cite{Eddy:1998ut}. 
-\seclabel{plan7}%
-\remark{Reference to \figref{plan7}??}%
-\ifpagetuning
-``Plan7'' implies that
-\else
-The model gets its name because
-\fi
- there are exactly~7 possible transitions
-into the states of any column~$j$.
-Each transition has its own probability:
-\begin{itemize} 
-\item
-A~transition into a match state 
-is more likely when column~$j$ is a consensus column.
-Depending on the predecessor state, 
-the probability of such a transition is 
-$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
-\item
-A~transition into a deletion state 
-is more likely when column~$j$ is a non-consensus column.
-The probability of such a transition is 
-$\txprobj M D$~or~$\txprobj D D$.
-\item
-A~transition into an insertion state 
-is more likely when column~$j$ is a non-consensus column.
-The probability of such a transition is 
-$\txprobjj M I$~or~$\txprobjj I I$.
-\end{itemize}
-
+We~group states that share a column~$j$; these states are referred
+to collectively as ``node~$j$,'' as shown in \figref{plan7}.
 This model
@@
 deletion state~$D$.
+\remark{We need to figure out if the begin state is really node 0}
+
+\begin{figure} 
+\ifpdfmadness
+\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
+\else
+\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
+\fi
+
+\caption{A hidden Markov model $(\alignwidth = 4)$}
+
+\figlabel{plan7}
+\end{figure}
+
+The model in \figref{plan7} is from a well-established family
+that permits
+transitions between match states and all other adjacent states,
+but forbids
+direct
+transitions between insertion 
+states and deletion states  \cite{Eddy:1998ut}. 
+The family is called the ``Plan7'' family because
+every model in the family has 7~transitions per node.
+Each transition has its own probability; the probabilities for
+transitions leaving states in node~$j$ are
+$\txprobjj I I$,
+$\txprobxj I M$,
+$\txprobjj M I$,
+$\txprobxj M M$,
+$\txprobxj M D$,
+$\txprobxj D M$,
+and
+$\txprobxj D D$.
+When column $j+1$ is a consensus column, the transitions that enter
+its match state 
+($\txprobxj I M$,
+$\txprobxj M M$,
+$\txprobxj D M$)
+are more likely.
+When column $j+1$ is a non-consensus column, the transitions that enter
+its insertion and deletion states are more likely.
 
@@
 that a new protein shares structure 
-\ifnotcutting and evolutionary history \fi
 with the proteins used to train the model.
-The computation finds the most likely path through the hidden Markov model.
-To~make best use of floating-point arithmetic, the software computes
-the \emph{logarithm} of the probability of each path, by~summing
-the logs of the 
-probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
-The~path that maximizes the
-log of the probability is the most likely path.
+The computation finds the most likely matching path through the hidden Markov model.
+In~any matching path, each match state and insertion state corresponds to a
+unique amino acid in the query sequence.
+The probability of the path is the product of all of the corresponding
+emission probabilities, together with the product of the transition
+probabilities along the edges of the path.
+
+Probabilities are small, and multiplying small numbers makes smaller
+numbers.
+Using floating-point arithmetic, multiplying probabilities amounts to
+storing data in the exponent of a floating-point number, which holds
+only 11~bits.
+In~order to store data in the significand, which holds 53~bits,
+we~instead sum the logarithms of probabilities \cite{Viterbi:1967hq}.
+Moreover, in~order to work with positive numbers, 
+our legacy file format also \emph{negates}
+the logarithm of each probability, making it a positive number.
+%%To~make these numbers more pleasant to read, we negate them,
+%%which eliminates hordes of minus signs from our file formats.
+The negated logarithm of a probability is called a \emph{score}.
+\smallverbatiminput{score}
+Type \texttt{Score} has a limited \texttt{Num} instance which permits
+scores to be added and subtracted but not multiplied.
 
@@
 
-\[
-\def\maxiquad{\hskip 1.2em\relax}
-%\mskip -5mu
-\begin{array}{@{}l@{}c@{}l}
-V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
-  \begin{array}{l@{}c@{}l}
-  \vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
-  \vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
-  \vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
-  \end{array} \right.\\[\goo]
-V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
-  \begin{array}{l@{}c@{}l}
-  \vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
-  \vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
-  \end{array} \right.\\[\gum]
-V_{j}^{D}(i) &=& \max \left\{
-  \begin{array}{l@{}c@{}l}
-  \vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
-  \vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
-  \end{array} \right.\\
-\end{array}
-\]
-
-\caption{Viterbi's original equations}
-\figlabel{viterbi-original}
-
-\bigskip
+%%  \[
+%%  \def\maxiquad{\hskip 1.2em\relax}
+%%  %\mskip -5mu
+%%  \begin{array}{@{}l@{}c@{}l}
+%%  V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
+%%    \begin{array}{l@{}c@{}l}
+%%    \vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
+%%    \vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
+%%    \vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
+%%    \end{array} \right.\\[\goo]
+%%  V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
+%%    \begin{array}{l@{}c@{}l}
+%%    \vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
+%%    \vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
+%%    \end{array} \right.\\[\gum]
+%%  V_{j}^{D}(i) &=& \max \left\{
+%%    \begin{array}{l@{}c@{}l}
+%%    \vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
+%%    \vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
+%%    \end{array} \right.\\
+%%  \end{array}
+%%  \]
+%%  
+%%  \caption{Viterbi's original equations}
+%%  \figlabel{viterbi-original}
+%%  
+%%  
+%%  \bigskip
 
@@
 
-\caption{Viterbi's equations transformed to work with scores}
+\caption{Viterbi's equations expressed using scores}
 \figlabel{viterbi-transformed}
@@
 
-The computation is specified in \figref{viterbi-original}.
-A~probability $V_j^M(i)$ represents the probability of the most
+The~most likely path through a model is the one with the
+smallest score.
+The~score of each path is computed by dynamic programming; the score associated with
+a path to a state~$s$ ($M$, $I$, or~$D$) at node~$j$ and position~$i$ in
+the query sequence is written $V_j^{\prime s}(i)$.
+For~example, 
+score~$V_j^{\prime M}(i)$ represents the probability of the most
 likely path of the first~$i$ amino acids in the query sequence,
 terminating with placement of amino acid $x_i$ in state~$M_j$.
-Probabilities $V_j^I(i)$ and $V_j^D(i)$ are similar.
-The
-equations are explained very clearly by
+Scores $V_j^{\prime I}(i)$~and~$V_j^{\prime D}(i)$ are similar.
+Scores are specified by Viterbi's equations, which are shown
+in \figref{viterbi-transformed}. 
+The~equations (expressed using probabilities) are explained very clearly by
 \citet[Chapter~5]{Durbin:1998wz}
@@
 %%    insertion emission vectors.}
-
-Our legacy file format also \emph{negates}
-the log of each probability, making it a positive number.
-%%To~make these numbers more pleasant to read, we negate them,
-%%which eliminates hordes of minus signs from our file formats.
-The negated logarithm of a probability is called a \emph{score}.
-\smallverbatiminput{score}
-Type \texttt{Score} has a limited \texttt{Num} instance which permits
-scores to be added and subtracted but not multiplied.
 
@@
   + \min\limits_s \left\{
-       \vsum{V_{j-1}^{\prime s}(i - 1)} {a^{\prime}_{s \,\hat
+       \vsum{V_{\mathit{pnode}\;\hat s\;j}^{\prime
+       s}(\mathit{presidue}\;\hat s\;i)} {a^{\prime}_{s \,\hat
        s}}, \mbox{ where $s$ can precede $\hat s$} \right \}
@@
 
-We also
+We also combined Viterbi's three equations into one.
+Each equation has the same form, but is parameterize over three
+values determined by the right-hand state~$\hat s$:
+the~index of the previous node ($j$~or~$j-1$),
+the~index of the previous residue ($i$~or~$i-1$),
+and the set of previous states~$s$ over which the minimum is taken.
+If~we abstract the index computations into functions
+$\mathit{pnode}$ 
+and 
+$\mathit{presidue}$
+respectively,
+we~get the combined, abstract form of Viterbi's equations shown in
+\figref{viterbi-abstracted}.
+The~implementation of \figref{viterbi-abstracted} looks like this:
+\smallverbatiminput{hoviterbi}
+Function \texttt{next} makes the recursive call to
 
@@
 Here is the general case in the new code:
-\smallverbatiminput{hoviterbi}
+
 THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.

### 2013-05-25 jfp.tex
tinking with end of sectino 3 intro

@@
 $\txprobxj D D$.
+(The transitions from match to insertion and from insertion to
+insertion leave states in 
+node~$j$ but also enter states in node~$j$.
+The other transitions enter states in node~$j+1$.)
 When column $j+1$ is a consensus column, the transitions that enter
 its match state 
-($\txprobxj I M$,
-$\txprobxj M M$,
-$\txprobxj D M$)
+%($\txprobxj I M$,
+%$\txprobxj M M$,
+%$\txprobxj D M$)
 are more likely.

### 2013-05-27 jfp.tex
line edits

@@
 %%
-In~our representation, each state is referred to by its position~$j$
-within the alignment and by one of these labels:
+In~our representation, 
+the begin and end states are known by their positions:
+in a $\alignwidth$-column model, the begin state is the only state in
+column~0 and the end state is the only state in column~$\alignwidth$.
+Every other state is referred to by its column~$j$
+and by one of these labels:
 \smallverbatiminput{statelabel}
@@
 \begin{figure} 
-\ifpdfmadness
-\centerline{\includegraphics[width=8cm]{Plan7.pdf}} 
-\else
-\centerline{\includegraphics[width=8cm]{Plan7.eps}} 
-\fi
+\centerline{\loadpdf[width=11cm]{Plan7_emit}}
 
@@
 we~instead sum the logarithms of probabilities \cite{Viterbi:1967hq}.
-Moreover, in~order to work with positive numbers, 
-our legacy file format also \emph{negates}
+Moreover, 
+ our models use a
+legacy file format which
+\emph{negates}
 the logarithm of each probability, making it a positive number.
@@
 \end{smallverbatim}
-But our models use a
-legacy file format in which each amino acid is a small integer
+But in our
+legacy file format, each amino acid is a small integer
 used only to index arrays.
@@
 
-In a real hidden Markov model, each probability is represented as a score.
-Our code implements a transformed version of Viterbi's equations
-which operates on scores.
-The~transformed equations are
-shown in \figref{viterbi-transformed}.
-They minimize the
-score (the negated log probability) for each combination of
-column~$j$, amino 
-acid~$x_i$, and state $M_j$, $I_j$, or~$D_j$.
-
-A~model is represented as a sequence of \emph{nodes};
-node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
-as well as the probabilities of transitions out of that node.
-Each node contains tables of emission scores
-$e'_{M_j}$~and~$e'_{I_j}$.
-These tables are read by function
+A~model is represented as a sequence of \emph{nodes}, 
+each of which includes three states:
+\smallverbatiminput{model3-node}
+The representation of a state~$s$ includes the scores of transitions leaving
+that state;
+if~state~$s$ is a match or insertion state, its representation also
+includes a table of emission scores.
+For~example, the representation of a match state is \texttt{MState}:
+\smallverbatiminput{model3-mstate}
+%
+The transition scores, which are shown as $a_{s\hat s}$
+in \figref{plan7}, are read by  
+function \texttt{aScore}, 
+%which is given states $s$~and~$\hat s$ and a node index~$j$; its
+whose
+specification is
+\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.
+The
+emission table is read by 
 \texttt{eScore}, whose specification
 is % parallel structure for line breaking below
-\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
-%
-We place the transition probabilities 
-into a record in which each field is labeled
-\texttt{$s$\_$\mskip 1mu\hat s$},
-where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
-labels:\label{sec:tprobs}
-\smallverbatiminput{tprob-tprobs}
-\remark{Need to explain those last two fields. HMMer?}
-The fields of this record are read by
-function \texttt{aScore}, 
-which is given states $s$~and~$\hat s$ and a node index; its %% fixes a bad line break
-specification is
-\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.
+\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$},
+where $x_i$ is the amino acid located at position~$i$ in the query
+sequence. 
+%
+%%  We place the transition probabilities 
+%%  into a record in which each field is labeled
+%%  \texttt{$s$\_$\mskip 1mu\hat s$},
+%%  where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
+%%  labels:\label{sec:tprobs}
+%%  \smallverbatiminput{tprob-tprobs}
+%%  \remark{Need to explain those last two fields. HMMer?}
 
@@
 
-Scores can be usefully attached to many types of values,
-so we have defined a small abstraction:
+All scores computed by \mrfy\ are sums of these transition and
+emission scores.
+Because these sums can be usefully attached to many types of values,
+we~have defined a small abstraction:
 \smallfuzzverbatiminput{10.8pt}{vscore}
@@
  an~``\texttt a'' with a score written on the side.
+\remark{Lazy in the first arg is about 1\%~slower}
 The \texttt{/+/} function adds to the score without touching the
@@
 Like a recursive Fibonacci function, Viterbi's functions,
-when implemented \naive ly,
+if~implemented \naive ly,
 take exponential time.
@@
 This call performs a computation \emph{not} shown
-in \figrefpair[or]{viterbi-original}{viterbi-transformed} :
+in \figref{viterbi-transformed}:
 \mrfy\ computes not only the {probability} of the most likely path but 
@@
 
-Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+.
-Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,
+Function~\verb+vee'+ makes a recursive call not to itself but
+to~\verb+vee''+. 
+Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+;
+it~produces the same result as~\verb+vee'+,
 but faster: 
@@
 The~combined model, an~example of which is shown in \figref{mrf}, is called a
-\textit{Markov random field}.
+\textit{Markov random~field}.
 
@@
 \begin{tabular}{cc}
-\loadpdf[width=5cm]{search-figs/search-graph1}&
-\loadpdf[width=5cm]{search-figs/search-graph5}\\
+\loadpdf[scale=0.35]{search-figs/search-graph1}&
+\loadpdf[scale=0.35]{search-figs/search-graph5}\\
 \begin{minipage}{4cm}
-\caption{Stochastic search with useful and useless steps}%
-\figlabel{search-steps}
+\caption{Stochastic search with useful and useless moves}%
+\figlabel{search-moves}
 \end{minipage}%
@@
 \begin{minipage}{4cm}
-\caption{Filtering out useless steps could cause nontermination}%
-\figlabel{filtered-search-steps}
+\caption{Filtering out useless moves could cause nontermination}%
+\figlabel{filtered-search-moves}
 \end{minipage}\\
@@
 previous point.
-Ideally, \mrfy\ would search by composing a \emph{generator}
+The~search is depicted graphically in \figref{search-moves}.
+The~solid lines represent useful moves; the dotted lines represent
+useless moves.
+The~points are laid out vertically in the order in which they are
+discovered;
+if~you like, time increases as you move downward.
+Horizontally, we~have organized the points into five columns; 
+the bottom point in each column is the result of a useful move, and we
+call it a useful point.
+Any points above it in the same column are useless.
+
+If~\mrfy\ could apply Hughes's technique, it~would search by composing
+a \emph{generator} 
 that produces an infinite sequence of moves, 
@@
 produce any values, and the search would diverge.
-
-We~address this problem by combining ``generate and filter'' into a
+This situation is depicted graphically in
+\figref{filtered-search-moves}:
+on the left-hand side, 
+the first move is useless; the second move is useful; and all
+subsequent moves are useless.
+Filtering out useless moves would produce the sequence shown on the
+right-hand side, which contains two points followed by a divergent
+computation.
+
+We~avoid the possibility of divergence by combining ``generate and
+filter'' into a 
 single abstraction, which has type \mono{SearchGen pt r}.

### 2013-05-27 hoviterbi.tex
code updates

vee' stateRight j i =
  internal [ child score state (vee'' state pj pi)
           | state <- preceders stateRight
           , let score = transition node state stateRight
                         + emission node state aa
           ]
  where pi = prevres  stateRight i
        pj = prevnode stateRight j
        node = middle pj
        aa = rs `rat` i
       vee' stateRight nodes@(node:ntail) aas =
         internal [ child score state (vee'' state (nextNodes stateRight)
                                         (nextAAs   stateRight))
        | state <- preceders stateRight
        , hasAA stateRight aas
        , let score = transition node state stateRight
                      + emission node state (head aas)
        ]
         where -- Del does not consume a residue
     -- Ins does not consume a node
     nextNodes Ins = nodes
     nextNodes _   = ntail
     nextAAs Del   = aas
     nextAAs _     = tail aas
     hasAA Del _   = True
     hasAA _  aas  = not (null aas)

### 2013-05-27 everygen.tex
paper changes

@@
 everyPt :: RandomGen r
-        => SearchGen pt r -> CCost -> Scored pt
-        -> Rand r [CCosted (Utility (Scored pt))]
-everyPt sg cost startPt = do
-  successors <- mapM (nextPt sg) (repeat startPt)
-  tagged <- zipWithM costedUtility successors [succ cost..]
-  let (useless, CCosted (Useful newPt) newCost : _) =
-                      span (isUseless . unCCosted) tagged
-  (++) (CCosted (Useful startPt) cost : useless) <$>
-                                  everyPt sg newCost newPt
- where costedUtility pt cost =
-         utility sg move >>= \u -> return $ CCosted u cost
-        where move = Move { older = startPt, younger = pt
-                          , youngerCCost = cost }
+        => SearchGen pt r -> CCosted (Scored pt)
+        -> Rand r [Utility (CCosted (Scored pt))]
+everyPt sg (CCosted cost startPt) = do
+  successors <- zipWithM costedUtility [succ cost..] =<<
+                mapM (nextPt sg) (repeat startPt)
+  let (useless, Useful pt : _) = span isUseless successors
+  (++) (Useful (CCosted cost startPt) : useless) <$> everyPt sg pt
+ where
+   costedUtility cost pt = return . fmap (CCosted cost) =<< utility sg move
+     where move = Move { older = startPt, younger = pt, youngerCCost = cost }

### 2013-05-27 hov-prevs.tex
paper changes

prevs preceders stateHat fj fi =
 internal [ edge score state (vee'' state pj pi)
          | state <- preceders
          , let pi = fi state
          , pj >= 0, pi >= 0
          , let score = transition (node pj) state stateHat
                        + emission (node pj) state (residue pi)
          ]
 where pj = fj stateHat

predUnless :: forall a . Enum a => a -> StateLabel -> StateLabel -> a
predUnless n don't_move s = if s == don't_move then n else pred n

### 2013-05-27 hov4.tex
paper changes

vee' Ins (NC 0) i = intoInsZero i
vee' Mat (NC 1) i = intoMatOne i
vee' Del (NC 1) i = intoDelOne i
vee' stateHat j i =
  internal [ edge score state (vee'' state pj pi)
           | state <- preceders stateHat
           , let pi = case state of { Del -> i ; _ -> pred i }
           , pj >= 0, pi >= 0
           , let score = transition (node pj) state stateHat
                         + emission (node pj) state (residue pi)
           ]
  where pj = case stateHat of { Ins -> j ; _ -> pred j }

### 2013-05-27 jfp.tex
paper changes

@@
 % \timestampfalse % it's post-submission time
+
+\newcommand\vfilbreak[1]{\vskip 0pt plus #1 \penalty-200 \vskip 0pt plus -#1 }
 
@@
 A~model is represented as a sequence of \emph{nodes}, 
-each of which includes three states:
+each of which includes three states:\label{code:model3-node}
 \smallverbatiminput{model3-node}
@@
 is computed by function \texttt{avSum}, in which
-the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
+the terms are computed by \texttt{aScore} and \texttt{vee''},
+respectively:\label{code:vee-prime} 
 \smallfuzzverbatiminput{2.9pt}{vfix}
@@
 
+Ideally, MRFy would use the now-classic, lazy, modular
+technique advocated by \citet{hughes:why}, in which one function
+computes an infinite sequence of points, and another function
+uses a finite prefix to decide on an approximation.
+But because MRFy's search is stochastic, 
+making MRFy's search modular is not so easy.
+
+To illustrate the difficulties, we~discuss our simplest search:
+random hill climbing.
+%The space of beta-strand placements has no notion of slope or
+%gradient,
+From any given point in the search space, this search moves a random
+distance in a random 
+direction.
+If~the move leads to a better point, 
+we call it \emph{useful};
+otherwise it is \emph{useless}.
+\smallverbatiminput{utility}
+(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
+With luck, an~infinite sequence of useful moves converges
+at a local optimum.
+
+\mrfy's search path follows only useful moves;
+if a move is useless, 
+\mrfy\ abandons it and moves again (in a new random direction) from the
+previous point.
+The~search is depicted graphically in \figref{search-moves}.
+The~solid lines represent useful moves; the dotted lines represent
+useless moves.
+The~points are laid out vertically in the order in which they are
+discovered;
+if~you like, time increases as you move downward.
+Horizontally, we~have organized the points into five columns; 
+the bottom point in each column is the result of a useful move, and we
+call it a useful point.
+Any points above it in the same column are useless.
+
 \begin{figure}
@@
 
-Ideally, MRFy would use the now-classic, lazy, modular
-technique advocated by \citet{hughes:why}, in which one function
-computes an infinite sequence of points, and another function
-uses a finite prefix to decide on an approximation.
-But because MRFy's search is stochastic, 
-making MRFy's search modular is not so easy.
-
-To illustrate the difficulties, we~discuss our simplest search:
-random hill climbing.
-%The space of beta-strand placements has no notion of slope or
-%gradient,
-From any given point in the search space, this search moves a random
-distance in a random 
-direction.
-If~the move leads to a better point, 
-we call it \emph{useful};
-otherwise it is \emph{useless}.
-\smallverbatiminput{utility}
-(We~also use \texttt{Useful} and \texttt{Useless} to tag \emph{points}.)
-With luck, an~infinite sequence of useful moves converges
-at a local optimum.
-
-\mrfy's search path follows only useful moves;
-if a move is useless, 
-\mrfy\ abandons it and moves again (in a new random direction) from the
-previous point.
-The~search is depicted graphically in \figref{search-moves}.
-The~solid lines represent useful moves; the dotted lines represent
-useless moves.
-The~points are laid out vertically in the order in which they are
-discovered;
-if~you like, time increases as you move downward.
-Horizontally, we~have organized the points into five columns; 
-the bottom point in each column is the result of a useful move, and we
-call it a useful point.
-Any points above it in the same column are useless.
-
 If~\mrfy\ could apply Hughes's technique, it~would search by composing
-a \emph{generator} 
+a~\emph{generator} 
 that produces an infinite sequence of moves, 
-a \emph{filter} that selects the useful moves, and 
-a \emph{test} function that enumerates finitely many useful moves
+a~\emph{filter} that selects the useful moves, and 
+a~\emph{test} function that enumerates finitely many useful moves
 and returns the final destination.
-But a generator may produce an infinite sequence of useless moves.
+But~a~generator may produce an infinite sequence of useless moves.
 (For example, if \mrfy\ should stumble upon a global optimum, every move from
 that point would be useless.)
-Given an infinite sequence of useless inputs, a filter would not
+Given an infinite sequence of useless inputs, a~filter would not
 produce any values, and the search would diverge.
@@
 and
-in~simulated annealing, for example, as the search ages,
+in~simulated annealing, for example, as the cumulative cost increases,
 the \texttt{utility} function becomes less likely to accept a move
 that worsens the score.
-
-\noindent
-\vbox{
+To~attach a cumulative cost to a point, we use type \texttt{CCosted}:
+\smallverbatiminput{aged}
+
 Using these pieces, function \texttt{everyPt} produces an infinite
-sequence containing a mix of useful and useless moves:
-\smallfuzzverbatiminput{10.8pt}{everygen}
-}
-\noindent
+sequence containing a mix of useful and useless points.
+Each useful point is tagged with cumulated cost and its score:
+\smallfuzzverbatiminput{6pt}{everygen}
 Both \texttt{nextPt} and \texttt{utility} are monadic, but 
@@
 tag each one with a cumulative cost and a utility.
-
-In Hughes's \citeyearpar{hughes:why} classic model of lazy search,
-we~might expect to use an infinite sequence of points obtained by
-applying \texttt{nextPt} to many points in succession, as done by
+Because we want to compose monadic computations in the same order we
+compose ordinary functions, 
+our code uses the monadic composition
+operator~\texttt{=<<}, which is the bind operator with its arguments
+swapped.
+
+If \mrfy\ used Hughes's classic model of lazy search,
+we~could possibly expect to see an infinite sequence of points obtained by
+applying \texttt{nextPt} to many points in succession, as~done by
 the Prelude function \texttt{iterate}.
-But~if you look carefully at how \texttt{successors} is computed,
-that's not what's happening:
-instead, we~compute an infinite list by applying \texttt{nextPt} to
-a~\emph{single} initial point using infinitely many distinct sources
-of random variation.
-\remark{This is the place to bring in a figure, I~think. ---NR}
-Then, when we find a useful \texttt{newPt}, a~recursive call
-to \texttt{everyPt} computes a \emph{fresh} infinite sequence of
-single jumps starting at~\texttt{newPt}.
-\mrfy's search may explore an arbitrary number of infinite
-lists of points, and the representation we~found easiest to think
-about is the pair of elements
-\texttt{pt0}~and~\texttt{nextPt} in the \texttt{SearchGen} record.
-
-Once the successors have been computed and tagged,
-\texttt{span} finds the first useful successor.
-In~case there \emph{is} no successor, \texttt{everyPt} also returns
-all the useless successors.
-If~we do find a useful successor, we start searching anew from that
-point, with a recursive call to \texttt{everyPt}.
-(Because \texttt{everyPt} is monadic, the points accumulated so far are
-appended to its result using the \verb+<$>+~operator.)
+But~that's not how \mrfy\ works;
+instead, 
+the~infinite list \texttt{successors}
+is computed by applying \texttt{nextPt} to
+the \emph{single} initial point \texttt{startPt}, \mbox{using} infinitely
+many distinct sources of random variation.
+To~understand the structure of
+ \mrfy's more complex search, it~may help to
+relate the code to \figref{search-moves}.
+Please assume that \texttt{startPt} is the point in the upper left
+corner of \figref{search-moves}.
+Then the first part of the \texttt{successors} list appears in the
+next column of \figref{search-moves}:
+the~first four elements are \texttt{useless};
+the next element, which is the first useful point, is~\texttt{pt};
+and~the remaining elements, which are never evaluated,
+match the
+underscore.
+The first useful point is located by \mono{span~isUseless}.
+After the search reaches~\texttt{pt},
+we~start the search anew with a recursive 
+call to \mono{everypt~sg}, 
+passing~\texttt{pt}.
 %
@@
 is followed by a (possibly infinite, possibly empty) list of useless
-points, and then continues recursively with another call to \texttt{everyPt}.
+points, and then continues with the results of a recursive call
+to \texttt{everyPt}.
+
+\mrfy's search may explore an arbitrary number of infinite
+lists of points, each reached using a different source of randomness.
+We~suppose this search space could be represented by a single
+value of type \mbox{\mono{Rand r [[Scored pt]]}}, but when writing
+client code, we~found it much easier to think about the
+representation we actually use in the
+\texttt{SearchGen} record:
+the~monadic value
+\texttt{pt0} together with the function \texttt{nextPt}.
 
@@
 %% \makeatletter\@sectionaboveskip=0.9\@sectionaboveskip\makeatother
+
+\vfilbreak{2.5\baselineskip}
+
 The rest of the search uses Hughes's
 classic composition of generator  and test function.
-Because our code is monadic, we~use the monadic composition
-operator~\texttt{=<<}, which is the bind operator with its arguments
-swapped: 
 \smallverbatiminput{search}
@@
 \texttt{parmap rseq} for \texttt{map}.
-\figref{fig:speedup} shows speedups when using from 
+\figref{speedup} shows speedups when using from 
  1~to~48 of the cores 
@@
 % And we expect that if we cross the street, we will get to the other side.
+
+\subsection{Improving \mrfy's sequential performance}
+
+Research software for
+computational biology doesn't have to be fast to be useful.
+But~when we say that \mrfy\ solves some problems 8~times slower then
+SMURF, we~felt it wasn't a good showing for Haskell.
+\mrfy\ spends 98\%\ of its time running Viterbi's algorithm, and
+we~felt sure that with some love and attention, we could make it
+more competitive with~C++.
+
+The code example for \verb+vee'+ on page~\pageref{code:vee-prime}
+shows the computation of the general case of~$V^{\prime M}_j(i)$ from 
+\figref{viterbi-transformed}.
+But the full story is not nearly as nice at one could hope for:
+in~the code that accompanies our 2012 ICFP paper,
+\verb+vee'+ has 25~cases overall.
+Many of these cases exist because we~did not have enough insight into
+Haskell, and we made some decisions that might have been more
+appropriate to a FORTRAN program:
+\begin{itemize}
+\item
+Recursive computations were done on smaller array indices, not smaller
+models.
+\item
+Array indices 1, 0, and $-1$ were all used to indicate special
+conditions.
+\item
+Senseless combinations of array indices and states were handled by
+assigning each such combination the largest representable score, which
+corresponds to the smallest representable probability.
+\end{itemize}
+Before trying to speed up the code, we~simplified~it.
+
+Our first step in simplification was to start thinking like Haskell
+programmers and use types:
+\begin{itemize}
+\item
+We~refactored the single, flat \texttt{Node} type from our ICFP paper
+into the three state types you see on page~\pageref{code:model3-node}.
+\item 
+We introduced a \texttt{Model} type that consists of a special Begin
+node containing a begin and an insertion state, an~unbounded sequence
+of ordinary \texttt{Nodes}, and an end state.
+\item
+We~observed that the main algorithm where \mrfy\ spends all its time
+uses only match, insertion, and deletion states, so we removed the
+value constructors corresponding to begin and end states.
+Those constructors were used only to prettyprint models, so~we made
+them part of a larger type.
+\end{itemize}
+As a result of these changes, in the revised code,
+\begin{itemize}
+\item
+Only array index~0 is used to indicate a special
+condition: the transition from the leftmost middle node to the unique
+begin node.
+\item
+All well-typed combinations of array indices and states make sense;
+senseless combinations are ruled out by the static type system.
+\end{itemize}
+
+\def\p#1{\mathit{p#1}}
+\begin{figure}
+\def\vsum#1#2{#1+#2}
+\[
+\begin{array}{r@{}c@{}l}
+\multicolumn3{c}{
+  V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
+    + \min\limits_s \big\{
+         \vsum{V_{\mathit{\p j}}^{\prime s}(\p i\; s)} 
+     {a^{\prime}_{s_{\p j} \,\hat s_j}}, 
+      \mbox{ where $s$ can precede $\hat s$}  \big\}
+}
+\\[10pt]
+\mbox{where }\p i\; s&{} = {}& 
+  \left\{ \begin{array}{l}
+          i \mbox{, if $s=D$}\\
+          i-1 \mbox{, otherwise}\\
+          \end{array}
+  \right.
+\\[10pt]
+\p j &{} = {}& 
+  \left\{ \begin{array}{l}
+          j \mbox{, if $\hat s=I$}\\
+          j-1 \mbox{, otherwise}\\
+          \end{array}
+  \right.
+\\
+\end{array}
+\]
+
+\caption{Viterbi's equations, abstracted over state~$\hat s$, 
+defining $e^{\prime}_{D_j}(x) = 0$}
+\figlabel{viterbi-abstracted}
+\end{figure}
+
+Our next step was to use more abstraction.
+In~Viterbi's equations,
+$V'$ is indexed by a residue index~$i$,\footnote
+{In~biology, when an amino acid enters a protein, it often loses a few
+atoms, and so within the protein, the amino acid is often called
+a \emph{residue}.}
+ a~node index~$j$,
+and
+a state $M$, $I$, or~$D$.
+Each state uses a slightly different equation, but it is possible to
+combine them.
+\figref{viterbi-abstracted} shows a single equation which, when
+state $\hat s$ is specialized to state $M$, $I$, or~$D$, recovers the
+original three equations from \figref{viterbi-transformed}.
+The figure sums the cost of emitting an amino acid at state~$\hat s$
+lowest of the costs of the paths entering~$\hat s$ from predecessor
+state~$s$.
+Index~$\p j$ is the index of the node containing the predecessor
+state, and $\p i; s$ is the index of the amino acid, if~any, to be
+emitted at state~$s$.
+
+NEED TO EXPLAIN PRECEDERS
+
+\figref{viterbi-abstracted} doesn't deal with the begin state.
+In~our legacy file format, and also in our code, the begin state is
+represented as a match state inside an ordinary node.
+In~a perfect world, then, we would add one case to say that the match
+state with $i=j=0$ is reachable at zero cost, and we'd be finished.
+But the world is imperfect, because although the begin state is
+represented as a match state, unlike a match state, the begin state
+does not emit.
+We~therefore need a special case for each of the three states that can
+follow the begin state.
+Here is the code:
+\smallverbatiminput{hov4}
+NEED TO EXPLAIN INTERNAL, EDGE, AND LEAF.
+NEED TO EXPLAIN [[transition]] and [[emission]]
+
+\smallverbatiminput{hov-prevs}
+
+The main benefit is that the Viterbi function is reduced from 90~lines
+of code to just 38~lines, of which 8~lines are type signatures.
+This change laid the groundwork for more experiments.
+
+If~we look not at the whole package but just at Viterbi's algorithm,
+the~performance story is even more grim:
+we~wrote a custom C~program that computes a score using Viterbi's
+algorithm and does nothing else, and it runs about a hundred times
+faster than \mrfy's implementation.
+
+BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
+
+THE STORY OF THE UNPACK PRAGMAS; CITING JOHAN TIBELL
 
@@
 
-\subsection{Competing with C++; less FORTRAN; more Haskell}
-
-EXPLAIN SECTION TITLE.
-
-WHY IS THIS SECTION RELEVANT?  TO SHOW TRAPS AND PITFALLS FOR NEW
-HASKELL PROGRAMMERS.
-
-FIRST STEP WAS TO SIMPLIFY THE CODE.
-
-As shown in \figrefpair{viterbi-original}{viterbi-transformed} and in
-the code for \verb+vee'+ on page~\pageref{code:vee-prime}, Viterbi's
-$V'$ is indexed by a residue index~$i$, a~node index~$j$,
-and
-a state $M$, $I$, or~$D$.
-A~query sequence is simply an array of~\texttt{AA} indexed by~$i$,
-and a model is an array of nodes indexed by~$j$.
-The node array and its indices are handled as in our group's C++ code:
-all nodes have the same representation, and every node can be
-interpreted in one of the special states Begin and End.
-(This possible interpretation causes the introduction of the
-mysterious probabilities \verb+b_m+ and \verb+m_e+ into the transition
-table on page~\pageref{sec:tprobs}.)
-These conventions of representation lead to a full implementation of
-Viterbi's algorithm that is more reminiscent of FORTRAN than of
-Haskell:
-\begin{itemize}
-\item
-Recursive computations are done on smaller array indices, not smaller
-models.
-\item
-Array indices 1, 0, and $-1$ are all used to indicate special
-conditions.
-\item
-Senseless combinations of array indices and states are handled by
-assigning each such combination the largest representable score, which
-corresponds to the smallest representable probability.
-\end{itemize}
-This is the code in which \mrfy\ spends 98\%\ of its time, and that
-we~wanted to speed up to make it more competitive with~C++.
-But the central function \verb+vee'+ has 25~cases overall.
-Before thinking seriously about how to tune it, we~decided to simplify~it.
-
-Our first step in simplification was to start thinking like Haskell
-programmers and make more effective use of both the static type system:
-(AND THE PATTERN-MATCH COMPILER)
-\begin{itemize}
-\item
-We introduced a \texttt{Model} type that consists of a single Begin
-node, an~unbounded sequence of Middle nodes, and a single End node.
-\item
-We~observed that the main algorithm where \mrfy\ spends all its time
-uses only Match, Insert, and Delete states.
-We~redefined the \texttt{StateLabel} type to include only these
-states; the Begin and End labels are now part of
-type \texttt{FullStateLabel}. 
-\end{itemize}
-As a result of these changes, in the revised code,
-\begin{itemize}
-\item
-Only array index~0 is used to indicate a special
-conditions: the transition from the leftmost middle node to the unique
-begin node.
-\item
-All well-typed combinations of array indices and states make sense;
-senseless combinations are ruled out by the static type system.
-\end{itemize}
-
-\begin{figure}
-\def\vsum#1#2{#1+#2}
-\[
-V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
-  + \min\limits_s \left\{
-       \vsum{V_{\mathit{pnode}\;\hat s\;j}^{\prime
-       s}(\mathit{presidue}\;\hat s\;i)} {a^{\prime}_{s \,\hat
-       s}}, \mbox{ where $s$ can precede $\hat s$} \right \}
-\]
-
-\caption{Viterbi's transformed equations, abstracted over states $s$
-and $\hat s$, defining $e^{\prime}_D(x) = 0$}
-\figlabel{viterbi-abstracted}
-\end{figure}
-
-We also combined Viterbi's three equations into one.
-Each equation has the same form, but is parameterize over three
-values determined by the right-hand state~$\hat s$:
-the~index of the previous node ($j$~or~$j-1$),
-the~index of the previous residue ($i$~or~$i-1$),
-and the set of previous states~$s$ over which the minimum is taken.
-If~we abstract the index computations into functions
-$\mathit{pnode}$ 
-and 
-$\mathit{presidue}$
-respectively,
-we~get the combined, abstract form of Viterbi's equations shown in
-\figref{viterbi-abstracted}.
-The~implementation of \figref{viterbi-abstracted} looks like this:
-\smallverbatiminput{hoviterbi}
-Function \texttt{next} makes the recursive call to
-
-The right-hand sideOur 
-Indices 1, 0, and $-1$ were all used to indicate special conditions,
-and
-
-\subsubsection{Reducing cases in Viterbi}
-
-\newcommand\pv{\textit{Predecessor Viterbi}}
-\newcommand\ov{{Original Viterbi}}
-
-Our first step in adapting the code was to create \pv.
-Like \ov, \pv\ takes a state and two indices as an argument.
-But at top level, there is no case analysis on the state, and the only
-indices considered are zero and nonzero, so there are only four cases.
-(The state is used in two ways: to compute transition probabilities,
-and to decide which states may permissibly precede the current state.
-The set of permissible preceding states is looked up in a data
-structure, not computed by case analysis.)
-
-Here is the general case in the new code:
-
-THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
-WE SHOULD PROBABLY FIX THAT.
-In the old code, there are three corresponding cases: one each for
-match, insert, and delete states.  
-Here is the code for the match case:
-\smallverbatiminput{viterbi}
-The other case are similar; the second argument to \texttt{aScore} is
-always equal to the first (state) argument to \texttt{vee'},
-and the argument to \texttt{map from} is different in each case.
-(In the new code, this argument is identical to \texttt{preceders !
-stateRight}.)
-NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(
-
-The main benefit is that the Viterbi function is reduced from 90~lines
-of code to just 40~lines.
-This change laid the groundwork for more experiments.
-
-BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
-
 \section{Our previous experience compared}
@@
 
-\section{Notes on the new paper}
+\section{THIS SECTION WILL BE DELETED.  SAUVE QUI PEUT!}
 

### 2013-05-27 jfp.tex
fix dangling \ref

@@
 and each column groups amino acids that play similar roles in
-different proteins (Figure~\ref{alignment}).
+different proteins (\figref{alignment}).
 

### 2013-05-27 hov4.tex
build instructions

@@
-vee' Ins (NC 0) i = intoInsZero i
-vee' Mat (NC 1) i = intoMatOne i
-vee' Del (NC 1) i = intoDelOne i
+vee' :: StateLabel -> NodeIndex -> ResidueCount -> a
+vee' Ins (NI 0) i = intoInsZero i
+vee' Mat (NI 1) i = intoMatOne i
+vee' Del (NI 1) i = intoDelOne i
 vee' stateHat j i =

### 2013-05-27 jfp.tex
build instructions

@@
 of ordinary \texttt{Nodes}, and an end state.
+And we represented the sequence as a list, not an array.
 \item
@@
 uses only match, insertion, and deletion states, so we removed the
-value constructors corresponding to begin and end states.
-Those constructors were used only to prettyprint models, so~we made
-them part of a larger type.
+\texttt{StateLabel} value constructors corresponding to begin and end
+states, leaving only the three constructors shown on
+page~\ref{code:state-label}. 
+(We~moved the begin and end labels to another type, which is used only
+to prettyprint models.) 
 \end{itemize}
-As a result of these changes, in the revised code,
-\begin{itemize}
-\item
-Only array index~0 is used to indicate a special
-condition: the transition from the leftmost middle node to the unique
-begin node.
-\item
-All well-typed combinations of array indices and states make sense;
-senseless combinations are ruled out by the static type system.
-\end{itemize}
+The immediate effect of these changes was to reduce case analysis on
+two dimensions:
+instead of five state labels, we got three;
+and instead of three special array indices, we got just one special
+list (the empty list).
+
+\begin{figure}
+\noindent\hspace*{1.5cm}\begin{minipage}{9cm}
+\smallverbatiminput{list-viterbi}
+\smallskip
+\end{minipage}
+
+\caption{Sketch of Viterbi's algorithm, using lists}
+\figlabel{list-viterbi}
+\end{figure}
+
+Our~next step was to use more abstraction---in~particular, to abstract
+over state labels.
+However, after abstracting over state labels, we~abandoned lists and
+went back to arrays.
+An~excerpt from our list-based code appears in \figref{list-viterbi}.
+The~problem is defining \verb+vee''+ as a memoized version
+of~\verb+vee'+.
+Although we experimented with specialized memo combinators that work
+with list prefixes, 
+we~were never confident that we could easily, efficiently, and
+correctly memoize~\verb+vee'+.
+Instead, we~wrote a new version, still using integer arithmetic, but
+guides by the idea $i$~is a \emph{count} of the number
+of amino acids in list~\texttt{aas}, but $j$~is the \emph{index}
+of~\texttt{node}.
+(We~also got
+an
+insight we got from
+the \texttt{next} function: Viterbi's algorithm is a form of minimum edit
+distance.
+As~we were trying to improve the code, this insight helped keep us on track.)
+Armed with this new, better way of thinking about $i$~and~$j$,
+we~turned back to arrays.
 
@@
 
-Our next step was to use more abstraction.
+Our final code, like the code in \figref{list-viterbi}, is abstracted
+over state labels.
+We~now turn your attention to that abstraction.
 In~Viterbi's equations,
-$V'$ is indexed by a residue index~$i$,\footnote
+$V'$ is indexed by a residue count~$i$,\footnote
 {In~biology, when an amino acid enters a protein, it often loses a few
@@
 
-NEED TO EXPLAIN PRECEDERS
+To~implement \figref{viterbi-abstracted}, we~define several
+auxiliary functions.
+The~set of states that can precede~$\hat s$
+is \mono{preceders~$\hat s$}; $\p i$~and~$\p j$ are computed using
+function \texttt{predUnless}:
+\smallverbatiminput{v4aux}
+We~also use \texttt{newtype} to define 
+a value constructor~\texttt{RC} for a residue count~$i$
+and
+a value constructor~\texttt{NI} for a node index~$j$.
+We~can then deliver the more abstract version of Viterbi's algorithm
+shown in \figref{viterbi-hov4}.
+
+\begin{figure}
+\noindent\hspace*{1.0cm}\begin{minipage}{11cm}
+\smallverbatiminput{hov4}
+\smallskip
+\end{minipage}
+
+\caption{Viterbi's algorithm abstracted over~$\hat s$}
+\figlabel{viterbi-hov4}
+\end{figure}
+
+The version of \verb+vee'+ shown in  \figref{viterbi-hov4} is abstract
+in another way: the type of the result is the unknown type~\texttt{a}.
+Type~\texttt{a} represents
 
@@
 follow the begin state.
+
 Here is the code:
-\smallverbatiminput{hov4}
 NEED TO EXPLAIN INTERNAL, EDGE, AND LEAF.
 NEED TO EXPLAIN [[transition]] and [[emission]]
+
+As a result of these changes, in the revised code,
+\begin{itemize}
+\item
+Only array index~0 is used to indicate a special
+condition: the transition from the leftmost middle node to the unique
+begin node.
+\item
+\end{itemize}
+
+\begin{verbatim}
+### 2013-05-27 01:23   Running med(vpasses=1): icfp-branch mrfy nr-jfp
+icfp-branch faab8dc       54.07s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
+mrfy d5604bc*             44.90s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
+nr-jfp 00ef54e*           47.59s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
+\end{verbatim}
 

### 2013-05-28 hov-prevs.tex
build instructions
more code snippets

@@
-prevs preceders stateHat fj fi =
+       prevs predecessors stateHat fj fi =
         internal [ edge score state (vee'' state pj pi)
-          | state <- preceders
+                 | state <- predecessors
                  , let pi = fi state
@@
         where pj = fj stateHat
-
-predUnless :: forall a . Enum a => a -> StateLabel -> StateLabel -> a
-predUnless n don't_move s = if s == don't_move then n else pred n
+prevs :: [StateLabel] -> StateLabel
+      -> (StateLabel -> NodeIndex) -> (StateLabel -> ResidueCount) -> a

### 2013-05-28 hov4.tex
more code snippets

--- 2013-05-27 hov4.tex
@@
-vee' Ins (NC 0) i = intoInsZero i
-vee' Mat (NC 1) i = intoMatOne i
-vee' Del (NC 1) i = intoDelOne i
+vee' :: StateLabel -> NodeIndex -> ResidueCount -> a
+vee' Del (NI 1) (RC 0) = leaf (transition (node 0) Mat Del)
+vee' Ins (NI 0) (RC 0) = leaf (transition (node 0) Mat Ins)
+vee' Mat (NI 1) (RC 0) = leaf (transition (node 0) Mat Mat)
 vee' stateHat j i = 
@@
            | state <- preceders stateHat
-           , let pi = case state of { Del -> i ; _ -> pred i }
+           , let pi = predUnless i Del state
            , pj >= 0, pi >= 0
@@
            ]
-  where pj = case stateHat of { Ins -> j ; _ -> pred j }
+  where pj = predUnless j Ins stateHat

### 2013-05-28 jfp.tex

@@
 \newcommand\secref[1]{Section~\ref{sec:#1}}
+\newcommand\secrefpair[3][and]{Sections \ref{sec:#2}~#1~\ref{sec:#3}}
 \newcommand\seclabel[1]{\label{sec:#1}}
@@
 improving
-performance (\secref{awkward-profiling}).
+performance (\secrefpair{improving-viterbi}{awkward-profiling}).
 %%  \item
@@
  (\secref{penumbra}).
-\item
-\emph{[POTENTIAL BULLETS ABOUT SPEED AND ``MORE FUNCTIONAL'']}
 \end{itemize}
@@
 an \emph{Experience Report} in ICFP~2012.
-This~paper adds to the original paper by describing our experience in
-trying to simplify and speed up the central algorithm.
-AND FIX THE CODE AND...
+This~paper adds a new technical section (\secref{improving-viterbi})
+which describes our experience 
+refactoring and improving our implementation of Viterbi's 
+\citeyearpar{Viterbi:1967hq}
+algorithm.
+We~have also simplified the stochastic search as well as other elements of
+the code.
+Finally, we~have improved and added to the figures and diagrams.
 
@@
 Every other state is referred to by its column~$j$
-and by one of these labels:
+and by one of these labels:\label{code:state-label}
 \smallverbatiminput{statelabel}
@@
  an~``\texttt a'' with a score written on the side.
-\remark{Lazy in the first arg is about 1\%~slower}
+%% \remark{Lazy in the first arg is about 1\%~slower}
 The \texttt{/+/} function adds to the score without touching the
@@
 \begin{figure}%%%[h!]
-\centerline{\includegraphics{speedup}}
+\centerline{\loadpdf[]{speedup}}
 \medskip
@@
 \subsection{Improving \mrfy's sequential performance}
+
+\seclabel{better-viterbi}
+\seclabel{improving-viterbi}
 
@@
 Before trying to speed up the code, we~simplified~it.
-
-Our first step in simplification was to start thinking like Haskell
+%
+%
+Our first step was to start thinking like Haskell
 programmers and use types:
@@
 states, leaving only the three constructors shown on
-page~\ref{code:state-label}. 
+page~\pageref{code:state-label}. 
 (We~moved the begin and end labels to another type, which is used only
@@
 The version of \verb+vee'+ shown in  \figref{viterbi-hov4} is abstract
-in another way: the type of the result is the unknown type~\texttt{a}.
-Type~\texttt{a} represents
-
-\figref{viterbi-abstracted} doesn't deal with the begin state.
+in another way: the type of the result is an unknown type~\texttt{a}.
+Type~\texttt{a} represents a tree-node in an abstract ``cost~tree;''
+we~instantiate it as a score, as a \mono{Scored~[StateLabel]},
+and as an explicit tree used to visualize and debug the computation.
+Values of type~\texttt{a} are built using three functions that are
+passed as parameters:
+\smallverbatiminput{hosigs}
+The \texttt{internal} function must treat an empty list of edges as an
+unreachable tree-node; for~example, if \texttt{a} is instantiated
+to \texttt{Score}, then \mono{internal~[]} must produce an infinite
+score (corresponding to zero probability).
+
+\figref{viterbi-hov4} is the first example in this paper that
+shows \emph{all} the cases for~\verb+vee'+.
+We~have reduced the original 25~cases to just~4.
+The~general case, which was shown in other examples, appears last.
+The three special cases above it handle the three states that can
+follow the begin state.
 In~our legacy file format, and also in our code, the begin state is
-represented as a match state inside an ordinary node.
-In~a perfect world, then, we would add one case to say that the match
-state with $i=j=0$ is reachable at zero cost, and we'd be finished.
-But the world is imperfect, because although the begin state is
-represented as a match state, unlike a match state, the begin state
-does not emit.
-We~therefore need a special case for each of the three states that can
-follow the begin state.
-
-Here is the code:
-NEED TO EXPLAIN INTERNAL, EDGE, AND LEAF.
-NEED TO EXPLAIN [[transition]] and [[emission]]
-
-As a result of these changes, in the revised code,
+identified with the match state in node~$j=0$.
+But although it is
+identified with a match state, the begin state does not behave like a
+match state: it~does not emit a residue.
+Therefore the three states that can follow it can be entered only when the
+count of residues emitted is~$i=0$.
+
+Using the code in \figref{viterbi-hov4}, plus the auxiliary functions
+shown above, the implementation of Viterbi's algorithm is just
+25~lines of code, of which 2~lines are type signatures.
+That's four times smaller than our original code.
+But~while we've gained something, we've also lost something---it's not
+at all obvious that the code in \figref{viterbi-hov4}  implements
+Viterbi's equations.
+
+Although the~new code is about 3\%~slower than our original code, 
+it~turned out to a good starting point for experiments.
+Of~the myriad variations we tried, 
+the best involved two transformations:
 \begin{itemize}
 \item
-Only array index~0 is used to indicate a special
-condition: the transition from the leftmost middle node to the unique
-begin node.
-\item
+We~abstracted the call to \texttt{internal} over the state~$\hat s$,
+the list of preceding states, and functions that
+compute \texttt{pi}~and~\texttt{pj}.
+\item
+We~then added a fourth case to \verb+vee'+, which handles the match
+state in node~1 when the residue count is nonzero.
 \end{itemize}
-
-\begin{verbatim}
-### 2013-05-27 01:23   Running med(vpasses=1): icfp-branch mrfy nr-jfp
-icfp-branch faab8dc       54.07s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
-mrfy d5604bc*             44.90s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
-nr-jfp 00ef54e*           47.59s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
-\end{verbatim}
-
-\smallverbatiminput{hov-prevs}
-
-The main benefit is that the Viterbi function is reduced from 90~lines
-of code to just 38~lines, of which 8~lines are type signatures.
+For reasons we don't understand, GHC specializes the resulting code
+very effectively, leading to 30\%~faster performance on our
+UNCHARACTERIZED MEDIUM benchmark.
+The~result is 26\%~faster than our original code.
+
 This change laid the groundwork for more experiments.
+The most interesting of these experiments was to follow the advice of
+\citet{tibell:high-performance} and use GHC's UNPACK pragma on scalar
+values---in our case, scores.
+Using the UNPACK pragma made our original code 7\%~faster but our new
+code 9\%~slower.
+
+We~then turned to GHC's profiler, running Viterbi's algorithm on a
+model of 343~nodes and a query of 592~residues.
+We~expect the memo table to have about 600,000 entries,
+and the profiler says that the program allocates 531MB, or 7.85KB
+per~entry.
+This seems excessive.
+\begin{itemize}
+\item
+One culprit is the function \verb+vee''+, which is charged with
+allocating about 66MB.
+To~estimate the amount of memory consumed by
+an entry, we~imagine that representing a boxed
+score on the heap might require three 32-bit words.
+A~thunk involving three parameters might be a bit like a four-element
+list, taking perhaps 12~words for the cons cells and another 8~words
+for the values?
+And then the pointer in the memo table itself would occupy a word.
+So~24~words per entry times 4~bytes per word would be 96B per
+entry for a total of about 59MB?
+Because the actual memo table involves arrays of arrays of arrays,
+we~won't quibble about a discrepancy of~10\%.
+\item
+The computation of \texttt{score} (the sum of transition 
+and emission scores) is charged with allocating 45MB total, or about
+4.5~words for each transition in the computation tree.
+We~have no idea why this combination is allocating, and various
+attempts to make the sum strict (e.g., by using a bang pattern in 
+\mono{let~!score = }\ldots) either have no effect or actually make
+performance worse.
+\item
+There is also some significant allocation in the parser, but the
+elephant in the room is the function \verb+vee'+ itself, which
+allocates 258MB, which is half the memory allocated by the entire
+program.
+We~have no idea what is happening.
+A~graphical heap profile shows a large amount of memory initially
+charged to \verb+vee''+ (the memo table), which then shrinks
+gradually, presumably as large thunks are replaced with smaller
+scores.
+The~amount of memory charged to \verb+vee'+
+increases roughly linearly with time, 
+so~perhaps we have a space leak, but even given the various profiling
+tools available, we~are not competent to find~it.
+\end{itemize}
+In~principle, we~could have diagnosed our problems by reading 
+GHC Core output from the compiler.
+We~found guidance in Don Stewart's invaluable blog posts,
+and we also used his \texttt{ghc-core} tool.
+But it wasn't enough.
+Even though one of us has worked on several optimizing compilers, 
+we~just don't have enough experience to use GHC Core to diagnose
+allocation problems in a 25-line function.
+
+%%\begin{verbatim}
+%%### 2013-05-27 01:23   Running med(vpasses=1): icfp-branch mrfy nr-jfp
+%%icfp-branch faab8dc       54.07s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
+%%mrfy d5604bc*             44.90s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
+%%nr-jfp 00ef54e*           47.59s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
+%%\end{verbatim}
+
+%%\smallverbatiminput{hov-prevs}
+
+We~succeeded in making Viterbi's algorithm a lot simpler and somewhat
+faster, but our original goal was to make it competetive with~C.
+The measurements using SMURF are depressing, and the SMURF code is
+even more depressing, so we reimplement Viterbi's algorithm from
+scratch in a clean C~program.
+To~avoid writing a parser, we extend \mrfy\ with code that reads a
+model and a query sequence and emits a C~file containing that model
+and query as initialized data.
 
@@
 faster than \mrfy's implementation.
-
-BENCHMARK RESULTS WERE WILDLY INCONSISTENT.
-
-THE STORY OF THE UNPACK PRAGMAS; CITING JOHAN TIBELL
-
-\subsection{Awkward debugging and testing}
+Ironically, we've~had a lot of trouble getting the code right, but it
+allocates only one double-precision floating-point number for each
+entry in the memo table; Haskell allocates about 30~times more.
+
+It's depressing to think that a language with a mature, optimizing,
+native-code compiler might not be usable for performance-critical
+work, but the difficulty of diagnosing allocation problems is a real
+obstacle.
+We~were lucky that we could get new results (CITE NOAH'S LATEST PAPER
+HERE?) even though our core algorithm is 100~times slower than the
+same algorithm written in~C.
+You~might not be so lucky.
+
+\subsection{Debugging and testing: awkward, then  less awkward}
 
@@
 really worked.
-But~we also encountered obstacles that prevented functional
+But~as discussed above, profiling didn't really work for~us.
+And we also encountered other obstacles that prevented functional
 programming from working as well as we would have liked.
-The~most significant obstacles were in debugging and testing.
+The~most significant such obstacles were in debugging and testing.
 
@@
 format are unclear.)
-But using Haskell, we~found the errors hard to diagnose.
+But~using Haskell, we~found the errors hard to diagnose.
 Calls to
@@
 Haskell is getting a stack trace on error,
-and Simon Marlow said that he may have figured out how to track call
-stacks properly in a lazy functional language.
-We~can't wait.
+and when GHC~7.4 was released in early~2012, it~included support for
+stack traces. 
+We~have used the stack-trace feature a couple of times, and it helps.
 
@@
 The examples and tutorials we found focused 
-predominantly on writing and testing properties using data types that
+on writing and testing properties using data types that
 already implemented class \texttt{Arbitrary}.
@@
 informative. 
+(To~be fair to ourselves, the \texttt{shrink} method is barely
+mentioned in the original 
+paper on QuickCheck.
+It~shows up under another name, as~an improvement made by Andy Gill.)
 \end{itemize}
@@
 The work got done only when the whole team pitched in to
-meet the deadlines for this paper.
-
-At the last minute, QuickCheck did find
-a bug in our implementation of Viterbi's algorithm: we had omitted
-the score for the transition from the final node of the hidden
-Markov model to the special ``end'' state.
-Without QuickCheck, we~probably wouldn't have known anything was wrong.
-%it is unlikely that we would have found this bug,
-%and if we had, determining the cause would have required much time.
+meet deadlines for ICFP~2012.
+
+After ICFP, 
+QuickCheck played a central role in further developments.
+Most important was the testing of the variations on Viterbi's
+algorithm described in \secref{better-viterbi}.
+We~used QuickCheck not only to shrink sequences of nodes and of amino
+acids, but also to shrink the number of nonzero transmission and
+emission probabilities in the models.
+We~were able to confirm that each new implementation computed the same
+paths with the same score as our original implementation.
+Almost all counterexamples were shrunk to include just two nodes 
+(the smallest model supported by our original implementation)
+and two or three amino acids.
+Moreover, the models contained just a handful of nonzero
+probabilities.
+QuickCheck played an invaluable role in helping us get edge cases
+right, and in confirming that most of the cases in the original code
+could be eliminated.
 
@@
 for medium-sized projects,
-but his contributions to MRFy have been limited, mostly to \emph{post hoc}
-refactoring and testing.
+but he has played a secondary role, primarily refactoring and testing
+others' code.
 
@@
 
-We~had quite some difficulty profiling, but we hope that
-this difficulty may be mitigated by new profiling tools released early
-in~2012 with
+Using GHC versions through~7.2, 
+we~had quite some difficulty profiling.
+These difficulties were technical in nature, and they have been
+completely mitigated by the new profiling tools released 
+with
 GHC~7.4.
-GHC assigns costs to ``cost centers'' \cite{sansom-pj},
-and
-in GHC~7.0, which we used 
-for most of MRFy's development, 
-cost-center annotations had to be added manually
-to nested functions.
-Although these annotations
-made our code so ugly that we felt compelled to
-remove them, they did enable us to improve the performance of~\verb+vee'+ as
-discussed in \secref{perf}.
-%We~miss valgrind, whose profiler gives useful information
-%without \emph{any} annotations.
-GHC~7.4 provides 
-more sophisticated profiling tools, which we look forward to using.
-Difficulties using Cabal to enable profiling of installed libraries
-may remain.
+The~most useful new feature is the automatic annotation of ``cost
+centers'' \cite{sansom-pj}. 
+Some technical difficulties remain, particularly with installation:
+the only reliable way we have found to enable compilation with
+profiling is to destroy all existing compiled libraries, 
+update our Cabal configuration to ask for libraries to be compiled
+with profiling, and reinstall all needed Cabal packages.
+As~of early 2013, the process of reinstallation, although not
+difficult, is time-consuming and tedious.
 
@@
 Ideally we would use QuickCheck to find bugs, but as we mention in
-\secref{awkward-quickcheck}, we found obstacles.
+\secref{awkward-quickcheck}, there are obstacles.
 We~have now overcome these obstacles, but
 we~sorely regret not doing so earlier.
-In~light of our experience, we will institute a new programming practice:
+In~light of our experience, we~have instituted a new programming practice:
 whenever we introduce a new data type,
-we~will write the instance of
+we~write the instance of
 \texttt{Arbitrary} right away, while
@@
 \mono{fmap (all p . shrink) arbitrary}.
+We~cannot overemphasize the value of QuickCheck!
 
@@
 Implementing Viterbi's equations in Haskell was pure joy.
+And boiling down the code to just four cases, with the help of
+QuickCheck, was also tremendous~fun.
 Writing an iterative search in purely functional style was~easy.
@@
 
-NEW PARAGRAPH: STEP SLOWLY AWAY FROM YOUR CODE, AND SEE IF YOU THINK
-IT RESEMBLES ANYTHING A HASKELL PROGRAMMER WOULD HAVE WRITTEN.
+On~reviewing our work, we~observed that we were a little too ready to
+use types and data structures better suited to C++ programming than to
+Haskell programming.
+Eventually we learned to step slowly away from the code and to seek
+out opportunities for abstraction and simplification.
+The~abstract, simple versions---supported by QuickCheck tests---are a
+lot more likely to be useful to the next generation of graduate students.
 
@@
 John Tibell taught us the \texttt{UNPACK} pragma.
-We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
+For~support and discussions 
+we also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
 Larsen, and Nathan Ricci.
@@
 
-\section{THIS SECTION WILL BE DELETED.  SAUVE QUI PEUT!}
-
-\input{jfpnotes}
+%\section{THIS SECTION WILL BE DELETED.  SAUVE QUI PEUT!}
+
+%\input{jfpnotes}
 

### 2013-05-28 acm-para.tex
eps of search graph 5. two not very good paragraphs on benchmark and acm submission.

A paper for the computational biology community, based on MRFy, has been
submitted to the ACM Conference on Bioinformatics, Computational Biology, and 
Biomedical Informatics.
This paper explores the effectiveness of various search strategies, and also
compares MRFy's accuracy at the task of remote homology detection to several
contemporary approaches.
MRFy outperforms the other software tested.
On these remote homology detection tasks, MRFy spends virtually all of its
time in the Viterbi algorithm, so improvements to this implementation will
measurably improve MRFy's effectiveness.

### 2013-05-28 medium-para.tex
eps of search graph 5. two not very good paragraphs on benchmark and acm submission.

We primarily benchmarked MRFy's run-time performance using a test case that
mimics real-world use of MRFy, but focuses on the implementation of the Viterbi
algorithm.
This ``medium-sized'' benchmark consists of computing the Viterbi score for
each of a set of protein sequences against each of a set of hidden Markov model
templates.
The medium-sized benchmark comprises 56 protein sequences from the deep-sea 
bacterium \emph{Thermotoga maritima},
ranging in length from 80 to approximately 1200 residues, and 9 hidden Markov
models, each built from a multiple structure alignment of the proteins in
a $\beta$-structural protein superfamily.
A single invocation of MRFy results from each hidden Markov model in the 
benchmark, and that invocation aligns each of the 56 protein sequences in the 
input file to that model. Thus, time spent in I/O and parsing of the model
is amortized over all protein sequences.

### 2013-05-29 localsearch-para.tex
2 paras about local search. Missing citation; I've emailed Richard to ask how he'd like it cited.

Richard Senington proposed and prototyped a ``local search'' model of
stochastic search, using his \texttt{local-search} library~\cite{}.
This library did not fit well with our stochastic search abstraction, and
thus we wrote an alternative MRFy implementation using \texttt{local-search}
as well as the majority of the MRFy modules.
While reconciling these two codebases will constitute a tedious task, and we are
not happy with this current state of MRFy, the results were terrific.
The local search approach outperformed random hill-climbing, simulated
annealing, and the genetic algorithm.

Our hasty modifications to MRFy, in order to incorporate local search, led to
good results quickly.
We did not need to change any interfaces in the various MRFy modules, though in
the interim, we have a separate \texttt{main} for local search.
This stands in contrast to our experiences modifying our C and C++-based
software, which inevitably led to headaches.

### 2013-05-29 jfp.tex
this draft now includes everything except the Senington experiment and AG's comments

@@
 
+\renewcommand\bibsection{\section*{\refname}}
+
 \IfFileExists{timestamp.tex}{\input{timestamp}}{}
@@
 the flexibility and rapid prototyping of a scripting
-language, plus the performance of native code.
-In~our experience, higher-order functions, lazy evaluation, and
-monads really worked, but
+language, and it compiles to native code.
+From experience using Haskell to implement new algorithms for
+computational biology, we~report where Haskell really worked for us
+and what obstacles prevented it from working better.
+Higher-order functions, lazy evaluation, and
+monads worked well, but
 profiling and debugging presented obstacles.
-Also, Haskell libraries vary greatly:
+And~because allocation costs are difficult to predict and control,
+compilation to native code did not guarantee great performance.
+Finally, our experience with Haskell libraries varied greatly:
 memoization combinators and parallel-evaluation
@@
 Haskell's ecosystem
-made it easy for us to develop new algorithms in computational
-biology.
+made it easy for us to explore and run new algorithms.
 \end{abstract}
@@
 the code.
-Finally, we~have improved and added to the figures and diagrams.
+Finally, we~have improved and added to explanations, figures, and diagrams.
 
@@
 in amino-acid sequences that are only distantly related.
+
+A~paper for the computational biology community, based on MRFy, has been
+submitted to the ACM Conference on Bioinformatics, Computational Biology, and 
+Biomedical Informatics.
+That paper explores the effectiveness of various search strategies,
+and it also
+compares MRFy with several other software tools.
+MRFy is more accurate than the other software tested.
 MRFy is available at \url{mrfy.cs.tufts.edu}.
@@
 where each $x_i$~is 
-an amino acid.
-The number of amino acids, $N$, can differ from the number of columns
+an amino-acid \emph{residue}.\footnote
+{In~biology, when an amino acid enters a protein, it often loses a few
+atoms, and so within the protein, what's left of the amino acid is called
+a \emph{residue}.
+In~this paper we use the terms ``amino acid'' and ``residue''
+interchangeably.
+}
+The number of residues, $N$, can differ from the number of columns
 in the alignment,~\alignwidth.
@@
 The match state contains a table $e_{M_j}(x)$ which gives the
- probability that a homologous protein has amino acid~$x$ in
+ probability that a homologous protein has residue~$x$ in
  column~$j$.
@@
 The insertion state contains a table $e_{I_j}(x)$ which gives the
-probability that a homologous protein has gained amino acid~$x$ by
+probability that a homologous protein has gained residue~$x$ by
 insertion at column~$j$.
@@
 The deletion state determines the probability that a homologous protein
-has lost an amino acid by deletion from column~$j$.
+has lost a~residue by deletion from column~$j$.
 A~deletion state contains no tables of probabilities.
@@
 sequence from the prototype: 
-match a prototypical amino~acid,
-insert an unexpected amino~acid,
+match a prototypical residue,
+insert an unexpected residue,
 or
-delete a prototypical amino~acid.
-
-A~full hidden Markov model also has distinguished ``begin'' and ``end''
+delete a prototypical residue.
+
+A~hidden Markov model also has distinguished ``begin'' and ``end''
 states; begin and end states are like match states, but they
-correspond to no amino acids and contain no emission probabilities.
+correspond to no residues and contain no emission probabilities.
 %%
@@
 the begin and end states are known by their positions:
-in a $\alignwidth$-column model, the begin state is the only state in
-column~0 and the end state is the only state in column~$\alignwidth$.
+in a $\alignwidth$-column model, the begin state shares
+column~0 with an insertion state, and the end state is the only state
+in column~$\alignwidth$. 
 Every other state is referred to by its column~$j$
@@
 This model
-has begin and end states $B$~and~$E$,
-as well as four
-nodes, each containing 
+has begin and end states $B$~and~$E$;
+an~insertion state in node~0; and
+four
+full nodes, each containing 
 an insertion state~$I$, 
@@
 deletion state~$D$.
-\remark{We need to figure out if the begin state is really node 0}
+(Node~0 may also be viewed as a full node, but its deletion state is
+not reachable from the begin state, so it is shown with only two
+states.)
 
@@
 states and deletion states  \cite{Eddy:1998ut}. 
-The family is called the ``Plan7'' family because
-every model in the family has 7~transitions per node.
+Models in this family are called ``Plan7'' models because
+every full node has 7~transitions.
 Each transition has its own probability; the probabilities for
@@
 The other transitions enter states in node~$j+1$.)
-When column $j+1$ is a consensus column, the transitions that enter
+When column $j+1$ is a consensus column, transitions that enter
 its match state 
@@
 are more likely.
-When column $j+1$ is a non-consensus column, the transitions that enter
+When column $j+1$ is a non-consensus column, transitions that enter
 its insertion and deletion states are more likely.
@@
 In~any matching path, each match state and insertion state corresponds to a
-unique amino acid in the query sequence.
+unique residue in the query sequence.
 The probability of the path is the product of all of the corresponding
@@
 
-Probabilities are small, and multiplying small numbers makes smaller
-numbers.
-Using floating-point arithmetic, multiplying probabilities amounts to
-storing data in the exponent of a floating-point number, which holds
-only 11~bits.
-In~order to store data in the significand, which holds 53~bits,
-we~instead sum the logarithms of probabilities \cite{Viterbi:1967hq}.
-Moreover, 
- our models use a
-legacy file format which
-\emph{negates}
-the logarithm of each probability, making it a positive number.
-%%To~make these numbers more pleasant to read, we negate them,
-%%which eliminates hordes of minus signs from our file formats.
-The negated logarithm of a probability is called a \emph{score}.
+Probabilities are small, and products of small numbers can become very
+small indeed.
+If~we were to multiply probabilities represented as floating-point
+numbers, all the useful information would wind up in exponents,
+and an exponent holds only
+11~bits.
+In~order to store useful information in the significand, which holds 53~bits,
+we~follow standard practice: instead of multiplying probabilities,
+we~add logarithms of probabilities \cite{Viterbi:1967hq}.
+Also following standard practice, we~work with
+\emph{negated}
+logarithms of probabilities, which are  positive numbers.
+The negated logarithm of a probability is called a \emph{score}:
 \smallverbatiminput{score}
-Type \texttt{Score} has a limited \texttt{Num} instance which permits
+We~have given type \texttt{Score} a limited \texttt{Num} instance which permits
 scores to be added and subtracted but not multiplied.
@@
 score~$V_j^{\prime M}(i)$ represents the probability of the most
-likely path of the first~$i$ amino acids in the query sequence,
-terminating with placement of amino acid $x_i$ in state~$M_j$.
+likely path of the first $i$~residues in the query sequence,
+terminating with placement of residue~$x_i$ in state~$M_j$.
 Scores $V_j^{\prime I}(i)$~and~$V_j^{\prime D}(i)$ are similar.
@@
 
-To~be able to use Haskell, we had to reimplement the standard
+To~use Haskell, we had to reimplement the standard
 algorithm for solving Viterbi's equations.
@@
 
-Our code represents a query sequence as an immutable array of amino
-acids.
+Our code represents a query sequence as an immutable array of residues.
 In~idiomatic Haskell, 
-we might represent an individual amino acid~$x_i$
-using a value of algebraic data type:
+we might represent an individual residue~$x_i$
+using a value of algebraic data type with constructors named for
+amino acids:
 \begin{smallverbatim}
@@
 But in our
-legacy file format, each amino acid is a small integer
+legacy file format, each amino acid (residue) is a small integer
 used only to index arrays.
 %which we use only as an array index.
-We therefore chose 
+We therefore chose a representation that is easier to make consistent
+with query sequences as represented on~disk:
 \smallverbatiminput{aa}
@@
 specification is
-\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.
+\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.\seclabel{aScore}
 The
@@
 \mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$},
-where $x_i$ is the amino acid located at position~$i$ in the query
+where $x_i$ is the residue located at position~$i$ in the query
 sequence. 
@@
 it~produces the same result as~\verb+vee'+,
-but faster: 
+but repeated calls return in constant time:
 \smallverbatiminput{memo}
@@
 the probability of finding amino acid~$x_i$ in column~$j$ 
-\ifpagetuning
+\iffalse
 depends on the amino acid~$x'$ in column~${\pairedwith i}$.
@@
 The~search is depicted graphically in \figref{search-moves}.
-The~solid lines represent useful moves; the dotted lines represent
+Solid lines represent useful moves; dashed lines represent
 useless moves.
@@
 discovered;
-if~you like, time increases as you move downward.
+as you move downward, time increases.
 Horizontally, we~have organized the points into five columns; 
-the bottom point in each column is the result of a useful move, and we
-call it a useful point.
-Any points above it in the same column are useless.
+the bottom point in each column is useful, and
+any points above it in the same column are useless.
 
@@
 But~a~generator may produce an infinite sequence of useless moves.
-(For example, if \mrfy\ should stumble upon a global optimum, every move from
-that point would be useless.)
+For example, if \mrfy\ should stumble upon a global optimum, every move from
+that point would be useless.
 Given an infinite sequence of useless inputs, a~filter would not
@@
 on the left-hand side, 
-the first move is useless; the second move is useful; and all
+the first move is useless, the second move is useful, and all
 subsequent moves are useless.
@@
 sequence containing a mix of useful and useless points.
-Each useful point is tagged with cumulated cost and its score:
+Each useful point is tagged with cumulative cost and its score:
 \smallfuzzverbatiminput{6pt}{everygen}
@@
 If \mrfy\ used Hughes's classic model of lazy search,
-we~could possibly expect to see an infinite sequence of points obtained by
+we~might expect to see an infinite sequence of points obtained by
 applying \texttt{nextPt} to many points in succession, as~done by
@@
 is computed by applying \texttt{nextPt} to
-the \emph{single} initial point \texttt{startPt}, \mbox{using} infinitely
-many distinct sources of random variation.
+the \emph{single} initial point \texttt{startPt}, 
+repeated infinitely many times.
+Each repetition uses a
+distinct source of random variation.
+
 To~understand the structure of
@@
 Then the first part of the \texttt{successors} list appears in the
-next column of \figref{search-moves}:
+next column:
 the~first four elements are \texttt{useless};
 the next element, which is the first useful point, is~\texttt{pt};
-and~the remaining elements, which are never evaluated,
+and~the remaining elements, which are never evaluated and are not
+shown in the figure,
 match the
 underscore.
-The first useful point is located by \mono{span~isUseless}.
+The first useful point \texttt{pt} is located by \mono{span~isUseless}.
 After the search reaches~\texttt{pt},
 we~start the search anew with a recursive 
-call to \mono{everypt~sg}, 
+call to \mono{everyPt~sg}, 
 passing~\texttt{pt}.
 %
-The most informative part of \texttt{everyPt} is last expression of
+The~most informative part of \texttt{everyPt} is last expression of
 the \texttt{do} block,
@@
 computational biology doesn't have to be fast to be useful.
-But~when we say that \mrfy\ solves some problems 8~times slower then
-SMURF, we~felt it wasn't a good showing for Haskell.
+But~seeing \mrfy\ solve some problems 8~times slower then
+SMURF felt like a poor showing for Haskell.
 \mrfy\ spends 98\%\ of its time running Viterbi's algorithm, and
@@
 more competitive with~C++.
+But~before we could try to make \mrfy\ faster, we~had to make it simpler.
 
@@
 \figref{viterbi-transformed}.
-But the full story is not nearly as nice at one could hope for:
+But the full story is not as nice at one could hope for:
 in~the code that accompanies our 2012 ICFP paper,
 \verb+vee'+ has 25~cases overall.
-Many of these cases exist because we~did not have enough insight into
-Haskell, and we made some decisions that might have been more
-appropriate to a FORTRAN program:
+Many of these cases exist because when we first wrote the code, we~did
+not have enough insight into 
+Haskell.
+We~made some decisions that might have been more
+appropriate to FORTRAN:
 \begin{itemize}
@@
 \end{itemize}
-Before trying to speed up the code, we~simplified~it.
-%
-%
-Our first step was to start thinking like Haskell
-programmers and use types:
+%
+%
+To~simplify this code, we started thinking more like Haskell
+programmers.
+In~particular, we used types more effectively:
 \begin{itemize}
@@
 \end{itemize}
-The immediate effect of these changes was to reduce case analysis on
+The immediate effect of these changes was to reduce case analysis along
 two dimensions:
@@
 \begin{figure}
-\noindent\hspace*{1.5cm}\begin{minipage}{9cm}
+\noindent\hspace*{1.5cm}\begin{minipage}{11cm}
 \smallverbatiminput{list-viterbi}
@@
 
-Our~next step was to use more abstraction---in~particular, to abstract
-over state labels.
-However, after abstracting over state labels, we~abandoned lists and
-went back to arrays.
-An~excerpt from our list-based code appears in \figref{list-viterbi}.
-The~problem is defining \verb+vee''+ as a memoized version
-of~\verb+vee'+.
-Although we experimented with specialized memo combinators that work
-with list prefixes, 
-we~were never confident that we could easily, efficiently, and
-correctly memoize~\verb+vee'+.
-Instead, we~wrote a new version, still using integer arithmetic, but
-guides by the idea $i$~is a \emph{count} of the number
-of amino acids in list~\texttt{aas}, but $j$~is the \emph{index}
-of~\texttt{node}.
-(We~also got
-an
-insight we got from
-the \texttt{next} function: Viterbi's algorithm is a form of minimum edit
+To~reduce case analysis even more, we~abstracted over state labels.
+\figref{list-viterbi} sketches the structure of the algorithm.
+\figref{viterbi-hov4} is the first example in this paper that
+shows \emph{all} the cases for~\verb+vee'+.
+It~reduces the original 25~cases to just~4.
+We~show only a sketch because we didn't use this version in
+production.
+The difficulty was in memoizing a \verb+vee'+ that takes lists as
+arguments. 
+We~built and tested specialized memo combinators that work
+with list prefixes, but 
+we~were never confident that the resulting code would be simple,
+correct, and efficient.
+Instead, we~used insights from the list-based code to create a new
+array-based version:
+\begin{itemize}
+\item
+We~kept the abstraction over \texttt{stateHat}, which we explain
+below.
+\item
+We~got the idea that if we have to compute with integers, we~want
+$i$ to \emph{count} the number
+of amino acids in the sequence~\texttt{aas}, 
+but we want $j$ to \emph{index} into an array of nodes.
+\item
+From
+the \texttt{next} function, we~got the idea that Viterbi's algorithm
+computes a minimum edit 
 distance.
-As~we were trying to improve the code, this insight helped keep us on track.)
-Armed with this new, better way of thinking about $i$~and~$j$,
-we~turned back to arrays.
+This idea didn't affect the code directly, but as we experimented with
+new variations, it~helped us keep track of the big picture.
+\end{itemize}
+Armed with these ideas, we~turned back to computing with arrays and
+integers.
 
@@
     + \min\limits_s \big\{
-         \vsum{V_{\mathit{\p j}}^{\prime s}(\p i\; s)} 
+         \vsum{V_{\mathit{\p j}}^{\prime s}(\p i)} 
      {a^{\prime}_{s_{\p j} \,\hat s_j}}, 
-      \mbox{ where $s$ can precede $\hat s$}  \big\}
+      \mbox{ where $s$ can precede $\hat s$ and $\p i = \widetilde{\p i}\; s$}  \big\}
 }
 \\[10pt]
-\mbox{where }\p i\; s&{} = {}& 
+\mbox{where }\widetilde{\p i}\; s&{} = {}& 
   \left\{ \begin{array}{l}
@@
 
-Our final code, like the code in \figref{list-viterbi}, is abstracted
-over state labels.
-We~now turn your attention to that abstraction.
+\figref{list-viterbi} shows a sketch in which the state
+label \texttt{stateHat} is abstract.
+We~kept that abstraction, and
+we~now turn your attention to how it works.
 In~Viterbi's equations,
-$V'$ is indexed by a residue count~$i$,\footnote
-{In~biology, when an amino acid enters a protein, it often loses a few
-atoms, and so within the protein, the amino acid is often called
-a \emph{residue}.}
+$V'$ is indexed by a residue count~$i$,
 a~node index~$j$,
 and
-a state $M$, $I$, or~$D$.
+a state label $M$, $I$, or~$D$.
 Each state uses a slightly different equation, but it is possible to
@@
 \figref{viterbi-abstracted} shows a single equation which, when
-state $\hat s$ is specialized to state $M$, $I$, or~$D$, recovers the
+state $\hat s$ is specialized to state $M$, $I$, or~$D$, becomes one of the
 original three equations from \figref{viterbi-transformed}.
-The figure sums the cost of emitting an amino acid at state~$\hat s$
-lowest of the costs of the paths entering~$\hat s$ from predecessor
+The figure adds the cost of emitting an amino acid at state~$\hat s$
+to the lowest of the costs of the paths entering~$\hat s$ from predecessor
 state~$s$.
 Index~$\p j$ is the index of the node containing the predecessor
-state, and $\p i; s$ is the index of the amino acid, if~any, to be
-emitted at state~$s$.
+state.
+Integer $\p i$~counts the number of amino acids emitted before~$\hat
+s$ and so is the index of the amino acid, if~any, to be
+emitted at state~$\hat s$.
 
@@
 a value constructor~\texttt{NI} for a node index~$j$.
-We~can then deliver the more abstract version of Viterbi's algorithm
+Given these tools, we~built the abstract, array-based version of
+Viterbi's algorithm 
 shown in \figref{viterbi-hov4}.
+Auxiliary functions \texttt{node} and \texttt{residue} do array
+indexing, and functions \texttt{transmission} and \texttt{emission}
+are variations of the functions \texttt{aScore} and \texttt{eScore}
+mentioned in \secref{aScore}.
 
@@
 
-\caption{Viterbi's algorithm abstracted over~$\hat s$}
+\caption{Viterbi's algorithm abstracted over~$\hat s$, with signatures
+of auxiliary functions}
 \figlabel{viterbi-hov4}
@@
 
-\figref{viterbi-hov4} is the first example in this paper that
-shows \emph{all} the cases for~\verb+vee'+.
-We~have reduced the original 25~cases to just~4.
-The~general case, which was shown in other examples, appears last.
-The three special cases above it handle the three states that can
+The~three base cases in \figref{viterbi-hov4} are not exactly the same
+as the base cases in \figref{list-viterbi}.
+In~\figref{viterbi-hov4}, the base cases
+handle the three states that can
 follow the begin state.
@@
 
-Although the~new code is about 3\%~slower than our original code, 
-it~turned out to a good starting point for experiments.
-Of~the myriad variations we tried, 
-the best involved two transformations:
+Our~original goal was to make Viterbi's algorithm faster.
+To~measure its speed, we~created a benchmark that runs only Viterbi's
+algorithm, but on inputs that mimic the way \mrfy\ is intended to be used.
+The benchmark contains 56 protein sequences and 9 hidden Markov
+models.
+The query sequences are taken from the from the deep-sea 
+bacterium \emph{Thermotoga maritima}, and they 
+range in length from 80 to 1200 residues.
+The models are each built from a multiple-structure alignment of the
+proteins in 
+a $\beta$-structural protein superfamily.
+To~run the benchmark, we read in the query sequences and the models,
+and we compute the score of the best alignment of each sequence
+against each model.
+Using GHC~7.4 on~an Intel Ivy Bridge Core i5-3570 CPU clocked at
+3.4GHz, this benchmark takes about 45~seconds to~run.
+\mrfy\ spends most of its time garbage collecting and 99\%\ of the
+remaining time running Viterbi's algorithm.
+Time~spent in I/O and parsing is only about 0.2\%~of the
+total.
+
+Using this benchmark, the new code wasn't much different from the old
+code---just about 3\%~slower.
+But it 
+was a good starting point for experiments.
+We~tried myriad variations, 
+the best of which involved two transformations:
 \begin{itemize}
@@
 \item
-We~then added a fourth case to \verb+vee'+, which handles the match
+We~added a fourth case to \verb+vee'+, which handles the match
 state in node~1 when the residue count is nonzero.
@@
 very effectively, leading to 30\%~faster performance on our
-UNCHARACTERIZED MEDIUM benchmark.
+benchmark.
 The~result is 26\%~faster than our original code.
 
-This change laid the groundwork for more experiments.
-The most interesting of these experiments was to follow the advice of
+One of the variations that didn't help, surprisingly, was
+to follow
 \citet{tibell:high-performance} and use GHC's UNPACK pragma on scalar
 values---in our case, scores.
-Using the UNPACK pragma made our original code 7\%~faster but our new
+Although using the UNPACK pragma made our original code 7\%~faster, it
+makes the new
 code 9\%~slower.
 
-We~then turned to GHC's profiler, running Viterbi's algorithm on a
-model of 343~nodes and a query of 592~residues.
-We~expect the memo table to have about 600,000 entries,
-and the profiler says that the program allocates 531MB, or 7.85KB
+To~develop a more detailed understanding of performance, we~turned
+to GHC's profiler, running Viterbi's algorithm on a single query sequence
+of 592~residues, against a
+model of 343~nodes.
+To~help you understand the results, we~observe that the memo table
+for \verb+vee''+ has 610,197 entries: one for each combination of
+$\hat s$, $j$, and~$i$.
+The~profiler told us that a lot of time was spent in memory
+management, and that \mrfy\ allocated 531MB, or 7.85KB
 per~entry.
 This seems excessive.
+%
+The profile contained some information we
+understood and some we didn't.
 \begin{itemize}
 \item
-One culprit is the function \verb+vee''+, which is charged with
+The most easily understood source of allocation
+is the function \verb+vee''+, which is charged with
 allocating about 66MB.
-To~estimate the amount of memory consumed by
-an entry, we~imagine that representing a boxed
-score on the heap might require three 32-bit words.
-A~thunk involving three parameters might be a bit like a four-element
+What \verb+vee''+ does is call memo combinators,
+so the 66MB might be the memo table.
+To~evaluate this hypothesis, we~estimated the amount of memory that
+might be consumed by a memo-table entry.
+
+Once evaluated, a memo-table entry will contain a pointer to a
+double-precision floating point number allocated on the heap.
+The~pointer itself should occupy a word, and representing a
+boxed \texttt{Double} on the heap might
+require three 32-bit words.
+
+Before it is evaluated, a memo-table entry should contain a thunk
+involving the parameters $\hat s$, $j$, and~$i$.
+Such 
+a~thunk might be a bit like a four-element
 list, taking perhaps 12~words for the cons cells and another 8~words
-for the values?
-And then the pointer in the memo table itself would occupy a word.
-So~24~words per entry times 4~bytes per word would be 96B per
-entry for a total of about 59MB?
+for  values.
+
+By this estimate we could expect
+24~words per entry times 4~bytes per word, which would be 96B per
+entry and a total of about 59MB.
 Because the actual memo table involves arrays of arrays of arrays,
-we~won't quibble about a discrepancy of~10\%.
-\item
-The computation of \texttt{score} (the sum of transition 
-and emission scores) is charged with allocating 45MB total, or about
+we~won't quibble about a discrepancy of~10\%; 
+our estimate of the memory taken by the memo table seems accurate
+enough, and we~can think of nothing we can do to improve it.
+\item
+Another source of allocation---one that surprised us---is the 
+computation of \texttt{score}, which is the let-bound sum of transition 
+and emission scores in \figref{viterbi-hov4}.
+This binding is charged with allocating 45MB total, or about
 4.5~words for each transition in the computation tree.
-We~have no idea why this combination is allocating, and various
-attempts to make the sum strict (e.g., by using a bang pattern in 
-\mono{let~!score = }\ldots) either have no effect or actually make
-performance worse.
-\item
-There is also some significant allocation in the parser, but the
-elephant in the room is the function \verb+vee'+ itself, which
-allocates 258MB, which is half the memory allocated by the entire
-program.
+The~\texttt{edge} function is strict in the first argument, which
+is \texttt{score}, and we don't know why the computation is allocating.
+As~an experiment, we tried to make
+the sum strict by using a bang pattern in 
+``\mono{let~!score = }\ldots''
+While this change eliminated \texttt{score} as a source of allocation,
+the allocations moved somewhere else, because the total number of
+bytes allocated increased very slightly, by just 12B.
+\item
+On~this much smaller input, 16\%\ of the allocation is done by the
+code that reads and parses the model and the query sequence.
+We~don't care about improving that part.
+\item
+The single most expensive part of the program
+is the function \verb+vee'+ itself, which
+allocates 258MB---half of all memory allocated.
 We~have no idea what is happening.
@@
 so~perhaps we have a space leak, but even given the various profiling
-tools available, we~are not competent to find~it.
+tools available, we~were not able to find~it.
+
+To~try to understand and eliminate the unwanted allocations,
+we~explored a many variations on the code, of~which the most ambitious (and
+most tedious) was to desugar the list comprehension by~hand and to
+replace the standard list operations with operations that participate
+in stream fusion \cite{coutts:fusion}.
+Nothing~helped.
 \end{itemize}
-In~principle, we~could have diagnosed our problems by reading 
-GHC Core output from the compiler.
-We~found guidance in Don Stewart's invaluable blog posts,
-and we also used his \texttt{ghc-core} tool.
-But it wasn't enough.
+In~desperation, we~tried to diagnose our problems by reading 
+Core output from~GHC.
+We~were guided by Don Stewart's invaluable blog posts,
+and we also used his \texttt{ghc-core} tool, 
+but it wasn't enough.
 Even though one of us has worked on several optimizing compilers, 
@@
 
-We~succeeded in making Viterbi's algorithm a lot simpler and somewhat
-faster, but our original goal was to make it competetive with~C.
-The measurements using SMURF are depressing, and the SMURF code is
-even more depressing, so we reimplement Viterbi's algorithm from
-scratch in a clean C~program.
-To~avoid writing a parser, we extend \mrfy\ with code that reads a
+Our efforts to improve \mrfy's sequential
+performance concluded unhappily.
+Although we made Viterbi's algorithm a lot simpler and somewhat
+faster, it~is still nowhere near competetive with~C++.
+The measurements using SMURF are depressing enough, but because we were
+not entirely sure what SMURF was doing, we wrote a fresh,
+``cleanroom'' implementation of Viterbi's algorithm in~C, using
+forward dynamic programming.
+To~avoid writing a parser, we extended \mrfy\ with a function that reads a
 model and a query sequence and emits a C~file containing that model
 and query as initialized data.
-
-If~we look not at the whole package but just at Viterbi's algorithm,
-the~performance story is even more grim:
-we~wrote a custom C~program that computes a score using Viterbi's
-algorithm and does nothing else, and it runs about a hundred times
-faster than \mrfy's implementation.
-Ironically, we've~had a lot of trouble getting the code right, but it
+On~a synthetic benchmark involving all pairs of three models and three
+query sequences, our new C~code runs about 45~times faster than \mrfy.
+Ironically, we've~had a lot of trouble getting the C~code right, but it
 allocates only one double-precision floating-point number for each
-entry in the memo table; Haskell allocates about 30~times more.
+entry in the memo table.
+Haskell allocates about 30~times more.
 
@@
 obstacle.
-We~were lucky that we could get new results (CITE NOAH'S LATEST PAPER
-HERE?) even though our core algorithm is 100~times slower than the
+We~were lucky that we could get new results
+even though our core algorithm is 45~times slower than the
 same algorithm written in~C.
@@
 
-\subsection{Debugging and testing: awkward, then  less awkward}
+\subsection{Debugging and testing: awkward at first, but then  less awkward}
 
@@
 
-Our experience writing code and trying new ideas was excellent, as~was
-the ease of parallelizing \mrfy.
-Higher-order functions, memoization, laziness, and parallel strategies
-really worked.
-But~as discussed above, profiling didn't really work for~us.
-And we also encountered other obstacles that prevented functional
+Because we got lucky, the difficulty of understanding Haskell's
+allocation behavior didn't present the obstacle it might have.
+But we did encounter other obstacles that prevented functional
 programming from working as well as we would have liked.
@@
 
-We had a hard time diagnosing run-time errors.
-We~expected some run-time errors;
+We~started \mrfy's development using GHC~7.0, and
+we had a hard time diagnosing run-time errors.
+Run-time errors were something we expected to deal with;
 our~group's legacy file
-format is poorly documented and hard to deal with.
+format is poorly documented, and code that reads it usually contains
+faults. 
 (When beta
@@
 format are unclear.)
-But~using Haskell, we~found the errors hard to diagnose.
-Calls to
-\texttt{trace} 
-littered our code,
-even when relegated to wrapper functions.
+But~even though we expected run-time errors, we~found them hard to diagnose.
+Our code became littered with
+calls to
+\texttt{trace}, 
+even when we relegated \texttt{trace} to wrapper functions.
 We~didn't know about the backtrace feature of GHC's profiler,
 and even after we learned about it, it didn't help:
-the profiler can be used only on very small test cases,
+that profiler could be used only on very small test cases,
 which didn't always trigger the errors.
-This same limitation affected GHCi's debugger; 
+
+The problems with run-time errors and profiling were all fixed in
+February~2012, when GHC~7.4 was released.
+By~that time our code had matured, and run-time errors were no longer
+a significant issue, but occasionally we still appreciate the ability 
+to get a stack trace on error.
+
+While we were wrestling with run-time errors, we~also tried to use
+GHCi's debugger, but it was too limited:
 in~GHCi, our \texttt{vee'} function
@@
 \texttt{vee''} function impractical.
-In~August~2011, 
-Lennart Augustsson said that the biggest advantage of Strict
-Haskell is getting a stack trace on error,
-and when GHC~7.4 was released in early~2012, it~included support for
-stack traces. 
-We~have used the stack-trace feature a couple of times, and it helps.
 
@@
 informative. 
-(To~be fair to ourselves, the \texttt{shrink} method is barely
+To~be fair to ourselves, the \texttt{shrink} method is barely
 mentioned in the original 
-paper on QuickCheck.
-It~shows up under another name, as~an improvement made by Andy Gill.)
+paper on QuickCheck \cite{claessen:quickcheck};
+it~shows up under another name and is described as~an improvement made
+by Andy Gill. 
 \end{itemize}
@@
 (the smallest model supported by our original implementation)
-and two or three amino acids.
+and two or three residues.
 Moreover, the models contained just a handful of nonzero
@@
 We~have experience modifying the older tools.
+ALL THESE TOOLS
 
@@
 and we had to give~up.
-We~believe we could write an equivalent tool in~Haskell,
-with most of Matt's performance, in at most nine months.
+We~believe we could write an equivalent tool in~Haskell
+% with most of Matt's performance, % HA!
+in at most nine months.
 
@@
 to implement.
+TALK ABOUT RICHARD SENINGTON'S STUFF HERE.
 
@@
 for medium-sized projects,
-but he has played a secondary role, primarily refactoring and testing
+but he has played a secondary role in \mrfy's development, primarily
+refactoring and testing 
 others' code.
@@
 Some technical difficulties remain, particularly with installation:
-the only reliable way we have found to enable compilation with
-profiling is to destroy all existing compiled libraries, 
+the only reliable way we have found to compile code with support for
+profiling is first to destroy all existing compiled libraries, 
 update our Cabal configuration to ask for libraries to be compiled
@@
 As~of early 2013, the process of reinstallation, although not
-difficult, is time-consuming and tedious.
+intellectually challenging, is time-consuming and tedious.
 
@@
 Bounds violations lead to run-time errors, which we have not been able
-to identify any systematic way to~debug.
-GHC's profiler can provide stack traces, but we~found this information
-difficult to discover, and as noted above,
-there are obstacles to profiling.
+to identify any systematic way to~debug, although the new stack traces help.
 We're~aware that debugging lazy functional programs has been a topic
@@
 
-Ideally we would use QuickCheck to find bugs, but as we mention in
+We~recommend that everyone learn to use QuickCheck to find bugs, but
+as we mention in 
 \secref{awkward-quickcheck}, there are obstacles.
 We~have now overcome these obstacles, but
-we~sorely regret not doing so earlier.
+we~regret not having done so earlier.
 In~light of our experience, we~have instituted a new programming practice:
@@
 When invariants are not enforced by Haskell's static type system,
-we~will write them as Haskell predicates.
+we~write them as Haskell predicates.
 We can then immediately run QuickCheck on each predicate, to verify
@@
 Haskell programming.
-Eventually we learned to step slowly away from the code and to seek
+Eventually we learned to step away from the code and to seek
 out opportunities for abstraction and simplification.
@@
 enticing tools, libraries, and packages,
-not all of them are worth using.
+not all are worth using.
 Some are not ready for prime time, and some were once great but are
@@
 
+%\begingroup
+%\tracingall
 \bibliography{mrfy_experience_report}
-\bibliographystyle{jfp}
+\bibliographystyle{plainnatx}
+%\endgroup
 

### 2013-05-29 jfp.tex
incorporates AG's comments and Noah's contribution on Senington
this draft might be submittable, although there is one nasty page break

@@
 
-\pagetuningfalse % trying to get page breaks in the best spots
+\pagetuningtrue % trying to get page breaks in the best spots
 
@@
 (\secref{perf}).
-\ifpagetuning
   And our new tool solves a problem that could not be solved by
   the~C++ tool which preceded~it.
-\else
-  \ifpagetuning
-  And overall,
-  \else
-  Moreover, measured as a whole, 
-  \fi
-  the software that incorporates our new
-  code 
-  performs significantly
-  better than the~C++ software that preceded~it.
-\fi
 \item
@@
 We couldn't tell in advance which 
-\ifpagetuning
 would be which
-\else
-experience we were likely to have
-\fi
  (\secref{penumbra}).
@@
 correct.
+
+\ifpagetuning\enlargethispage{1.2\baselineskip}\fi
 
@@
 using Viterbi's algorithm.
-\ifpagetuning
+\iffalse
 This likelihood is \emph{conditioned} on the placement.
@@
 and a full day to implement.
-{\ifpagetuning\looseness=1 \par\fi}
 
@@
 SMURF.
-\ifpagetuning\vadjust{\break}\fi
 For example, on a microbenchmark that searches for a
@@
 
+\begin{figure}%%%[h!]
+\centerline{\loadpdf[]{speedup}}
+\medskip
+\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
+\figlabel{speedup}
+\end{figure}
+
 We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
@@
 this benchmark has a lot of parallelism,
-which Haskell made it easy to exploit.
+which Haskell made easy to exploit.
 Using
@@
 
-\begin{figure}%%%[h!]
-\centerline{\loadpdf[]{speedup}}
-\medskip
-\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
-\figlabel{speedup}
-\end{figure}
-
 %%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
@@
 
-Research software for
-computational biology doesn't have to be fast to be useful.
-But~seeing \mrfy\ solve some problems 8~times slower then
-SMURF felt like a poor showing for Haskell.
+Software for
+computational-biology research doesn't have to be fast to be useful.
+But~if \mrfy\ solves some problems 8~times slower then
+SMURF, that's a poor showing for Haskell.
 \mrfy\ spends 98\%\ of its time running Viterbi's algorithm, and
@@
 From
-the \texttt{next} function, we~got the idea that Viterbi's algorithm
+\figref{list-viterbi}'s \texttt{next} function, we~got the idea that
+Viterbi's algorithm 
 computes a minimum edit 
@@
 To~develop a more detailed understanding of performance, we~turned
-to GHC's profiler, running Viterbi's algorithm on a single query sequence
+to GHC's profiler.
+We ran Viterbi's algorithm on a single query sequence
 of 592~residues, against a
 model of 343~nodes.
-To~help you understand the results, we~observe that the memo table
-for \verb+vee''+ has 610,197 entries: one for each combination of
+The~profiler told us that a lot of time was spent in memory
+management, and that \mrfy\ allocated a total of 531MB.
+\mrfy's main data structure is the memo table used for \verb+vee''+, 
+which has one entry for each combination of
 $\hat s$, $j$, and~$i$.
-The~profiler told us that a lot of time was spent in memory
-management, and that \mrfy\ allocated 531MB, or 7.85KB
-per~entry.
-This seems excessive.
-%
-The profile contained some information we
+In~this run, that's 610,197 entries.
+We~expected \mrfy\ to use an amount of memory proportional to the
+size of the memo table.
+But the constant of proportionality is 7.85KB, which seems excessive.
+We~hoped that the detailed profile would explain, but it
+contained some information we
 understood and some we didn't.
@@
 
-To~try to understand and eliminate the unwanted allocations,
-we~explored a many variations on the code, of~which the most ambitious (and
+To~try to eliminate the unwanted allocations,
+we~explored many variations on the code, of~which the most ambitious (and
 most tedious) was to desugar the list comprehension by~hand and to
-replace the standard list operations with operations that participate
+replace standard list operations with counterparts that participate
 in stream fusion \cite{coutts:fusion}.
@@
 %%%% \remark{Others were mostly full-time work, but no need to distinguish.}
-{\ifpagetuning\looseness=1 \par\fi}
 \end{itemize}
@@
  in~haste.
-We~have experience modifying the older tools.
-ALL THESE TOOLS
+We~have experience modifying all three tools.
+
+\ifpagetuning\enlargethispage{\baselineskip}\fi
 
@@
 nearly every change resulted in new~segfaults.
-{\ifpagetuning\looseness=1 \par\fi}
 
@@
 
-We~built MRFy quickly, and we expect that
-higher-order functions will make MRFy easy to extend.
-Each new addition to MRFy's stochastic search has taken at most a day
+We~built MRFy quickly, and we expected that
+higher-order functions would make MRFy easy to extend.
+Each new addition to MRFy's stochastic search took at most a day
 to implement.
-TALK ABOUT RICHARD SENINGTON'S STUFF HERE.
+However, the next extension we wished to make
+was based on
+Richard Senington's model of \emph{local
+search} \cite{senington:combinators-local}, 
+which does not fit our abstraction of stochastic search.
+We~wound up writing a new program that combines parts
+of MRFy with Senington's  \texttt{local-search} package.
+This~task took only three weeks, and the technical results were terrific.
+Local search outperforms random hill-climbing, simulated
+annealing, and the genetic algorithm.
+Even though we're not entirely happy with the
+state of the resulting code, we~count the story as a success.
 
@@
 our hastily written Ruby and~C++.
-{\ifpagetuning\looseness=1 \par\fi}
 

### 2013-05-29 jfp.tex
tighten the Senington story

@@
  in~haste.
-We~have experience modifying all three tools.
+And all three have later been modified.
 
@@
 which does not fit our abstraction of stochastic search.
-We~wound up writing a new program that combines parts
-of MRFy with Senington's  \texttt{local-search} package.
-This~task took only three weeks, and the technical results were terrific.
+We~wound up writing a new program that adapts parts
+of MRFy to work with Senington's  \texttt{local-search} package.
+We're~disappointed that such similar algorithms have entirely
+different interfaces,
+but the adaptation took only three weeks, and the
+technical results were terrific. 
 Local search outperforms random hill-climbing, simulated
 annealing, and the genetic algorithm.
-Even though we're not entirely happy with the
-state of the resulting code, we~count the story as a success.
+We~therefore count the story as a success.
 

### 2013-05-29 jfp.tex
residue foot note improved

@@
 an amino-acid \emph{residue}.\footnote
-{In~biology, when an amino acid enters a protein, it often loses a few
-atoms, and so within the protein, what's left of the amino acid is called
+{When an amino acid combines with other amino acids to form a protein, it gives up
+two hydrogen atoms and an oxygen atom.
+What's left is called
 a \emph{residue}.
-In~this paper we use the terms ``amino acid'' and ``residue''
+In~this paper we use the terms ``residue'' and ``amino acid'' 
 interchangeably.

### 2013-05-29 jfp.tex
tuning page breaks
corrected and improved citations

@@
 
+\ifpagetuning\enlargethispage{1.2\baselineskip}\fi
+
 The~most likely path through a model is the one with the
@@
 correct.
-
-\ifpagetuning\enlargethispage{1.2\baselineskip}\fi
 
@@
 
-\ifpagetuning\enlargethispage{\baselineskip}\fi
+% \ifpagetuning\enlargethispage{\baselineskip}\fi
 
@@
 Richard Senington's model of \emph{local
-search} \cite{senington:combinators-local}, 
+search} \cite{senington:hybrid}, 
 which does not fit our abstraction of stochastic search.
@@
 our hastily written Ruby and~C++.
+
+\ifpagetuning\enlargethispage{\baselineskip}\fi
 

### 2013-05-29 jfp.tex
omit needless words

@@
 But~while we've gained something, we've also lost something---it's not
-at all obvious that the code in \figref{viterbi-hov4}  implements
+obvious that the code in \figref{viterbi-hov4}  implements
 Viterbi's equations.

### 2015-11-29 jfp.tex
Added MRFy journal paper to bib, correct typo

--- 2013-05-29 jfp.tex
@@
 
-\pagetuningtrue % trying to get page breaks in the best spots
+\pagetuningfalse % trying to get page breaks in the best spots
 
@@
 
-\renewcommand\bibsection{\section*{\refname}}
-
 \IfFileExists{timestamp.tex}{\input{timestamp}}{}
@@
 the flexibility and rapid prototyping of a scripting
-language, and it compiles to native code.
-From experience using Haskell to implement new algorithms for
-computational biology, we~report where Haskell really worked for us
-and what obstacles prevented it from working better.
-Higher-order functions, lazy evaluation, and
-monads worked well, but
+language, plus the performance of native code.
+In~our experience, higher-order functions, lazy evaluation, and
+monads really worked, but
 profiling and debugging presented obstacles.
-And~because allocation costs are difficult to predict and control,
-compilation to native code did not guarantee great performance.
-Finally, our experience with Haskell libraries varied greatly:
+Also, Haskell libraries vary greatly:
 memoization combinators and parallel-evaluation
@@
 Haskell's ecosystem
-made it easy for us to explore and run new algorithms.
+made it easy for us to develop new algorithms in computational
+biology.
 \end{abstract}
@@
 (\secref{perf}).
+\ifpagetuning
   And our new tool solves a problem that could not be solved by
   the~C++ tool which preceded~it.
+\else
+  \ifpagetuning
+  And overall,
+  \else
+  Moreover, measured as a whole, 
+  \fi
+  the software that incorporates our new
+  code 
+  performs significantly
+  better than the~C++ software that preceded~it.
+\fi
 \item
@@
 We couldn't tell in advance which 
+\ifpagetuning
 would be which
+\else
+experience we were likely to have
+\fi
  (\secref{penumbra}).
@@
 the code.
-Finally, we~have improved and added to explanations, figures, and diagrams.
+Finally, we~have improved and added to the figures and diagrams.
 
@@
 in amino-acid sequences that are only distantly related.
-
-A~paper for the computational biology community, based on MRFy, has been
-submitted to the ACM Conference on Bioinformatics, Computational Biology, and 
-Biomedical Informatics.
-That paper explores the effectiveness of various search strategies,
-and it also
-compares MRFy with several other software tools.
-MRFy is more accurate than the other software tested.
 MRFy is available at \url{mrfy.cs.tufts.edu}.
@@
 where each $x_i$~is 
-an amino-acid \emph{residue}.\footnote
-{When an amino acid combines with other amino acids to form a protein, it gives up
-two hydrogen atoms and an oxygen atom.
-What's left is called
-a \emph{residue}.
-In~this paper we use the terms ``residue'' and ``amino acid'' 
-interchangeably.
-}
-The number of residues, $N$, can differ from the number of columns
+an amino acid.
+The number of amino acids, $N$, can differ from the number of columns
 in the alignment,~\alignwidth.
@@
 The match state contains a table $e_{M_j}(x)$ which gives the
- probability that a homologous protein has residue~$x$ in
+ probability that a homologous protein has amino acid~$x$ in
  column~$j$.
@@
 The insertion state contains a table $e_{I_j}(x)$ which gives the
-probability that a homologous protein has gained residue~$x$ by
+probability that a homologous protein has gained amino acid~$x$ by
 insertion at column~$j$.
@@
 The deletion state determines the probability that a homologous protein
-has lost a~residue by deletion from column~$j$.
+has lost an amino acid by deletion from column~$j$.
 A~deletion state contains no tables of probabilities.
@@
 sequence from the prototype: 
-match a prototypical residue,
-insert an unexpected residue,
+match a prototypical amino~acid,
+insert an unexpected amino~acid,
 or
-delete a prototypical residue.
-
-A~hidden Markov model also has distinguished ``begin'' and ``end''
+delete a prototypical amino~acid.
+
+A~full hidden Markov model also has distinguished ``begin'' and ``end''
 states; begin and end states are like match states, but they
-correspond to no residues and contain no emission probabilities.
+correspond to no amino acids and contain no emission probabilities.
 %%
@@
 the begin and end states are known by their positions:
-in a $\alignwidth$-column model, the begin state shares
-column~0 with an insertion state, and the end state is the only state
-in column~$\alignwidth$. 
+in a $\alignwidth$-column model, the begin state is the only state in
+column~0 and the end state is the only state in column~$\alignwidth$.
 Every other state is referred to by its column~$j$
@@
 This model
-has begin and end states $B$~and~$E$;
-an~insertion state in node~0; and
-four
-full nodes, each containing 
+has begin and end states $B$~and~$E$,
+as well as four
+nodes, each containing 
 an insertion state~$I$, 
@@
 deletion state~$D$.
-(Node~0 may also be viewed as a full node, but its deletion state is
-not reachable from the begin state, so it is shown with only two
-states.)
+\remark{We need to figure out if the begin state is really node 0}
 
@@
 states and deletion states  \cite{Eddy:1998ut}. 
-Models in this family are called ``Plan7'' models because
-every full node has 7~transitions.
+The family is called the ``Plan7'' family because
+every model in the family has 7~transitions per node.
 Each transition has its own probability; the probabilities for
@@
 The other transitions enter states in node~$j+1$.)
-When column $j+1$ is a consensus column, transitions that enter
+When column $j+1$ is a consensus column, the transitions that enter
 its match state 
@@
 are more likely.
-When column $j+1$ is a non-consensus column, transitions that enter
+When column $j+1$ is a non-consensus column, the transitions that enter
 its insertion and deletion states are more likely.
@@
 In~any matching path, each match state and insertion state corresponds to a
-unique residue in the query sequence.
+unique amino acid in the query sequence.
 The probability of the path is the product of all of the corresponding
@@
 
-Probabilities are small, and products of small numbers can become very
-small indeed.
-If~we were to multiply probabilities represented as floating-point
-numbers, all the useful information would wind up in exponents,
-and an exponent holds only
-11~bits.
-In~order to store useful information in the significand, which holds 53~bits,
-we~follow standard practice: instead of multiplying probabilities,
-we~add logarithms of probabilities \cite{Viterbi:1967hq}.
-Also following standard practice, we~work with
-\emph{negated}
-logarithms of probabilities, which are  positive numbers.
-The negated logarithm of a probability is called a \emph{score}:
+Probabilities are small, and multiplying small numbers makes smaller
+numbers.
+Using floating-point arithmetic, multiplying probabilities amounts to
+storing data in the exponent of a floating-point number, which holds
+only 11~bits.
+In~order to store data in the significand, which holds 53~bits,
+we~instead sum the logarithms of probabilities \cite{Viterbi:1967hq}.
+Moreover, 
+ our models use a
+legacy file format which
+\emph{negates}
+the logarithm of each probability, making it a positive number.
+%%To~make these numbers more pleasant to read, we negate them,
+%%which eliminates hordes of minus signs from our file formats.
+The negated logarithm of a probability is called a \emph{score}.
 \smallverbatiminput{score}
-We~have given type \texttt{Score} a limited \texttt{Num} instance which permits
+Type \texttt{Score} has a limited \texttt{Num} instance which permits
 scores to be added and subtracted but not multiplied.
@@
 score~$V_j^{\prime M}(i)$ represents the probability of the most
-likely path of the first $i$~residues in the query sequence,
-terminating with placement of residue~$x_i$ in state~$M_j$.
+likely path of the first~$i$ amino acids in the query sequence,
+terminating with placement of amino acid $x_i$ in state~$M_j$.
 Scores $V_j^{\prime I}(i)$~and~$V_j^{\prime D}(i)$ are similar.
@@
 
-To~use Haskell, we had to reimplement the standard
+To~be able to use Haskell, we had to reimplement the standard
 algorithm for solving Viterbi's equations.
@@
 
-\ifpagetuning\enlargethispage{1.2\baselineskip}\fi
-
-Our code represents a query sequence as an immutable array of residues.
+Our code represents a query sequence as an immutable array of amino
+acids.
 In~idiomatic Haskell, 
-we might represent an individual residue~$x_i$
-using a value of algebraic data type with constructors named for
-amino acids:
+we might represent an individual amino acid~$x_i$
+using a value of algebraic data type:
 \begin{smallverbatim}
@@
 But in our
-legacy file format, each amino acid (residue) is a small integer
+legacy file format, each amino acid is a small integer
 used only to index arrays.
 %which we use only as an array index.
-We therefore chose a representation that is easier to make consistent
-with query sequences as represented on~disk:
+We therefore chose 
 \smallverbatiminput{aa}
@@
 specification is
-\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.\seclabel{aScore}
+\mbox{$\mathtt{aScore}\;s\;\hat s\;j = \txprobxj s {\hat s}$}.
 The
@@
 \mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$},
-where $x_i$ is the residue located at position~$i$ in the query
+where $x_i$ is the amino acid located at position~$i$ in the query
 sequence. 
@@
 it~produces the same result as~\verb+vee'+,
-but repeated calls return in constant time:
+but faster: 
 \smallverbatiminput{memo}
@@
 the probability of finding amino acid~$x_i$ in column~$j$ 
-\iffalse
+\ifpagetuning
 depends on the amino acid~$x'$ in column~${\pairedwith i}$.
@@
 using Viterbi's algorithm.
-\iffalse
+\ifpagetuning
 This likelihood is \emph{conditioned} on the placement.
@@
 The~search is depicted graphically in \figref{search-moves}.
-Solid lines represent useful moves; dashed lines represent
+The~solid lines represent useful moves; the dotted lines represent
 useless moves.
@@
 discovered;
-as you move downward, time increases.
+if~you like, time increases as you move downward.
 Horizontally, we~have organized the points into five columns; 
-the bottom point in each column is useful, and
-any points above it in the same column are useless.
+the bottom point in each column is the result of a useful move, and we
+call it a useful point.
+Any points above it in the same column are useless.
 
@@
 But~a~generator may produce an infinite sequence of useless moves.
-For example, if \mrfy\ should stumble upon a global optimum, every move from
-that point would be useless.
+(For example, if \mrfy\ should stumble upon a global optimum, every move from
+that point would be useless.)
 Given an infinite sequence of useless inputs, a~filter would not
@@
 on the left-hand side, 
-the first move is useless, the second move is useful, and all
+the first move is useless; the second move is useful; and all
 subsequent moves are useless.
@@
 sequence containing a mix of useful and useless points.
-Each useful point is tagged with cumulative cost and its score:
+Each useful point is tagged with cumulated cost and its score:
 \smallfuzzverbatiminput{6pt}{everygen}
@@
 If \mrfy\ used Hughes's classic model of lazy search,
-we~might expect to see an infinite sequence of points obtained by
+we~could possibly expect to see an infinite sequence of points obtained by
 applying \texttt{nextPt} to many points in succession, as~done by
@@
 is computed by applying \texttt{nextPt} to
-the \emph{single} initial point \texttt{startPt}, 
-repeated infinitely many times.
-Each repetition uses a
-distinct source of random variation.
-
+the \emph{single} initial point \texttt{startPt}, \mbox{using} infinitely
+many distinct sources of random variation.
 To~understand the structure of
@@
 Then the first part of the \texttt{successors} list appears in the
-next column:
+next column of \figref{search-moves}:
 the~first four elements are \texttt{useless};
 the next element, which is the first useful point, is~\texttt{pt};
-and~the remaining elements, which are never evaluated and are not
-shown in the figure,
+and~the remaining elements, which are never evaluated,
 match the
 underscore.
-The first useful point \texttt{pt} is located by \mono{span~isUseless}.
+The first useful point is located by \mono{span~isUseless}.
 After the search reaches~\texttt{pt},
 we~start the search anew with a recursive 
-call to \mono{everyPt~sg}, 
+call to \mono{everypt~sg}, 
 passing~\texttt{pt}.
 %
-The~most informative part of \texttt{everyPt} is last expression of
+The most informative part of \texttt{everyPt} is last expression of
 the \texttt{do} block,
@@
 
-\begin{figure}%%%[h!]
-\centerline{\loadpdf[]{speedup}}
-\medskip
-\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
-\figlabel{speedup}
-\end{figure}
-
 We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
@@
 this benchmark has a lot of parallelism,
-which Haskell made easy to exploit.
+which Haskell made it easy to exploit.
 Using
@@
 
+\begin{figure}%%%[h!]
+\centerline{\loadpdf[]{speedup}}
+\medskip
+\caption{\mrfy's parallel speedup on an 8-bladed beta propeller}
+\figlabel{speedup}
+\end{figure}
+
 %%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
@@
 
-Software for
-computational-biology research doesn't have to be fast to be useful.
-But~if \mrfy\ solves some problems 8~times slower then
-SMURF, that's a poor showing for Haskell.
+Research software for
+computational biology doesn't have to be fast to be useful.
+But~when we say that \mrfy\ solves some problems 8~times slower then
+SMURF, we~felt it wasn't a good showing for Haskell.
 \mrfy\ spends 98\%\ of its time running Viterbi's algorithm, and
@@
 more competitive with~C++.
-But~before we could try to make \mrfy\ faster, we~had to make it simpler.
 
@@
 \figref{viterbi-transformed}.
-But the full story is not as nice at one could hope for:
+But the full story is not nearly as nice at one could hope for:
 in~the code that accompanies our 2012 ICFP paper,
 \verb+vee'+ has 25~cases overall.
-Many of these cases exist because when we first wrote the code, we~did
-not have enough insight into 
-Haskell.
-We~made some decisions that might have been more
-appropriate to FORTRAN:
+Many of these cases exist because we~did not have enough insight into
+Haskell, and we made some decisions that might have been more
+appropriate to a FORTRAN program:
 \begin{itemize}
@@
 \end{itemize}
-%
-%
-To~simplify this code, we started thinking more like Haskell
-programmers.
-In~particular, we used types more effectively:
+Before trying to speed up the code, we~simplified~it.
+%
+%
+Our first step was to start thinking like Haskell
+programmers and use types:
 \begin{itemize}
@@
 \end{itemize}
-The immediate effect of these changes was to reduce case analysis along
+The immediate effect of these changes was to reduce case analysis on
 two dimensions:
@@
 \begin{figure}
-\noindent\hspace*{1.5cm}\begin{minipage}{11cm}
+\noindent\hspace*{1.5cm}\begin{minipage}{9cm}
 \smallverbatiminput{list-viterbi}
@@
 
-To~reduce case analysis even more, we~abstracted over state labels.
-\figref{list-viterbi} sketches the structure of the algorithm.
-\figref{viterbi-hov4} is the first example in this paper that
-shows \emph{all} the cases for~\verb+vee'+.
-It~reduces the original 25~cases to just~4.
-We~show only a sketch because we didn't use this version in
-production.
-The difficulty was in memoizing a \verb+vee'+ that takes lists as
-arguments. 
-We~built and tested specialized memo combinators that work
-with list prefixes, but 
-we~were never confident that the resulting code would be simple,
-correct, and efficient.
-Instead, we~used insights from the list-based code to create a new
-array-based version:
-\begin{itemize}
-\item
-We~kept the abstraction over \texttt{stateHat}, which we explain
-below.
-\item
-We~got the idea that if we have to compute with integers, we~want
-$i$ to \emph{count} the number
-of amino acids in the sequence~\texttt{aas}, 
-but we want $j$ to \emph{index} into an array of nodes.
-\item
-From
-\figref{list-viterbi}'s \texttt{next} function, we~got the idea that
-Viterbi's algorithm 
-computes a minimum edit 
+Our~next step was to use more abstraction---in~particular, to abstract
+over state labels.
+However, after abstracting over state labels, we~abandoned lists and
+went back to arrays.
+An~excerpt from our list-based code appears in \figref{list-viterbi}.
+The~problem is defining \verb+vee''+ as a memoized version
+of~\verb+vee'+.
+Although we experimented with specialized memo combinators that work
+with list prefixes, 
+we~were never confident that we could easily, efficiently, and
+correctly memoize~\verb+vee'+.
+Instead, we~wrote a new version, still using integer arithmetic, but
+guides by the idea $i$~is a \emph{count} of the number
+of amino acids in list~\texttt{aas}, but $j$~is the \emph{index}
+of~\texttt{node}.
+(We~also got
+an
+insight we got from
+the \texttt{next} function: Viterbi's algorithm is a form of minimum edit
 distance.
-This idea didn't affect the code directly, but as we experimented with
-new variations, it~helped us keep track of the big picture.
-\end{itemize}
-Armed with these ideas, we~turned back to computing with arrays and
-integers.
+As~we were trying to improve the code, this insight helped keep us on track.)
+Armed with this new, better way of thinking about $i$~and~$j$,
+we~turned back to arrays.
 
@@
     + \min\limits_s \big\{
-         \vsum{V_{\mathit{\p j}}^{\prime s}(\p i)} 
+         \vsum{V_{\mathit{\p j}}^{\prime s}(\p i\; s)} 
      {a^{\prime}_{s_{\p j} \,\hat s_j}}, 
-      \mbox{ where $s$ can precede $\hat s$ and $\p i = \widetilde{\p i}\; s$}  \big\}
+      \mbox{ where $s$ can precede $\hat s$}  \big\}
 }
 \\[10pt]
-\mbox{where }\widetilde{\p i}\; s&{} = {}& 
+\mbox{where }\p i\; s&{} = {}& 
   \left\{ \begin{array}{l}
@@
 
-\figref{list-viterbi} shows a sketch in which the state
-label \texttt{stateHat} is abstract.
-We~kept that abstraction, and
-we~now turn your attention to how it works.
+Our final code, like the code in \figref{list-viterbi}, is abstracted
+over state labels.
+We~now turn your attention to that abstraction.
 In~Viterbi's equations,
-$V'$ is indexed by a residue count~$i$,
+$V'$ is indexed by a residue count~$i$,\footnote
+{In~biology, when an amino acid enters a protein, it often loses a few
+atoms, and so within the protein, the amino acid is often called
+a \emph{residue}.}
 a~node index~$j$,
 and
-a state label $M$, $I$, or~$D$.
+a state $M$, $I$, or~$D$.
 Each state uses a slightly different equation, but it is possible to
@@
 \figref{viterbi-abstracted} shows a single equation which, when
-state $\hat s$ is specialized to state $M$, $I$, or~$D$, becomes one of the
+state $\hat s$ is specialized to state $M$, $I$, or~$D$, recovers the
 original three equations from \figref{viterbi-transformed}.
-The figure adds the cost of emitting an amino acid at state~$\hat s$
-to the lowest of the costs of the paths entering~$\hat s$ from predecessor
+The figure sums the cost of emitting an amino acid at state~$\hat s$
+lowest of the costs of the paths entering~$\hat s$ from predecessor
 state~$s$.
 Index~$\p j$ is the index of the node containing the predecessor
-state.
-Integer $\p i$~counts the number of amino acids emitted before~$\hat
-s$ and so is the index of the amino acid, if~any, to be
-emitted at state~$\hat s$.
+state, and $\p i; s$ is the index of the amino acid, if~any, to be
+emitted at state~$s$.
 
@@
 a value constructor~\texttt{NI} for a node index~$j$.
-Given these tools, we~built the abstract, array-based version of
-Viterbi's algorithm 
+We~can then deliver the more abstract version of Viterbi's algorithm
 shown in \figref{viterbi-hov4}.
-Auxiliary functions \texttt{node} and \texttt{residue} do array
-indexing, and functions \texttt{transmission} and \texttt{emission}
-are variations of the functions \texttt{aScore} and \texttt{eScore}
-mentioned in \secref{aScore}.
 
@@
 
-\caption{Viterbi's algorithm abstracted over~$\hat s$, with signatures
-of auxiliary functions}
+\caption{Viterbi's algorithm abstracted over~$\hat s$}
 \figlabel{viterbi-hov4}
@@
 
-The~three base cases in \figref{viterbi-hov4} are not exactly the same
-as the base cases in \figref{list-viterbi}.
-In~\figref{viterbi-hov4}, the base cases
-handle the three states that can
+\figref{viterbi-hov4} is the first example in this paper that
+shows \emph{all} the cases for~\verb+vee'+.
+We~have reduced the original 25~cases to just~4.
+The~general case, which was shown in other examples, appears last.
+The three special cases above it handle the three states that can
 follow the begin state.
@@
 
-Our~original goal was to make Viterbi's algorithm faster.
-To~measure its speed, we~created a benchmark that runs only Viterbi's
-algorithm, but on inputs that mimic the way \mrfy\ is intended to be used.
-The benchmark contains 56 protein sequences and 9 hidden Markov
-models.
-The query sequences are taken from the from the deep-sea 
-bacterium \emph{Thermotoga maritima}, and they 
-range in length from 80 to 1200 residues.
-The models are each built from a multiple-structure alignment of the
-proteins in 
-a $\beta$-structural protein superfamily.
-To~run the benchmark, we read in the query sequences and the models,
-and we compute the score of the best alignment of each sequence
-against each model.
-Using GHC~7.4 on~an Intel Ivy Bridge Core i5-3570 CPU clocked at
-3.4GHz, this benchmark takes about 45~seconds to~run.
-\mrfy\ spends most of its time garbage collecting and 99\%\ of the
-remaining time running Viterbi's algorithm.
-Time~spent in I/O and parsing is only about 0.2\%~of the
-total.
-
-Using this benchmark, the new code wasn't much different from the old
-code---just about 3\%~slower.
-But it 
-was a good starting point for experiments.
-We~tried myriad variations, 
-the best of which involved two transformations:
+Although the~new code is about 3\%~slower than our original code, 
+it~turned out to a good starting point for experiments.
+Of~the myriad variations we tried, 
+the best involved two transformations:
 \begin{itemize}
@@
 \item
-We~added a fourth case to \verb+vee'+, which handles the match
+We~then added a fourth case to \verb+vee'+, which handles the match
 state in node~1 when the residue count is nonzero.
@@
 very effectively, leading to 30\%~faster performance on our
-benchmark.
+UNCHARACTERIZED MEDIUM benchmark.
 The~result is 26\%~faster than our original code.
 
-One of the variations that didn't help, surprisingly, was
-to follow
+This change laid the groundwork for more experiments.
+The most interesting of these experiments was to follow the advice of
 \citet{tibell:high-performance} and use GHC's UNPACK pragma on scalar
 values---in our case, scores.
-Although using the UNPACK pragma made our original code 7\%~faster, it
-makes the new
+Using the UNPACK pragma made our original code 7\%~faster but our new
 code 9\%~slower.
 
-To~develop a more detailed understanding of performance, we~turned
-to GHC's profiler.
-We ran Viterbi's algorithm on a single query sequence
-of 592~residues, against a
-model of 343~nodes.
-The~profiler told us that a lot of time was spent in memory
-management, and that \mrfy\ allocated a total of 531MB.
-\mrfy's main data structure is the memo table used for \verb+vee''+, 
-which has one entry for each combination of
-$\hat s$, $j$, and~$i$.
-In~this run, that's 610,197 entries.
-We~expected \mrfy\ to use an amount of memory proportional to the
-size of the memo table.
-But the constant of proportionality is 7.85KB, which seems excessive.
-We~hoped that the detailed profile would explain, but it
-contained some information we
-understood and some we didn't.
+We~then turned to GHC's profiler, running Viterbi's algorithm on a
+model of 343~nodes and a query of 592~residues.
+We~expect the memo table to have about 600,000 entries,
+and the profiler says that the program allocates 531MB, or 7.85KB
+per~entry.
+This seems excessive.
 \begin{itemize}
 \item
-The most easily understood source of allocation
-is the function \verb+vee''+, which is charged with
+One culprit is the function \verb+vee''+, which is charged with
 allocating about 66MB.
-What \verb+vee''+ does is call memo combinators,
-so the 66MB might be the memo table.
-To~evaluate this hypothesis, we~estimated the amount of memory that
-might be consumed by a memo-table entry.
-
-Once evaluated, a memo-table entry will contain a pointer to a
-double-precision floating point number allocated on the heap.
-The~pointer itself should occupy a word, and representing a
-boxed \texttt{Double} on the heap might
-require three 32-bit words.
-
-Before it is evaluated, a memo-table entry should contain a thunk
-involving the parameters $\hat s$, $j$, and~$i$.
-Such 
-a~thunk might be a bit like a four-element
+To~estimate the amount of memory consumed by
+an entry, we~imagine that representing a boxed
+score on the heap might require three 32-bit words.
+A~thunk involving three parameters might be a bit like a four-element
 list, taking perhaps 12~words for the cons cells and another 8~words
-for  values.
-
-By this estimate we could expect
-24~words per entry times 4~bytes per word, which would be 96B per
-entry and a total of about 59MB.
+for the values?
+And then the pointer in the memo table itself would occupy a word.
+So~24~words per entry times 4~bytes per word would be 96B per
+entry for a total of about 59MB?
 Because the actual memo table involves arrays of arrays of arrays,
-we~won't quibble about a discrepancy of~10\%; 
-our estimate of the memory taken by the memo table seems accurate
-enough, and we~can think of nothing we can do to improve it.
-\item
-Another source of allocation---one that surprised us---is the 
-computation of \texttt{score}, which is the let-bound sum of transition 
-and emission scores in \figref{viterbi-hov4}.
-This binding is charged with allocating 45MB total, or about
+we~won't quibble about a discrepancy of~10\%.
+\item
+The computation of \texttt{score} (the sum of transition 
+and emission scores) is charged with allocating 45MB total, or about
 4.5~words for each transition in the computation tree.
-The~\texttt{edge} function is strict in the first argument, which
-is \texttt{score}, and we don't know why the computation is allocating.
-As~an experiment, we tried to make
-the sum strict by using a bang pattern in 
-``\mono{let~!score = }\ldots''
-While this change eliminated \texttt{score} as a source of allocation,
-the allocations moved somewhere else, because the total number of
-bytes allocated increased very slightly, by just 12B.
-\item
-On~this much smaller input, 16\%\ of the allocation is done by the
-code that reads and parses the model and the query sequence.
-We~don't care about improving that part.
-\item
-The single most expensive part of the program
-is the function \verb+vee'+ itself, which
-allocates 258MB---half of all memory allocated.
+We~have no idea why this combination is allocating, and various
+attempts to make the sum strict (e.g., by using a bang pattern in 
+\mono{let~!score = }\ldots) either have no effect or actually make
+performance worse.
+\item
+There is also some significant allocation in the parser, but the
+elephant in the room is the function \verb+vee'+ itself, which
+allocates 258MB, which is half the memory allocated by the entire
+program.
 We~have no idea what is happening.
@@
 so~perhaps we have a space leak, but even given the various profiling
-tools available, we~were not able to find~it.
-
-To~try to eliminate the unwanted allocations,
-we~explored many variations on the code, of~which the most ambitious (and
-most tedious) was to desugar the list comprehension by~hand and to
-replace standard list operations with counterparts that participate
-in stream fusion \cite{coutts:fusion}.
-Nothing~helped.
+tools available, we~are not competent to find~it.
 \end{itemize}
-In~desperation, we~tried to diagnose our problems by reading 
-Core output from~GHC.
-We~were guided by Don Stewart's invaluable blog posts,
-and we also used his \texttt{ghc-core} tool, 
-but it wasn't enough.
+In~principle, we~could have diagnosed our problems by reading 
+GHC Core output from the compiler.
+We~found guidance in Don Stewart's invaluable blog posts,
+and we also used his \texttt{ghc-core} tool.
+But it wasn't enough.
 Even though one of us has worked on several optimizing compilers, 
@@
 
-Our efforts to improve \mrfy's sequential
-performance concluded unhappily.
-Although we made Viterbi's algorithm a lot simpler and somewhat
-faster, it~is still nowhere near competetive with~C++.
-The measurements using SMURF are depressing enough, but because we were
-not entirely sure what SMURF was doing, we wrote a fresh,
-``cleanroom'' implementation of Viterbi's algorithm in~C, using
-forward dynamic programming.
-To~avoid writing a parser, we extended \mrfy\ with a function that reads a
+We~succeeded in making Viterbi's algorithm a lot simpler and somewhat
+faster, but our original goal was to make it competetive with~C.
+The measurements using SMURF are depressing, and the SMURF code is
+even more depressing, so we reimplement Viterbi's algorithm from
+scratch in a clean C~program.
+To~avoid writing a parser, we extend \mrfy\ with code that reads a
 model and a query sequence and emits a C~file containing that model
 and query as initialized data.
-On~a synthetic benchmark involving all pairs of three models and three
-query sequences, our new C~code runs about 45~times faster than \mrfy.
-Ironically, we've~had a lot of trouble getting the C~code right, but it
+
+If~we look not at the whole package but just at Viterbi's algorithm,
+the~performance story is even more grim:
+we~wrote a custom C~program that computes a score using Viterbi's
+algorithm and does nothing else, and it runs about a hundred times
+faster than \mrfy's implementation.
+Ironically, we've~had a lot of trouble getting the code right, but it
 allocates only one double-precision floating-point number for each
-entry in the memo table.
-Haskell allocates about 30~times more.
+entry in the memo table; Haskell allocates about 30~times more.
 
@@
 obstacle.
-We~were lucky that we could get new results
-even though our core algorithm is 45~times slower than the
+We~were lucky that we could get new results (CITE NOAH'S LATEST PAPER
+HERE?) even though our core algorithm is 100~times slower than the
 same algorithm written in~C.
@@
 
-\subsection{Debugging and testing: awkward at first, but then  less awkward}
+\subsection{Debugging and testing: awkward, then  less awkward}
 
@@
 
-Because we got lucky, the difficulty of understanding Haskell's
-allocation behavior didn't present the obstacle it might have.
-But we did encounter other obstacles that prevented functional
+Our experience writing code and trying new ideas was excellent, as~was
+the ease of parallelizing \mrfy.
+Higher-order functions, memoization, laziness, and parallel strategies
+really worked.
+But~as discussed above, profiling didn't really work for~us.
+And we also encountered other obstacles that prevented functional
 programming from working as well as we would have liked.
@@
 
-We~started \mrfy's development using GHC~7.0, and
-we had a hard time diagnosing run-time errors.
-Run-time errors were something we expected to deal with;
+We had a hard time diagnosing run-time errors.
+We~expected some run-time errors;
 our~group's legacy file
-format is poorly documented, and code that reads it usually contains
-faults. 
+format is poorly documented and hard to deal with.
 (When beta
@@
 format are unclear.)
-But~even though we expected run-time errors, we~found them hard to diagnose.
-Our code became littered with
-calls to
-\texttt{trace}, 
-even when we relegated \texttt{trace} to wrapper functions.
+But~using Haskell, we~found the errors hard to diagnose.
+Calls to
+\texttt{trace} 
+littered our code,
+even when relegated to wrapper functions.
 We~didn't know about the backtrace feature of GHC's profiler,
 and even after we learned about it, it didn't help:
-that profiler could be used only on very small test cases,
+the profiler can be used only on very small test cases,
 which didn't always trigger the errors.
-
-The problems with run-time errors and profiling were all fixed in
-February~2012, when GHC~7.4 was released.
-By~that time our code had matured, and run-time errors were no longer
-a significant issue, but occasionally we still appreciate the ability 
-to get a stack trace on error.
-
-While we were wrestling with run-time errors, we~also tried to use
-GHCi's debugger, but it was too limited:
+This same limitation affected GHCi's debugger; 
 in~GHCi, our \texttt{vee'} function
@@
 \texttt{vee''} function impractical.
+In~August~2011, 
+Lennart Augustsson said that the biggest advantage of Strict
+Haskell is getting a stack trace on error,
+and when GHC~7.4 was released in early~2012, it~included support for
+stack traces. 
+We~have used the stack-trace feature a couple of times, and it helps.
 
@@
 informative. 
-To~be fair to ourselves, the \texttt{shrink} method is barely
+(To~be fair to ourselves, the \texttt{shrink} method is barely
 mentioned in the original 
-paper on QuickCheck \cite{claessen:quickcheck};
-it~shows up under another name and is described as~an improvement made
-by Andy Gill. 
+paper on QuickCheck.
+It~shows up under another name, as~an improvement made by Andy Gill.)
 \end{itemize}
@@
 (the smallest model supported by our original implementation)
-and two or three residues.
+and two or three amino acids.
 Moreover, the models contained just a handful of nonzero
@@
 \item
-MRFy is used to detect homologous proteins in the presence of paired
+MRFy~\cite{daniels2015mrfy} is used to detect homologous proteins in the 
+presence of paired
 beta strands; it effectively supplants SMURF.
@@
  in~haste.
-And all three have later been modified.
-
-\ifpagetuning\enlargethispage{\baselineskip}\fi
+We~have experience modifying the older tools.
 
@@
 nearly every change resulted in new~segfaults.
+{\ifpagetuning\looseness=1 \par\fi}
 
@@
 and we had to give~up.
-We~believe we could write an equivalent tool in~Haskell
-% with most of Matt's performance, % HA!
-in at most nine months.
+We~believe we could write an equivalent tool in~Haskell,
+with most of Matt's performance, in at most nine months.
 
@@
 
-We~built MRFy quickly, and we expected that
-higher-order functions would make MRFy easy to extend.
-Each new addition to MRFy's stochastic search took at most a day
+We~built MRFy quickly, and we expect that
+higher-order functions will make MRFy easy to extend.
+Each new addition to MRFy's stochastic search has taken at most a day
 to implement.
-However, the next extension we wished to make
-was based on
-Richard Senington's model of \emph{local
-search} \cite{senington:combinators-local}, 
-which does not fit our abstraction of stochastic search.
-We~wound up writing a new program that adapts parts
-of MRFy to work with Senington's  \texttt{local-search} package.
-We're~disappointed that such similar algorithms have entirely
-different interfaces,
-but the adaptation took only three weeks, and the
-technical results were terrific. 
-Local search outperforms random hill-climbing, simulated
-annealing, and the genetic algorithm.
-We~therefore count the story as a success.
 
@@
 for medium-sized projects,
-but he has played a secondary role in \mrfy's development, primarily
-refactoring and testing 
+but he has played a secondary role, primarily refactoring and testing
 others' code.
@@
 Some technical difficulties remain, particularly with installation:
-the only reliable way we have found to compile code with support for
-profiling is first to destroy all existing compiled libraries, 
+the only reliable way we have found to enable compilation with
+profiling is to destroy all existing compiled libraries, 
 update our Cabal configuration to ask for libraries to be compiled
@@
 As~of early 2013, the process of reinstallation, although not
-intellectually challenging, is time-consuming and tedious.
+difficult, is time-consuming and tedious.
 
@@
 Bounds violations lead to run-time errors, which we have not been able
-to identify any systematic way to~debug, although the new stack traces help.
+to identify any systematic way to~debug.
+GHC's profiler can provide stack traces, but we~found this information
+difficult to discover, and as noted above,
+there are obstacles to profiling.
 We're~aware that debugging lazy functional programs has been a topic
@@
 
-We~recommend that everyone learn to use QuickCheck to find bugs, but
-as we mention in 
+Ideally we would use QuickCheck to find bugs, but as we mention in
 \secref{awkward-quickcheck}, there are obstacles.
 We~have now overcome these obstacles, but
-we~regret not having done so earlier.
+we~sorely regret not doing so earlier.
 In~light of our experience, we~have instituted a new programming practice:
@@
 When invariants are not enforced by Haskell's static type system,
-we~write them as Haskell predicates.
+we~will write them as Haskell predicates.
 We can then immediately run QuickCheck on each predicate, to verify
@@
 Haskell programming.
-Eventually we learned to step away from the code and to seek
+Eventually we learned to step slowly away from the code and to seek
 out opportunities for abstraction and simplification.
@@
 enticing tools, libraries, and packages,
-not all are worth using.
+not all of them are worth using.
 Some are not ready for prime time, and some were once great but are
@@
 implementation of the \texttt{Rand} monad.
-John Tibell taught us the \texttt{UNPACK} pragma.
+Johan Tibell taught us the \texttt{UNPACK} pragma.
 For~support and discussions 
@@
 
-%\begingroup
-%\tracingall
 \bibliography{mrfy_experience_report}
-\bibliographystyle{plainnatx}
-%\endgroup
+\bibliographystyle{jfp}
 
